### 2018-06-27 paper/metric_space_exploration.tex
added paper

\documentclass[10pt]{article}
\usepackage[margin=26mm]{geometry}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathbbol}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{algorithm}
\usepackage{url}
%\usepackage{floatflt}
%\usepackage{mathptmx}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{color}

\usepackage{subfig}

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{comment}

% \usepackage{ccaption}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\S}{\mathbb{S}}
%\newcommand{\Wlog}{W.l.o.g.\ }
%\newcommand{\wlog}{w.l.o.g.\ }
\newcommand{\eps}{\epsilon}
\newcommand{\orient}{O}
\newcommand{\cpro}{\Pi}
\newcommand{\fib}{\mathrm{Fib}}

\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diameter}{\mathrm{diam}}
\newcommand{\pointset}{P}
\newcommand{\metricspace}{\mathcal{M}}
\newcommand{\metric}{\delta}
\newcommand{\approxmetric}{\gamma}
\newcommand{\complexity}{C_{\metric}}
\newcommand{\doublingdimension}{\Delta}

\newcommand{\remark}[1]{\textbf{[#1]}}

%\newtheorem{theorem}{Theorem}
%\newdef{definition}{Definition}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{corollary}[theorem]{Corollary}

\newcommand{\myparagraph}[1]{\textbf{#1.}}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
%\newtheorem{remark}[theorem]{Remark}
\newtheorem{note}[theorem]{Note}
\newtheorem{problem}{Problem}

%\numberwithin{equation}{section}
%\numberwithin{figure}{section}

\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}

\title{Geometric Filters for Metric Spaces}
\author{Michael Kerber, Arnur Nigmetov}
%
\date{\today}

\begin{document}
\maketitle

\section{Introduction}

\paragraph{Motivation.}
%
Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
in a metric space $(\metricspace,\metric)$, 
consider the following standard operations:

\begin{description}
\item[Nearest Neighbor] Given a point $q\in\metricspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[d(q,p_i)\leq d(q,p_j)\]

\item[Metric approximation] Given $\eps>0$, compute a metric $\approxmetric$
such that for all $1\leq i,j\leq n$,
%
\[\metric(p_i,p_j)\leq\approxmetric(p_i,p_j)\leq (1+\eps)\metric(p_i,p_j).\]

By ``computing'' $\approxmetric$, we mean to compute a $n\times n$-matrix $E$ 
such that $E_{ij}=\approxmetric(p_i,p_j)$.
\end{description}

Assume that an algorithm is given to compute $\metric(\cdot,\cdot)$
for any two points in $\metricspace$ in worst-case complexity $\complexity$.
Then trivially, the above problems can be solved (exactly) in
$O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
For low-dimensional\footnote{we will make this notion precise
later in the text} metric spaces, we can improve these bounds:
Using cover trees~\cite{cover-trees}, we can preprocess the point set
in $O(\complexity n\log n)$ time to answer nearest neighbor queries
in $O(\complexity \log n)$ time. Combining cover trees with well-separated
pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
induced by $\metric$ in $O(\complexity n\log n)$ time and
of linear size (assuming, for simplicity, that $\eps$ is a constant).
Out of this spanner,
we can construct an approximate metric
with an all-pair shortest path computation,
resulting in a total running time of
\[O(\complexity n\log n+ n^2\log n)\]
using Johnson's algorithm~\cite{johnson}.

In the aforementioned techniques, $\complexity$ is usually considered
as a constant-time operation and hence ignored in the analysis.
There are good reasons for that: the most common case of a metric space
is $\metricspace=\R^d$ with $d$ some constant, in which case $\metric$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivating examples come from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed,
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.

In the light of the above examples, we pose the question: 
how can we practically reduce the number of distance computations 
in algorithms for metric spaces?
Note that for low-dimensional spaces, the aforementioned techniques
already provide a partial solution: for approximating the metric, for instance,
we only compute $O(n\log n)$ distances instead of $O(n^2)$.

\paragraph{Contribution.}
%
We demonstrate that the number of distance computations can be further
reduced by a careful re-implementation of the underlying geometric algorithms.
Our improvement is based on the following simple observation: many geometric
constructions (such as cover tree and WSPD constructions) require distance
computations only in the form of the following predicate: 
``Is $\metric(p_i,p_j)\leq t$'', where $t$ is a real value.
We show that using the information of the already computed distances and
the triangle inequality, we can answer the above predicate in many cases
without computing the actual distance.

We describe our idea with an example. In Figure~\ref{fig:1st_example},
we see a weighted graph on $P$ whose edges are the distances that have
been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\metric(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
distance computation is necessary. This approach is reminiscent to the usage
of \emph{numerical filters} in geometric computing in the CGAL library,
where geometric predicates in exact number types are first evaluated using 
certified floating-point arithmetic, and exact computation is only
invoked if the numerical evaluation did not arrive at a decisive result
(e.g.~\cite{bbp-interval,kerber-phd}).

\begin{figure}[h]

\centering
\includegraphics[width=6cm]{intro_example.eps}
\caption{An example of a partially revealed metric.}
\label{fig:1st_example}
\end{figure}

In our implementation, we maintain the best lower and upper bounds of all
pairwise distances in an $n\times n$ matrix of intervals. We show how to
update this matrix in $O(n^2)$ time when a new edge is added.
%We point out once more that a quadratic update time only pays off
%if distance computations are expensive, say $\complexity=\Omega(n^2)$.

We perform an extensive experimental evaluation of our idea.
In the first part, we concentrate on point clouds in Euclidean space~--
of course, the approach does not yield any performance benefits in that
situation because distance computations are cheap; however, the Euclidean
case provides a simple setup for experiments.
We investigate how the dimensionality, the clustering density, 
the approximation quality, and noise 
in the point cloud affect the number of distance computations.
Our result show that $\ldots$

We also provide experimental results for the case of persistence diagrams
which motivated this line of research. We point out that the space of
persistence diagrams fails to be ``low-dimensional''. To be precise,
its doubling dimension is $\infty$, and a finite set of $n$ diagram
can have the worst-case doublng dimension $\log n$. This worst-case
behavior is also the expected one if the diagrams are chosen randomly
(in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
However, we demonstrate that
for natural collections of persistence diagrams, a substantial fraction
of distance computations can be avoided. In detail, $\ldots$

We also have derived two minor technical results necessary for 
our implementation. First of all, since our underlying distance computation
algorithm for persistence diagrams only returns approximate answers,
we had to adapt all algorithms to cope with approximate distances.
Moreover, we show how a WSPDs can be constructed
out of cover trees (we only found 
the theoretically more efficient version of
net trees used in the literature, but net trees are more difficult to implement
efficiently). Both results can be obtained using standard methods, 
but seem to be not present in prior work.

\section{Geometric concepts}

We revise the concepts that were designed for nearest neighbor queries
and metric approximation in metric spaces with low dimensionality.
We will also review the computation methods and how they depend
on the predicate ``$\metric(p_i,p_j)\leq t$?'' mentioned above.

\paragraph{Notion of Dimensionality.}
%
Exansion constant, doubling dimension.

\subsection{Cover trees}

\subsection{WSPD}

\subsection{Spanners}

\section{Distance Bounds in Partially Revealed Metrics}
%
We describe our algorithm to maintain lower and upper bounds for the
distance of any two points in our (finite) metric space $(P=\{p_1,\ldots,p_n\},\metric)$. We assume for simplicity that all points in $P$ are distinct.

We maintain an $n\times n$ matrix $M$ whose entries
are intervals $M_{ij}=[a_{ij},b_{ij}]$, meaning that the distance
of $p_i$ and $p_j$ lies in that interval. 
We initialize $M_{ij}=[0,\infty]$ for all $i\neq j$
and $M_{ij}=[0,0]$ for $i=j$.
We describe next how to update $M$ when a new distance has been computed.

Note that an explicit graph representation of the partially revealed metric,
such as in Figure~\ref{fig:1st_example} is not necessary. However, for 
the sake of clarity,
this graph is helpful. So, let $G=(P,E)$ be the undirected weighted graph
where an edge between $p_i$ and $p_j$ with weight $\metric(p_i,p_j)$
exists if that distance has been computed. 
We maintain the invariant that the upper bound $b_{ij}$ equals the 
shortest path from $p_i$ to $p_j$ in $G$.

Assume that $\metric(p_i,p_j)=v\in\R$ has been computed during the algorithm.
First, we reset $M_{i,j}$ and $M_{j,i}$ to $[v,v]$. 
In $G$, the edge $(p_i,p_j)$ is added with weight $v$.
To update the upper bound of some entry $M_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
%
\[b_{k,\ell}\gets \min\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}\]
%
Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
time.

For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
%
\[v-b_{k,i}-b_{\ell,j}\]
%
is a lower bound for $\metric(p_k,p_\ell)$. Indeed, this follows from
the triangle inequality
%
\[\metric(p_i,p_j)\leq \metric(p_i,p_k)+\metric(p_k,p_\ell)+\metric(p_\ell,p_j)\]
by rearranging terms and plugging in the upper bounds for $\metric(p_i,p_k)$
and $\metric(p_k,p_\ell)$. An analogue bound holds for the roles of $i$
and $j$ swapped. Hence, we update
%
\[a_{k,\ell}\gets \max\{a_{k,\ell},v-b_{k,i}-b_{\ell,j},v-b_{k,j}-b_{\ell,i}\}\]
%
for all $k,\ell$ and obtain the updated $M$ in $O(n^2)$ time.

\medskip

With the matrix $M$, we can define filtered versions of our geometric
algorithms very easily: whenever the algorithm makes a query of the form
``Is $\metric(p_i,p_j)\leq t$?'', it is checked whether $a_{ij}> t$,
in which case the answer is ``yes'', or whether $b_{ij}\leq t$,
in which case the answer is ``no''. If none of the cases applies, 
$v\gets\metric(p_i,p_j)$ is computed, 
the matrix $M$ is updated as described above, and the predicate is answered
by comparing $v$ and $t$.

\section{Experiments on Point Clouds}
\label{sec:experiments_point_cloud}

\section{Metric spaces of Topological Summaries}
\label{sec:metric_spaces_top_summaries}

\section{Experiments on Persistence Diagrams}
\label{sec:experiments_persistence_diagrams}

\section{Conclusion}
\label{sec:conclusion}

\bibliography{bib}
\bibliographystyle{plain}

\end{document}

### 2018-06-27 cpp_version/include/cover_tree/boost_graph_import.h
cpp code added

@@ -1,0 +1,46 @@
+#ifndef SPANNER_WASSERSTEIN_BOOST_GRAPH_IMPORT_H
+#define SPANNER_WASSERSTEIN_BOOST_GRAPH_IMPORT_H
+
+#include <boost/config.hpp>
+#include <boost/graph/graph_traits.hpp>
+#include <boost/graph/adjacency_list.hpp>
+#include <boost/graph/dijkstra_shortest_paths.hpp>
+#include <boost/graph/incremental_components.hpp>
+
+using boost::graph_traits;
+using boost::adjacency_list;
+using boost::disjoint_sets;
+using boost::property_map;
+using boost::edge_weight_t;
+using boost::property;
+using boost::no_property;
+using boost::vecS;
+using boost::undirectedS;
+using boost::initialize_incremental_components;
+using boost::incremental_components;
+using boost::dijkstra_shortest_paths;
+
+
+using EdgeWeightProperty = property<edge_weight_t, double>;
+using Graph = adjacency_list<vecS, vecS, undirectedS, no_property, EdgeWeightProperty>;
+using VertexDescriptor = typename graph_traits<Graph>::vertex_descriptor;
+using EdgeDescriptor = typename graph_traits<Graph>::edge_descriptor;
+using EdgeIterator = typename graph_traits<Graph>::edge_iterator;
+using Edge = typename std::pair<VertexDescriptor, VertexDescriptor>;
+using DisjointSets = boost::disjoint_sets<size_t*, size_t*>;
+using WeightMap = boost::property_map<Graph, edge_weight_t>::type;
+
+
+struct EdgeHashStruct
+{
+    size_t operator()(const Edge& p) const
+    {
+        size_t seed = 0;
+        boost::hash_combine(seed, p.first);
+        boost::hash_combine(seed, p.second);
+        return seed;
+    }
+};
+
+
+#endif //SPANNER_WASSERSTEIN_BOOST_GRAPH_IMPORT_H

### 2018-06-27 cpp_version/include/cover_tree/cover_tree.h

@@ -1,0 +1,118 @@
+#pragma once
+
+#include <iostream>
+#include <map>
+#include <memory>
+#include <set>
+#include <unordered_set>
+
+#include "spdlog/spdlog.h"
+#include "wasserstein/wasserstein_space_point.h"
+
+namespace spd = spdlog;
+
+namespace wasser_spanner {
+
+    using DynamicSpannerR = DynamicSpanner<double>;
+    using AuctionParamsR = hera::AuctionParams<double>;
+    using Point = DynamicSpannerR::WassersteinSpacePoint;
+    using MatrixR = DynamicSpannerR::MatrixR;
+
+
+    struct CoverTree
+    {
+        static constexpr int k_max_lvl = std::numeric_limits<int>::max() - 10;
+        static constexpr int k_invalid_idx = -1;
+
+        struct Node;
+
+        using NodeVec = std::vector<Node*>;
+
+        struct Node
+        {
+            // data
+            int m_level { k_max_lvl + 1 };
+            int m_point_idx { k_invalid_idx };
+            // m_children vector does not include the node itself, it is implicit
+            std::unordered_map<int, std::vector<std::unique_ptr<Node>>> m_children;
+
+            // methods
+            Node(int _point_idx, int _level);
+
+            int get_point_idx() const
+            { return m_point_idx; }
+
+            int get_level() const
+            { return m_level; }
+
+            int get_num_children() const;
+
+            NodeVec get_children(int level) const;
+
+            friend std::ostream& operator<<(std::ostream& os, const Node& node);
+
+            std::shared_ptr<spd::logger> console { spd::get("console") };
+        };
+
+        // data members
+        static constexpr double m_base { 2.0 };
+        std::vector<Point> m_points;
+        unsigned int m_num_points;
+        int m_max_level;
+        int m_min_level;
+        std::unique_ptr<Node> m_root;
+        DynamicSpannerR m_dspanner;
+        unsigned int m_num_nodes { 0 };
+        std::unordered_map<int, Node*> m_point_to_node;
+
+        std::shared_ptr<spd::logger> console { spd::get("console") };
+
+
+        // member functions
+        CoverTree(double max_dist,
+                  const std::vector<Point>& points,
+                  const MatrixR& distance_matrix = MatrixR(),
+                  const AuctionParamsR& _ap = AuctionParamsR());
+
+        void insert(const Point& new_point);
+
+        bool insert_rec(const Point& p,
+                        const std::vector<Node*>& Qi,
+                        const int level,
+                        bool& inserted);
+
[…]

### 2018-06-27 cpp_version/include/cover_tree/wspd.h

@@ -1,0 +1,68 @@
+#pragma once
+
+#include <utility>
+#include <unordered_set>
+
+#include <boost/config.hpp>
+#include <boost/functional/hash.hpp>
+#include <boost/graph/graph_traits.hpp>
+#include <boost/graph/adjacency_list.hpp>
+#include <boost/graph/dijkstra_shortest_paths.hpp>
+#include <boost/graph/incremental_components.hpp>
+
+#include "spdlog/spdlog.h"
+
+#include "cover_tree/cover_tree.h"
+#include "cover_tree/wspd_node.h"
+#include "cover_tree/boost_graph_import.h"
+
+namespace spd = spdlog;
+
+namespace wasser_spanner {
+
+    struct WSPD
+    {
+        // types
+        using Node = WspdNode;
+        using NodePair = std::pair<const Node, const Node>;
+
+        struct NodePairHasher
+        {
+            std::size_t operator()(const NodePair &p) const
+            {
+                std::size_t result = 0;
+                boost::hash_combine(result, hash_value(p.first));
+                boost::hash_combine(result, hash_value(p.second));
+                return result;
+            }
+        };
+        using NodePairSet = std::unordered_set<NodePair, NodePairHasher>;
+
+        // data
+
+        CoverTree* m_cover_tree;
+        int m_num_points;
+        double m_epsilon;
+        double m_tau;
+        DynamicSpannerR* m_spanner;
+        NodePairSet m_wspd;
+
+        std::shared_ptr<spd::logger> console{spd::get("console")};
+
+        std::unique_ptr<Graph> m_graph { nullptr };
+        std::vector<size_t> m_rank;
+        std::vector<size_t> m_parent;
+        DisjointSets m_conn_comps;
+        std::unordered_map<Edge, double, EdgeHashStruct> m_distance_cache;
+
+        // methods
+
+        WSPD(CoverTree &ct, double eps);
+        void gen_wspd(NodePairSet& result, const Node &a, const Node &b, int rec_depth);
+        bool is_valid() const;
+        size_t size() const;
+        void make_spanner();
+        double get_spanner_distance(VertexDescriptor i, VertexDescriptor j);
+        double get_max_relative_error();
+    };
+}

### 2018-06-27 cpp_version/include/cover_tree/wspd_node.h

@@ -1,0 +1,67 @@
+#ifndef SPANNER_WASSERSTEIN_WSPD_NODE_H
+#define SPANNER_WASSERSTEIN_WSPD_NODE_H
+
+#include <iostream>
+#include <vector>
+#include <memory>
+#include <unordered_set>
+#include <unordered_map>
+
+#include <boost/functional/hash.hpp>
+
+#include "cover_tree/cover_tree.h"
+
+namespace wasser_spanner {
+
+    class WspdNode
+    {
+    private:
+        // members
+        CoverTree::Node* m_ct_node { nullptr };
+        const int m_level { CoverTree::k_max_lvl };
+        const int m_point_idx { CoverTree::k_invalid_idx };
+
+        std::unordered_map<int, CoverTree::NodeVec>* m_children { nullptr };
+
+        void get_all_descendants_indices_helper(std::unordered_set<int>& res) const;
+
+        WspdNode get_self_as_child() const;
+
+    public:
+        static DynamicSpannerR* dspanner;
+        static double constexpr base { 2.0 };
+
+        WspdNode(CoverTree::Node* node);
+
+        WspdNode(CoverTree::Node* node, int level);
+
+        int get_level() const
+        { return m_level; }
+
+        int get_point_idx() const
+        { return m_point_idx; }
+
+        CoverTree::Node* get_node_ptr() const
+        { return m_ct_node; }
+
+        bool is_leaf_node() const;
+
+        std::unordered_set<int> get_all_descendants() const;
+
+        double delta() const;
+
+        double get_distance_to_node(const WspdNode& other) const;
+
+        double diameter() const;
+
+        std::vector<WspdNode> get_next_layer_children() const;
+    };
+
+    std::size_t hash_value(const WspdNode& node);
+
+    bool operator==(const WspdNode& a, const WspdNode& b);
+
+    std::ostream& operator<<(std::ostream& s, const WspdNode& n);
+
+} // namespace wasser_spanner
+#endif //SPANNER_WASSERSTEIN_WSPD_NODE_H

### 2018-06-27 cpp_version/include/dnn/local/kd-tree.hpp

@@ -1,0 +1,330 @@
+#include <boost/range/counting_range.hpp>
+#include <boost/range/algorithm_ext/push_back.hpp>
+#include <boost/range.hpp>
+
+#include <queue>
+#include <stack>
+
+#include "../parallel/tbb.h"
+#include "def_debug_ws.h"
+
+template<class T>
+hera::ws::dnn::KDTree<T>::
+KDTree(const Traits& traits, HandleContainer&& handles, double _wassersteinPower):
+    traits_(traits), tree_(std::move(handles)), wassersteinPower(_wassersteinPower)
+{ assert(wassersteinPower >= 1.0); init(); }
+
+template<class T>
+template<class Range>
+hera::ws::dnn::KDTree<T>::
+KDTree(const Traits& traits, const Range& range, double _wassersteinPower):
+    traits_(traits), wassersteinPower(_wassersteinPower)
+{
+    assert( wassersteinPower >= 1.0);
+    init(range);
+}
+
+template<class T>
+template<class Range>
+void
+hera::ws::dnn::KDTree<T>::
+init(const Range& range)
+{
+    size_t sz = std::distance(std::begin(range), std::end(range));
+    tree_.reserve(sz);
+    weights_.resize(sz, 0);
+    subtree_weights_.resize(sz, 0);
+    for (PointHandle h : range)
+        tree_.push_back(h);
+    init();
+}
+
+template<class T>
+void
+hera::ws::dnn::KDTree<T>::
+init()
+{
+    if (tree_.empty())
+        return;
+
+#if defined(TBB)
+    task_group g;
+    g.run(OrderTree(tree_.begin(), tree_.end(), 0, traits()));
+    g.wait();
+#else
+    OrderTree(tree_.begin(), tree_.end(), 0, traits()).serial();
+#endif
+
+    for (size_t i = 0; i < tree_.size(); ++i)
+        indices_[tree_[i]] = i;
+}
+
+template<class T>
+struct
+hera::ws::dnn::KDTree<T>::OrderTree
+{
+                OrderTree(HCIterator b_, HCIterator e_, size_t i_, const Traits& traits_):
+                    b(b_), e(e_), i(i_), traits(traits_)            {}
+
+    void        operator()() const
+    {
+        if (e - b < 1000)
+        {
+            serial();
+            return;
+        }
+
+        HCIterator m = b + (e - b)/2;
+        CoordinateComparison cmp(i, traits);
[…]

### 2018-06-27 cpp_version/include/dynamic_spanner.h

@@ -1,0 +1,167 @@
+#ifndef DYNAMIC_SPANNER_H
+#define DYNAMIC_SPANNER_H
+/*
+#include <vector>
+#include <utility>
+
+#include <boost/config.hpp>
+#include <boost/functional/hash.hpp>
+
+#include <boost/graph/graph_traits.hpp>
+#include <boost/graph/adjacency_list.hpp>
+#include <boost/graph/dijkstra_shortest_paths.hpp>
+#include <boost/graph/incremental_components.hpp>
+
+using namespace boost;
+
+// class Point must provide function distance
+template<class Point, class Real = double>
+class DynamicSpanner
+{
+public:
+
+    using EdgeWeightProperty = property<edge_weight_t, double>;
+    using Graph = adjacency_list<vecS, vecS,undirectedS, no_property, EdgeWeightProperty>;
+    using VertexDescriptor = typename graph_traits<Graph>::vertex_descriptor;
+    using EdgeDescriptor = typename graph_traits<Graph>::edge_descriptor;
+    using EdgeIterator = typename graph_traits<Graph>::edge_iterator;
+    using Edge = typename std::pair<VertexDescriptor, VertexDescriptor>;
+    using DisjointSets = disjoint_sets<size_t*, size_t*>;
+    using WeightMap = property_map<Graph, edge_weight_t>::type;
+
+    struct EdgeHashStruct {
+        size_t operator()(const Edge& p) const
+        {
+            size_t seed = 0;
+            boost::hash_combine(seed, p.first);
+            boost::hash_combine(seed, p.second);
+            return seed;
+        }
+    };
+
+    DynamicSpanner(std::vector<Point>& _pts) :
+        points_(_pts),
+        num_points_(_pts.size()),
+        graph_(_pts.size()),
+        rank_(num_points_),
+        parent_(num_points_),
+        conn_comps_(&rank_[0], &parent_[0])
+    {
+        initialize_incremental_components(graph_, conn_comps_);
+        incremental_components(graph_, conn_comps_);
+    }
+
+    std::pair<Real, Real> distance_estimate(VertexDescriptor i, VertexDescriptor j)
+    {
+        Real d;
+        if (is_distance_cached(i, j, d)) {
+            return std::make_pair(d, d);
+        } else {
+            return std::make_pair(lower_bound(i,j), upper_bound(i,j));
+        }
+    }
+
+    // upper bound for distance between points_[i] and points_[j]
+    Real upper_bound(VertexDescriptor i, VertexDescriptor j)
+    {
[…]

### 2018-06-27 cpp_version/include/experiment_helper.h

@@ -1,0 +1,82 @@
+#ifndef SPANNER_WASSERSTEIN_EXPERIMENT_HELPER_H
+#define SPANNER_WASSERSTEIN_EXPERIMENT_HELPER_H
+
+#include <vector>
+#include <ostream>
+
+#include "wasserstein/wasserstein_space_point.h"
+
+extern std::mt19937_64 twister;
+
+namespace wasser_spanner {
+
+    using DynamicSpannerR = DynamicSpanner<double>;
+    using WassersteinSpacePointR = DynamicSpannerR::WassersteinSpacePoint;
+    using DiagramPointR = DynamicSpannerR::DiagramPointR;
+    using DiagramR = DynamicSpannerR::DiagramR;
+
+    using DgmVec = std::vector<WassersteinSpacePointR>;
+    using AuctionParamsR = hera::AuctionParams<double>;
+    using MatrixR= DynamicSpannerR::MatrixR;
+
+    struct RandomDiagParams
+    {
+        enum class Strategy { NORMAL, CLUSTERED };
+        friend std::ostream& operator<<(std::ostream& os, const RandomDiagParams& params);
+
+        Strategy strategy { Strategy::CLUSTERED };
+        int random_generator_seed { 1 };
+        int n_diagrams { 10 }; // for NORMAL only
+        int min_diagram_size { 10 };
+        int max_diagram_size { 1000 };
+        double std_dev { 100.0 };
+        double lower_bound_hor { 0.0 };
+        double upper_bound_hor { 100.0 };
+        double lower_bound_vert { 0.0 };
+        double upper_bound_vert { 100.0 };
+
+        // for clustering
+        int n_seeds { 30 };
+        double survive_prob { 1.0 };
+        int n_hierarchy_levels { 3 };
+        int branching_factor{ 3 };
+        double std_dev_decay { 0.1 };
+
+        AuctionParamsR auction_params;
+    };
+
+
+    void generate_random_diag(DgmVec& dgms, int n_diag, int n_points, int seed = 1);
+
+    // check triangle inequality for all possible combinations of three m_points
+    // (note that a priori approximate distance computation function is not symmetric)
+    bool is_triangle_ineq_satisfied(int a_idx, int b_idx, int c_idx, DynamicSpannerR& ds);
+
+    bool is_triangle_ineq_satisfied(DgmVec& dgms, DynamicSpannerR& ds);
+
+    double get_expansion_constant(const std::vector<std::vector<double>>& dist_matrix, double base = 2.0);
+
[…]

### 2018-06-27 cpp_version/include/wasserstein/auction_oracle_kdtree_restricted.hpp

@@ -1,0 +1,602 @@
+/*
+
+Copyright (c) 2015, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
+into other computer software, distribute, and sublicense such enhancements or
[…]

### 2018-06-27 cpp_version/include/wasserstein/auction_runner_gs.h

@@ -1,0 +1,124 @@
+/*
+
+Copyright (c) 2016, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
+into other computer software, distribute, and sublicense such enhancements or
[…]

### 2018-06-27 cpp_version/include/wasserstein/auction_runner_gs.hpp

@@ -1,0 +1,538 @@
+/*
+
+Copyright (c) 2016, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
[…]

### 2018-06-27 cpp_version/include/wasserstein/basic_defs_ws.h

@@ -1,0 +1,322 @@
+/*
+
+Copyright (c) 2015, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
[…]

### 2018-06-27 cpp_version/include/wasserstein/diagram_reader.h

@@ -1,0 +1,369 @@
+/*
+
+Copyright (c) 2015, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
[…]

### 2018-06-27 cpp_version/include/wasserstein/dist_bounds.h

@@ -1,0 +1,162 @@
+/*
+
+Copyright (c) 2015, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
[…]

### 2018-06-27 cpp_version/include/wasserstein/lower_bound.h

@@ -1,0 +1,82 @@
+#ifndef SPANNER_WASSERSTEIN_LOWER_BOUND_H
+#define SPANNER_WASSERSTEIN_LOWER_BOUND_H
+
+#include "basic_defs_ws.h"
+#include "dnn/geometry/euclidean-fixed.h"
+#include "dnn/local/kd-tree.h"
+
+namespace hera {
+    namespace ws {
+
+        // match every bidder to the nearest neigbour to get lower bound for Wasserstein distance
+        template<class Real, class PointContainer>
+        Real
+        lower_bound_(const PointContainer& bidders, const PointContainer& items, const AuctionParams <Real>& params)
+        {
+
+            const Real q = params.wasserstein_power;
+
+            const bool is_bottleneck = (q == std::numeric_limits<Real>::infinity() or q == hera::get_infinity());
+
+            Real result { 0.0 };
+
+            using DnnPoint          = dnn::Point<2, Real>;
+            using DnnTraits         = dnn::PointTraits<DnnPoint>;
+
+            std::vector<DnnPoint> dnn_points;
+            std::vector<DnnPoint*> dnn_point_handles;
+
+            IdxType true_idx { 0 };
+
+            dnn_points.clear();
+            dnn_points.reserve(items.size());
+            // store normal items in kd-tree
+            for (const DiagramPoint<Real>& g : items) {
+                if (g.is_normal()) {
+                    // index of items is id of dnn-point
+                    DnnPoint p(true_idx);
+                    p[0] = g.getRealX();
+                    p[1] = g.getRealY();
+                    dnn_points.push_back(p);
+                }
+                true_idx++;
+            }
+
+            for (size_t i = 0; i < dnn_points.size(); ++i) {
+                dnn_point_handles.push_back(&dnn_points[i]);
+            }
+
+            DnnTraits traits;
+            traits.internal_p = params.internal_p;
+            dnn::KDTree<DnnTraits> kdtree { traits, dnn_point_handles, is_bottleneck ? 1.0 : q };
+
+            for (const DiagramPoint<Real>& bidder : bidders) {
+                if (bidder.is_diagonal())
+                    continue;
+                DnnPoint bidder_dnn;
+                bidder_dnn[0] = bidder.getRealX();
+                bidder_dnn[1] = bidder.getRealY();
+                auto best_item = kdtree.findK(bidder_dnn, 1);
+                if (is_bottleneck)
[…]

### 2018-06-27 cpp_version/include/wasserstein/wasserstein.h

@@ -1,0 +1,381 @@
+/*
+
+Copyright (c) 2015, M. Kerber, D. Morozov, A. Nigmetov
+All rights reserved.
+
+Redistribution and use in source and binary forms, with or without modification, are permitted provided that the following conditions are met:
+
+1. Redistributions of source code must retain the above copyright notice, this list of conditions and the following disclaimer.
+
+2. Redistributions in binary form must reproduce the above copyright notice, this list of conditions and the following disclaimer in the documentation and/or other materials provided with the distribution.
+
+3. Neither the name of the copyright holder nor the names of its contributors may be used to endorse or promote products derived from this software without specific prior written permission.
+
+THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+
+
+You are under no obligation whatsoever to provide any bug fixes, patches, or
+upgrades to the features, functionality or performance of the source code
+(Enhancements) to anyone; however, if you choose to make your Enhancements
+available either publicly, or directly to copyright holder,
+without imposing a separate written license agreement for such Enhancements,
+then you hereby grant the following license: a  non-exclusive, royalty-free
+perpetual license to install, use, modify, prepare derivative works, incorporate
+into other computer software, distribute, and sublicense such enhancements or
[…]

### 2018-06-27 cpp_version/include/wasserstein/wasserstein_space_point.h

@@ -1,0 +1,548 @@
+#ifndef WASSERSTEIN_SPACE_POINT_H
+#define WASSERSTEIN_SPACE_POINT_H
+
+#include <unordered_map>
+#include <iostream>
+#include <utility>
+#include <memory>
+
+#include <boost/config.hpp>
+#include <boost/functional/hash.hpp>
+
+#include "spdlog/spdlog.h"
+
+#include "wasserstein/wasserstein.h"
+#include "bottleneck/bottleneck.h"
+#include "cover_tree/boost_graph_import.h"
+
+namespace spd = spdlog;
+
+constexpr int INVALID_ID = -1;
+
+// class Point must provide function distance
+template<class Real_ = double>
+class DynamicSpanner
+{
+public:
+
+// types
+    using Real = Real_;
+    using DiagramPointR = std::pair<Real, Real>;
+    using DiagramR = std::vector<DiagramPointR>;
+    using AuctionParamsR = typename hera::AuctionParams<Real>;
+    using MatrixR = typename std::vector<std::vector<Real>>;
+
+    struct WassersteinSpacePoint
+    {
+        // diagrams are stored without any diagonal m_points
+        // Hera will add projections itself
+        //
+        using Real = Real_;
+        using IdType = int;
+        //using CacheKeyType = std::pair<IdType, IdType>;
+        //using CacheType = std::unordered_map<CacheKeyType, Real, boost::hash<CacheKeyType>>;
+
+        DiagramR dgm;
+        IdType id;
+        std::string comment;
+
+        DynamicSpanner<Real>* dyn_spanner;
+
+        WassersteinSpacePoint() :
+                id(INVALID_ID)
+        {
+        }
+
+
+        WassersteinSpacePoint(const DiagramR& _dgm, int _id) :
+                dgm(_dgm), id(_id)
+        {
+        }
+
+        // read diagram from text file
+        WassersteinSpacePoint(const std::string& dgm_file, int _id) :
+                id(_id),
+                comment(dgm_file)
+        {
+            if (!hera::read_diagram_point_set(dgm_file, dgm)) {
+                throw std::runtime_error(std::string("Cannot read file ") + dgm_file);
+            }
+        }
+
+        std::string to_string() const
+        {
+            std::string result = "Point(id = ";
+            return result;
+        }
+
+        void print() const
+        {
+        };
+
+        bool operator==(const WassersteinSpacePoint& other) const
+        {
[…]

### 2018-06-27 cpp_version/src/cover_tree.cpp

@@ -1,0 +1,417 @@
+#include <boost/functional/hash.hpp>
+
+#include "cover_tree/cover_tree.h"
+
+namespace wasser_spanner {
+
+    CoverTree::Node::Node(int _point_idx, int _level) :
+            m_level(_level),
+            m_point_idx(_point_idx)
+    {
+        if (console) { console->debug("Created Node, level = {}, point_idx = {}", m_level, m_point_idx); }
+    }
+
+    bool operator==(const CoverTree::Node& a, const CoverTree::Node& b)
+    {
+        assert(a.m_point_idx != b.m_point_idx or (a.m_level == b.m_level and a.m_children == b.m_children));
+        return a.m_level == b.m_level and a.m_point_idx == b.m_point_idx;
+    }
+
+    int CoverTree::Node::get_num_children() const
+    {
+        int result = 0;
+        for (const auto& lev_ch_pair : m_children)
+            result += lev_ch_pair.second.size();
+        return result;
+    }
+
+    void CoverTree::add_child(Node* node, int level, int point_idx)
+    {
+        if (console) {
+            console->debug("Enter add child, level = {}, point_idx = {}, parent node = {}", level, point_idx,
+                           node->m_point_idx);
+        }
+        assert(m_point_to_node.find(point_idx) == m_point_to_node.end());
+        assert(node->get_level() > level);
+
+        node->m_children[level].emplace_back(new Node(point_idx, level));
+        m_point_to_node[point_idx] = node->m_children[level].back().get();
+    }
+
+
+    CoverTree::NodeVec CoverTree::Node::get_children(int level) const
+    {
+        std::vector<CoverTree::Node*> result;
+
+        auto find_res = m_children.find(level);
+        if (find_res != m_children.end()) {
+            for (const auto& c : find_res->second) {
+                result.push_back(c.get());
+            }
+        }
+
+        return result;
+    }
+
+    bool CoverTree::are_all_children_further(const Point& p,
+                                             const CoverTree::NodeVec& Qi,
+                                             double threshold,
+                                             int level)
+    {
+        if (console) { console->debug("Enter are_all_children_further, id = {}, threshold = {}", p.id, threshold); }
+
+        bool use_lazy = true;
+
+        if (use_lazy) {
+            bool lazy_failed = false;
+
+            for (const auto node : Qi) {
+                if (console) { console->debug("lazy considering node {}", *node); }
+
+                // node with point p has point p as a child - check separately
[…]

### 2018-06-27 cpp_version/src/experiment_helper.cpp

@@ -1,0 +1,731 @@
+#include <random>
+#include <fstream>
+#include <iostream>
+#include <dirent.h>
+#include <sys/stat.h>
+
+#include "boost/filesystem/path.hpp"
+#include "boost/filesystem/operations.hpp"
+
+#include "experiment_helper.h"
+
+namespace fs = boost::filesystem;
+
+namespace wasser_spanner {
+
+
+//bool get_random_diagram_UnifShear(DiagramR& diag,
+//                                 const int diagram_size,
+//                                 const double lower_bound_hor,
+//                                 const double upper_bound_hor,
+//                                 const double lower_bound_vert,
+//                                 const double upper_bound_vert)
+//{
+//    if (diagram_size <= 0) {
+//        std::cerr << "Error! diagram_size is not positive: " << diagram_size << std::endl;
+//        return false;
+//    }
+//    static std::uniform_real_distribution<double> unif_hor(lower_bound_hor, upper_bound_hor);
+//    static std::default_random_engine rand_engineHor;
+//    static std::uniform_real_distribution<double> unifVert(lower_bound_vert, upper_bound_vert);
+//    static std::default_random_engine rand_engineVert;
+//    diag.clear();
+//    diag.reserve(diagram_size);
+//
+//    for(int k = 0; k < diagram_size; ++k) {
+//        double rand_x = unif_hor(rand_engineHor);
+//        double rand_y = rand_x + unifVert(rand_engineVert);
+//        diag.push_back(Point(rand_x, rand_y));
+//    }
+//    return true;
+//}
+
+
+//bool getRandomDiagramUnif(DiagramR& diag,
+//                            const int diagram_size,
+//                            const double lower_bound_hor,
+//                            const double upper_bound_hor,
+//                            const double lower_bound_vert,
+//                            const double upper_bound_vert)
+//{
+//    if (diagram_size <= 0) {
+//        std::cerr << "Error! diagram_size is not positive: " << diagram_size << std::endl;
+//        return false;
+//    }
+//    static std::default_random_engine rand_engine;
+//    static std::uniform_real_distribution<double> unif_hor(lower_bound_hor, upper_bound_hor);
+//    static std::uniform_real_distribution<double> unifVert(lower_bound_vert, upper_bound_vert);
+//    //static std::default_random_engine rand_engineVert;
+//    diag.clear();
+//    diag.reserve(diagram_size);
+//
+//
+//    size_t pointsGenerated { 0 };
+//
+//    while(pointsGenerated < diagram_size) {
[…]

### 2018-06-27 cpp_version/src/main.cpp

@@ -1,0 +1,154 @@
+#include <iostream>
+
+//#define LOG_AUCTION
+
+#include "spdlog/spdlog.h"
+
+#include "wasserstein/wasserstein_space_point.h"
+#include "cover_tree/cover_tree.h"
+#include "cover_tree/wspd.h"
+#include "experiment_helper.h"
+
+std::mt19937_64 twister;
+
+namespace spd = spdlog;
+using namespace wasser_spanner;
+
+int main(int argc, char** argv)
+{
+    auto console = spd::stdout_color_mt("console");
+    console->set_level(spd::level::info);
+
+    DgmVec dgms;
+    MatrixR dist_matrix;
+    RandomDiagParams par;
+    double max_dist = -1.0;
+    double min_dist = std::numeric_limits<double>::max();
+
+//    par.auction_params.wasserstein_power = atof(argv[2]);
+//    create_dist_matr_from_dir(argv[1], par.auction_params, dist_matrix, max_dist, min_dist);
+//    return 0;
+
+    if (false) {
+
+        if (argc < 10) {
+            std::cout << "Usage: " << argv[0]
+                      << " dir_name wasser_power number_of_diagrams min_diagram_size max_diagram_size max_x std_dev std_dev_decay WSPD_epsilon [seed]"
+                      << std::endl;
+            return 0;
+        }
+
+        int arg_idx = 1;
+        std::string dir_name = argv[arg_idx++];
+        create_dir_if_not_exists(dir_name);
+
+        par.auction_params.wasserstein_power = atof(argv[arg_idx++]);
+        par.n_diagrams = atoi(argv[arg_idx++]);
+        par.min_diagram_size = atoi(argv[arg_idx++]);
+        par.max_diagram_size = atoi(argv[arg_idx++]);
+        par.upper_bound_hor = atof(argv[arg_idx++]);
+        par.std_dev = atof(argv[arg_idx++]);
+        par.std_dev_decay = atof(argv[arg_idx++]);
+        double eps = atof(argv[arg_idx++]);
+        par.random_generator_seed = (argc > arg_idx) ? atoi(argv[arg_idx++]) : 1;
+
+        par.strategy = RandomDiagParams::Strategy::CLUSTERED;
+
+        twister.seed(par.random_generator_seed);
+
+    }
+
+    if (argc < 5) {
+        std::cout << "Usage: " << argv[0]
+                  << " dir_name wasser_power WSPD_epsilon subsample_prob"
[…]

### 2018-06-27 cpp_version/src/wspd.cpp

@@ -1,0 +1,192 @@
+#include "cover_tree/wspd.h"
+
+namespace wasser_spanner {
+
+    WSPD::WSPD(CoverTree& ct, double eps) :
+            m_cover_tree(&ct),
+            m_num_points(ct.m_num_points),
+            m_epsilon(eps),
+            m_tau(ct.m_base),
+            m_spanner(&ct.m_dspanner),
+            m_rank(ct.m_num_points),
+            m_parent(ct.m_num_points),
+            m_conn_comps(&m_rank[0], &m_parent[0])
+    {
+        Node root_init { m_cover_tree->get_root() };
+        gen_wspd(m_wspd, root_init, root_init, 0);
+    }
+
+    void WSPD::gen_wspd(NodePairSet& result, const Node& a, const Node& b, int rec_depth)
+    {
+//        std::string prefix = "[%H:%M:%S.%e]";
+//        for(int i = 0; i < rec_depth; ++i) prefix += "  ";
+//        prefix += "%v";
+//        spd::set_pattern(prefix);
+
+        if (console) { console->debug("Entered gen_wspd, a = {}, b = {}, rec_depth = {}", a, b, rec_depth); }
+
+        if (a.is_leaf_node() and a.get_node_ptr() == b.get_node_ptr()) {
+            if (console) { console->info("a is leaf, a == b, return empty set. Exit"); }
+            return;
+        }
+
+        double a_delta = a.delta();
+        double b_delta = b.delta();
+        bool swap_needed = a_delta < b_delta or (a_delta == b_delta and a.get_point_idx() > b.get_point_idx());
+        if (swap_needed) {
+            if (console) {
+                console->debug("swap needed, a_delta = {}, b_delta = {}. Recursing, same depth", a_delta, b_delta);
+            }
+            gen_wspd(result, b, a, rec_depth);
+            return;
+        }
+
+        assert(a_delta >= b_delta);
+
+        double threshold = (1.0 + m_epsilon) * a_delta / m_epsilon;
+        if (m_spanner->is_distance_greater(a.get_point_idx(), b.get_point_idx(), threshold, true)) {
+            // TODO - remove debug code
+//            auto a_desc = a.get_all_descendants();
+//            auto b_desc = b.get_all_descendants();
+//            if (a_desc.count(2) == 1 and b_desc.count(3) == 1)
[…]

### 2018-06-27 cpp_version/src/wspd_node.cpp

@@ -1,0 +1,131 @@
+#include "cover_tree/wspd_node.h"
+
+namespace wasser_spanner {
+
+// WspdNode
+
+    DynamicSpannerR* WspdNode::dspanner = nullptr;
+
+    WspdNode::WspdNode(CoverTree::Node* node) :
+            m_ct_node(node),
+            m_level(node->get_level()),
+            m_point_idx(node->get_point_idx())
+    {
+    }
+
+    WspdNode::WspdNode(CoverTree::Node* node, int level) :
+            m_ct_node(node),
+            m_level(level),
+            m_point_idx(node->get_point_idx())
+    {
+        assert(level <= node->get_level());
+    }
+
+
+    void WspdNode::get_all_descendants_indices_helper(std::unordered_set<int>& res) const
+    {
+        res.insert(m_point_idx);
+        for (const auto& lev_ch_pair : m_ct_node->m_children) {
+            if (lev_ch_pair.first < get_level()) {
+                for (const auto& ch : lev_ch_pair.second) {
+                    WspdNode(ch.get()).get_all_descendants_indices_helper(res);
+                }
+            }
+        }
+    }
+
+    std::unordered_set<int> WspdNode::get_all_descendants() const
+    {
+        std::unordered_set<int> res;
+        get_all_descendants_indices_helper(res);
+        return res;
+    }
+
+    double WspdNode::diameter() const
+    {
+        double res = 0.0;
+        auto indices = get_all_descendants();
+        for (auto p_1 = indices.cbegin(); p_1 != indices.cend(); ++p_1) {
+            for (auto p_2 = std::next(p_1); p_2 != indices.cend(); ++p_2) {
+                res = std::max(res, dspanner->get_distance_no_cache(*p_1, *p_2));
+            }
+        }
+        return res;
+    }
+
+    std::vector<WspdNode> WspdNode::get_next_layer_children() const
+    {
+        std::vector<WspdNode> result;
+        result.push_back(get_self_as_child());
+        auto find_res = m_ct_node->m_children.find(get_level() - 1);
+        if (find_res != m_ct_node->m_children.cend()) {
+            for (const auto& c : find_res->second) {
+                result.push_back(c.get());
+            }
+        }
+        return result;
+    }
+
+    WspdNode WspdNode::get_self_as_child() const
+    {
+        return WspdNode(m_ct_node, get_level() - 1);
+    }
+
+    bool WspdNode::is_leaf_node() const
+    {
[…]

### 2018-10-17 paper/metric_space_exploration.tex
Copy proof from Bose et al for approximate greedy spanner

@@
 \renewcommand{\S}{\mathbb{S}}
+\newcommand{\dappr}{\tilde{d}}
 %\newcommand{\Wlog}{W.l.o.g.\ }
@@
 \newcommand{\level}{\ell}
-\newcommand{\diameter}{\mathrm{diam}}
+\newcommand{\diam}{\mathrm{diam}}
 \newcommand{\pointset}{P}
@@
 \subsection{WSPD}
+Given $s > 1$, two disjoint subsets $A, B$ of a metric space $(X, d)$ are called $s$-\textit{well-separated},
+if 
+\[
+\forall a \in A \, \forall b \in B \, d(a, b) \geq s \max(\diam(A), \diam(B))
+\]
+A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $X$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
+The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd}.
 
 \subsection{Spanners}
+
+\subsection{Approximate Greedy Spannner}
+
+Suppose that we have fast algorithms for computing $\lambda$-approximation of $d(x,y)$.
+In other words, for every pair of points $x,y$ we can compute an interval $[c, \lambda c]$ that is guaranteed to contain 
+the true distance $d(x,y)$.  Note that for every interval of this form there exists 
+a unique $k \in \{1, \dots, N-1 \}$ such that $[c, \lambda c] \subset [\lambda^k, \lambda^{k+2}]$. 
+Indeed, $k = \lceil \log_{\lambda}(c) \rceil$.
+We also assume that we know the range of all possible values of $k$ and they are all non-negative,
+$k = 1\dots N$. We call the intervals $[\lambda^k, \lambda^{k+2}]$ \textit{buckets}, even though they are overlapping.
+Now we can formulate the algorithm for finding a spanner.
+
+\begin{algorithmic}
+\label{alg:local_init}
+\Function{Approximate greedy spanner}{$rho$}
+    \ForAll {pairs $x,y \in X$}
+    \State {Place $x, y$ into corresponding bucket }
+    \EndFor
+\EndFunction
+\end{algorithmic}
+
+We can use essentially the same proof as in \cite{bose2010computing}.
+Claim 1. If $s > \lambda^2 $ and $A, B$ are $s$-separated, then all pairs $(a, a')$, $(b, b')$ with $a, a' \in A$
+and $b, b' \in B$ are processed before any of the pairs $(a, b)$.
+
+Indeed, if $s > \lambda^2$, then for every points $a_1, a_2 \in A$ and $b \in B$
+we have $\lambda^2 d(a_1, a_2) < d(a_1, b)$. If $\dappr(a_1, a_2)$ and $\dappr(a_1, b)$ denote the $\lambda$-approximations
+of the corresponding distances, then the inequality $\frac{\dappr(a_1, b)}{\dappr(a_1, a_2)} > \lambda$
+holds , hence $(a_1, a_2)$ will be placed in the bucket that precedes the bucket of $(a_1, b)$.
+
+Claim 2. If $s > \max(\lambda^2, \frac{2t+2}{t-1})$ and $A, B$ are two $s$-well-separated subsets of $X$,
+then in the approximate greedy spanner computed by \ref{alg:approx_greedy_spanner} there is at most one edge between $A$ and $B$.
+
+Suppose that the pair $(a_1, b_1)$ with $a_1 \in A$, $b_1 \in B$ was processed and the edge $(a, b)$
+was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
+Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
+to the spanner. By claim 1
+$d_G(a_2, a_1) \leq td(a_2, a_1)$ and $d_G(b_1, b_2) \leq t d(b_1, b_2)$.
+By properties of WSPD, $d_G(a_1, b_1) = d(a_1, b_1) \leq (1 + \frac{2}{s})d(a_2, b_2)$
+and $t d(a_2, a_1) \leq \frac{t}{s} d(a_2, b_2)$, $t d(b_1, b_2) \leq \frac{t}{s} d(a_2, b_2)$.
+
+Combining these inequalities, we get
+
+\begin{equation}
+    \begin{split}
+        d_G(a_2, b_2) & \leq  d_G(a_2, a_1) + d_G(a_1, b_1) + d_G(b_1, b_2)  \\
+                      & \leq  \frac{t}{s} d(a_2, b_2) + (1 + \frac{2}{s})d(a_2, b_2) + \frac{t}{s} d(a_2, b_2) \\
+                      & =     (1 + \frac{2t + 2}{s}) d(a_2, b_2)
+    \end{split}
+\end{equation}
+
+Solving for $s$ the inequality $1 + \frac{2t + 2}{s} < t$ gives $s > \frac{2t+2}{t-1}$.
 

### 2018-10-17 python_version/greedy_spanner.py
python script for approximate greedy spanner

@@ -1,0 +1,143 @@
+#!/usr/bin/env python3
+
+import networkx as nx
+import numpy as np
+import random
+import sys
+
+inf_dist = 10000000000
+
+dim = 2
+
+
+def get_points(n_points):
+    return np.random.randn(n_points, dim)
+
+
+def distance_matrix(points):
+    n_points = len(points)
+    result = np.zeros((n_points, n_points))
+    for i in range(n_points):
+        for j in range(n_points):
+            if i <= j:
+                continue
+            result[i][j] = result[j][i] = get_distance(points[i, :], points[j, :])
+    return result
+
+
+def distances_list(points):
+    n_points = len(points)
+    dm = distance_matrix(points)
+    return [(dm[i][j], i, j) for i in range(n_points) for j in range(i + 1, n_points)]
+
+
+def sorted_distances(points):
+    return sorted(distances_list(points))
+
+
+def upper_bound(g, i, j):
+    try:
+        return nx.shortest_path_length(g, i, j, weight='weight')
+    except nx.NetworkXNoPath:
+        return inf_dist
+
+
+def get_distance(a, b):
+    return np.linalg.norm(a - b)
+
+
+# def is_distance_larger(a, b, t):
+#     """Return true, if d(a, b) > t;
+#        return false, if d(a,b) < t/2; if d(a,b) is between t/2 and t,
+#        return either true or false"""
+#     d = get_distance(a, b)
+#     if d > t:
+#         return True
+#     elif d < t / 2:
+#         return False
+#     else:
+#         return np.random.uniform() < 0.5
+
+
+def get_buckets(max_dist, n_buckets):
+    scaling_factor = max_dist / (2 ** (n_buckets))
+    result = [(0.0, scaling_factor * 2.0)]
+    result.extend([(scaling_factor * (2 ** k), 4.0 * scaling_factor * (2 ** k)) for k in range(n_buckets - 1)])
+    return result
+
+
+def get_bucket_index(buckets, d):
+    dist = d[0]
+    cand_list = []
+    for i in range(len(buckets)):
+        lb, ub = buckets[i]
+        if dist >= lb and dist <= ub:
+            cand_list.append(i)
+    if len(cand_list) == 0:
+        b0 = buckets[0]
+        b1 = buckets[-1]
+        print(f"ACHTUNG! dist = {dist}, {b0} - {b1}  bucket not found")
+    assert (len(cand_list) > 0)
+    if len(cand_list) == 1:
[…]

### 2018-11-19 cpp_version/include/dynamic_spanner.h
Interim

@@ -6,6 +6,8 @@
 #include <vector>
 #include <limits>
 #include <algorithm>
+#include <math.h>
+#include <random>
 
 #include <cassert>
 
@@ -54,12 +56,18 @@ public:
 
     size_t m_num_points;
     Matrix m_matrix;
+    double m_min_dist;
+    double m_max_dist;
+    std::mt19937 m_twister;
 
 public:
 
     DynamicSpanner(const MatrixReal& distance_matrix)
+            :
+            m_min_dist(std::numeric_limits<double>::max()),
+            m_max_dist(0.0),
+            m_twister(std::random_device()())
     {
-
         m_num_points = distance_matrix.size();
         m_matrix.resize(m_num_points);
 
@@ -76,6 +84,69 @@ public:
             m_matrix[i][i].lower_bound = m_matrix[i][i].upper_bound = 0.0;
             m_matrix[i][i].exact_distance_used = true; // not sure that is ever queried
         }
+
+        for (int i = 0; i < m_num_points; ++i) {
+            for (int j = i + 1; j < m_num_points; ++j) {
+                m_max_dist = std::max(m_max_dist, distance_matrix[i][j]);
+                m_min_dist = std::min(m_min_dist, distance_matrix[i][j]);
+            }
+        }
+    }
+
+    std::vector<std::pair<double, double>> get_buckets()
+    {
+        int n_buckets = ceil(log2(m_max_dist / m_min_dist));
+        std::vector<std::pair<double, double>> result;
+        result.reserve(n_buckets);
+        double scaling_factor = m_min_dist;
+        for (int k = 0; k < n_buckets; ++k) {
+            result.emplace_back(scaling_factor * std::pow(2.0, k), 4.0 * scaling_factor * std::pow(2.0, k));
+        }
+        return result;
+    }
+
+    int get_bucket_index(const std::vector<std::pair<double, double>>& buckets, double value)
+    {
+        std::vector<int> cand_list;
+        for (int i = 0; i < buckets.size(); ++i) {
+            auto bucket = buckets[i];
+            if (value >= buckets[i].first and value <= buckets[i].second) {
+                cand_list.push_back(i);
+            }
+        }
+        assert(not cand_list.empty());
+        if (cand_list.size() == 1)
+            return cand_list[0];
+        else {
+            std::random_shuffle(cand_list.begin(), cand_list.end());
+            return cand_list[0];
+        }
+    }
+
+    decltype(auto) weakly_sorted_edges()
+    {
+        auto buckets = get_buckets();
[…]

### 2018-11-19 python_version/blind_spanner.py

@@ -218,13 +218,12 @@ class BlindSpanner:
                 break
             if go_up:
                 _, i, j = sd[idx_min]
-                idx_min = idx_min + 1
+                idx_min += 1
             else:
                 _, i, j = sd[idx_max]
-                idx_max = idx_max - 1
+                idx_max -= 1
             go_up = not go_up
             self.add_edge_to_spanner(i,j)
-
 
 
 def distance_matrix(points):
@@ -299,6 +298,7 @@ if __name__ == "__main__":
         for n_points in n_pointses
         for epsilon in epsilons
         for (ps_gen_method, ps_arg) in zip(ps_gen_methods, ps_gen_args))
+
     print("################################################################################")
 
     df_arg = [{"Dim": er.dim, "N_points": er.n_points,

### 2018-11-19 python_version/save_in_text_files.py

@@ -1,0 +1,32 @@
+#!/usr/bin/env python3
+
+import sys
+import numpy as np
+from utils_euclidean import *
+
+if __name__ == "__main__":
+    if len(sys.argv) < 5:
+        print("Parameters: point_method(unif or normal) dim n_points points_file [matrix_file]")
+        exit(1)
+    arg_idx = 1
+    point_method = sys.argv[arg_idx]
+    arg_idx += 1
+    dim = int(sys.argv[arg_idx])
+    arg_idx += 1
+    n_points = int(sys.argv[arg_idx])
+    arg_idx += 1
+    point_file_name = sys.argv[arg_idx]
+    arg_idx += 1
+    if len(sys.argv) == arg_idx:
+        matrix_file_name = "dist_matrix_" + point_file_name
+    else:
+        matrix_file_name = sys.argv[arg_idx]
+
+    pm_dic = { "unif" : get_uniform_points, "normal" : get_points }
+    if not point_method in pm_dic:
+        print("Unknown point_method")
+        exit(1)
+    points = pm_dic[point_method](n_points, dim)
+    np.savetxt(point_file_name, points, "%15.7f")
+    dist_matr = distance_matrix(points)
+    np.savetxt(matrix_file_name, dist_matr, "%15.7f")

### 2018-11-21 paper/draft.tex
Interim

--- 2018-10-17 paper/metric_space_exploration.tex
@@
 \usepackage{comment}
+\usepackage{braket}
 
@@
 \newcommand{\pointset}{P}
-\newcommand{\metricspace}{\mathcal{M}}
-\newcommand{\metric}{\delta}
-\newcommand{\approxmetric}{\gamma}
-\newcommand{\complexity}{C_{\metric}}
+\newcommand{\distspace}{\mathcal{M}}
+\newcommand{\dist}{\delta}
+\newcommand{\adist}{\gamma}
+\newcommand{\complexity}{C_{\dist}}
 \newcommand{\doublingdimension}{\Delta}
@@
 Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
-in a metric space $(\metricspace,\metric)$, 
+in a metric space $(\distspace,\dist)$, 
 consider the following standard operations:
@@
 \begin{description}
-\item[Nearest Neighbor] Given a point $q\in\metricspace$,
+\item[Nearest Neighbor] Given a point $q\in\distspace$,
 find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
@@
 
-\item[Metric approximation] Given $\eps>0$, compute a metric $\approxmetric$
+\item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
 such that for all $1\leq i,j\leq n$,
 %
-\[\metric(p_i,p_j)\leq\approxmetric(p_i,p_j)\leq (1+\eps)\metric(p_i,p_j).\]
+\[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]
 
-By ``computing'' $\approxmetric$, we mean to compute a $n\times n$-matrix $E$ 
-such that $E_{ij}=\approxmetric(p_i,p_j)$.
+By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
+such that $E_{ij}=\adist(p_i,p_j)$.
 \end{description}
 
-Assume that an algorithm is given to compute $\metric(\cdot,\cdot)$
-for any two points in $\metricspace$ in worst-case complexity $\complexity$.
+Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
+for any two points in $\distspace$ in worst-case complexity $\complexity$.
 Then trivially, the above problems can be solved (exactly) in
@@
 pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
-induced by $\metric$ in $O(\complexity n\log n)$ time and
+induced by $\dist$ in $O(\complexity n\log n)$ time and
 of linear size (assuming, for simplicity, that $\eps$ is a constant).
@@
 There are good reasons for that: the most common case of a metric space
-is $\metricspace=\R^d$ with $d$ some constant, in which case $\metric$ can be evaluated in $O(d)=O(1)$ time.
+is $\distspace=\R^d$ with $d$ some constant, in which case $\dist$ can be evaluated in $O(d)=O(1)$ time.
 Even if $d$ is considered non-constant,
@@
 However, we argue that in some situations, distance computations
-in $\metricspace$ can be costly and $\complexity$ might be incomparable
+in $\distspace$ can be costly and $\complexity$ might be incomparable
 with $n$. Our motivating examples come from topological summaries
@@
 computations only in the form of the following predicate: 
-``Is $\metric(p_i,p_j)\leq t$'', where $t$ is a real value.
+``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
 We show that using the information of the already computed distances and
@@
 we see a weighted graph on $P$ whose edges are the distances that have
-been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
+been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
 The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
 constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\metric(p_1,p_2)\geq 3$:
+However, we can also infer that $\dist(p_1,p_2)\geq 3$:
 otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
@@
 triangle inequality.
-Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
+Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
 for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
@@
 pairwise distances in an $n\times n$ matrix of intervals. We show how to
-update this matrix in $O(n^2)$ time when a new edge is added.
+update this matrix in $O(n^4)$ time when a new edge is added.
 %We point out once more that a quadratic update time only pays off
@@
 
-We also provide experimental results for the case of persistence diagrams
+We also provide some experimental results for the case of persistence diagrams
 which motivated this line of research. We point out that the space of
@@
 
+\section{Metric Approximation}
+
+It is known that asymptotically optimal spanner can be constructed by the following greedy algorithm: 
+start with an empty graph on vertices $p_i$,
+sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
+until it becomes a $1+\eps$-spanner. In our setting this is the most expensive approach, because it requires
+computing all pairwise distances first.
+
+ In our setting this is the most expensive approach, because it requires
+ computing all pairwise distances first. We propose an alternative approach, which we call
+ \textit{blind} spanner, meaning that the algorithm does not have access to $\dist(p_i, p_j)$
+ before it decides to add the edge $(p_i, p_j)$. 
+ The algorithm also maintains lower and upper bounds
+ for all pairwise distances, which are updated after a new edge is added. and the stopping criterion
+ is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
+ for the distance between $p_i$ and $p_j$, respectively, then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
+ for all $i \neq j$.
+
+\begin{algorithmic}
+\label{alg:blind_spanner}
+\Function{BlindSpanner}{$P, \eps$}
+    \State {$E \gets \emptyset$}
+    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
+    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
+    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
+    \State {$(i,j) \gets GetEdgeToAdd$}
+    \State {$v \gets \dist(p_i, pj)$}
+    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
+    \State {$UpdateBounds(i, j, v)$}
+    \EndWhile
+\EndFunction
+\end{algorithmic}
+
+In this pseudocode we adopt the convention $ \textrm{positive number} / 0 = \infty$ and $\infty > \textrm{any real number}$,
+thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection, 
+now we only mentioning that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
+and the update is monotonic: no upper bound can increase, no lower bound can decrease (in particular, $a_{i,j} = b_{i,j} = v$).
+The most interesting detail of this algorithm is the way of choosing the next edge to be added, and here 
+there are several possibilities, which give very different results.
+
+The first possibility will be called $BlindGreedy$, where we choose a pair $p_i, p_j$ that maximizes
+the ratio $b_{i,j} / a_{i,j}$. If the number of maximizers is more than 1, we choose one of them at random.
+Note that our conventions imply that in $BlindGreedy$ the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
+have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
+lower bounds for every edge before it will start adding any other edges.
+
+The second possibility will be referred to as $BlindRandom$, where among all edges $(p_i, p_j)$
+for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach
+
+ \paragraph{Maintaining lower and upper bounds}
+ More specifically,
+ we maintain an $n\times n$ matrix $M$ whose entries
+are intervals $M_{ij}=[a_{ij},b_{ij}]$, meaning that the distance
+between $p_i$ and $p_j$ lies in that interval. 
+We initialize $M_{ij}=[0,\infty]$ for all $i\neq j$
+and $M_{ij}=[0,0]$ for $i=j$.  Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
+First, we reset $M_{i,j}$ and $M_{j,i}$ to $[v,v]$. 
+To update the upper bound of some entry $M_{k,\ell}$,
+we observe that the shortest path from $p_k$ to $p_\ell$ might now
+go through the new edge. Hence, we update
+%
+\[b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}\]
+%
+Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
+time per entry.
+
+For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
+%
+\[v-b_{k,i}-b_{\ell,j}\]
+%
+is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
+the triangle inequality
+%
+\[\dist(p_i,p_j)\leq \dist(p_i,p_k)+\dist(p_k,p_\ell)+\dist(p_\ell,p_j)\]
+by rearranging terms and plugging in the upper bounds for $\dist(p_i,p_k)$
+and $\dist(p_k,p_\ell)$. An analogue bound holds with $i$ and $j$ swapped. Hence, we update
+%
+\[a_{k,\ell}\gets \max\{a_{k,\ell},v-b_{k,i}-b_{\ell,j},v-b_{k,j}-b_{\ell,i}\}\]
+%
+for all $k,\ell$ and obtain the updated $M$ in $O(n^4)$ time.
+
+\section{Approximate Nearest Neighbor}
+
+In this section we consider the standard problem of finding an approximate nearest neighbor: given
+$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
+find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. We assume for simplicity
+that all exact pairwise distances $\dist(p_i, p_j)$ are already computed.
+
+\begin{lemma}
+    Let $p_i, p_j \in P$ be such that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$,
+    and denote $r = \dist(p_i, q) / (1 + \eps)$, $r_j = \dist(p_j, q)$.
+    Then for every $x \in B(p_j, r - r_j)$ we have $\dist(x, q) \geq r$
+    and for every $y \notin B(p_j, r + r_j)$ we have $\dist(y, q) \geq r$.
+\end{lemma}
+Proof. The triangle inequality.
+
+\begin{lemma}
+\end{lemma}
+
+The algorithm can be informally described as follows. Pick a random candidate $p_i$ and try to verify that
+it is an approximate nearest neighbor by scanning all other points. Pick the next random point $p_j$
+and assume that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$. This allows us to discard not just the point
+$p_j$, but also the points that are very close to $p_j$ and very far from $p_j$ using triangle inequality.
+Indeed, consider the ball of radius $ \dist(p_j, q) - \frac{ \dist(p_i, q) }{ 1+\eps} $ centered at $p_j$.
+The triangle inequality implies that all points in this ball are at the distance at least $\frac{ \dist(p_i, q) }{ 1+\eps}$
+from $q$. Now consider the ball of radius $\dist(p_j, q) + 2 \frac{\dist(p_i, q)}{1+ \eps}$ centered at $p_j$. 
+Every point outside of this ball is a
+
+The algorithm that we analyze is based on the following simple observation.
+Suppose that we want to to prove that a candidate for the approximate nearest neighbor is $p_1$,
+
+\begin{algorithmic}
+\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
+\For {$i \in \{\,1, \dots, |P|\,\}$}
+\State {$a_i \gets 0$}
+\State {$b_i \gets \infty$}
+\EndFor
+\State {$i_0 \gets \argmin u $}
+\State {$B  \gets \{\, i \mid l_i < u[i_0]\,\}$}
+\If {$B = \emptyset$}
+    \State {\Return {$p_{i_0}$}}
+\Else
+    \State{$j \gets \mbox{ random element of } C$}
+    \State{$u_j \gets d(p_j, q)$}
+    \State{$l[j] \gets d(p_j, q)$}
+    \State{Mark distance from $q$ to $p_j$ as computed}
+    \State{\Call{UpdateBounds}{$l, u, d(p_j, q), j$}}
+\EndIf
+\EndFunction
+\end{algorithmic}
+
+The goal is to compute
+as few $\dist(q, p_k)$ as possible.
+
 \section{Geometric concepts}
@@
 We will also review the computation methods and how they depend
-on the predicate ``$\metric(p_i,p_j)\leq t$?'' mentioned above.
+on the predicate ``$\dist(p_i,p_j)\leq t$?'' mentioned above.
 
@@
 
-\section{Distance Bounds in Partially Revealed Metrics}
-%
-We describe our algorithm to maintain lower and upper bounds for the
-distance of any two points in our (finite) metric space $(P=\{p_1,\ldots,p_n\},\metric)$. We assume for simplicity that all points in $P$ are distinct.
-
-We maintain an $n\times n$ matrix $M$ whose entries
-are intervals $M_{ij}=[a_{ij},b_{ij}]$, meaning that the distance
-of $p_i$ and $p_j$ lies in that interval. 
-We initialize $M_{ij}=[0,\infty]$ for all $i\neq j$
-and $M_{ij}=[0,0]$ for $i=j$.
-We describe next how to update $M$ when a new distance has been computed.
-
-Note that an explicit graph representation of the partially revealed metric,
-such as in Figure~\ref{fig:1st_example} is not necessary. However, for 
-the sake of clarity,
-this graph is helpful. So, let $G=(P,E)$ be the undirected weighted graph
-where an edge between $p_i$ and $p_j$ with weight $\metric(p_i,p_j)$
-exists if that distance has been computed. 
-We maintain the invariant that the upper bound $b_{ij}$ equals the 
-shortest path from $p_i$ to $p_j$ in $G$.
-
-Assume that $\metric(p_i,p_j)=v\in\R$ has been computed during the algorithm.
-First, we reset $M_{i,j}$ and $M_{j,i}$ to $[v,v]$. 
-In $G$, the edge $(p_i,p_j)$ is added with weight $v$.
-To update the upper bound of some entry $M_{k,\ell}$,
-we observe that the shortest path from $p_k$ to $p_\ell$ might now
-go through the new edge. Hence, we update
-%
-\[b_{k,\ell}\gets \min\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}\]
-%
-Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
-time.
-
-For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
-%
-\[v-b_{k,i}-b_{\ell,j}\]
-%
-is a lower bound for $\metric(p_k,p_\ell)$. Indeed, this follows from
-the triangle inequality
-%
-\[\metric(p_i,p_j)\leq \metric(p_i,p_k)+\metric(p_k,p_\ell)+\metric(p_\ell,p_j)\]
-by rearranging terms and plugging in the upper bounds for $\metric(p_i,p_k)$
-and $\metric(p_k,p_\ell)$. An analogue bound holds for the roles of $i$
-and $j$ swapped. Hence, we update
-%
-\[a_{k,\ell}\gets \max\{a_{k,\ell},v-b_{k,i}-b_{\ell,j},v-b_{k,j}-b_{\ell,i}\}\]
-%
-for all $k,\ell$ and obtain the updated $M$ in $O(n^2)$ time.
-
-\medskip
-
-With the matrix $M$, we can define filtered versions of our geometric
-algorithms very easily: whenever the algorithm makes a query of the form
-``Is $\metric(p_i,p_j)\leq t$?'', it is checked whether $a_{ij}> t$,
-in which case the answer is ``yes'', or whether $b_{ij}\leq t$,
-in which case the answer is ``no''. If none of the cases applies, 
-$v\gets\metric(p_i,p_j)$ is computed, 
-the matrix $M$ is updated as described above, and the predicate is answered
-by comparing $v$ and $t$.
-
-\section{Experiments on Point Clouds}
-\label{sec:experiments_point_cloud}
-
-\section{Metric spaces of Topological Summaries}
-\label{sec:metric_spaces_top_summaries}
-
-\section{Experiments on Persistence Diagrams}
-\label{sec:experiments_persistence_diagrams}
-
 \section{Conclusion}

### 2018-11-21 cpp_version/include/dynamic_spanner.h

@@ -59,6 +59,8 @@ public:
     double m_min_dist;
     double m_max_dist;
     std::mt19937 m_twister;
+    double m_epsilon { 0.0 };
+    std::string m_strategy;
 
 public:
 
@@ -454,6 +456,15 @@ public:
 
     }
 
+    std::string get_statistics() const
+    {
+        double n_pairs = m_num_points * (m_num_points - 1) / 2;
+        double distances_computed = get_number_of_computed_distances();
+        double sparseness_1 = distances_computed / n_pairs;
+        double sparseness_2 = distances_computed / m_num_points;
+        return fmt::format("{};{};{};{};{};{}", m_strategy, m_epsilon, m_num_points, distances_computed, sparseness_1, sparseness_2);
+    }
+
     bool find_random_bad_ratio(int& i, int& j, bool connect_first, bool lower_bound_first, double epsilon)
     {
         std::vector<std::pair<size_t, size_t>> candidates;
@@ -552,6 +563,8 @@ public:
 
     void construct_blind_greedy_eps_spanner(double eps)
     {
+        m_epsilon = eps;
+        m_strategy = "blind-greedy";
         size_t i, j;
         double ratio;
 
@@ -565,29 +578,38 @@ public:
 
     void construct_blind_random_eps_spanner(double eps)
     {
+        m_epsilon = eps;
+        m_strategy = "blind-random";
+
         size_t i, j;
         double ratio;
 
         ratio = find_worst_ratio();
 
-        std::vector<Value_with_index> vec;
+        std::vector<std::pair<size_t, size_t>> vec;
 
         for (size_t i = 0; i < m_num_points; i++) {
             for (size_t j = i + 1; j < m_num_points; j++) {
-                vec.push_back(Value_with_index(m_matrix[i][j].distance, i, j));
+                vec.emplace_back(i, j);
             }
         }
 
         std::random_shuffle(vec.begin(), vec.end());
+        auto vec_iter = vec.begin();
 
         while (ratio > (1 + eps)) {
+            std::tie(i, j) = *vec_iter++;
             get_distance(i, j);
+            std::cout << "adding i = " << i << ", j = " << j << std::endl;
             ratio = find_worst_ratio();
         }
     }
 
     void construct_blind_quasi_sorted_greedy_eps_spanner(double eps)
     {
+        m_epsilon = eps;
+        m_strategy = "blind-quasi-sorted-greedy";
+
[…]

### 2018-11-21 cpp_version/src/main.cpp

@@ -184,8 +184,9 @@ MatrixR get_distance_matrix(const arma::mat& points)
 int main(int argc, char** argv)
 {
     auto console = spd::stdout_color_mt("console");
+    auto exp_logger = spd::basic_logger_st("experiment_logger", "experiment_log-blind-greedy.txt");
     console->set_level(spd::level::info);
-
+    exp_logger->set_pattern("%v");
 
 #if 0
     if (argc < 4) {
@@ -313,68 +314,66 @@ int main(int argc, char** argv)
 
 #else
 
-
+//    {
+//        DynamicSpannerR greedy_spanner(dist_matrix);
+//        greedy_spanner.construct_greedy_eps_spanner(eps);
+//        console->info("{};{}", dist_name, greedy_spanner.get_statistics());
+//        exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
+//    }
+//
+//    {
+//        DynamicSpannerR quasi_greedy_spanner(dist_matrix);
+//        quasi_greedy_spanner.construct_blind_quasi_sorted_greedy_eps_spanner(eps);
+//        console->info("{};{}", dist_name, quasi_greedy_spanner.get_statistics());
+//        exp_logger->info("{};{}", dist_name, quasi_greedy_spanner.get_statistics());
+//    }
+//
+//    {
+//        DynamicSpannerR quasi_shaker_spanner(dist_matrix);
+//        quasi_shaker_spanner.construct_blind_quasi_sorted_shaker_eps_spanner(eps);
+//        exp_logger->info("{};{}", dist_name, quasi_shaker_spanner.get_statistics());
+//    }
+//
+//    {
+//        DynamicSpannerR blind_random_spanner_1_1(dist_matrix);
+//        blind_random_spanner_1_1.construct_blind_random_ratio_eps_spanner(true, true, eps);
+//        console->info("{};{}", dist_name, blind_random_spanner_1_1.get_statistics());
+//        exp_logger->info("{};{}", dist_name, blind_random_spanner_1_1.get_statistics());
+//    }
+//
+//    {
+//        DynamicSpannerR blind_random_spanner_0_0(dist_matrix);
+//        blind_random_spanner_0_0.construct_blind_random_ratio_eps_spanner(false, false, eps);
+//        console->info("{};{}", dist_name, blind_random_spanner_0_0.get_statistics());
[…]

### 2018-11-21 python_version/save_in_text_files.py

@@ -4,9 +4,35 @@ import sys
 import numpy as np
 from utils_euclidean import *
 
+def save_to_files(point_method, dim, n_points, point_file_name, matrix_file_name = None):
+    if matrix_file_name == None:
+        matrix_file_name = "dist_matrix_" + point_file_name
+    pm_dic = {"unif": get_uniform_points,
+              "normal": get_points,
+              "clustered": get_clustered_points,
+              "exp": get_exponential_points}
+    if not point_method in pm_dic:
+        print("Unknown point_method")
+        exit(1)
+    points = pm_dic[point_method](n_points, dim)
+    np.savetxt(point_file_name, points, "%15.7f")
+    dist_matr = distance_matrix(points)
+    np.savetxt(matrix_file_name, dist_matr, "%15.7f")
+
+
 if __name__ == "__main__":
+
+
+    for n_points in range(50, 850, 50):
+        for dim in [2, 3, 10]:
+            for point_method in ["unif", "normal", "exp", "clustered"]:
+                point_file_name = "points_{}_{}_{}.txt".format(dim, n_points, point_method)
+                np.random.seed(1)
+                save_to_files(point_method, dim, n_points, point_file_name)
+    sys.exit(0)
+
     if len(sys.argv) < 5:
-        print("Parameters: point_method(unif or normal) dim n_points points_file [matrix_file]")
+        print("Parameters: point_method(unif, normal, exp, clustered) dim n_points points_file [matrix_file]")
         exit(1)
     arg_idx = 1
     point_method = sys.argv[arg_idx]
@@ -18,15 +44,10 @@ if __name__ == "__main__":
     point_file_name = sys.argv[arg_idx]
     arg_idx += 1
     if len(sys.argv) == arg_idx:
-        matrix_file_name = "dist_matrix_" + point_file_name
+        matrix_file_name = ""
     else:
         matrix_file_name = sys.argv[arg_idx]
 
-    pm_dic = { "unif" : get_uniform_points, "normal" : get_points }
-    if not point_method in pm_dic:
-        print("Unknown point_method")
-        exit(1)
-    points = pm_dic[point_method](n_points, dim)
-    np.savetxt(point_file_name, points, "%15.7f")
-    dist_matr = distance_matrix(points)
-    np.savetxt(matrix_file_name, dist_matr, "%15.7f")
+    save_to_files(point_method, dim, n_points, point_file_name, matrix_file_name)
+
+

### 2018-11-21 python_version/utils_euclidean.py

@@ -22,6 +22,22 @@ def get_exponential_points(n_points, dim):
     return np.power(2.0, res)
 
 
+def get_clustered_points(n_points, dim, n_clusters = -1):
+    if n_clusters == -1:
+        n_clusters = max(5, int(n_points / 50))
+    assert(n_points % n_clusters == 0)
+    points_per_cluster = int(n_points / n_clusters)
+    a = get_uniform_points(n_clusters, dim, 10000.0)
+    cluster_centres = np.repeat(a, points_per_cluster, axis = 0)
+    # print("********************************************************************************")
+    # print(a)
+    # print("********************************************************************************")
+    # print(cluster_centres)
+    # print("********************************************************************************")
+    diffs = np.random.normal(scale = 10.0, size = (n_points, dim))
+    return diffs + cluster_centres
+
+
 def distance_matrix(points):
     n_points = len(points)
     result = np.zeros((n_points, n_points))

### 2018-11-22 paper/draft.tex
Interim

@@
-\documentclass[10pt]{article}
-\usepackage[margin=26mm]{geometry}
+\documentclass[a4paper,USenglish]{socg-lipics-v2018}
 \usepackage{amsthm}
@@
 \usepackage{graphicx}
-%\usepackage{subfigure}
 \usepackage{algorithm}
@@
 \usepackage{color}
-
-\usepackage{subfig}
 
@@
 
-\newcommand{\remark}[1]{\textbf{[#1]}}
+%\newcommand{\remark}[1]{\textbf{[#1]}}
 
@@
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{definition}[theorem]{Definition}
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{proposition}[theorem]{Proposition}
+%\newtheorem{corollary}[theorem]{Corollary}
+%\newtheorem{definition}[theorem]{Definition}
 %\newtheorem{remark}[theorem]{Remark}
-\newtheorem{note}[theorem]{Note}
-\newtheorem{problem}{Problem}
+%\newtheorem{note}[theorem]{Note}
 
@@
 
-\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
-\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
+%\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
+%\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
 
 \title{Geometric Filters for Metric Spaces}
-\author{Michael Kerber, Arnur Nigmetov}
-%
-\date{\today}
+\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
+\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
 
@@
 We perform an extensive experimental evaluation of our idea.
-In the first part, we concentrate on point clouds in Euclidean space~--
+We mostly concentrate on point clouds in Euclidean space~--
 of course, the approach does not yield any performance benefits in that
@@
 
-We also have derived two minor technical results necessary for 
-our implementation. First of all, since our underlying distance computation
-algorithm for persistence diagrams only returns approximate answers,
-we had to adapt all algorithms to cope with approximate distances.
-Moreover, we show how a WSPDs can be constructed
-out of cover trees (we only found 
-the theoretically more efficient version of
-net trees used in the literature, but net trees are more difficult to implement
-efficiently). Both results can be obtained using standard methods, 
-but seem to be not present in prior work.
+%We also have derived two minor technical results necessary for 
+%our implementation. First of all, since our underlying distance computation
+%algorithm for persistence diagrams only returns approximate answers,
+%we had to adapt all algorithms to cope with approximate distances.
+%Moreover, we show how a WSPDs can be constructed
+%out of cover trees (we only found 
+%the theoretically more efficient version of
+%net trees used in the literature, but net trees are more difficult to implement
+%efficiently). Both results can be obtained using standard methods, 
+%but seem to be not present in prior work.
+
+\section{Notation and Preliminary Notions}
+
+A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
+if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
+and $k$ is the smallest number having that property.
+\textit{Doubling dimension} of a doubling space is defined as $\log k$
+(since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
+we always use $\log$ to denote the logarithm with base 2).
+It is easy to see that a subspace of a space with doubling dimension $d$ 
+is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).
+
+In this paper we consider some fixed metric space $(X, \dist)$, and, to get interesting
+results we must assume doubling property.
+
+We shall need the following lemma.
+
+\begin{lemma}
+\label{lem:packing_lemma_dd}
+Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
+Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $. 
+\end{lemma}
+
+\begin{proof}
+We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
+balls of radius $R/4$, etc. Repeating this process $\lceil \log \frac{R}{r/2} \rceil = m $ times, 
+we cover
+$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can contain at most one point from $P$, the claim follows.
+\end{proof}
 
@@
 computing all pairwise distances first.
+the true distance belongs.
 
  In our setting this is the most expensive approach, because it requires
- computing all pairwise distances first. We propose an alternative approach, which we call
- \textit{blind} spanner, meaning that the algorithm does not have access to $\dist(p_i, p_j)$
+ computing all pairwise distances first. We propose an alternative approach (or rather a family of algorithms), which we call
+ \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
  before it decides to add the edge $(p_i, p_j)$.
@@
 
-In this pseudocode we adopt the convention $ \textrm{positive number} / 0 = \infty$ and $\infty > \textrm{any real number}$,
+In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
+and $\infty$ is larger than any real number,
 thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection,
-now we only mentioning that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
-and the update is monotonic: no upper bound can increase, no lower bound can decrease (in particular, $a_{i,j} = b_{i,j} = v$).
-The most interesting detail of this algorithm is the way of choosing the next edge to be added, and here 
+now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
+and the update is monotonic: no upper bound can increase, no lower bound can decrease, and, after the distance
+between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
+The most important part of this algorithm is the way of choosing the next edge to be added, and
 there are several possibilities, which give very different results.
 
-The first possibility will be called $BlindGreedy$, where we choose a pair $p_i, p_j$ that maximizes
-the ratio $b_{i,j} / a_{i,j}$. If the number of maximizers is more than 1, we choose one of them at random.
-Note that our conventions imply that in $BlindGreedy$ the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
+The first possibility will be called \textsc{BlindGreedy}, where we choose a pair $(p_i, p_j)$ that maximizes
+the ratio $b_{i,j} / a_{i,j}$. If the number of maximizing pairs is more than 1, we choose one of them at random.
+Note that our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
 have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
@@
 
-The second possibility will be referred to as $BlindRandom$, where among all edges $(p_i, p_j)$
+The second possibility will be referred to as \textsc{BlindRandom}, where among all edges $(p_i, p_j)$
 for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach
+by insisting that we must first create a connected graph, or that we must first have some positive lower bound
+for all edges, and only after these requirements are satisfied, we consider other pairs.
 
- \paragraph{Maintaining lower and upper bounds}
+The next option is to make an additional assumption: while
+computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a cheap approximation algorithm
+that gives, say, a $2$-approximation of $\dist$ much faster (the output of this approximation algorithm is guaranteed to
+lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
+that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
+We now introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
+these intervals are not disjoint, but, by calling our algorithm we can assign
+each pair $(p_i, p_j)$ to one of the buckets.
+Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
+then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
+that fully contains this interval, thus we need to find $k$ such that
+$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
+and we take $k = \lfloor \log v \rfloor - 1$.
+If we find a bucket for each pair $(p_i, p_j)$, we have some sort of a weak ordering: we cannot say anything
+about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
+are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
+traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
+we travers the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
+we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
+order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
+in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
+We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.
+
+ \paragraph{Maintaining Lower and Upper Bounds}
  More specifically,
@@
 
+\paragraph{Experimental Results}
+
 \section{Approximate Nearest Neighbor}
@@
 find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. We assume for simplicity
-that all exact pairwise distances $\dist(p_i, p_j)$ are already computed.
+that all exact pairwise distances $\dist(p_i, p_j)$ are already computed. Our goal is to reduce the number
+of computed distances $d(p_i, q)$.
+
+The algorithm that we analyze can be roughly described as follows. Pick a random point $p_i$,
+compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
+to update lower bounds for other distances $\dist(p_j, q)$.
+If we see that the lower bound for $\dist(p_j, q)$ is larger than $\dist(p_i, q) / (1+\eps)$, then 
+we can discard the point $p_j$. If all points were discarded, we found an approximate nearest neighbor;
+otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
+There are two possibilities: if $\dist(p_j, q) \leq \dist(p_i, q) / (1+\eps)$,
+then we choose $p_j$ as the current candidate for an approximate nearest neighbor,
+otherwise we keep $p_i$ in this role. In any case, the new computed distance
+may also allow us to discard more than one point, as figure \ref{fig:discarded_balls} shows. In
+this figure points inside the shaded smaller ball cannot be a better candidate,
+and the same is true for the points in the shaded area outside of the larger ball.
+The precise radii of these balls are given in the following lemma.
 
@@
     Let $p_i, p_j \in P$ be such that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$,
-    and denote $r = \dist(p_i, q) / (1 + \eps)$, $r_j = \dist(p_j, q)$.
-    Then for every $x \in B(p_j, r - r_j)$ we have $\dist(x, q) \geq r$
-    and for every $y \notin B(p_j, r + r_j)$ we have $\dist(y, q) \geq r$.
+    and denote $r_i = \dist(p_i, q)$, $r_j = \dist(p_j, q)$.
+    Then for every $x \in B(p_j, r_j - r_i / (1+\eps))$ we have $\dist(x, q) \geq r$
+    and for every $y \notin B(p_j, r_j + r_i / (1+ \eps))$ we have $\dist(y, q) \geq r$,
+    so all points of $P$ inside the smaller ball and all points of $P$ outside the larger ball
+    can be discarded.
 \end{lemma}
-Proof. The triangle inequality.
+\begin{proof}
+    The triangle inequality.
+\end{proof}
 
-\begin{lemma}
-\end{lemma}
-
-The algorithm can be informally described as follows. Pick a random candidate $p_i$ and try to verify that
-it is an approximate nearest neighbor by scanning all other points. Pick the next random point $p_j$
-and assume that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$. This allows us to discard not just the point
-$p_j$, but also the points that are very close to $p_j$ and very far from $p_j$ using triangle inequality.
-Indeed, consider the ball of radius $ \dist(p_j, q) - \frac{ \dist(p_i, q) }{ 1+\eps} $ centered at $p_j$.
-The triangle inequality implies that all points in this ball are at the distance at least $\frac{ \dist(p_i, q) }{ 1+\eps}$
-from $q$. Now consider the ball of radius $\dist(p_j, q) + 2 \frac{\dist(p_i, q)}{1+ \eps}$ centered at $p_j$. 
-Every point outside of this ball is a
-
-The algorithm that we analyze is based on the following simple observation.
-Suppose that we want to to prove that a candidate for the approximate nearest neighbor is $p_1$,
+We can hope to avoid computation of $O(n)$ distances, if we assume low doubling dimension
+and solve approximate neighbor problem. The necessity of the first assumption is easily justified: if all $d(p_i, p_j)$
+and $d(p_i, q)$ are approximately the same, at each iteration of our algorithm we will be able to discard only one point.
+The necessity of the second requirement can be seen from the figure \ref{fig:bad_exact_nn_example},
+where a set of points in the plane is in general position, and yet, regardless of the order we choose them,
+at each iteration of our algorithm we are able to discard only one point. This example does not work in the
+approximate formulation.
 
@@
 \EndFor
-\State {$i_0 \gets \argmin u $}
+%\State {$i_0 \gets \argmin u $}
 \State {$B  \gets \{\, i \mid l_i < u[i_0]\,\}$}
@@
 
-The goal is to compute
-as few $\dist(q, p_k)$ as possible.
+\begin{theorem}
+    If $(X, \dist)$ is a doubling space, then, for any fixed $\eps > 0$ the
+    algorithm \ref{alg:ann_blind} computes $O(\log n)$ distances $\dist(p_i, q)$ in expectation.
+\end{theorem}
 
-\section{Geometric concepts}
+In order to prove this theorem, we can think of the algorithm as
 
-We revise the concepts that were designed for nearest neighbor queries
-and metric approximation in metric spaces with low dimensionality.
-We will also review the computation methods and how they depend
-on the predicate ``$\dist(p_i,p_j)\leq t$?'' mentioned above.
+\begin{lemma}
+\end{lemma}
 
-\paragraph{Notion of Dimensionality.}
-%
-Exansion constant, doubling dimension.
+\begin{lemma}
+Consider the following random procedure: we start with $n$ numbers $[n] = \{\,1,\dots, n\,\}$ and 
+choose $\xi_1$ uniformly at random from $[n]$. If $\xi_1 = n$, then we stop,
+otherwise we remove $\xi_1$ and all numbers preceding it and from $\{\,\xi_1, \dots, n\,\}$
+we choose $\xi_2$. If $\xi_2 = n$, we stop, if $\xi_2 < n$, we repeat the process removing all numbers not greater than $\xi_2$ and
+drawing $\xi_3$ from the remaining numbers uniformly at random. The expected number of steps of this procedure
+is $O(\log n)$.
+\end{lemma}
+\begin{proof}
+We can rephrase the procedure as a Markov chain with set of states $[n]$ and transition matrix
+\begin{equation}
+    P = \begin{pmatrix}
+         0      & \frac{1}{n-1} & \frac{1}{n-1} & \dots  & \frac{1}{n-1} & \frac{1}{n-1} \\
+         0      &  0            & \frac{1}{n-2} & \dots  & \frac{1}{n-2} & \frac{1}{n-2} \\
+         0      &  0            & 0             & \dots  & \frac{1}{n-3} & \frac{1}{n-3} \\
+         \vdots &  \vdots       & \vdots        & \ddots & \vdots        & \vdots \\
+         0      &  0            & 0             & \dots  & \frac{1}{2} & \frac{1}{2} \\
+         0      &  0            & 0             & \dots  & 0           & 1           \\
+        \end{pmatrix};
+\end{equation}
+the chain starts from state $1$.
+Note that there is one absorbing state $n$, and all other states are transient, and the matrix is already in the canonical
+form, meaning that all transient states precede all absorbing states. 
+We denote the square submatrix formed by the first $n-1$
+rows and $n-1$ columns by $Q$, it is the submatrix corresponding to all transient states. It is known
+(\cite{absorbing_markov_chain}) that the matrix $I-Q$ is invertible, let $N$ be its inverse,
+\begin{equation}
+N = (I- Q)^{-1} = 
+\begin{pmatrix}
+         1      & -\frac{1}{n-1} & -\frac{1}{n-1} & \dots  & -\frac{1}{n-1}  \\
+         0      &  1             & -\frac{1}{n-2} & \dots  & -\frac{1}{n-2}  \\
+         0      &  0             & 1             & \dots  & -\frac{1}{n-3}   \\
+         \vdots &  \vdots       & \vdots        & \ddots & \vdots            \\
+         0      &  0            & 0             & \dots  & -\frac{1}{2}      \\
+         0      &  0            & 0             & \dots  & 0           & 1   \\
+        \end{pmatrix}^{-1}
+\end{equation}
+Let $t_i$ denote the expected time
+until absorption, if chain starts from a transient state $i = 1, \dots, n-1$; 
+vector of $t_i$'s is given by the formula
+\[
+\begin{pmatrix}
+      t_1 \\
+      t_2 \\
+      \dots \\
+      t_{n-1} \\
+      \end{pmatrix}
+= N \begin{pmatrix}
+      1 \\
+      1 \\
+      \dots \\
+      1 \\
+      \end{pmatrix}.
+\]
+The expected number of steps that we are interested in is $t_1$, so we need to calculate the matrix $N$
+and prove that the sum of the elements in its first row is $O(\log n)$.
+One can easily check that
+\[
+N  =  \begin{pmatrix}
+      1 & \frac{1}{n-1} & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
+      0 &             1 & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
+      0 &             0 &            1  & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
+ \vdots &        \vdots &  \vdots       & \vdots        & \ddots & \frac{1}{3} & \frac{1}{2} \\
+      0 &             0 &            0  &            0  &  \dots &            1 & \frac{1}{2} \\
+      0 &             0 &            0  &            0  &  \dots &            0 &           1 \\
+      \end{pmatrix},
+\]
+thus $t_1 = 1 + \frac{1}{n-1} + \frac{1}{n-2} + \dots + \frac{1}{2} = O(\log n)$.
+\end{proof}
 
-\subsection{Cover trees}
+\paragraph{Experimental Results}
 
-\subsection{WSPD}
-Given $s > 1$, two disjoint subsets $A, B$ of a metric space $(X, d)$ are called $s$-\textit{well-separated},
-if 
-\[
-\forall a \in A \, \forall b \in B \, d(a, b) \geq s \max(\diam(A), \diam(B))
-\]
-A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $X$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
-The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd}.
+%The algorithm can be informally described as follows. Pick a random candidate $p_i$ and try to verify that
+%it is an approximate nearest neighbor by scanning all other points. Pick the next random point $p_j$
+%and assume that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$. This allows us to discard not just the point
+%$p_j$, but also the points that are very close to $p_j$ and very far from $p_j$ using triangle inequality.
+%Indeed, consider the ball of radius $ \dist(p_j, q) - \frac{ \dist(p_i, q) }{ 1+\eps} $ centered at $p_j$.
+%The triangle inequality implies that all points in this ball are at the distance at least $\frac{ \dist(p_i, q) }{ 1+\eps}$
+%from $q$. Now consider the ball of radius $\dist(p_j, q) + 2 \frac{\dist(p_i, q)}{1+ \eps}$ centered at $p_j$. 
+%Every point outside of this ball is a
 
-\subsection{Spanners}
+%The algorithm that we analyze is based on the following simple observation.
+%Suppose that we want to to prove that a candidate for the approximate nearest neighbor is $p_1$,
 
-\subsection{Approximate Greedy Spannner}
+%The goal is to compute
+%as few $\dist(q, p_k)$ as possible.
 
-Suppose that we have fast algorithms for computing $\lambda$-approximation of $d(x,y)$.
-In other words, for every pair of points $x,y$ we can compute an interval $[c, \lambda c]$ that is guaranteed to contain 
-the true distance $d(x,y)$.  Note that for every interval of this form there exists 
-a unique $k \in \{1, \dots, N-1 \}$ such that $[c, \lambda c] \subset [\lambda^k, \lambda^{k+2}]$. 
-Indeed, $k = \lceil \log_{\lambda}(c) \rceil$.
-We also assume that we know the range of all possible values of $k$ and they are all non-negative,
-$k = 1\dots N$. We call the intervals $[\lambda^k, \lambda^{k+2}]$ \textit{buckets}, even though they are overlapping.
-Now we can formulate the algorithm for finding a spanner.
+%\section{Geometric concepts}
 
-\begin{algorithmic}
-\label{alg:local_init}
-\Function{Approximate greedy spanner}{$rho$}
-    \ForAll {pairs $x,y \in X$}
-    \State {Place $x, y$ into corresponding bucket }
-    \EndFor
-\EndFunction
-\end{algorithmic}
+%We revise the concepts that were designed for nearest neighbor queries
+%and metric approximation in metric spaces with low dimensionality.
+%We will also review the computation methods and how they depend
+%on the predicate ``$\dist(p_i,p_j)\leq t$?'' mentioned above.
 
-We can use essentially the same proof as in \cite{bose2010computing}.
-Claim 1. If $s > \lambda^2 $ and $A, B$ are $s$-separated, then all pairs $(a, a')$, $(b, b')$ with $a, a' \in A$
-and $b, b' \in B$ are processed before any of the pairs $(a, b)$.
+%\paragraph{Notion of Dimensionality.}
+%%
+%Exansion constant, doubling dimension.
 
-Indeed, if $s > \lambda^2$, then for every points $a_1, a_2 \in A$ and $b \in B$
-we have $\lambda^2 d(a_1, a_2) < d(a_1, b)$. If $\dappr(a_1, a_2)$ and $\dappr(a_1, b)$ denote the $\lambda$-approximations
-of the corresponding distances, then the inequality $\frac{\dappr(a_1, b)}{\dappr(a_1, a_2)} > \lambda$
-holds , hence $(a_1, a_2)$ will be placed in the bucket that precedes the bucket of $(a_1, b)$.
+%\subsection{Cover trees}
 
-Claim 2. If $s > \max(\lambda^2, \frac{2t+2}{t-1})$ and $A, B$ are two $s$-well-separated subsets of $X$,
-then in the approximate greedy spanner computed by \ref{alg:approx_greedy_spanner} there is at most one edge between $A$ and $B$.
+%\subsection{WSPD}
+%Given $s > 1$, two disjoint subsets $A, B$ of a metric space $(X, d)$ are called $s$-\textit{well-separated},
+%if 
+%\[
+%\forall a \in A \, \forall b \in B \, d(a, b) \geq s \max(\diam(A), \diam(B))
+%\]
+%A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
+%\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $X$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
+%The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd}.
 
-Suppose that the pair $(a_1, b_1)$ with $a_1 \in A$, $b_1 \in B$ was processed and the edge $(a, b)$
-was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
-Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
-to the spanner. By claim 1
-$d_G(a_2, a_1) \leq td(a_2, a_1)$ and $d_G(b_1, b_2) \leq t d(b_1, b_2)$.
-By properties of WSPD, $d_G(a_1, b_1) = d(a_1, b_1) \leq (1 + \frac{2}{s})d(a_2, b_2)$
-and $t d(a_2, a_1) \leq \frac{t}{s} d(a_2, b_2)$, $t d(b_1, b_2) \leq \frac{t}{s} d(a_2, b_2)$.
+%\subsection{Spanners}
 
-Combining these inequalities, we get
+%\subsection{Approximate Greedy Spannner}
 
-\begin{equation}
-    \begin{split}
-        d_G(a_2, b_2) & \leq  d_G(a_2, a_1) + d_G(a_1, b_1) + d_G(b_1, b_2)  \\
-                      & \leq  \frac{t}{s} d(a_2, b_2) + (1 + \frac{2}{s})d(a_2, b_2) + \frac{t}{s} d(a_2, b_2) \\
-                      & =     (1 + \frac{2t + 2}{s}) d(a_2, b_2)
-    \end{split}
-\end{equation}
+%Suppose that we have fast algorithms for computing $\lambda$-approximation of $d(x,y)$.
+%In other words, for every pair of points $x,y$ we can compute an interval $[c, \lambda c]$ that is guaranteed to contain 
+%the true distance $d(x,y)$.  Note that for every interval of this form there exists 
+%a unique $k \in \{1, \dots, N-1 \}$ such that $[c, \lambda c] \subset [\lambda^k, \lambda^{k+2}]$. 
+%Indeed, $k = \lceil \log_{\lambda}(c) \rceil$.
+%We also assume that we know the range of all possible values of $k$ and they are all non-negative,
+%$k = 1\dots N$. We call the intervals $[\lambda^k, \lambda^{k+2}]$ \textit{buckets}, even though they are overlapping.
+%Now we can formulate the algorithm for finding a spanner.
 
-Solving for $s$ the inequality $1 + \frac{2t + 2}{s} < t$ gives $s > \frac{2t+2}{t-1}$.
+%\begin{algorithmic}
+%\label{alg:local_init}
+%\Function{Approximate greedy spanner}{$rho$}
+%    \ForAll {pairs $x,y \in X$}
+%    \State {Place $x, y$ into corresponding bucket }
+%    \EndFor
+%\EndFunction
+%\end{algorithmic}
+
+%We can use essentially the same proof as in \cite{bose2010computing}.
+%Claim 1. If $s > \lambda^2 $ and $A, B$ are $s$-separated, then all pairs $(a, a')$, $(b, b')$ with $a, a' \in A$
+%and $b, b' \in B$ are processed before any of the pairs $(a, b)$.
+
+%Indeed, if $s > \lambda^2$, then for every points $a_1, a_2 \in A$ and $b \in B$
+%we have $\lambda^2 d(a_1, a_2) < d(a_1, b)$. If $\dappr(a_1, a_2)$ and $\dappr(a_1, b)$ denote the $\lambda$-approximations
+%of the corresponding distances, then the inequality $\frac{\dappr(a_1, b)}{\dappr(a_1, a_2)} > \lambda$
+%holds , hence $(a_1, a_2)$ will be placed in the bucket that precedes the bucket of $(a_1, b)$.
+
+%Claim 2. If $s > \max(\lambda^2, \frac{2t+2}{t-1})$ and $A, B$ are two $s$-well-separated subsets of $X$,
+%then in the approximate greedy spanner computed by \ref{alg:approx_greedy_spanner} there is at most one edge between $A$ and $B$.
+
+%Suppose that the pair $(a_1, b_1)$ with $a_1 \in A$, $b_1 \in B$ was processed and the edge $(a, b)$
+%was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
+%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
+%to the spanner. By claim 1
+%$d_G(a_2, a_1) \leq td(a_2, a_1)$ and $d_G(b_1, b_2) \leq t d(b_1, b_2)$.
+%By properties of WSPD, $d_G(a_1, b_1) = d(a_1, b_1) \leq (1 + \frac{2}{s})d(a_2, b_2)$
+%and $t d(a_2, a_1) \leq \frac{t}{s} d(a_2, b_2)$, $t d(b_1, b_2) \leq \frac{t}{s} d(a_2, b_2)$.
+
+%Combining these inequalities, we get
+
+%\begin{equation}
+%    \begin{split}
+%        d_G(a_2, b_2) & \leq  d_G(a_2, a_1) + d_G(a_1, b_1) + d_G(b_1, b_2)  \\
+%                      & \leq  \frac{t}{s} d(a_2, b_2) + (1 + \frac{2}{s})d(a_2, b_2) + \frac{t}{s} d(a_2, b_2) \\
+%                      & =     (1 + \frac{2t + 2}{s}) d(a_2, b_2)
+%    \end{split}
+%\end{equation}
+
+%Solving for $s$ the inequality $1 + \frac{2t + 2}{s} < t$ gives $s > \frac{2t+2}{t-1}$.
 

### 2018-11-22 paper/draft.tex
Interim

@@
  \paragraph{Maintaining Lower and Upper Bounds}
- More specifically,
- we maintain an $n\times n$ matrix $M$ whose entries
-are intervals $M_{ij}=[a_{ij},b_{ij}]$, meaning that the distance
-between $p_i$ and $p_j$ lies in that interval. 
-We initialize $M_{ij}=[0,\infty]$ for all $i\neq j$
-and $M_{ij}=[0,0]$ for $i=j$.  Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
-First, we reset $M_{i,j}$ and $M_{j,i}$ to $[v,v]$. 
-To update the upper bound of some entry $M_{k,\ell}$,
+
+ In this subsection we explain the update procedure that we use. Suppose that
+ $a_{k,l}$ and $b_{k,l}$ are valid lower and upper bounds, so that
+\[
+     \forall 1 \leq k < l \leq n \quad a_{k,l} \leq \dist(p_k, p_l) \leq b_{k,l}.
+\]
+Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
+First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$. 
+To update the upper bound of some entry $b_{k,\ell}$,
 we observe that the shortest path from $p_k$ to $p_\ell$ might now
 go through the new edge. Hence, we update
-%
-\[b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}\]
-%
+\[
+    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
+\]
 Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
@@
 For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
-%
-\[v-b_{k,i}-b_{\ell,j}\]
-%
+\[
+    v-b_{k,i}-b_{\ell,j}
+\]
 is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
@@
 %
-for all $k,\ell$ and obtain the updated $M$ in $O(n^4)$ time.
+for all $k,\ell$.
+
+Note that after we added the edge $(p_i, p_j)$, the ratio of the upper and lower bound
+for this pair becomes 1, so a blind spanner algorithm always terminates (of course,
+we assume that the procedure \textsc{GetEdgeToAdd} picks an edge only once).
 
 \paragraph{Experimental Results}
+At the moment we cannot provide any theoretical guarantees for the heuristics that we propose to compute blind spanners,
+and it can happen that it will yield the complete graph as a spanner.
+We run extensive experiments on the points sampled from the low-dimensional Euclidean space to investigate
+experimentally the performance of these heuristics. Clearly, this has no practical meaning for $\R^d$, but
+this space is well-suited for controlled experiments.
+
+We look at points that are sampled uniformly at random from a cube, points that are sampled
+from a standard normal distribution, points that are clustered (we sample cluster centers uniformly at random
+and then we sample roughly the same number of points which are normally distributed around each of the cluster
+centers), and at points that are obtained from the uniform sample by exponentiating the coordinates:
+$(e^{\xi_1}, \dots, e^{\xi_d})$, where $\xi_i$ are i.i.d. with uniform distribution.
+Our findings are summarized in the plots \ref{fig:spanner_plots}.
+
+Our experiments show that the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
+algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
+first. The disappointing fact is that quasi-sorted variants produce spanners which are close to the complete graph
+(\textsc{BlindQuasiSortedGreedy} is much worse, requiring all the edges). It would seem plausible that, if we have
+access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
+find a working heuristics.
+
+It does not seem that any of the blind variants provides a spanner that is asymptotically comparable
+with the greedy spanner. Regression analysis shows that in the datasets examined the best \textsc{BlindGreedy}
+variant produces spanners which are asymptotically $O(??)$.
 
@@
 
-We can hope to avoid computation of $O(n)$ distances, if we assume low doubling dimension
+We may hope to avoid computation of $O(n)$ distances, if we assume low doubling dimension
 and solve approximate neighbor problem. The necessity of the first assumption is easily justified: if all $d(p_i, p_j)$

### 2018-11-24 paper/draft.tex
Interim

@@
 $n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
-find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. We assume for simplicity
+find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
+section, and we shall also use the shorthand notation
+\[
+    r_i := \dist(p_i, q).
+\]
+We assume for simplicity
 that all exact pairwise distances $\dist(p_i, p_j)$ are already computed. Our goal is to reduce the number
@@
 
-In order to prove this theorem, we can think of the algorithm as
+The proof is based on the following lemma.
+\begin{lemma}
+    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
+    such that after computing at most $C$ distances $\dist(p_k, q)$
+    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
+    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
+\end{lemma}
+\begin{proof}
+    First, all points that are not contained in the ball $B(p_i, 2r_i)$
+    are discarded without computing the distance to them.
+    Now we have to examine all points $p_k$ inside this ball. 
+    There are three possibilities for $r_k$:
+    \begin{enumerate}
+        \item $r_i < r_1 / (1 + \eps)$. This outcome refutes the statement $p_1$ is an approximate
+            nearest neighbor and we can stop verification.
 
-\begin{lemma}
-\end{lemma}
+    Second, after we computed the distance $\dist(p_i, q)$
+    and checked that it is greater than $r_1 / (1 + \eps)$, we can discard all points inside
+    the ball $B(p_1, r_i - r_1 / (1+ \eps))$
+\end{proof}
+
+First we consider a simplified version of the problem, where the candidate for being an approximate nearest neighbor
+is given, and we only have to verify this. Without loss of generality, assume that the candidate is $p_1$.
 

### 2018-11-24 python_version/spanner_experiment_data/plots.py

@@ -1,29 +1,90 @@
 import pandas as pd
-import matplotlib.pyplot as plt
+import sys
 
-ds = pd.read_csv("all_data.txt", header=0, sep=';')
+# ds = pd.read_csv("all_data.txt", header=0, sep=';')
+#
+# index_cols = ['dim', 'n_points', 'point_method', 'epsilon']
+# ds_greedy = ds.loc[(ds['spanner_method'] == 'greedy')].set_index(index_cols, drop=False)
+# ds_random = ds.loc[(ds['spanner_method'] == 'random')].set_index(index_cols, drop=False)
+# ds_quasi = ds.loc[(ds['spanner_method'] == 'quasi')].set_index(index_cols, drop=False)
+# ds_blind_greedy = ds.loc[(ds['spanner_method'] == 'blind_greedy')].set_index(index_cols, drop=False)
+#
+# ds_greedy.drop(columns=['spanner_method'], inplace=True)
+# ds_random.drop(columns=['spanner_method'], inplace=True)
+# ds_quasi.drop(columns=['spanner_method'], inplace=True)
+# ds_blind_greedy.drop(columns=['spanner_method'], inplace=True)
+#
+# ds_all_dims = ds_greedy.join(ds_random, lsuffix='_greedy', rsuffix='_random').join(ds_quasi, rsuffix='_quasi').join(
+#     ds_blind_greedy, rsuffix='_blind_greedy')
+#
+# print(ds_all_dims.head())
+# print(ds_all_dims.describe(include='all').to_string())
+#
+# ycols = ["sparseness_random", "sparseness_greedy", "sparseness", "sparseness_blind_greedy"]
+# # ycols = ["sparseness_random", "sparseness_greedy", "sparseness_blind_greedy"]
+# mask_1 = (ds_all_dims['dim'] == 2) & (ds_all_dims['point_method'] == 'get_exponential_points') & (ds_all_dims['epsilon'] == 0.01)
+# ds_all_dims.loc[mask_1].plot(x="n_points", y=ycols)
+# # mask_1 = (ds_all_dims['n_points'] == 200) & (ds_all_dims['point_method'] == 'get_exponential_points') & (ds_all_dims['epsilon'] == 0.01)
+# # ds_all_dims.loc[mask_1].plot(x="dim", y=ycols)
+# plt.show()
 
-index_cols = ['dim', 'n_points', 'point_method', 'epsilon']
-ds_greedy = ds.loc[(ds['spanner_method'] == 'greedy')].set_index(index_cols, drop=False)
[…]

### 2018-11-25 cpp_version/src/main.cpp
Interim

@@ -369,12 +369,12 @@ int main(int argc, char** argv)
         exp_logger->info("{};{}", dist_name, blind_greedy_spanner.get_statistics());
     }
 
-    {
-        DynamicSpannerR blind_random_spanner(dist_matrix);
-        blind_random_spanner.construct_blind_random_eps_spanner(eps);
-        console->info("{};{}", dist_name, blind_random_spanner.get_statistics());
-        exp_logger->info("{};{}", dist_name, blind_random_spanner.get_statistics());
-    }
+    //{
+    //    DynamicSpannerR blind_random_spanner(dist_matrix);
+    //    blind_random_spanner.construct_blind_random_eps_spanner(eps);
+    //    console->info("{};{}", dist_name, blind_random_spanner.get_statistics());
+    //    exp_logger->info("{};{}", dist_name, blind_random_spanner.get_statistics());
+    //}
 
 //    for (size_t i = 0; i < queries.size(); i++) {
 //        std::vector<double>& query = queries[i];

### 2018-11-25 python_version/spanner_experiment_data/plots.py

@@ -1,5 +1,6 @@
 import pandas as pd
 import sys
+import matplotlib.pyplot as plt
 
 # ds = pd.read_csv("all_data.txt", header=0, sep=';')
 #
@@ -31,45 +32,39 @@ import sys
 
 if __name__ == "__main__":
     ds = pd.read_csv("spanner_cpp_logs.txt", header=0, sep=';')
-    # print(ds.point_gen_method.unique())
-    # print(ds.spanner_method.unique())
-    # index_cols = ['dim', 'point_gen_method', 'input_file', 'spanner_method', 'epsilon', 'n_points', 'n_edges', 'sparseness', 'edges_to_points_ratio']
-    index_cols = ['dim', 'point_gen_method', 'input_file', 'spanner_method', 'epsilon', 'n_points']
-    # ds.drop(columns=['sparseness', 'edges_to_points_ratio'], inplace=True)
-    # index_cols = ['dim', 'point_gen_method', 'input_file', 'spanner_method', 'epsilon', 'n_points']
+    index_cols = ['dim', 'point_gen_method', 'epsilon', 'input_file', 'n_points']
 
-    ds_non_blind_greedy = ds.loc[(ds['spanner_method'] == 'greedy-non-blind')].set_index(index_cols, drop=False)
-    ds_blind_greedy = ds.loc[(ds['spanner_method'] == 'blind-greedy')].set_index(index_cols, drop=False)
-    ds_blind_random = ds.loc[(ds['spanner_method'] == 'blind-random')].set_index(index_cols, drop=False)
-    ds_quasi_greedy = ds.loc[(ds['spanner_method'] == 'blind-quasi-sorted-greedy')].set_index(index_cols, drop=False)
-    ds_quasi_shaker = ds.loc[(ds['spanner_method'] == 'blind-quasi-sorted-shaker')].set_index(index_cols, drop=False)
+    ds_non_blind_greedy = ds.loc[(ds['spanner_method'] == 'greedy-non-blind')].set_index(index_cols, drop=True)
+    ds_blind_greedy = ds.loc[(ds['spanner_method'] == 'blind-greedy')].set_index(index_cols, drop=True)
+    ds_blind_random = ds.loc[(ds['spanner_method'] == 'blind-random')].set_index(index_cols, drop=True)
+    ds_quasi_greedy = ds.loc[(ds['spanner_method'] == 'blind-quasi-sorted-greedy')].set_index(index_cols, drop=True)
[…]

### 2018-11-25 python_version/spanner_experiment_data/plots.py
Interim

@@ -32,7 +32,7 @@ import matplotlib.pyplot as plt
 
 if __name__ == "__main__":
     ds = pd.read_csv("spanner_cpp_logs.txt", header=0, sep=';')
-    index_cols = ['dim', 'point_gen_method', 'epsilon', 'input_file', 'n_points']
+    index_cols = ['Dim', 'point_gen_method', 'epsilon', 'input_file', 'Points']
 
     ds_non_blind_greedy = ds.loc[(ds['spanner_method'] == 'greedy-non-blind')].set_index(index_cols, drop=True)
     ds_blind_greedy = ds.loc[(ds['spanner_method'] == 'blind-greedy')].set_index(index_cols, drop=True)
@@ -64,26 +64,26 @@ if __name__ == "__main__":
     ds_blind_random_bad_ratio_cf.drop(columns=['spanner_method'], inplace=True)
     ds_blind_random_bad_ratio_cf_lbf.drop(columns=['spanner_method'], inplace=True)
 
-    ds_all_dims = ds_non_blind_greedy.join(ds_blind_greedy, lsuffix='_nbg', rsuffix='_bg', how="outer")
-    ds_all_dims = ds_all_dims.join(ds_blind_random, rsuffix='_br')
-    ds_all_dims = ds_all_dims.join(ds_quasi_greedy, rsuffix='_qg').join(ds_quasi_shaker, rsuffix='_qs')
-    ds_all_dims = ds_all_dims.join(ds_blind_random_bad_ratio, rsuffix='_rbr')
-    ds_all_dims = ds_all_dims.join(ds_blind_random_bad_ratio_lbf, rsuffix='_rbr_lbf')
-    ds_all_dims = ds_all_dims.join(ds_blind_random_bad_ratio_cf, rsuffix='_rbr_cf')
-    ds_all_dims = ds_all_dims.join(ds_blind_random_bad_ratio_cf_lbf, rsuffix='_rbr_cf_lbf')
+    ds_all_dims = ds_non_blind_greedy.join(ds_blind_greedy, lsuffix=' (non-blind greedy)', rsuffix=' (blind greedy)', how="outer")
+    ds_all_dims = ds_all_dims.join(ds_blind_random, rsuffix=' (blind random)')
+    ds_all_dims = ds_all_dims.join(ds_quasi_greedy, rsuffix=' (quasi-sorted greedy').join(ds_quasi_shaker, rsuffix=' (quasi-sorted shaker)')
[…]

### 2018-11-25 python_version/spanner_experiment_data/plots.py
Iterim

@@ -77,10 +77,10 @@ if __name__ == "__main__":
     # ycols = ["edges_to_points_ratio_nbg", "edges_to_points_ratio_bg", "edges_to_points_ratio_rbr_lbf",
     #          "edges_to_points_ratio_qs", "edges_to_points_ratio_qg"]
 
-    # ycols = ["edges_to_points_ratio_nbg", "edges_to_points_ratio_bg", "edges_to_points_ratio_rbr_lbf" ]
     ycols = ["Edges (non-blind greedy)", "Edges (blind greedy)", "Edges (blind random bad ratio, force lower bound)", "Edges (blind random bad ratio)"]
+    ycols = ["Edges to points ratio (non-blind greedy)", "Edges to points ratio (blind greedy)", "Edges to points ratio (blind random bad ratio, force lower bound)", "Edges to points ratio (blind random bad ratio)"]
     # # ycols = ["sparseness_random", "sparseness_greedy", "sparseness_blind_greedy"]
-    ds_to_plot = ds_all_dims.loc[(2, "normal", 0.1)]
+    ds_to_plot = ds_all_dims.loc[(2, "exp", 0.1)]
     ds_to_plot.reset_index(inplace=True)
     ds_to_plot.drop(columns="input_file", inplace=True)
     ds_to_plot = ds_to_plot.set_index(["Points"])

### 2018-11-25 paper/abstract.tex
add abstract

\documentclass[a4paper,USenglish]{socg-lipics-v2018}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathbbol}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{url}
%\usepackage{floatflt}
%\usepackage{mathptmx}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{color}

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{comment}
\usepackage{braket}

% \usepackage{ccaption}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\dappr}{\tilde{d}}
%\newcommand{\Wlog}{W.l.o.g.\ }
%\newcommand{\wlog}{w.l.o.g.\ }
\newcommand{\eps}{\epsilon}
\newcommand{\orient}{O}
\newcommand{\cpro}{\Pi}
\newcommand{\fib}{\mathrm{Fib}}

\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\pointset}{P}
\newcommand{\distspace}{\mathcal{M}}
\newcommand{\dist}{\delta}
\newcommand{\adist}{\gamma}
\newcommand{\complexity}{C_{\dist}}
\newcommand{\doublingdimension}{\Delta}

%\newcommand{\remark}[1]{\textbf{[#1]}}

%\newtheorem{theorem}{Theorem}
%\newdef{definition}{Definition}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{corollary}[theorem]{Corollary}

\newcommand{\myparagraph}[1]{\textbf{#1.}}

%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{proposition}[theorem]{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}
%\newtheorem{definition}[theorem]{Definition}
%\newtheorem{remark}[theorem]{Remark}
%\newtheorem{note}[theorem]{Note}

%\numberwithin{equation}{section}
%\numberwithin{figure}{section}

%\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
%\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}

\title{Geometric Filters for Metric Spaces}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}

\begin{document}
\maketitle

\section{Abstract}
Often the complexity bounds for algorithms in metric spaces
assume that the metric can be computed in constant time and
subsume this constant in $O()$ notation. 
Motivated by examples of metrics from Topological Data Analysis,
in this paper we consider
the case of metric spaces where each distance computation
is an expensive operation.
We consider two classical problems: computing a spanner and finding approximate
nearest neighbor in the setting when the number of input points $n$
is small enough to consider operations with polynomial complexity in $n$ (even of
relatively high degree) as something cheap compared to the computation of an individual distance,
and aim to reduce the number of computed distances.
An example of such a situation is having about a hundred of persistence
diagrams, each containing hundreds of thousands of points.

\remark{start of the short version}

We propose simple algorithms  for each of the two problems,
and for the approximate nearest neighbor we are able to 
prove theoretically that the algorithm
reduces the number of computed distances to $O(\log n)$,
if we assume doubling metric space.
We also provide experimental evidence supporting this bound.

For spanners we can only provide experimental results for
different heuristics that we tried; but the results look
convincing in the sense that on many different random point
sets we are able to compute the spanner 
avoiding computation of all $n \choose 2$ pairwise distances.

The main idea of our approach to both problems is to use
the triangle inequality to extract as much information as possible
from the distances that we have computed so far, and
it turns out that in case of low doubling dimension
this idea really works, sparing us from the computation
of many distances we would have needed, if we
chose some of the existing algorithms. To the best of our knowledge,
this is the first paper in which this setting is considered,
and we believe that these results are just first
steps in this direction, which is theoretically
appealing as well as practically important.

\remark{end of the short version, start of the long version}

The main idea of our approach to the both problems is to maintain
lower and upper bounds
for each of the distances we might need to compute
and use these bounds whenever possible instead
of the true distance. For example, when the algorithm
needs to decide whether the distance between two given points
is less than some $t$, and the upper bound for this distance is
less than $t$ or the lower bound is larger than $t$,
we avoid computing the actual distance;
otherwise we are forced to compute it, but then
we use it to update our bounds for all other distances (the
only tool that we can use here is, of course, the triangle inequality).
For spanners we propose what we call the \textit{blind} approach,
where we add an edge to the spanner for each of the distances we computed 
(hence the name: the
algorithm has no access to exact distance between two points before it decides to connect them
with an edge), and we continue adding edges until the lower and upper bounds
for all pairs of points become sufficiently close to each other.
We experimentally demonstrate that in case of low dimension
some of the heuristics we tried help to significantly reduce the number of
computed distances.

For nearest neighbor problem we want to minimize the number
of computed distances to the query point. Here we are able to theoretically prove that, 
if we assume doubling property,
it is possible to find the approximate nearest neighbor computing
only $O(\log n)$ distances to the query point.
We also provide experimental evidence showing that this approach
reduces the number of distance computations in practice.

As far as we know, the setting that we consider is new, and the results in this paper
are just the first steps in this direction, which, we believe, has large
theoretical potential and is of great practical importance.

\end{document}

### 2018-11-26 paper/draft.tex
Inerim

@@
 \author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
+\authorrunning{M. Kerber and A. Nigmetov}
+
+\Copyright{Michael Kerber and Arnur Nigmetov}
 
@@
 \paragraph{Motivation.}
-%
+
+In this paper we look at algorithms in metric spaces,
+in which computing a single distance between points can be very costly.
+More specifically,
+we consider the problem of finding an approximate nearest neighbor
+of some point $q$ among the given $n$ points, and the problem of finding a $t$-spanner,
+i.e., a graph that approximates the metric on the given set of points,
+and we not so much interested in reducing the complexity of the algorithm,
+but rather in reducing the number of distances computed.
+The motivating example for us was Topological Data Analysis, where computation
+of bottleneck and Wasserstein distances between two persistence diagrams
+can be done in time polynomial in the number of points of the input diagrams.
+
+We believe that the scenario when the input consists of a relatively small number of points, 
+e.g., of order $10^2-10^3$, but the complexity of computing a distance is rather large,
+is plausible (for the persistence diagrams, the size of 0-dimensional ones can be huge,
+because in dimension 0 there are especially efficient algorithms that can handle much
+larger inputs, so it is possible to produce diagrams with millions of  points in each).
+In such a situation, the natural approach would be to avoid computing the
+distance between two points if we do not need it,
+
+We demonstrate that the number of distance computations can be further
+reduced by a careful re-implementation of the underlying geometric algorithms.
+Our improvement is based on the following simple observation: many geometric
+constructions (such as cover tree and WSPD constructions) require distance
+computations only in the form of the following predicate: 
+``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
+We show that using the information of the already computed distances and
+the triangle inequality, we can answer the above predicate in many cases
+without computing the actual distance.
+
+We describe our idea with an example. In Figure~\ref{fig:1st_example},
+we see a weighted graph on $P$ whose edges are the distances that have
+been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
+The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
+constitutes an upper bound on the distance by triangle inequality.
+However, we can also infer that $\dist(p_1,p_2)\geq 3$:
+otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
+would be shorter than the edge $(p_3,p_4)$, again contradicting
+triangle inequality.
+Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
+for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
+distance computation is necessary. This approach is reminiscent to the usage
+of \emph{numerical filters} in geometric computing in the CGAL library,
+where geometric predicates in exact number types are first evaluated using 
+certified floating-point arithmetic, and exact computation is only
+invoked if the numerical evaluation did not arrive at a decisive result
+(e.g.~\cite{bbp-interval,kerber-phd}).
+
 Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
@@
 $O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
-For low-dimensional\footnote{we will make this notion precise
-later in the text} metric spaces, we can improve these bounds:
-Using cover trees~\cite{cover-trees}, we can preprocess the point set
-in $O(\complexity n\log n)$ time to answer nearest neighbor queries
-in $O(\complexity \log n)$ time. Combining cover trees with well-separated
-pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
-induced by $\dist$ in $O(\complexity n\log n)$ time and
-of linear size (assuming, for simplicity, that $\eps$ is a constant).
-Out of this spanner,
-we can construct an approximate metric
-with an all-pair shortest path computation,
-resulting in a total running time of
-\[O(\complexity n\log n+ n^2\log n)\]
-using Johnson's algorithm~\cite{johnson}.
+% For low-dimensional\footnote{we will make this notion precise
+% later in the text} metric spaces, we can improve these bounds:
+% Using cover trees~\cite{cover-trees}, we can preprocess the point set
+% in $O(\complexity n\log n)$ time to answer nearest neighbor queries
+% in $O(\complexity \log n)$ time. Combining cover trees with well-separated
+% pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
+% induced by $\dist$ in $O(\complexity n\log n)$ time and
+% of linear size (assuming, for simplicity, that $\eps$ is a constant).
+% Out of this spanner,
+% we can construct an approximate metric
+% with an all-pair shortest path computation,
+% resulting in a total running time of
+% \[O(\complexity n\log n+ n^2\log n)\]
+% using Johnson's algorithm~\cite{johnson}.
 
@@
 in algorithms for metric spaces?
-Note that for low-dimensional spaces, the aforementioned techniques
-already provide a partial solution: for approximating the metric, for instance,
-we only compute $O(n\log n)$ distances instead of $O(n^2)$.
+% Note that for low-dimensional spaces, the aforementioned techniques
+% already provide a partial solution: for approximating the metric, for instance,
+% we only compute $O(n\log n)$ distances instead of $O(n^2)$.
 
 \paragraph{Contribution.}
+
 %
@@
 sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
-until it becomes a $1+\eps$-spanner. In our setting this is the most expensive approach, because it requires
-computing all pairwise distances first.
-the true distance belongs.
+until it becomes a $1+\eps$-spanner.
 
@@
 
-We look at points that are sampled uniformly at random from a cube, points that are sampled
-from a standard normal distribution, points that are clustered (we sample cluster centers uniformly at random
-and then we sample roughly the same number of points which are normally distributed around each of the cluster
-centers), and at points that are obtained from the uniform sample by exponentiating the coordinates:
-$(e^{\xi_1}, \dots, e^{\xi_d})$, where $\xi_i$ are i.i.d. with uniform distribution.
-Our findings are summarized in the plots \ref{fig:spanner_plots}.
+We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
+\begin{enumerate}
+    \item In the \textbf{uniform} test set points are sampled uniformly at
+        random from the unit cube in $\mathbb{R}^d$.
+    \item In the \textbf{normal} test set points are sampled from the standard
+        normal distribution in $\mathbb{R}^d$.
+    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
+        at random from $[0,10000]^d$, and then we add normally distributed noise around
+        each of the centers. The number of clusters is chosen so that each cluster
+        contains 50 points.
+    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
+        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
+\end{enumerate}
+In all experiments the algorithms that we tested compared in the same way,
+so we only present results for the \textbf{uniform} point set in dimension 2.
+%Let $e$ denote the number of edges of a spanner on $n$ vertices;
+%in the plots we visualize the ratios $e / { n \choose 2}$ and $e / n$.
+%We refer to the first ratio as \textit{density} and to the second one
+%as \textit{growth}. Note that 
+The results are summarized in the plots \ref{plot1}, \ref{fig:spanner_plot2}.
+The first plot shows the number of edges of the spanner.
+We can see that, even though none of the blind spanners can produce
+spanners of the same quality (i.e., sparseness) as the standard greedy algorithm,
+\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
+perform significantly better than both variants of \textsc{BlindQuasiSorted}.
+it is important to remember that for all blind spanners the number of edges
+is equal to the number of distances that were actually computed, while
+for non-blind greedy spanner the latter number is always $n \choose 2$.
+The second plot shows the ratio of the number of edges to the number of points.
+The ideal behavior is demonstrated by the non-blind greedy spanner,
+for which this ratio stays practically constant, confirming the linear growth.
+None of the blind algorithms seems to have this property, but among them
+the blind greedy spanner is the best one. If we assume that the number of edges
+is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
+linear regression.  We give in the table \ref{tbl:regr_coeff_spanner}
+the maximal exponents $\alpha$ for each of the algorithms,
+and it is worth mentioning that even for the quasi-sorted shaker variant 
+the exponent seems to be bounded away from 2.
 
-Our experiments show that the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
+Summing up, we can conclude from the experiments that 
+ the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
 algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
-first. The disappointing fact is that quasi-sorted variants produce spanners which are close to the complete graph
+first. The disappointing fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
 (\textsc{BlindQuasiSortedGreedy} is much worse, requiring all the edges). It would seem plausible that, if we have
@@
 
-It does not seem that any of the blind variants provides a spanner that is asymptotically comparable
-with the greedy spanner. Regression analysis shows that in the datasets examined the best \textsc{BlindGreedy}
-variant produces spanners which are asymptotically $O(??)$.
+%We look at points that are sampled uniformly at random from a cube, points that are sampled
+%from a standard normal distribution, points that are clustered (we sample cluster centers uniformly at random
+%and then we sample roughly the same number of points which are normally distributed around each of the cluster
+%centers), and at points that are obtained from the uniform sample by exponentiating the coordinates:
+%$(e^{\xi_1}, \dots, e^{\xi_d})$, where $\xi_i$ are i.i.d. with uniform distribution.
+%Our findings are summarized in the plots \ref{fig:spanner_plots}.
 
@@
 We assume for simplicity
-that all exact pairwise distances $\dist(p_i, p_j)$ are already computed. Our goal is to reduce the number
-of computed distances $d(p_i, q)$.
+that all exact pairwise distances $\dist(p_i, p_j)$ are already computed.
+Our goal is to reduce the number of computed distances $\dist(p_i, q)$. We shall also assume w.l.o.g.
+that all $\dist(p_i, q)$ are distinct (we can always break ties lexicographically).
 
@@
 \begin{algorithmic}
+\Function{DiscardPoints}{$L, p_i, r_i, \eps$}
+    \ForAll { $p_j \in L$}
+       \If { $ \dist(p_i, p_j) < r_j - r_i / ( 1 + \eps) $ }
+           \State {Remove $p_j$ from $L$}
+       \EndIf
+
+       \If {  $ \dist(p_i, p_j) > r_j + r_i / ( 1 + \eps) $ }
+           \State {Remove $p_j$ from $L$}
+       \EndIf
+
+    \EndFor
+\EndFunction
+\end{algorithmic}
+
+\begin{algorithmic}
+\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
+    \State {Initialize list $L$ with a random permutation of $P$}
+    \While {$L$ is not empty}
+        \State {Set current candidate $p_i$ to be the first element of $L$}
+        \State {Compute $r_i = \dist(q, p_i)$}
+        \State \Call{DiscardPoints}{$L, p_i, r_i, \eps$}
+        \ForAll {$p_j \in L $}
+            \State {Compute $r_j = \dist(p_j, q)$}
+            \State \Call{DiscardPoints}{$L, p_j, r_j$}
+        \EndFor
+    \EndWhile
+\EndFunction
+\end{algorithmic}
+
+\begin{algorithmic}
 \Function{ApproximateNearestNeighbor}{$P, q, \eps$}
@@
     are discarded without computing the distance to them.
-    Now we have to examine all points $p_k$ inside this ball. 
-    There are three possibilities for $r_k$:
-    \begin{enumerate}
-        \item $r_i < r_1 / (1 + \eps)$. This outcome refutes the statement $p_1$ is an approximate
-            nearest neighbor and we can stop verification.
-
-    Second, after we computed the distance $\dist(p_i, q)$
-    and checked that it is greater than $r_1 / (1 + \eps)$, we can discard all points inside
-    the ball $B(p_1, r_i - r_1 / (1+ \eps))$
+    Now we have to examine all points $p_j$ inside this ball. 
+    Suppose that we cannot discard the current point $p_j$ based
+    on the lower bound for $r_j$, then we must compute the distance $r_j$.
+    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
+    The latter case means that we discard the point $p_j$,
+    but we also discard all points of $P$ that are contained in some ball around it.
+    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
+    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
+    In the figure this radius corresponds to the width of the annulus surrounding $q$,
+    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
+    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
+    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
+    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
+    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
+    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
+    This means that even in the worst case after we have computed $T$ distances, we have discarded all
+    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
+    do not require distance computation. Indeed, each of these points lies in
+    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
+    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
+    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
+    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
+    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
+    because some of the points that are inside the annulus can be discarded, 
+    if there is a point from $P$ close to them but outside of the annulus. 
 \end{proof}
 
-First we consider a simplified version of the problem, where the candidate for being an approximate nearest neighbor
-is given, and we only have to verify this. Without loss of generality, assume that the candidate is $p_1$.
+Now the theorem follows from the lemma by standard argument
+of the backwards analysis.
+
+% First we consider a simplified version of the problem, where the candidate for being an approximate nearest neighbor
+% is given, and we only have to verify this. Without loss of generality, assume that the candidate is $p_1$.
 
@@
 \paragraph{Experimental Results}
+We use the same
 

### 2018-11-26 paper/draft.tex
interim

@@
 \begin{algorithmic}
-\Function{DiscardPoints}{$L, p_i, r_i, \eps$}
+\Procedure{Discard\_1}{$L, p_i$}
+    \State {$r_i = \dist(q, p_i)$}
     \ForAll { $p_j \in L$}
-       \If { $ \dist(p_i, p_j) < r_j - r_i / ( 1 + \eps) $ }
+       \If {  $ \dist(p_i, p_j) > r_i \left(1 + \frac{1}{ 1 + \eps} \right)$ }
            \State {Remove $p_j$ from $L$}
        \EndIf
-
-       \If {  $ \dist(p_i, p_j) > r_j + r_i / ( 1 + \eps) $ }
-           \State {Remove $p_j$ from $L$}
-       \EndIf
-
     \EndFor
-\EndFunction
+\EndProcedure
 \end{algorithmic}
@@
 \begin{algorithmic}
-\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
-    \State {Initialize list $L$ with a random permutation of $P$}
-    \While {$L$ is not empty}
-        \State {Set current candidate $p_i$ to be the first element of $L$}
-        \State {Compute $r_i = \dist(q, p_i)$}
-        \State \Call{DiscardPoints}{$L, p_i, r_i, \eps$}
-        \ForAll {$p_j \in L $}
-            \State {Compute $r_j = \dist(p_j, q)$}
-            \State \Call{DiscardPoints}{$L, p_j, r_j$}
+\Procedure{Discard\_2}{$L, w, c$}
+    \State {$r_w = \dist(q, w)$}
+    \State {$r_c = \dist(q, w)$}
+    \Comment {Distances $r_w$ and $r_c$ are cached}
+    \ForAll { $p_j \in L$}
+       \If { $ \dist(w, p_j) < r_w - r_c / ( 1 + \eps) $ }
+           \State {Remove $p_j$ from $L$}
+           \Comment {In particular, the worse candidate $w$ will be removed here}
+       \EndIf
+       \If {  $ \dist(w, p_j) > r_w + r_c / ( 1 + \eps) $ }
+           \State {Remove $p_j$ from $L$}
+       \EndIf
     \EndFor
-    \EndWhile
-\EndFunction
+\EndProcedure
 \end{algorithmic}
@@
 \begin{algorithmic}
+\label{alg:ann_blind}
 \Function{ApproximateNearestNeighbor}{$P, q, \eps$}
-\For {$i \in \{\,1, \dots, |P|\,\}$}
-\State {$a_i \gets 0$}
-\State {$b_i \gets \infty$}
+    \State {Initialize array $L$ with a random permutation of $P$}
+    \State {$ c \gets \mbox{the first element of }L$}
+    \State {Compute $r_i = \dist(q, p_i)$}
+    \While {$|L| > 1$}
+        \State \Call{Discard\_1}{$L, c$}
+        \ForAll {$p_j \in L $}
+            \State {Compute $r_j = \dist(p_j, q)$}
+            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
+            \If {$r_j < r_i$ }
+                \State {$ c \gets p_j $}
+                \State {$ w \gets p_i $}
+            \Else
+                \State {$ w \gets p_i $}
+            \EndIf
+            \State \Call{Discard\_2}{$L, w, c$}
         \EndFor
-%\State {$i_0 \gets \argmin u $}
-\State {$B  \gets \{\, i \mid l_i < u[i_0]\,\}$}
-\If {$B = \emptyset$}
-    \State {\Return {$p_{i_0}$}}
-\Else
-    \State{$j \gets \mbox{ random element of } C$}
-    \State{$u_j \gets d(p_j, q)$}
-    \State{$l[j] \gets d(p_j, q)$}
-    \State{Mark distance from $q$ to $p_j$ as computed}
-    \State{\Call{UpdateBounds}{$l, u, d(p_j, q), j$}}
-\EndIf
+    \EndWhile
+    \State \Return {the unique element left in $L$}
 \EndFunction

### 2018-11-26 python_version/ann_experiment_data/plots.py
Interim

@@ -1,0 +1,53 @@
+import pandas as pd
+import sys
+import matplotlib.pyplot as plt
+
+
+if __name__ == "__main__":
+    # ds = pd.read_csv("ann_results.txt", header=0, sep=';', dtype = { 'Epsilon' : object })
+    ds = pd.read_csv("ann_results.txt", header=0, sep=';', dtype = { 'Epsilon' : object })
+    index_cols = ['Dimension', 'Epsilon',  'point_gen_method', 'query_method', 'Npoints']
+
+    ds.drop(columns=['MinFraction', 'MaxFraction', 'MinNumberOfDistances', 'MaxNumberOfDistances'], inplace=True)
+
+    print(ds.describe())
+    print("********************************************************************************")
+    print(ds.head(20))
+    print("********************************************************************************")
+    ds.set_index(index_cols, inplace=True)
+    # sys.exit(0)
+
+
+    # ds_non_blind_greedy.drop(columns=['spanner_method'], inplace=True)
+    # ds_blind_greedy.drop(columns=['spanner_method'], inplace=True)
+    # ds_blind_random.drop(columns=['spanner_method'], inplace=True)
+    # ds_quasi_greedy.drop(columns=['spanner_method'], inplace=True)
+    # ds_quasi_shaker.drop(columns=['spanner_method'], inplace=True)
+    # ds_blind_random_bad_ratio.drop(columns=['spanner_method'], inplace=True)
+    # ds_blind_random_bad_ratio_lbf.drop(columns=['spanner_method'], inplace=True)
+    # ds_blind_random_bad_ratio_cf.drop(columns=['spanner_method'], inplace=True)
+    # ds_blind_random_bad_ratio_cf_lbf.drop(columns=['spanner_method'], inplace=True)
+    #
+    # ds_all_dims = ds_non_blind_greedy.join(ds_blind_greedy, lsuffix=' (non-blind greedy)', rsuffix=' (blind greedy)', how="outer")
+    # ds_all_dims = ds_all_dims.join(ds_blind_random, rsuffix=' (blind random)')
+    # ds_all_dims = ds_all_dims.join(ds_quasi_greedy, rsuffix=' (quasi-sorted greedy').join(ds_quasi_shaker, rsuffix=' (quasi-sorted shaker)')
[…]

### 2018-11-27 paper/draft.tex
interim

@@
 
-The algorithm that we analyze can be roughly described as follows. Pick a random point $p_i$,
+The algorithm that we analyze can be roughly described as follows. 
+Pick a random point $p_i$,
 compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
@@
     \State {Compute $r_i = \dist(q, p_i)$}
+    \State \Call{Discard\_1}{$L, c$}
     \While {$|L| > 1$}
-        \State \Call{Discard\_1}{$L, c$}
         \ForAll {$p_j \in L $}
@@
                 \State {$ c \gets p_j $}
+                \State \Call{Discard\_1}{$L, c$}
                 \State {$ w \gets p_i $}
@@
 
-% First we consider a simplified version of the problem, where the candidate for being an approximate nearest neighbor
-% is given, and we only have to verify this. Without loss of generality, assume that the candidate is $p_1$.
-
-\begin{lemma}
-Consider the following random procedure: we start with $n$ numbers $[n] = \{\,1,\dots, n\,\}$ and 
-choose $\xi_1$ uniformly at random from $[n]$. If $\xi_1 = n$, then we stop,
-otherwise we remove $\xi_1$ and all numbers preceding it and from $\{\,\xi_1, \dots, n\,\}$
-we choose $\xi_2$. If $\xi_2 = n$, we stop, if $\xi_2 < n$, we repeat the process removing all numbers not greater than $\xi_2$ and
-drawing $\xi_3$ from the remaining numbers uniformly at random. The expected number of steps of this procedure
-is $O(\log n)$.
-\end{lemma}
-\begin{proof}
-We can rephrase the procedure as a Markov chain with set of states $[n]$ and transition matrix
-\begin{equation}
-    P = \begin{pmatrix}
-         0      & \frac{1}{n-1} & \frac{1}{n-1} & \dots  & \frac{1}{n-1} & \frac{1}{n-1} \\
-         0      &  0            & \frac{1}{n-2} & \dots  & \frac{1}{n-2} & \frac{1}{n-2} \\
-         0      &  0            & 0             & \dots  & \frac{1}{n-3} & \frac{1}{n-3} \\
-         \vdots &  \vdots       & \vdots        & \ddots & \vdots        & \vdots \\
-         0      &  0            & 0             & \dots  & \frac{1}{2} & \frac{1}{2} \\
-         0      &  0            & 0             & \dots  & 0           & 1           \\
-        \end{pmatrix};
-\end{equation}
-the chain starts from state $1$.
-Note that there is one absorbing state $n$, and all other states are transient, and the matrix is already in the canonical
-form, meaning that all transient states precede all absorbing states. 
-We denote the square submatrix formed by the first $n-1$
-rows and $n-1$ columns by $Q$, it is the submatrix corresponding to all transient states. It is known
-(\cite{absorbing_markov_chain}) that the matrix $I-Q$ is invertible, let $N$ be its inverse,
-\begin{equation}
-N = (I- Q)^{-1} = 
-\begin{pmatrix}
-         1      & -\frac{1}{n-1} & -\frac{1}{n-1} & \dots  & -\frac{1}{n-1}  \\
-         0      &  1             & -\frac{1}{n-2} & \dots  & -\frac{1}{n-2}  \\
-         0      &  0             & 1             & \dots  & -\frac{1}{n-3}   \\
-         \vdots &  \vdots       & \vdots        & \ddots & \vdots            \\
-         0      &  0            & 0             & \dots  & -\frac{1}{2}      \\
-         0      &  0            & 0             & \dots  & 0           & 1   \\
-        \end{pmatrix}^{-1}
-\end{equation}
-Let $t_i$ denote the expected time
-until absorption, if chain starts from a transient state $i = 1, \dots, n-1$; 
-vector of $t_i$'s is given by the formula
-\[
-\begin{pmatrix}
-      t_1 \\
-      t_2 \\
-      \dots \\
-      t_{n-1} \\
-      \end{pmatrix}
-= N \begin{pmatrix}
-      1 \\
-      1 \\
-      \dots \\
-      1 \\
-      \end{pmatrix}.
-\]
-The expected number of steps that we are interested in is $t_1$, so we need to calculate the matrix $N$
-and prove that the sum of the elements in its first row is $O(\log n)$.
-One can easily check that
-\[
-N  =  \begin{pmatrix}
-      1 & \frac{1}{n-1} & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
-      0 &             1 & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
-      0 &             0 &            1  & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
- \vdots &        \vdots &  \vdots       & \vdots        & \ddots & \frac{1}{3} & \frac{1}{2} \\
-      0 &             0 &            0  &            0  &  \dots &            1 & \frac{1}{2} \\
-      0 &             0 &            0  &            0  &  \dots &            0 &           1 \\
-      \end{pmatrix},
-\]
-thus $t_1 = 1 + \frac{1}{n-1} + \frac{1}{n-2} + \dots + \frac{1}{2} = O(\log n)$.
-\end{proof}
+%\begin{lemma}
+%Consider the following random procedure: we start with $n$ numbers $[n] = \{\,1,\dots, n\,\}$ and 
+%choose $\xi_1$ uniformly at random from $[n]$. If $\xi_1 = n$, then we stop,
+%otherwise we remove $\xi_1$ and all numbers preceding it and from $\{\,\xi_1, \dots, n\,\}$
+%we choose $\xi_2$. If $\xi_2 = n$, we stop, if $\xi_2 < n$, we repeat the process removing all numbers not greater than $\xi_2$ and
+%drawing $\xi_3$ from the remaining numbers uniformly at random. The expected number of steps of this procedure
+%is $O(\log n)$.
+%\end{lemma}
+%\begin{proof}
+%We can rephrase the procedure as a Markov chain with set of states $[n]$ and transition matrix
+%\begin{equation}
+%    P = \begin{pmatrix}
+%         0      & \frac{1}{n-1} & \frac{1}{n-1} & \dots  & \frac{1}{n-1} & \frac{1}{n-1} \\
+%         0      &  0            & \frac{1}{n-2} & \dots  & \frac{1}{n-2} & \frac{1}{n-2} \\
+%         0      &  0            & 0             & \dots  & \frac{1}{n-3} & \frac{1}{n-3} \\
+%         \vdots &  \vdots       & \vdots        & \ddots & \vdots        & \vdots \\
+%         0      &  0            & 0             & \dots  & \frac{1}{2} & \frac{1}{2} \\
+%         0      &  0            & 0             & \dots  & 0           & 1           \\
+%        \end{pmatrix};
+%\end{equation}
+%the chain starts from state $1$.
+%Note that there is one absorbing state $n$, and all other states are transient, and the matrix is already in the canonical
+%form, meaning that all transient states precede all absorbing states. 
+%We denote the square submatrix formed by the first $n-1$
+%rows and $n-1$ columns by $Q$, it is the submatrix corresponding to all transient states. It is known
+%(\cite{absorbing_markov_chain}) that the matrix $I-Q$ is invertible, let $N$ be its inverse,
+%\begin{equation}
+%N = (I- Q)^{-1} = 
+%\begin{pmatrix}
+%         1      & -\frac{1}{n-1} & -\frac{1}{n-1} & \dots  & -\frac{1}{n-1}  \\
+%         0      &  1             & -\frac{1}{n-2} & \dots  & -\frac{1}{n-2}  \\
+%         0      &  0             & 1             & \dots  & -\frac{1}{n-3}   \\
+%         \vdots &  \vdots       & \vdots        & \ddots & \vdots            \\
+%         0      &  0            & 0             & \dots  & -\frac{1}{2}      \\
+%         0      &  0            & 0             & \dots  & 0           & 1   \\
+%        \end{pmatrix}^{-1}
+%\end{equation}
+%Let $t_i$ denote the expected time
+%until absorption, if chain starts from a transient state $i = 1, \dots, n-1$; 
+%vector of $t_i$'s is given by the formula
+%\[
+%\begin{pmatrix}
+%      t_1 \\
+%      t_2 \\
+%      \dots \\
+%      t_{n-1} \\
+%      \end{pmatrix}
+%= N \begin{pmatrix}
+%      1 \\
+%      1 \\
+%      \dots \\
+%      1 \\
+%      \end{pmatrix}.
+%\]
+%The expected number of steps that we are interested in is $t_1$, so we need to calculate the matrix $N$
+%and prove that the sum of the elements in its first row is $O(\log n)$.
+%One can easily check that
+%\[
+%N  =  \begin{pmatrix}
+%      1 & \frac{1}{n-1} & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
+%      0 &             1 & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
+%      0 &             0 &            1  & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
+% \vdots &        \vdots &  \vdots       & \vdots        & \ddots & \frac{1}{3} & \frac{1}{2} \\
+%      0 &             0 &            0  &            0  &  \dots &            1 & \frac{1}{2} \\
+%      0 &             0 &            0  &            0  &  \dots &            0 &           1 \\
+%      \end{pmatrix},
+%\]
+%thus $t_1 = 1 + \frac{1}{n-1} + \frac{1}{n-2} + \dots + \frac{1}{2} = O(\log n)$.
+%\end{proof}
 
 \paragraph{Experimental Results}
-We use the same
+
+In order to experimentally evaluate the performance of our algorithm,
+we generate random point sets and random query points, and for each query point
+run the algorithm 10 times. The average number of distances to the query point
+that were actually computed is the measure that we are interested in.
+
+\begin{figure}[ht]
+    \label{fig:ann_const_ratio}
+    \includegraphics[width=0.7\textwidth]{log_dependency_constant_ratio.png}
+    \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
+    points.}
+\end{figure}
+
+We used the following methods of generating random points:
+\begin{enumerate}
+    \item Uniform. Points are sampled uniformly at random from the unit cube in $\R^d$.
+    \item Normal. Points are sampled from the standard normal distribution.
+\end{enumerate}
+Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
+and from the normal distribution centered at the origin with scale 100,
+thus we get query points that are "inside" the point set and also "outside".
+We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
+the maximal number of points is $30,000$.
+In high dimensions the algorithm does not perform as well for very small epsilon,
+but for $\eps = 0.01$ even in dimension 20 just about 10 percent of the distances
+are computed for 7,000 points (while for 1,000 points more than 0.3 of the distances must be computed).
+We note, however, that the constants in asymptotics are very different for different combinations
+of distributions of query points and points in the point set.
 

### 2018-11-27 paper/draft.tex
interim

@@
 
+\begin{figure}[ht]
+    \label{fig:spanner-sparseness}
+    \includegraphics[width=0.7\textwidth]{edges_to_points_ratio_dim_2_normal_points.png}
+    \caption{Number of edges in blind spanners generated by different variants of the blind
+    algorithm. Greedy non-blind algorithm is included for comparison. The plot is for normally distributed points
+    in dimension 2, $\eps = 0.1$.}
+\end{figure}
+
+\begin{figure}[ht]
+    \label{fig:blind-greedy-only}
+    \includegraphics[width=0.7\textwidth]{blind-greedy-dims.png}
+    \caption{}
+\end{figure}
+
 We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
@@
 %as \textit{growth}. Note that
-The results are summarized in the plots \ref{plot1}, \ref{fig:spanner_plot2}.
+
+The results are summarized in the plot \ref{fig:spanner-sparseness}.
 The first plot shows the number of edges of the spanner.
@@
 the exponent seems to be bounded away from 2.
+
+As for different variants of the \textsc{BlindRandomBadRatio} algorithm,
+we note that their performance is almost the same, and the algorithm works
+significantly better than \textsc{QuasiSorted} variants, but obviously
+worse than the blind greedy variant. There is a consistent, though small,
+difference between the variants that do not force lower bounds first and
+the other two variants of the BlindRandomBadRatio.
+
+We also look at higher dimensions, but the situation there is more
+disappointing. We show the results for the best algorithm, \textsc{BlindGreedy},
+in dimensions 2, 3 and 4 in the plot \ref{fig:blind-greedy-only}.
+We can see that already in dimension 4 the best available algorithm
+for the uniformly distributed point sets the blind greedy variant
+produces almost a complete graph up to 700 points.
 

### 2018-11-27 python_version/nearest_neighbor1.py
Interim

@@ -1,0 +1,232 @@
+#!/usr/bin/env python3
+
+import joblib as jl
+import numpy as np
+import random
+from utils_euclidean import *
+from blind_spanner import *
+
+inf_dist = 10000000000.0
+
+
+class ExactSpanner:
+    def __init__(self, points):
+        self.points = points
+        self.n_points = len(points)
+        # self.dist_matrix = distance_matrix(points)
+
+    def get_distance(self, p_a, p_b):
+        return euclidean_distance(p_a, p_b)
+
+    def upper_bound_by_index(self, i, j):
+        return euclidean_distance(self.points[i], self.points[j])
+
+
+class AnnFinder:
+    def __init__(self, points, dyn_spanner):
+        self.points = points
+        self.n_points = len(points)
+        self.dyn_spanner = dyn_spanner
+
+    def discard_1(self, candidate_index):
+        discard_outer_dist = self.distance_to_query(candidate_index) * (1.0 + 1.0 / (1.0 + self.epsilon))
+        self.candidate_list = list(filter(
+            lambda idx: self.pwdist(candidate_index, idx) < discard_outer_dist, self.candidate_list))
+
+    def discard_2(self, worse_cand_index, candidate_index):
+        dw = self.distance_to_query(worse_cand_index)
+        dce = self.distance_to_query(candidate_index) / (1.0 + self.epsilon)
+        discard_small_radius = dw - dce
+        assert (discard_small_radius > 0.0)
+        discard_large_radius = dw + dce
+        discard_predicate = lambda idx: self.pwdist(worse_cand_index, idx) > discard_small_radius and self.pwdist(
+            worse_cand_index, idx) < discard_large_radius
+        self.candidate_list = list(filter(discard_predicate, self.candidate_list))
+
+    def pwdist(self, idx1, idx2):
+        return self.dyn_spanner.get_distance(self.points[idx1], self.points[idx2])
+
+    def distance_to_query(self, idx):
+        dist = self.dyn_spanner.get_distance(self.points[idx], self.query_point)
+        self.distance_computed[idx] = True
+        return dist
+
+    def find_ann(self, query_point, epsilon):
+        self.distance_computed = np.zeros(self.n_points, dtype=np.bool)
+        self.query_point = query_point
+        self.epsilon = epsilon
+        self.candidate_list = [i for i in range(self.n_points)]
+        random.shuffle(self.candidate_list)
+        cand_idx = self.candidate_list.pop()
[…]

### 2018-11-28 paper/draft.tex
add Michaels proof

@@
 \usepackage{algorithm}
-\usepackage{algpseudocode}
+\usepackage[noend]{algpseudocode}
 \usepackage{comment}
@@
 
-The algorithm that we analyze can be roughly described as follows. 
-Pick a random point $p_i$,
-compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
-to update lower bounds for other distances $\dist(p_j, q)$.
-If we see that the lower bound for $\dist(p_j, q)$ is larger than $\dist(p_i, q) / (1+\eps)$, then 
-we can discard the point $p_j$. If all points were discarded, we found an approximate nearest neighbor;
-otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
-There are two possibilities: if $\dist(p_j, q) \leq \dist(p_i, q) / (1+\eps)$,
-then we choose $p_j$ as the current candidate for an approximate nearest neighbor,
-otherwise we keep $p_i$ in this role. In any case, the new computed distance
-may also allow us to discard more than one point, as figure \ref{fig:discarded_balls} shows. In
-this figure points inside the shaded smaller ball cannot be a better candidate,
-and the same is true for the points in the shaded area outside of the larger ball.
-The precise radii of these balls are given in the following lemma.
+\begin{algorithmic}
+\label{alg:ann_blind}
+\Procedure{UpdateBounds}{$p_i, r_i$}
+    \For{$k = i + 1, \dots, n$}
+        \State {$a_k \gets \max(a_k, |\dist(p_i, p_k) - r_i|)$}
+    \EndFor
+\EndProcedure
 
-\begin{lemma}
-    Let $p_i, p_j \in P$ be such that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$,
-    and denote $r_i = \dist(p_i, q)$, $r_j = \dist(p_j, q)$.
-    Then for every $x \in B(p_j, r_j - r_i / (1+\eps))$ we have $\dist(x, q) \geq r$
-    and for every $y \notin B(p_j, r_j + r_i / (1+ \eps))$ we have $\dist(y, q) \geq r$,
-    so all points of $P$ inside the smaller ball and all points of $P$ outside the larger ball
-    can be discarded.
-\end{lemma}
-\begin{proof}
-    The triangle inequality.
-\end{proof}
+\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
+    \State {$[p_1, \dots, p_n] \gets \mbox{random permutation of }P$}
+    \State { $a_i \gets 0$}
+    \Comment {$a_i$ is lower bound for $\dist(p_i, q)$}
+    \State {$ c \gets p_1, \quad v \gets \dist(p_1, q)$}
+    \Comment {$c$ keeps the current candidate}
+    \State \Call {UpdateBounds}{$p_1, v$}
+    \For{$i = 2\dots n$}
+        \If {$a_i \geq \frac{v}{1+\eps}$}
+            \State {\textbf{continue}}
+        \Else
+            \State {Compute $r_i = \dist(p_i, q)$}
+            \State \Call{UpdateBounds}{$p_i, r_i$}
+            \If {$r_i < v$}
+                \State {$c \gets p_i,\quad v\gets r_i$}
+            \EndIf
+        \EndIf
+    \EndFor
+    \State \Return {$c, v$}
+\EndFunction
+\end{algorithmic}
+
+The algorithm that we analyze can be described as follows. 
+Fix a random permutation of the points of $P$ (to simplify notation,
+we re-index them, so the order is again $p_1, \dots, p_n$). Pick $p_1$
+as the current candidate, denote $v = r_1$ and start traversing the list of $p_i$'s.
+At each element $p_i$ look at the lower bound for $r_i$,
+if it is larger than $v / (1+\eps)$, then we can safely skip $p_i$.
+Otherwise we must compute the distance $r_i$, and we immediately
+ use it to update the lower bounds for all the next points $p_k$'s. 
+If $p_i$ turns out to be a better candidate (we compare $r_ir$ with $v$, not with $v/(1+\eps)$),
+we update the current candidate $c$ and the distance to the query point $v$ accordingly, 
+otherwise $c$ and $v$ remain unchanged.
 
@@
 
-\begin{algorithmic}
-\Procedure{Discard\_1}{$L, p_i$}
-    \State {$r_i = \dist(q, p_i)$}
-    \ForAll { $p_j \in L$}
-       \If {  $ \dist(p_i, p_j) > r_i \left(1 + \frac{1}{ 1 + \eps} \right)$ }
-           \State {Remove $p_j$ from $L$}
-       \EndIf
-    \EndFor
-\EndProcedure
-\end{algorithmic}
+%\begin{algorithmic}
+%\Procedure{Discard\_1}{$L, p_i$}
+%    \State {$r_i = \dist(q, p_i)$}
+%    \ForAll { $p_j \in L$}
+%       \If {  $ \dist(p_i, p_j) > r_i \left(1 + \frac{1}{ 1 + \eps} \right)$ }
+%           \State {Remove $p_j$ from $L$}
+%       \EndIf
+%    \EndFor
+%\EndProcedure
+%\end{algorithmic}
 
-\begin{algorithmic}
-\Procedure{Discard\_2}{$L, w, c$}
-    \State {$r_w = \dist(q, w)$}
-    \State {$r_c = \dist(q, w)$}
-    \Comment {Distances $r_w$ and $r_c$ are cached}
-    \ForAll { $p_j \in L$}
-       \If { $ \dist(w, p_j) < r_w - r_c / ( 1 + \eps) $ }
-           \State {Remove $p_j$ from $L$}
-           \Comment {In particular, the worse candidate $w$ will be removed here}
-       \EndIf
-       \If {  $ \dist(w, p_j) > r_w + r_c / ( 1 + \eps) $ }
-           \State {Remove $p_j$ from $L$}
-       \EndIf
-    \EndFor
-\EndProcedure
-\end{algorithmic}
+%\begin{algorithmic}
+%\Procedure{Discard\_2}{$L, w, c$}
+%    \State {$r_w = \dist(q, w)$}
+%    \State {$r_c = \dist(q, w)$}
+%    \Comment {Distances $r_w$ and $r_c$ are cached}
+%    \ForAll { $p_j \in L$}
+%       \If { $ \dist(w, p_j) < r_w - r_c / ( 1 + \eps) $ }
+%           \State {Remove $p_j$ from $L$}
+%           \Comment {In particular, the worse candidate $w$ will be removed here}
+%       \EndIf
+%       \If {  $ \dist(w, p_j) > r_w + r_c / ( 1 + \eps) $ }
+%           \State {Remove $p_j$ from $L$}
+%       \EndIf
+%    \EndFor
+%\EndProcedure
+%\end{algorithmic}
 
-\begin{algorithmic}
-\label{alg:ann_blind}
-\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
-    \State {Initialize array $L$ with a random permutation of $P$}
-    \State {$ c \gets \mbox{the first element of }L$}
-    \State {Compute $r_i = \dist(q, p_i)$}
-    \State \Call{Discard\_1}{$L, c$}
-    \While {$|L| > 1$}
-        \ForAll {$p_j \in L $}
-            \State {Compute $r_j = \dist(p_j, q)$}
-            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
-            \If {$r_j < r_i$ }
-                \State {$ c \gets p_j $}
-                \State \Call{Discard\_1}{$L, c$}
-                \State {$ w \gets p_i $}
-            \Else
-                \State {$ w \gets p_i $}
-            \EndIf
-            \State \Call{Discard\_2}{$L, w, c$}
-        \EndFor
-    \EndWhile
-    \State \Return {the unique element left in $L$}
-\EndFunction
-\end{algorithmic}
+%\begin{algorithmic}
+%\label{alg:ann_blind}
+%\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
+%    \State {Initialize array $L$ with a random permutation of $P$}
+%    \State {$ c \gets \mbox{the first element of }L$}
+%    \State {Compute $r_i = \dist(q, p_i)$}
+%    \State \Call{Discard\_1}{$L, c$}
+%    \While {$|L| > 1$}
+%        \ForAll {$p_j \in L $}
+%            \State {Compute $r_j = \dist(p_j, q)$}
+%            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
+%            \If {$r_j < r_i$ }
+%                \State {$ c \gets p_j $}
+%                \State \Call{Discard\_1}{$L, c$}
+%                \State {$ w \gets p_i $}
+%            \Else
+%                \State {$ w \gets p_i $}
+%            \EndIf
+%            \State \Call{Discard\_2}{$L, w, c$}
+%        \EndFor
+%    \EndWhile
+%    \State \Return {the unique element left in $L$}
+%\EndFunction
+%\end{algorithmic}
 
@@
 
-The proof is based on the following lemma.
-\begin{lemma}
-    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
-    such that after computing at most $C$ distances $\dist(p_k, q)$
-    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
-    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
-\end{lemma}
 \begin{proof}
-    First, all points that are not contained in the ball $B(p_i, 2r_i)$
-    are discarded without computing the distance to them.
-    Now we have to examine all points $p_j$ inside this ball. 
-    Suppose that we cannot discard the current point $p_j$ based
-    on the lower bound for $r_j$, then we must compute the distance $r_j$.
-    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
-    The latter case means that we discard the point $p_j$,
-    but we also discard all points of $P$ that are contained in some ball around it.
-    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
-    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
-    In the figure this radius corresponds to the width of the annulus surrounding $q$,
-    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
-    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
-    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
-    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
-    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
-    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
-    This means that even in the worst case after we have computed $T$ distances, we have discarded all
-    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
-    do not require distance computation. Indeed, each of these points lies in
-    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
-    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
-    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
-    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
-    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
-    because some of the points that are inside the annulus can be discarded, 
-    if there is a point from $P$ close to them but outside of the annulus. 
+Let $i_1,...,i_k$ denote the indices of the minima in the sequence $r_1, \dots, r_n$
+    (recall that this is a random sequence). Then $k=O(log n)$,
+    by the standard backwards analysis argument.
+
+    Consider two consecutive minima  $s:=i_j$ and $t:=i_{j+1}$. We claim that in the sequence
+$p_s,...,p_{t-1}$
+only a constant number of distances are computed, irrespective of its length.
+
+To see that, let $c_1,...,c_n$ be the candidates after the $i$-th step of
+the for loop of the algorithm. We have no guarantee that $c_s=p_s$, because $p_s$ it is possible
+    that $p_s$ was discarded without computing $r_s$ (see the picture). However, we know that
+$d(c_s,q)<=(1+eps) r_s$.
+
+Indeed, suppose that $c_s \neq p_s$ and $r_s < d(c_s,q)/(1+eps)$. In this
+    case $p_s$ cannot be discarded 
+by its lower bound, and hence the distance to $p_s$ would
+have been queried, but this distance is a minimum among $r_1, \dots, r_s$,
+    and then we must reset $c_s$ to be $p_s$, a contradiction.
+
+It follows that any point in the sequence $p_s,...,p_{t-1}$ for which a
+distance is queried must lie in the ball around $c_s$ of radius at most $2(1+eps)r_s$.
+
+Moreover, we know that for any index $j=s,\dots, t-1$ it holds
+$d(c_j,q)>=r_s$.
+
+Indeed, assume the opposite. $c_j$ is equal to $p_k$ for some $k<t$. If $k<s$,
+that is, the candidate came from before, than $p_s$ would not be in the
+minimum sequence. If $s<k<t$, then it follows that at least one of the
+elements in $p_{s+1},\dots,p_k$ is in the minimum sequence, contradicting
+the assumption that $p_s$ and $p_t$ are consecutive minima.
+
+Hence, it follows that whenever a distance is queried, we can exclude a
+ball of radius $\eps r_s$ around the query point. All queried points must
+hence have pairwise distance of at least $ \eps r_s / 2$. Hence, the number of
+queried points is by the packing lemma bounded by some constant that depends only on $\eps$ and the dimension. 
 \end{proof}
 
-Now the theorem follows from the lemma by standard argument
-of the backwards analysis.
+%\begin{proof}
+%    We start with the lemma of purely combinatorial (or probabilistic) nature.
+%\begin{lemma}
+%    Let $z$ be a random permutation of $n$ distinct real numbers,
+%    $z = X_{1}, \dots, X_{n}$, and denote $x_t = \min \Set{X_1, \dots, X_t}$.
+%    The expected number of positions $t$ where the sequence $X_t$ changes is $O(n)$.
+%\end{lemma}
+%    \begin{proof}
+%        The proof is easily obtained with the standard backwards analysis technique and can be found in the appendix.
+%        %Define random variable $Y_i$ to be 1, if $X_{i-1} = X_{i}$, and 0 otherwise.
+%        %We must estimate $Y = \sum_{i = 2}^n Y_i$, and 
+%        %\[
+%        %    E[Y] = \sum_{i = 2}^n E[Y_i] = \sum_{i = 2}^n P(Y_i = 1).
+%        %\]
+%        %It remains to observe than $P(Y_i) = 1 / (n - i)$, so the expectation
+%        %of $Y$ is the sum of the 
+%    \end{proof}
+
+%\begin{lemma}
+%    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
+%    such that after computing at most $C$ distances $\dist(p_k, q)$
+%    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
+%    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
+%\end{lemma}
+%\begin{proof}
+%    First, all points that are not contained in the ball $B(p_i, 2r_i)$
+%    are discarded without computing the distance to them.
+%    Now we have to examine all points $p_j$ inside this ball. 
+%    Suppose that we cannot discard the current point $p_j$ based
+%    on the lower bound for $r_j$, then we must compute the distance $r_j$.
+%    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
+%    The latter case means that we discard the point $p_j$,
+%    but we also discard all points of $P$ that are contained in some ball around it.
+%    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
+%    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
+%    In the figure this radius corresponds to the width of the annulus surrounding $q$,
+%    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
+%    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
+%    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
+%    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
+%    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
+%    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
+%    This means that even in the worst case after we have computed $T$ distances, we have discarded all
+%    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
+%    do not require distance computation. Indeed, each of these points lies in
+%    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
+%    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
+%    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
+%    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
+%    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
+%    because some of the points that are inside the annulus can be discarded, 
+%    if there is a point from $P$ close to them but outside of the annulus. 
+%\end{proof}
+
+%Now the theorem follows from the lemma by standard argument
+%of the backwards analysis.
+
+%\end{proof}
 

### 2018-11-29 paper/draft.tex
interim

--- 2018-11-27 paper/draft.tex
@@
 
-It is known that asymptotically optimal spanner can be constructed by the following greedy algorithm: 
-start with an empty graph on vertices $p_i$,
-sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
-until it becomes a $1+\eps$-spanner.
+In this section $P = \Set{p_1, \dots, p_n}$ is a given set of points in a metric space;
+our goal is to construct a $(1+\eps)$-spanner on $P$. All algorithms that we consider 
+build a spanner by starting with an empty graph on vertices $P$, to which they iteratively
+add edges. We shall denote this graph (at any stage of the algorithm) by $G$.
 
- In our setting this is the most expensive approach, because it requires
- computing all pairwise distances first. We propose an alternative approach (or rather a family of algorithms), which we call
+It is known that an asymptotically optimal spanner can be constructed by the following greedy algorithm: 
+sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to $G$
+until it becomes a $1+\eps$-spanner. This algorithm in its most naive form is expensive in terms of $n$,
+because it requires computing many shortest paths, and it is also the most expensive approach
+in our setting, because it requires
+ computing all pairwise distances first. We propose an alternative algorithm (or rather a family of algorithms), 
+ which we call
  \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
- before it decides to add the edge $(p_i, p_j)$.
+ before it decides to add the edge $(p_i, p_j)$ to $G$.
  The algorithm also maintains lower and upper bounds
- for all pairwise distances, which are updated after a new edge is added. and the stopping criterion
+ for all pairwise distances, which are updated after a new edge is added, and the stopping criterion
  is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
@@
 \Function{BlindSpanner}{$P, \eps$}
-    \State {$E \gets \emptyset$}
+    \State {$G \gets (P, \emptyset)$}
     \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
@@
     \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
-    \State {$(i,j) \gets GetEdgeToAdd$}
+        \State {$(i,j) \gets$ \Call {GetEdgeToAdd}{}
         \State {$v \gets \dist(p_i, pj)$}
-    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
+        \State {Add weighted edge $(p_i, p_j, v)$ to $G$}
         \State {$UpdateBounds(i, j, v)$}
     \EndWhile
+    \State \Return {$G$}
 \EndFunction
@@
 and $\infty$ is larger than any real number,
-thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection,
+thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is described in the next subsection,
 now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
@@
 between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
+Combined with the natural requirement that the function \textsc{GetEdgeToAdd} never
+returns the same edge twice, these properties ensure that the algorithm terminates.
 The most important part of this algorithm is the way of choosing the next edge to be added, and
@@
 First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$. 
+Now observe that for a pair $(k, \ell)$ there are three possibilities:
+either the shortest path from $p_k$ to $p_\ell$ in $G$
 To update the upper bound of some entry $b_{k,\ell}$,
 we observe that the shortest path from $p_k$ to $p_\ell$ might now
-go through the new edge. Hence, we update
+go through the new edge.
+
+Hence, we update
 \[
@@
 Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
-time per entry.
+time.
 
@@
 so we only present results for the \textbf{uniform} point set in dimension 2.
-%Let $e$ denote the number of edges of a spanner on $n$ vertices;
-%in the plots we visualize the ratios $e / { n \choose 2}$ and $e / n$.
-%We refer to the first ratio as \textit{density} and to the second one
-%as \textit{growth}. Note that
 
@@
 otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
-There are two possibilities: if $\dist(p_j, q) \leq \dist(p_i, q) / (1+\eps)$,
+There are two possibilities: if $\dist(p_j, q) < \dist(p_i, q) / (1+\eps)$,
 then we choose $p_j$ as the current candidate for an approximate nearest neighbor,

### 2018-11-29 cpp_version/src/main.cpp

@@ -184,9 +184,7 @@ MatrixR get_distance_matrix(const arma::mat& points)
 int main(int argc, char** argv)
 {
     auto console = spd::stdout_color_mt("console");
-    auto exp_logger = spd::basic_logger_st("experiment_logger", "experiment_log-blind-greedy.txt");
     console->set_level(spd::level::info);
-    exp_logger->set_pattern("%v");
 
 #if 0
     if (argc < 4) {
@@ -222,7 +220,7 @@ int main(int argc, char** argv)
 
     if (argc < 3) {
         std::cout << "Usage: " << argv[0]
-                  << " input_name epsilon"
+                  << " input_name epsilon [logname]"
                   << std::endl;
         return 0;
     }
@@ -232,6 +230,11 @@ int main(int argc, char** argv)
 
     double eps = atof(argv[arg_idx++]);
     console->info("Reading from file {}, epsilon = {}", argv[1], argv[2]);
+
+    std::string log_name  = (argc > 3) ? argv[arg_idx++] : "experiment_log.txt";
+    console->info("log_name = {}, epsilon = {}", log_name, eps);
+    auto exp_logger = spd::basic_logger_st("experiment_logger", log_name);
+    exp_logger->set_pattern("%v");
 
     read_distance_matrix(dist_matrix, max_dist, min_dist, dist_name);
 
@@ -314,59 +317,66 @@ int main(int argc, char** argv)
 
 #else
 
-//    {
-//        DynamicSpannerR greedy_spanner(dist_matrix);
-//        greedy_spanner.construct_greedy_eps_spanner(eps);
-//        console->info("{};{}", dist_name, greedy_spanner.get_statistics());
-//        exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
-//    }
-//
-//    {
-//        DynamicSpannerR quasi_greedy_spanner(dist_matrix);
-//        quasi_greedy_spanner.construct_blind_quasi_sorted_greedy_eps_spanner(eps);
-//        console->info("{};{}", dist_name, quasi_greedy_spanner.get_statistics());
-//        exp_logger->info("{};{}", dist_name, quasi_greedy_spanner.get_statistics());
-//    }
-//
-//    {
-//        DynamicSpannerR quasi_shaker_spanner(dist_matrix);
-//        quasi_shaker_spanner.construct_blind_quasi_sorted_shaker_eps_spanner(eps);
[…]

### 2018-11-29 python_version/ann_experiment_data/plots.py

@@ -1,53 +1,116 @@
+#!/usr/bin/env python3
+
+import numpy as np
 import pandas as pd
 import sys
 import matplotlib.pyplot as plt
 
-
 if __name__ == "__main__":
-    # ds = pd.read_csv("ann_results.txt", header=0, sep=';', dtype = { 'Epsilon' : object })
-    ds = pd.read_csv("ann_results.txt", header=0, sep=';', dtype = { 'Epsilon' : object })
-    index_cols = ['Dimension', 'Epsilon',  'point_gen_method', 'query_method', 'Npoints']
+    if True:
+        ds = pd.read_csv("averaged_ann_data.csv", header=0, sep=';', dtype={'epsilon': object})
+        # ds.drop(columns=["epsilon", "point_gen_method", "query_method"], inplace=True)
+        index_cols = ['dim']
+        ds.set_index(index_cols, inplace=True)
 
-    ds.drop(columns=['MinFraction', 'MaxFraction', 'MinNumberOfDistances', 'MaxNumberOfDistances'], inplace=True)
+        ds_to_plot2 = ds.loc[(2)]
+        ds_to_plot5 = ds.loc[(5)]
+        ds_to_plot10 = ds.loc[(10)]
+        # ds_to_plot20 = ds.loc[(20)]
+
+        ds_to_plot2.set_index(["Points"], inplace=True)
+        ds_to_plot5.set_index(["Points"], inplace=True)
+        ds_to_plot10.set_index(["Points"], inplace=True)
+
+
+        ds_to_plot2.rename(index=str, columns={"Computed distances": "Computed distances (dim 2)"}, inplace=True)
+        ds_to_plot5.rename(index=str, columns={"Computed distances": "Computed distances (dim 5)"}, inplace=True)
+        ds_to_plot10.rename(index=str, columns={"Computed distances": "Computed distances (dim 10)"}, inplace=True)
+        # ds_to_plot20.set_index(["Points"], inplace=True)
+
+        ds_all_dims = ds_to_plot2.join(ds_to_plot5)
+        ds_all_dims = ds_all_dims.join(ds_to_plot10)
+        # ds_all_dims = ds_all_dims.join(ds_to_plot20, rsuffix=" (dim 20)")
+
+        print(ds_all_dims.index.values)
+        ds_all_dims.plot(use_index = True)
+
+        plt.show()
+        sys.exit(0)
+
[…]

### 2018-11-29 python_version/ann_experiment_data/plots1.py

@@ -1,0 +1,56 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import pandas as pd
+import sys
+import matplotlib.pyplot as plt
+
+
+if __name__ == "__main__":
+    ds = pd.read_csv("more_ann_results.txt", header=0, sep=';', dtype = { 'Epsilon' : object })
+    # ds = pd.read_csv("ann_results_single.txt", header=0, sep=';', dtype = { 'Epsilon' : object })
+    index_cols = ['Dimension', 'Epsilon',  'point_gen_method', 'query_method', 'Npoints']
+
+    ds.drop(columns=['MinFraction', 'MaxFraction'], inplace=True)
+
+    print(ds.describe())
+    print("********************************************************************************")
+    print(ds.head(20))
+    print("********************************************************************************")
+    ds.set_index(index_cols, inplace=True)
+
+    ds_to_plot = ds.loc[(6, '0.01', 'get_uniform_points', 'query_2')]
+    ds_to_plot.reset_index(inplace=True)
+    x = ds_to_plot.Npoints
+    y6_01 = ds_to_plot.AvgNumberOfDistances
+
+    ds_to_plot = ds.loc[(6, '0.005', 'get_uniform_points', 'query_2')]
+    ds_to_plot.reset_index(inplace=True)
+    y6_005 = ds_to_plot.AvgNumberOfDistances
+
+    ds_to_plot = ds.loc[(6, '0.001', 'get_uniform_points', 'query_2')]
+    ds_to_plot.reset_index(inplace=True)
+    y6_001 = ds_to_plot.AvgNumberOfDistances
+
+    ds_to_plot = ds.loc[(6, '0.0005', 'get_uniform_points', 'query_2')]
+    ds_to_plot.reset_index(inplace=True)
+    y6_0005 = ds_to_plot.AvgNumberOfDistances
+
+    ds_to_plot = ds.loc[(6, '0.0001', 'get_uniform_points', 'query_2')]
+    ds_to_plot.reset_index(inplace=True)
+    y6_0001 = ds_to_plot.AvgNumberOfDistances
+
+    # fit = np.polyfit(x, y, 1)
+    # fit_fn = np.poly1d(fit)
+    # plt.plot(x, y, 'x', x, fit_fn(x), '--r')
+
+    plt.plot(x, y6_01, x, y6_005, x, y6_001, x, y6_0005, x, y6_0001)
[…]

### 2018-11-29 python_version/blind_spanner.py

@@ -233,14 +233,14 @@ def distance_matrix(points):
         for j in range(n_points):
             if i <= j:
                 continue
-            result[i][j] = result[j][i] = euclidean_distance(points[i, :], points[j, :])
+            result[i][j] = result[j][i] = get_lp_distance(points[i, :], points[j, :])
     return result
 
 
 def run_experiment(dim, n_points, epsilon, points_generator, ps_gen_args):
     points = points_generator(n_points, dim, *ps_gen_args)
     result = []
-    gs = BlindSpanner(points, inf_dist, dist_function=euclidean_distance, dist_matrix=None)
+    gs = BlindSpanner(points, inf_dist, dist_function=get_lp_distance, dist_matrix=None)
     gs.build_random(epsilon)
     n_gs_edges = gs.number_of_edges()
     result.append(ExperimentResult(dim, n_points, epsilon, points_generator.__name__, "blind_random", n_gs_edges))
@@ -251,7 +251,7 @@ def run_experiment(dim, n_points, epsilon, points_generator, ps_gen_args):
 def run_experiment_quasi_greedy(dim, n_points, epsilon, points_generator, ps_gen_args):
     points = points_generator(n_points, dim, *ps_gen_args)
     result = []
-    gs = BlindSpanner(points, inf_dist, dist_function=euclidean_distance, dist_matrix=None)
+    gs = BlindSpanner(points, inf_dist, dist_function=get_lp_distance, dist_matrix=None)
     gs.build_quasi_sorted_greedy(epsilon)
     n_gs_edges = gs.number_of_edges()
     result.append(ExperimentResult(dim, n_points, epsilon, points_generator.__name__, "blind_quasi_sorted_greedy", n_gs_edges))
@@ -268,7 +268,7 @@ def run_experiment_quasi_greedy(dim, n_points, epsilon, points_generator, ps_gen
 def run_experiment_quasi_shaker(dim, n_points, epsilon, points_generator, ps_gen_args):
     points = points_generator(n_points, dim, *ps_gen_args)
     result = []
-    gs = BlindSpanner(points, inf_dist, dist_function=euclidean_distance, dist_matrix=None)
+    gs = BlindSpanner(points, inf_dist, dist_function=get_lp_distance, dist_matrix=None)
     gs.build_quasi_sorted_shaker(epsilon)
     n_gs_edges = gs.number_of_edges()
     result.append(ExperimentResult(dim, n_points, epsilon, points_generator.__name__, "blind_quasi_sorted_shaker", n_gs_edges))

### 2018-11-29 python_version/nearest_neighbor.py

@@ -4,7 +4,7 @@ import joblib as jl
 import numpy as np
 import random
 from utils_euclidean import *
-from blind_greedy_spanner import *
+from blind_spanner import *
 
 inf_dist = 10000000000.0
 
@@ -16,10 +16,74 @@ class ExactSpanner:
         # self.dist_matrix = distance_matrix(points)
 
     def get_distance(self, p_a, p_b):
-        return euclidean_distance(p_a, p_b)
+        return get_lp_distance(p_a, p_b)
 
     def upper_bound_by_index(self, i, j):
-        return euclidean_distance(self.points[i], self.points[j])
+        return get_lp_distance(self.points[i], self.points[j])
+
+
+class AnnFinder:
+    def __init__(self, points, dyn_spanner):
+        self.points = points
+        self.n_points = len(points)
+        self.dyn_spanner = dyn_spanner
+
+    def discard_1(self, candidate_index):
+        discard_outer_dist = self.distance_to_query(candidate_index) * (1.0 + 1.0 / (1.0 + self.epsilon))
+        self.candidate_list = list(filter(
+            lambda idx: self.pwdist(candidate_index, idx) < discard_outer_dist, self.candidate_list))
+
+    def discard_2(self, worse_cand_index, candidate_index):
+        dw = self.distance_to_query(worse_cand_index)
+        dce = self.distance_to_query(candidate_index) / (1.0 + self.epsilon)
+        discard_small_radius = dw - dce
+        assert (discard_small_radius > 0.0)
+        discard_large_radius = dw + dce
+        discard_predicate = lambda idx: self.pwdist(worse_cand_index, idx) > discard_small_radius and self.pwdist(
+            worse_cand_index, idx) < discard_large_radius
+        self.candidate_list = list(filter(discard_predicate, self.candidate_list))
+
+    def pwdist(self, idx1, idx2):
+        return self.dyn_spanner.get_distance(self.points[idx1], self.points[idx2])
+
+    def distance_to_query(self, idx):
+        dist = self.dyn_spanner.get_distance(self.points[idx], self.query_point)
+        self.distance_computed[idx] = True
+        return dist
+
+    def find_ann(self, query_point, epsilon):
+        self.distance_computed = np.zeros(self.n_points, dtype=np.bool)
+        self.query_point = query_point
+        self.epsilon = epsilon
+        self.candidate_list = [i for i in range(self.n_points)]
[…]

### 2018-11-29 python_version/nearest_neighbor1.py

@@ -16,10 +16,10 @@ class ExactSpanner:
         # self.dist_matrix = distance_matrix(points)
 
     def get_distance(self, p_a, p_b):
-        return euclidean_distance(p_a, p_b)
+        return get_lp_distance(p_a, p_b)
 
     def upper_bound_by_index(self, i, j):
-        return euclidean_distance(self.points[i], self.points[j])
+        return get_lp_distance(self.points[i], self.points[j])
 
 
 class AnnFinder:

### 2018-11-29 python_version/save_in_text_files.py

@@ -3,14 +3,24 @@
 import sys
 import numpy as np
 from utils_euclidean import *
+import joblib as jl
 
-def save_to_files(point_method, dim, n_points, point_file_name, matrix_file_name = None):
+
+def save_to_files(point_method, dim, n_points, point_file_name=None, matrix_file_name=None, norm = None):
+    if point_file_name == None:
+        point_file_name = "points_{}_{}_{}.txt".format(dim, n_points, point_method)
     if matrix_file_name == None:
         matrix_file_name = "dist_matrix_" + point_file_name
     pm_dic = {"unif": get_uniform_points,
               "normal": get_points,
               "clustered": get_clustered_points,
-              "exp": get_exponential_points}
+              "exp": get_exponential_points,
+              "beta_05_05": get_beta_05_05,
+              "beta_5_1": get_beta_5_1,
+              "beta_1_3": get_beta_1_3,
+              "beta_2_2": get_beta_2_2,
+              "beta_2_5": get_beta_2_5,
+              }
     if not point_method in pm_dic:
         print("Unknown point_method")
         exit(1)
@@ -19,17 +29,70 @@ def save_to_files(point_method, dim, n_points, point_file_name, matrix_file_name
     dist_matr = distance_matrix(points)
     np.savetxt(matrix_file_name, dist_matr, "%15.7f")
 
+def save_to_files(point_method, dim, n_points, norm = 2.0, point_file_name=None, matrix_file_name=None):
+    if point_file_name == None:
+        point_file_name = "points_{}_{}_{}.txt".format(dim, n_points, point_method)
+    if matrix_file_name == None:
+        matrix_file_name = "dist_matrix_norm_{}_{}".format(norm,  point_file_name)
+    pm_dic = {"unif": get_uniform_points,
+              "normal": get_points,
+              "clustered": get_clustered_points,
+              "exp": get_exponential_points,
+              "beta_05_05": get_beta_05_05,
+              "beta_5_1": get_beta_5_1,
+              "beta_1_3": get_beta_1_3,
+              "beta_2_2": get_beta_2_2,
+              "beta_2_5": get_beta_2_5,
+              }
+    if not point_method in pm_dic:
+        print("Unknown point_method")
+        exit(1)
+    points = pm_dic[point_method](n_points, dim)
+    np.savetxt(point_file_name, points, "%15.7f")
[…]

### 2018-11-29 python_version/spanner_experiment_data/plots.py

@@ -31,6 +31,16 @@ import matplotlib.pyplot as plt
 
 
 if __name__ == "__main__":
+
+    # ds = pd.read_csv("n_edges_blind_greedy_dimensions.csv", header  = 0, sep = ';')
+    # ds = ds.set_index(["# points"], drop=True)
+    # ds = ds.sort_index(ascending=True)
+    # print(ds.index)
+    # print(ds.head(n=20))
+    # ds.plot()
+    # plt.show()
+    # sys.exit(0)
+
     ds = pd.read_csv("spanner_cpp_logs.txt", header=0, sep=';')
     index_cols = ['Dim', 'point_gen_method', 'epsilon', 'input_file', 'Points']
 
@@ -55,7 +65,22 @@ if __name__ == "__main__":
     # sys.exit(0)
 
     ds_non_blind_greedy.drop(columns=['spanner_method'], inplace=True)
-    ds_blind_greedy.drop(columns=['spanner_method'], inplace=True)
+    ds_blind_greedy.drop(columns=['spanner_met
+        x = ds_to_plot2.n_points
+        y = ds_to_plot2.dist_num
+
+
+        y = y / np.log(x)
+        # x = np.log(x)
+        print(x)
+        print(y)
+        fit = np.polyfit(x, y, 1)
+        fit_fn= np.poly1d(fit)
+        print(fit_fn(x))
+        plt.xlabel("#points")
+        plt.ylabel("#computed distances / log(#points)")
+        plt.plot(x, y)
+        # plt.plot(x, y, 'ro', x, fit_fn(x), '--k')hod'], inplace=True)
     ds_blind_random.drop(columns=['spanner_method'], inplace=True)
     ds_quasi_greedy.drop(columns=['spanner_method'], inplace=True)
     ds_quasi_shaker.drop(columns=['spanner_method'], inplace=True)
@@ -78,9 +103,14 @@ if __name__ == "__main__":
     #          "edges_to_points_ratio_qs", "edges_to_points_ratio_qg"]
 
     ycols = ["Edges (non-blind greedy)", "Edges (blind greedy)", "Edges (blind random bad ratio, force lower bound)", "Edges (blind random bad ratio)"]
-    ycols = ["Edges to points ratio (non-blind greedy)", "Edges to points ratio (blind greedy)", "Edges to points ratio (blind random bad ratio, force lower bound)", "Edges to points ratio (blind random bad ratio)"]
+    ycols = ["Edges (blind random bad ratio, force lower bound)",
[…]

### 2018-11-29 python_version/utils_euclidean.py

@@ -3,8 +3,8 @@
 import sys
 import numpy as np
 
-def euclidean_distance(a, b):
-    return np.linalg.norm(a - b)
+def get_lp_distance(a, b, norm = 2.0):
+    return np.linalg.norm(a - b, ord=norm)
 
 def get_points(n_points, dim):
     return np.random.randn(n_points, dim)
@@ -38,14 +38,29 @@ def get_clustered_points(n_points, dim, n_clusters = -1):
     return diffs + cluster_centres
 
 
-def distance_matrix(points):
+def get_beta_05_05(n_points, dim):
+    return np.random.beta(0.5, 0.5, (n_points, dim))
+
+def get_beta_5_1(n_points, dim):
+    return np.random.beta(5, 1, (n_points, dim))
+
+def get_beta_1_3(n_points, dim):
+    return np.random.beta(1, 3, (n_points, dim))
+
+def get_beta_2_2(n_points, dim):
+    return np.random.beta(2, 2, (n_points, dim))
+
+def get_beta_2_5(n_points, dim):
+    return np.random.beta(2, 5, (n_points, dim))
+
+def distance_matrix(points, norm = 2.0):
     n_points = len(points)
     result = np.zeros((n_points, n_points))
     for i in range(n_points):
         for j in range(n_points):
             if i <= j:
                 continue
-            result[i][j] = result[j][i] = euclidean_distance(points[i, :], points[j, :])
+            result[i][j] = result[j][i] = get_lp_distance(points[i, :], points[j, :], norm)
     return result
 
 

### 2018-11-29 paper/draft.tex

--- 2018-11-28 paper/draft.tex
@@
 
-\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
-    \State {$[p_1, \dots, p_n] \gets \mbox{random permutation of }P$}
-    \State { $a_i \gets 0$}
-    \Comment {$a_i$ is lower bound for $\dist(p_i, q)$}
-    \State {$ c \gets p_1, \quad v \gets \dist(p_1, q)$}
-    \Comment {$c$ keeps the current candidate}
-    \State \Call {UpdateBounds}{$p_1, v$}
-    \For{$i = 2\dots n$}
-        \If {$a_i \geq \frac{v}{1+\eps}$}
-            \State {\textbf{continue}}
-        \Else
-            \State {Compute $r_i = \dist(p_i, q)$}
-            \State \Call{UpdateBounds}{$p_i, r_i$}
-            \If {$r_i < v$}
-                \State {$c \gets p_i,\quad v\gets r_i$}
-            \EndIf
-        \EndIf
-    \EndFor
-    \State \Return {$c, v$}
-\EndFunction
-\end{algorithmic}
+The algorithm that we analyze can be roughly described as follows. 
+Pick a random point $p_i$,
+compute $\dist(p_i, q)$ and use this distance combined with pairwise distances $\dist(p_i, p_j)$
+to update lower bounds for other distances $\dist(p_j, q)$.
+If we see that the lower bound for $\dist(p_j, q)$ is larger than $\dist(p_i, q) / (1+\eps)$, then 
+we can discard the point $p_j$. If all points were discarded, we found an approximate nearest neighbor;
+otherwise pick a random point $p_j$ from the remaining ones and compute $\dist(p_j, q)$.
+There are two possibilities: if $\dist(p_j, q) < \dist(p_i, q) / (1+\eps)$,
+then we choose $p_j$ as the current candidate for an approximate nearest neighbor,
+otherwise we keep $p_i$ in this role. In any case, the new computed distance
+may also allow us to discard more than one point, as figure \ref{fig:discarded_balls} shows. In
+this figure points inside the shaded smaller ball cannot be a better candidate,
+and the same is true for the points in the shaded area outside of the larger ball.
+The precise radii of these balls are given in the following lemma.
 

### 2018-11-29 paper/draft.tex
paper

--- 2018-11-28 paper/draft.tex
@@
 %\newcommand{\wlog}{w.l.o.g.\ }
-\newcommand{\eps}{\epsilon}
+\newcommand{\eps}{\varepsilon}
 \newcommand{\orient}{O}
@@
 
+\newcommand{\metricspace}{\mathcal{M}}
+\newcommand{\approxmetric}{\gamma}
 \newcommand{\parent}{\mathrm{par}}
@@
 
-\title{Geometric Filters for Metric Spaces}
+\title{Metric Spaces with Expensive Distances}
 \author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
@@
 
+\begin{abstract}
+
+In algorithms for finite metric spaces, it is common
+to assume that the distance between two points
+can be computed in constant time, and complexity
+bounds are expressed only in terms of the number
+of points of the metric space.
+We introduce a different model where we assume
+that the computation of a single distance is
+an expensive operation and consequently, the goal is
+to minimize the number of such distance queries.
+This model is motivated by metric spaces
+that appear in the context of topological data analysis.
+
+We consider two standard operations on metric spaces,
+namely the construction of an $\epsilon$-spanner
+and the computation of an approximate nearest
+neighbor for a given query point.
+In both cases, we partially explore the metric space
+through distance queries and infer lower and upper bounds
+for yet unexplored distances through triangle inequality.
+For spanners, we evaluate several exploration strategies
+through extensive experimental evaluation.
+For approximate nearest neighbors, we prove that our
+strategy returns an approximate nearest neighbor
+after a logarithmic number of distance queries.
+\end{abstract}
+
 \section{Introduction}
 
-\paragraph{Motivation.}
+\subsection{Motivation.}
 
@@
 in which computing a single distance between points can be very costly.
-More specifically,
-we consider the problem of finding an approximate nearest neighbor
-of some point $q$ among the given $n$ points, and the problem of finding a $t$-spanner,
-i.e., a graph that approximates the metric on the given set of points,
-and we not so much interested in reducing the complexity of the algorithm,
-but rather in reducing the number of distances computed.
 The motivating example for us was Topological Data Analysis, where computation
@@
 
-We believe that the scenario when the input consists of a relatively small number of points, 
-e.g., of order $10^2-10^3$, but the complexity of computing a distance is rather large,
-is plausible (for the persistence diagrams, the size of 0-dimensional ones can be huge,
-because in dimension 0 there are especially efficient algorithms that can handle much
-larger inputs, so it is possible to produce diagrams with millions of  points in each).
-In such a situation, the natural approach would be to avoid computing the
-distance between two points if we do not need it,
-
-We demonstrate that the number of distance computations can be further
-reduced by a careful re-implementation of the underlying geometric algorithms.
-Our improvement is based on the following simple observation: many geometric
-constructions (such as cover tree and WSPD constructions) require distance
-computations only in the form of the following predicate: 
-``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
-We show that using the information of the already computed distances and
-the triangle inequality, we can answer the above predicate in many cases
-without computing the actual distance.
-
-We describe our idea with an example. In Figure~\ref{fig:1st_example},
-we see a weighted graph on $P$ whose edges are the distances that have
-been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
-The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\dist(p_1,p_2)\geq 3$:
-otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-would be shorter than the edge $(p_3,p_4)$, again contradicting
-triangle inequality.
-Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
-for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
-distance computation is necessary. This approach is reminiscent to the usage
-of \emph{numerical filters} in geometric computing in the CGAL library,
-where geometric predicates in exact number types are first evaluated using 
-certified floating-point arithmetic, and exact computation is only
-invoked if the numerical evaluation did not arrive at a decisive result
-(e.g.~\cite{bbp-interval,kerber-phd}).
-
 Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
-in a metric space $(\distspace,\dist)$, 
+in a metric space $(\metricspace,\dist)$, 
 consider the following standard operations:
@@
 \begin{description}
-\item[Nearest Neighbor] Given a point $q\in\distspace$,
+\item[Nearest Neighbor] Given a point $q\in\metricspace$,
 find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
@@
 
-\item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
+\item[Metric approximation] Given $\eps>0$, compute a metric $\approxmetric$
 such that for all $1\leq i,j\leq n$,
 %
-\[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]
+\[\dist(p_i,p_j)\leq\approxmetric(p_i,p_j)\leq (1+\eps)\dist(p_i,p_j).\]
 
-By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
-such that $E_{ij}=\adist(p_i,p_j)$.
+By ``computing'' $\approxmetric$, we mean to compute a $n\times n$-matrix $E$ 
+such that $E_{ij}=\approxmetric(p_i,p_j)$.
 \end{description}
 
-Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
-for any two points in $\distspace$ in worst-case complexity $\complexity$.
-Then trivially, the above problems can be solved (exactly) in
-$O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
+% We believe that the scenario when the input consists of a relatively small number of points, 
+% e.g., of order $10^2-10^3$, but the complexity of computing a distance is rather large,
+% is plausible (for the persistence diagrams, the size of 0-dimensional ones can be huge,
+% because in dimension 0 there are especially efficient algorithms that can handle much
+% larger inputs, so it is possible to produce diagrams with millions of  points in each).
+% In such a situation, the natural approach would be to avoid computing the
+% distance between two points if we do not need it,
+
+% We demonstrate that the number of distance computations can be further
+% reduced by a careful re-implementation of the underlying geometric algorithms.
+% Our improvement is based on the following simple observation: many geometric
+% constructions (such as cover tree and WSPD constructions) require distance
+% computations only in the form of the following predicate: 
+% ``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
+% We show that using the information of the already computed distances and
+% the triangle inequality, we can answer the above predicate in many cases
+% without computing the actual distance.
+
+% We describe our idea with an example. In Figure~\ref{fig:1st_example},
+% we see a weighted graph on $P$ whose edges are the distances that have
+% been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
+% The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
+% constitutes an upper bound on the distance by triangle inequality.https://v2.overleaf.com/project/5bf6fa26032afa24d94383a5
+% However, we can also infer that $\dist(p_1,p_2)\geq 3$:
+% otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
+% would be shorter than the edge $(p_3,p_4)$, again contradicting
+% triangle inequality.
+% Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
+% for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
+% distance computation is necessary. This approach is reminiscent to the usage
+% of \emph{numerical filters} in geometric computing in the CGAL library,
+% where geometric predicates in exact number types are first evaluated using 
+% certified floating-point arithmetic, and exact computation is only
+% invoked if the numerical evaluation did not arrive at a decisive result
+% (e.g.~\cite{bbp-interval,kerber-phd}).
+
+% Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
+% in a metric space $(\distspace,\dist)$, 
+% consider the following standard operations:
+
+% \begin{description}
+% \item[Nearest Neighbor] Given a point $q\in\distspace$,
+% find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
+% \[d(q,p_i)\leq d(q,p_j)\]
+
+% \item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
+% such that for all $1\leq i,j\leq n$,
+% %
+% \[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]
+
+% By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
+% such that $E_{ij}=\adist(p_i,p_j)$.
+% \end{description}
+
+% Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
+% for any two points in $\distspace$ in worst-case complexity $\complexity$.
+% Then trivially, the above problems can be solved (exactly) in
+% $O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
 % For low-dimensional\footnote{we will make this notion precise
@@
 
-\paragraph{Contribution.}
+\subsection{Contribution.}
+We consider the two well-known problems: finding a $1 + \eps$-spanner
+and finding an approximate nearest neighbor.
+%
+% We demonstrate that the number of distance computations can be further
+% reduced by a careful re-implementation of the underlying geometric algorithms.
+% Our improvement is based on the following simple observation: many geometric
+% constructions (such as cover tree and WSPD constructions) require distance
+% computations only in the form of the following predicate: 
+% ``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
+% We show that using the information of the already computed distances and
+% the triangle inequality, we can answer the above predicate in many cases
+% without computing the actual distance.
 
-%
-We demonstrate that the number of distance computations can be further
-reduced by a careful re-implementation of the underlying geometric algorithms.
-Our improvement is based on the following simple observation: many geometric
-constructions (such as cover tree and WSPD constructions) require distance
-computations only in the form of the following predicate: 
-``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
-We show that using the information of the already computed distances and
-the triangle inequality, we can answer the above predicate in many cases
-without computing the actual distance.
+% We describe our idea with an example. In Figure~\ref{fig:1st_example},
+% we see a weighted graph on $P$ whose edges are the distances that have
+% been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
+% The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
+% constitutes an upper bound on the distance by triangle inequality.
+% However, we can also infer that $\dist(p_1,p_2)\geq 3$:
+% otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
+% would be shorter than the edge $(p_3,p_4)$, again contradicting
+% triangle inequality.
+% Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
+% for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
+% distance computation is necessary. This approach is reminiscent to the usage
+% of \emph{numerical filters} in geometric computing in the CGAL library,
+% where geometric predicates in exact number types are first evaluated using 
+% certified floating-point arithmetic, and exact computation is only
+% invoked if the numerical evaluation did not arrive at a decisive result
+% (e.g.~\cite{bbp-interval,kerber-phd}).
 
-We describe our idea with an example. In Figure~\ref{fig:1st_example},
-we see a weighted graph on $P$ whose edges are the distances that have
-been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
-The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\dist(p_1,p_2)\geq 3$:
-otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-would be shorter than the edge $(p_3,p_4)$, again contradicting
-triangle inequality.
-Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
-for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
-distance computation is necessary. This approach is reminiscent to the usage
-of \emph{numerical filters} in geometric computing in the CGAL library,
-where geometric predicates in exact number types are first evaluated using 
-certified floating-point arithmetic, and exact computation is only
-invoked if the numerical evaluation did not arrive at a decisive result
-(e.g.~\cite{bbp-interval,kerber-phd}).
+% \begin{figure}[h]
 
-\begin{figure}[h]
+% \centering
+% \includegraphics[width=6cm]{intro_example.eps}
+% \caption{An example of a partially revealed metric.}
+% \label{fig:1st_example}
+% \end{figure}
 
-\centering
-\includegraphics[width=6cm]{intro_example.eps}
-\caption{An example of a partially revealed metric.}
-\label{fig:1st_example}
-\end{figure}
+% In our implementation, we maintain the best lower and upper bounds of all
+% pairwise distances in an $n\times n$ matrix of intervals. We show how to
+% update this matrix in $O(n^2)$ time when a new edge is added.
+% %We point out once more that a quadratic update time only pays off
+% %if distance computations are expensive, say $\complexity=\Omega(n^2)$.
 
-In our implementation, we maintain the best lower and upper bounds of all
-pairwise distances in an $n\times n$ matrix of intervals. We show how to
-update this matrix in $O(n^4)$ time when a new edge is added.
-%We point out once more that a quadratic update time only pays off
-%if distance computations are expensive, say $\complexity=\Omega(n^2)$.
+% We perform an extensive experimental evaluation of our idea.
+% We mostly concentrate on point clouds in Euclidean space~--
+% of course, the approach does not yield any performance benefits in that
+% situation because distance computations are cheap; however, the Euclidean
+% case provides a simple setup for experiments.
+% We investigate how the dimensionality, the clustering density, 
+% the approximation quality, and noise 
+% in the point cloud affect the number of distance computations.
+% Our result show that $\ldots$
 
-We perform an extensive experimental evaluation of our idea.
-We mostly concentrate on point clouds in Euclidean space~--
-of course, the approach does not yield any performance benefits in that
-situation because distance computations are cheap; however, the Euclidean
-case provides a simple setup for experiments.
-We investigate how the dimensionality, the clustering density, 
-the approximation quality, and noise 
-in the point cloud affect the number of distance computations.
-Our result show that $\ldots$
-
-We also provide some experimental results for the case of persistence diagrams
-which motivated this line of research. We point out that the space of
-persistence diagrams fails to be ``low-dimensional''. To be precise,
-its doubling dimension is $\infty$, and a finite set of $n$ diagram
-can have the worst-case doublng dimension $\log n$. This worst-case
-behavior is also the expected one if the diagrams are chosen randomly
-(in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
-However, we demonstrate that
-for natural collections of persistence diagrams, a substantial fraction
-of distance computations can be avoided. In detail, $\ldots$
+% We also provide some experimental results for the case of persistence diagrams
+% which motivated this line of research. We point out that the space of
+% persistence diagrams fails to be ``low-dimensional''. To be precise,
+% its doubling dimension is $\infty$, and a finite set of $n$ diagram
+% can have the worst-case doublng dimension $\log n$. This worst-case
+% behavior is also the expected one if the diagrams are chosen randomly
+% (in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
+% However, we demonstrate that
+% for natural collections of persistence diagrams, a substantial fraction
+% of distance computations can be avoided. In detail, $\ldots$
 
@@
 \section{Notation and Preliminary Notions}
+Let $(X, \dist)$ be a finite metric space with $n$. Let $G$ be a graph with set of vertices $X$,
+its edges have natural weighting $w(x,y) = \dist(x, y)$ which induces the graph metric $\dist_G$ on $X$.
+Graph $X$ is called a $1 + \eps$-spanner of $P$, if $\dist_G$ is a good approximation of $\dist$:
+$\dist(x,y) \leq \dist_G(x, y) \leq (1+\eps) \dist(x, y)$. Obviously, $G$ must have $\Omega(n)$ edges,
+and we consider $G$ optimal, if it has $O(n)$ edges (for each edge of $G$ we must compute the corresponding
+distance, if we want to use the spanner instead of the original metric, which dictates this
+choice of optimality measure).
 
@@
 
-We shall need the following lemma.
+We shall need the following lemma, which is just a reformulation of the well-known
+packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).
 
 \begin{lemma}
-\label{lem:packing_lemma_dd}
-Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
-Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $. 
+\label{lem:packing_lemma}
+ Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
+ $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
+ Then $P$ is finite and its cardinality is bounded by some $C$ that depends only on $d$
+ and the ratio $r/R$.
+% Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $.
+
+% Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
+% Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $. 
 \end{lemma}
-
 \begin{proof}
 We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
-balls of radius $R/4$, etc. Repeating this process $\lceil \log \frac{R}{r/2} \rceil = m $ times, 
+balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
 we cover
-$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can contain at most one point from $P$, the claim follows.
+$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
+contain at most one point from $P$, the claim follows.
 \end{proof}
+If the doubling dimension of $(X, \dist)$ is larger than $d$, then
+there exists a ball of radius $R$ such that there are at least $2^d$ points with pairwise distance at least $R$. On the other hand,
+the maximal distance between any two of these points is at most $2R$, and we have a lot of points with approximately the same distance
+between them. If we want a high-quality spanner on these points, then we can expect that its size will be quadratic, not linear;
+so the space of high doubling dimension cannot have optimal spanners. Thus the question of approximating
+the non-doubling metric without computing many distances cannot be solved positively, and we should
+restrict our attention to the case of doubling spaces.
 
@@
 
-It is known that asymptotically optimal spanner can be constructed by the following greedy algorithm: 
+It is known that optimal spanner can be constructed by the following greedy algorithm: 
 start with an empty graph on vertices $p_i$,
@@
  The algorithm also maintains lower and upper bounds
- for all pairwise distances, which are updated after a new edge is added. and the stopping criterion
- is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
- for the distance between $p_i$ and $p_j$, respectively, then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
+ for all pairwise distances, which are updated after a new edge is added.
+ These bounds are used in the stopping criterion, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
+ for the distance between $p_i$ and $p_j$, respectively, 
+ then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
  for all $i \neq j$.
@@
     \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
-    \State {$(i,j) \gets GetEdgeToAdd$}
+    \State {$(i,j) \gets $} \Call{GetEdgeToAdd}{}
     \State {$v \gets \dist(p_i, pj)$}
     \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
-    \State {$UpdateBounds(i, j, v)$}
+    \State \Call{UpdateBounds}{$i, j, v$}
     \EndWhile
@@
 and $\infty$ is larger than any real number,
-thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection,
+thus making the predicate in the while loop well-defined.  The procedure \textsc{UpdateBounds} is described in the next subsection,
 now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
@@
 
- \paragraph{Maintaining Lower and Upper Bounds}
+ \subsection{Maintaining Lower and Upper Bounds}
 
@@
 
-\paragraph{Experimental Results}
+\subsection{Experimental Results}
 At the moment we cannot provide any theoretical guarantees for the heuristics that we propose to compute blind spanners,
 and it can happen that it will yield the complete graph as a spanner.
-We run extensive experiments on the points sampled from the low-dimensional Euclidean space to investigate
+We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
 experimentally the performance of these heuristics. Clearly, this has no practical meaning for $\R^d$, but
@@
     \label{fig:spanner-sparseness}
-    \includegraphics[width=0.7\textwidth]{edges_to_points_ratio_dim_2_normal_points.png}
+    \includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
     \caption{Number of edges in blind spanners generated by different variants of the blind
@@
     \label{fig:blind-greedy-only}
-    \includegraphics[width=0.7\textwidth]{blind-greedy-dims.png}
+    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
     \caption{}
+\end{figure}
+
+\begin{figure}[ht]
+    \label{fig:blind-rbr-variants}
+    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
+    \caption{Comparison of the four variants of \textsc{BlindRandomBadRatio} algorithm.}
 \end{figure}
@@
 so we only present results for the \textbf{uniform} point set in dimension 2.
-%Let $e$ denote the number of edges of a spanner on $n$ vertices;
-%in the plots we visualize the ratios $e / { n \choose 2}$ and $e / n$.
-%We refer to the first ratio as \textit{density} and to the second one
-%as \textit{growth}. Note that
 
-The results are summarized in the plot \ref{fig:spanner-sparseness}.
-The first plot shows the number of edges of the spanner.
+The plot \ref{fig:spanner-sparseness}.
+shows the number of edges of the spanner.
 We can see that, even though none of the blind spanners can produce
-spanners of the same quality (i.e., sparseness) as the standard greedy algorithm,
+spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
 \textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
@@
 is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
-linear regression.  We give in the table \ref{tbl:regr_coeff_spanner}
-the maximal exponents $\alpha$ for each of the algorithms,
-and it is worth mentioning that even for the quasi-sorted shaker variant 
-the exponent seems to be bounded away from 2.
+linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
+the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
+greedy algorithms. Note that even for the greedy algorithm these estimated
+exponents can be significantly larger than 1,
+which is explained by the fact that the number of points
+on which we computed is not large enough to clearly see
+the linear dependency.
+
+\begin{table}[]
+\label{tbl:regr_coeff_spanner}
+\begin{tabular}{|l|l|l|}
+\hline
+dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
+2         &         1.08                &  1.12                 \\ \hline
+3         &         1.24                &  1.41                \\ \hline
+4         &         1.42                &  1.77                 \\ \hline
+\end{tabular}
+\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependency of the number of edges
+on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
+\end{table}
 
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the BlindRandomBadRatio.
+the other two variants of the \textsc{BlindRandomBadRatio} (see \ref{fig:blind-rbr-variants}).
 
@@
 
+The plot \ref{fig:spanner-eps-dependence}
+compares the \textsc{BlindGreedy} and \textsc{BlindRandomRatioForceLowerBound} algorithms
+on \textbf{uniform} point sets for different choices of $\eps$.
+
 Summing up, we can conclude from the experiments that 
@@
 algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
-first. The disappointing fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
+first. The data that we observed can serve as weak evidence
+of the conjecture: the blind greedy algorithm computes
+a spanner with $O(n)$ edges. The disappointing fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
 (\textsc{BlindQuasiSortedGreedy} is much worse, requiring all the edges). It would seem plausible that, if we have
@@
 find a working heuristics.
-
-%We look at points that are sampled uniformly at random from a cube, points that are sampled
-%from a standard normal distribution, points that are clustered (we sample cluster centers uniformly at random
-%and then we sample roughly the same number of points which are normally distributed around each of the cluster
-%centers), and at points that are obtained from the uniform sample by exponentiating the coordinates:
-%$(e^{\xi_1}, \dots, e^{\xi_d})$, where $\xi_i$ are i.i.d. with uniform distribution.
-%Our findings are summarized in the plots \ref{fig:spanner_plots}.
 
@@
 Our goal is to reduce the number of computed distances $\dist(p_i, q)$. We shall also assume w.l.o.g.
-that all $\dist(p_i, q)$ are distinct (we can always break ties lexicographically).
+that all $\dist(p_i, q)$ are distinct.
 
@@
 At each element $p_i$ look at the lower bound for $r_i$,
-if it is larger than $v / (1+\eps)$, then we can safely skip $p_i$.
+if it is larger than $v / (1+\eps)$, then we can safely discard $p_i$.
 Otherwise we must compute the distance $r_i$, and we immediately
  use it to update the lower bounds for all the next points $p_k$'s. 
-If $p_i$ turns out to be a better candidate (we compare $r_ir$ with $v$, not with $v/(1+\eps)$),
+If, after computing $r_i$, we see that $p_i$ is a better candidate (we compare $r_ir$ with $v$, not with $v/(1+\eps)$),
 we update the current candidate $c$ and the distance to the query point $v$ accordingly, 
 otherwise $c$ and $v$ remain unchanged.
-
-We may hope to avoid computation of $O(n)$ distances, if we assume low doubling dimension
-and solve approximate neighbor problem. The necessity of the first assumption is easily justified: if all $d(p_i, p_j)$
-and $d(p_i, q)$ are approximately the same, at each iteration of our algorithm we will be able to discard only one point.
-The necessity of the second requirement can be seen from the figure \ref{fig:bad_exact_nn_example},
-where a set of points in the plane is in general position, and yet, regardless of the order we choose them,
-at each iteration of our algorithm we are able to discard only one point. This example does not work in the
-approximate formulation.
-
-%\begin{algorithmic}
-%\Procedure{Discard\_1}{$L, p_i$}
-%    \State {$r_i = \dist(q, p_i)$}
-%    \ForAll { $p_j \in L$}
-%       \If {  $ \dist(p_i, p_j) > r_i \left(1 + \frac{1}{ 1 + \eps} \right)$ }
-%           \State {Remove $p_j$ from $L$}
-%       \EndIf
-%    \EndFor
-%\EndProcedure
-%\end{algorithmic}
-
-%\begin{algorithmic}
-%\Procedure{Discard\_2}{$L, w, c$}
-%    \State {$r_w = \dist(q, w)$}
-%    \State {$r_c = \dist(q, w)$}
-%    \Comment {Distances $r_w$ and $r_c$ are cached}
-%    \ForAll { $p_j \in L$}
-%       \If { $ \dist(w, p_j) < r_w - r_c / ( 1 + \eps) $ }
-%           \State {Remove $p_j$ from $L$}
-%           \Comment {In particular, the worse candidate $w$ will be removed here}
-%       \EndIf
-%       \If {  $ \dist(w, p_j) > r_w + r_c / ( 1 + \eps) $ }
-%           \State {Remove $p_j$ from $L$}
-%       \EndIf
-%    \EndFor
-%\EndProcedure
-%\end{algorithmic}
-
-%\begin{algorithmic}
-%\label{alg:ann_blind}
-%\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
-%    \State {Initialize array $L$ with a random permutation of $P$}
-%    \State {$ c \gets \mbox{the first element of }L$}
-%    \State {Compute $r_i = \dist(q, p_i)$}
-%    \State \Call{Discard\_1}{$L, c$}
-%    \While {$|L| > 1$}
-%        \ForAll {$p_j \in L $}
-%            \State {Compute $r_j = \dist(p_j, q)$}
-%            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
-%            \If {$r_j < r_i$ }
-%                \State {$ c \gets p_j $}
-%                \State \Call{Discard\_1}{$L, c$}
-%                \State {$ w \gets p_i $}
-%            \Else
-%                \State {$ w \gets p_i $}
-%            \EndIf
-%            \State \Call{Discard\_2}{$L, w, c$}
-%        \EndFor
-%    \EndWhile
-%    \State \Return {the unique element left in $L$}
-%\EndFunction
-%\end{algorithmic}
 
@@
 
+Before we start the proof, let us make some remarks. First, we want
+to rephrase what discarding a point means. 
+Let $v$ be the distance from the current candidate $c$ to $q$;
+suppose that we have computed $r_i = \dist(p_i, q)$ which is larger than $v$,
+and want to use it to estimate $r_j = \dist(p_j, q)$ for some other $p_j$. If we assume 
+that $\dist(p_i, p_j) < r_i - v/(1+\eps)$,
+and substitute this in the triangle inequality $\dist(p_j, q) \geq \dist(p_i, q) - \dist(p_i, p_j)$, then we obtain $\dist(p_j, q) \geq v / (1 + \eps)$.
+Thus, if $p_i$ fails to be closer to $q$ than the current candidate, then all points in the ball of radius $r_i - v / (1 + \eps)$,
+even those who are closer to $q$ than $c$, can be discarded, because we are looking for an approximate nearest neighbor.
+Figure \ref{fig:ann_discard_circles} illustrates this.
+Note that with the triangle inequality we can also discard points that are too far from $p_i$, that is the shaded exterior of the bigger balls in the figure.
+
+Secondly,
+we could modify the algorithm to search for the exact neighbor, just by setting $\eps = 0$. In this case, however, 
+the figure \ref{fig:bad_exact_nn_example},
+shows a set of points in the plane for which, regardless of the order we choose them,
+at each iteration of our algorithm we are able to discard only one point (the shaded balls do not contain any other point from the point set), 
+therefore we may not hope to avoid computation of $O(n)$ distances, if we
+are trying to solve the exact nearest neighbor problem.
+The necessity of the doubling property assumption is also easily justified: if all $d(p_i, p_j)$
+and $d(p_i, q)$ are approximately the same, at each iteration of our algorithm we will be able to discard only one point.
+
 \begin{proof}
-Let $i_1,...,i_k$ denote the indices of the minima in the sequence $r_1, \dots, r_n$
-    (recall that this is a random sequence). Then $k=O(log n)$,
-    by the standard backwards analysis argument.
+Let $i_k$ be the {$k$-th} position where the sequence  $m_i := \min \Set{m_1, \dots, m_i}$
+changes, so that
+\[
+m_1 = m_2 = \dots = m_{i_1 - 1} > m_{i_1} = m_{i_1 + 1} = \dots = m_{i_2 - 1} > m_{i_2} = \dots. 
+\]
+In other words, if we traversed the list $p_1, \dots, p_n$ and used the bruteforce algorithm
+for exact nearest neighbor, then $i_k$ would the indices where the current candidate
+changes, because we found a point that is closer to $q$ than all the previously seen points.
+We shall refer to $i_k$ as the \textit{change position}.
+Standard backwards analysis argument (e.g., \cite{dubch}, Sect. 4.4) shows that $k=O(log n)$.
 
-    Consider two consecutive minima  $s:=i_j$ and $t:=i_{j+1}$. We claim that in the sequence
+    Consider two consecutive change positions $s:=i_j$ and $t:=i_{j+1}$. We claim that in the sequence
 $p_s,...,p_{t-1}$
-only a constant number of distances are computed, irrespective of its length.
+only a constant number of distances are computed by our algorithm, irrespective of its length. This claim immediately implies the theorem.
 
-To see that, let $c_1,...,c_n$ be the candidates after the $i$-th step of
+To see that it true, let $c_1,...,c_n$ be the candidates after the $i$-th step of
 the for loop of the algorithm. We have no guarantee that $c_s=p_s$, because $p_s$ it is possible
     that $p_s$ was discarded without computing $r_s$ (see the picture). However, we know that
-$d(c_s,q)<=(1+eps) r_s$.
-
-Indeed, suppose that $c_s \neq p_s$ and $r_s < d(c_s,q)/(1+eps)$. In this
+$d(c_s,q)<=(1+ \eps) r_s$.
+Indeed, suppose that $c_s \neq p_s$ and $r_s < d(c_s,q)/(1+\eps)$. In this
     case $p_s$ cannot be discarded 
@@
 It follows that any point in the sequence $p_s,...,p_{t-1}$ for which a
-distance is queried must lie in the ball around $c_s$ of radius at most $2(1+eps)r_s$.
+distance is queried must lie in the ball around $c_s$ of radius at most $2(1+\eps)r_s$.
 
@@
 that is, the candidate came from before, than $p_s$ would not be in the
-minimum sequence. If $s<k<t$, then it follows that at least one of the
-elements in $p_{s+1},\dots,p_k$ is in the minimum sequence, contradicting
-the assumption that $p_s$ and $p_t$ are consecutive minima.
+change position. If $s<k<t$, then at least one of the
+elements in $p_{s+1},\dots,p_k$ is a change position, contradicting
+the assumption that $p_s$ and $p_t$ were chosen to be consecutive change positions.
 
@@
 ball of radius $\eps r_s$ around the query point. All queried points must
-hence have pairwise distance of at least $ \eps r_s / 2$. Hence, the number of
-queried points is by the packing lemma bounded by some constant that depends only on $\eps$ and the dimension. 
+hence have pairwise distance of at least $ \eps r_s / 2$. 
+Now we can apply the packing lemma \ref{lem:packing_lemma} to bound the number of
+queried points by some constant that depends only on $\eps$ and the dimension. Indeed,
+they all are contained in a ball of radius $R = 2(1+\eps)r_s$, and, if we divide this by $\eps r_s$, the ratio depends only on $\eps$.
 \end{proof}
+
+We should note that the constant factor that is obtained in this proof can be very pessimistic,
+because in practice the algorithm can discard much more points.
 
@@
 
-%\begin{lemma}
-%Consider the following random procedure: we start with $n$ numbers $[n] = \{\,1,\dots, n\,\}$ and 
-%choose $\xi_1$ uniformly at random from $[n]$. If $\xi_1 = n$, then we stop,
-%otherwise we remove $\xi_1$ and all numbers preceding it and from $\{\,\xi_1, \dots, n\,\}$
-%we choose $\xi_2$. If $\xi_2 = n$, we stop, if $\xi_2 < n$, we repeat the process removing all numbers not greater than $\xi_2$ and
-%drawing $\xi_3$ from the remaining numbers uniformly at random. The expected number of steps of this procedure
-%is $O(\log n)$.
-%\end{lemma}
-%\begin{proof}
-%We can rephrase the procedure as a Markov chain with set of states $[n]$ and transition matrix
-%\begin{equation}
-%    P = \begin{pmatrix}
-%         0      & \frac{1}{n-1} & \frac{1}{n-1} & \dots  & \frac{1}{n-1} & \frac{1}{n-1} \\
-%         0      &  0            & \frac{1}{n-2} & \dots  & \frac{1}{n-2} & \frac{1}{n-2} \\
-%         0      &  0            & 0             & \dots  & \frac{1}{n-3} & \frac{1}{n-3} \\
-%         \vdots &  \vdots       & \vdots        & \ddots & \vdots        & \vdots \\
-%         0      &  0            & 0             & \dots  & \frac{1}{2} & \frac{1}{2} \\
-%         0      &  0            & 0             & \dots  & 0           & 1           \\
-%        \end{pmatrix};
-%\end{equation}
-%the chain starts from state $1$.
-%Note that there is one absorbing state $n$, and all other states are transient, and the matrix is already in the canonical
-%form, meaning that all transient states precede all absorbing states. 
-%We denote the square submatrix formed by the first $n-1$
-%rows and $n-1$ columns by $Q$, it is the submatrix corresponding to all transient states. It is known
-%(\cite{absorbing_markov_chain}) that the matrix $I-Q$ is invertible, let $N$ be its inverse,
-%\begin{equation}
-%N = (I- Q)^{-1} = 
-%\begin{pmatrix}
-%         1      & -\frac{1}{n-1} & -\frac{1}{n-1} & \dots  & -\frac{1}{n-1}  \\
-%         0      &  1             & -\frac{1}{n-2} & \dots  & -\frac{1}{n-2}  \\
-%         0      &  0             & 1             & \dots  & -\frac{1}{n-3}   \\
-%         \vdots &  \vdots       & \vdots        & \ddots & \vdots            \\
-%         0      &  0            & 0             & \dots  & -\frac{1}{2}      \\
-%         0      &  0            & 0             & \dots  & 0           & 1   \\
-%        \end{pmatrix}^{-1}
-%\end{equation}
-%Let $t_i$ denote the expected time
-%until absorption, if chain starts from a transient state $i = 1, \dots, n-1$; 
-%vector of $t_i$'s is given by the formula
-%\[
-%\begin{pmatrix}
-%      t_1 \\
-%      t_2 \\
-%      \dots \\
-%      t_{n-1} \\
-%      \end{pmatrix}
-%= N \begin{pmatrix}
-%      1 \\
-%      1 \\
-%      \dots \\
-%      1 \\
-%      \end{pmatrix}.
-%\]
-%The expected number of steps that we are interested in is $t_1$, so we need to calculate the matrix $N$
-%and prove that the sum of the elements in its first row is $O(\log n)$.
-%One can easily check that
-%\[
-%N  =  \begin{pmatrix}
-%      1 & \frac{1}{n-1} & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
-%      0 &             1 & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
-%      0 &             0 &            1  & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
-% \vdots &        \vdots &  \vdots       & \vdots        & \ddots & \frac{1}{3} & \frac{1}{2} \\
-%      0 &             0 &            0  &            0  &  \dots &            1 & \frac{1}{2} \\
-%      0 &             0 &            0  &            0  &  \dots &            0 &           1 \\
-%      \end{pmatrix},
-%\]
-%thus $t_1 = 1 + \frac{1}{n-1} + \frac{1}{n-2} + \dots + \frac{1}{2} = O(\log n)$.
-%\end{proof}
-
-\paragraph{Experimental Results}
+\subsection{Experimental Results}
 
@@
 that were actually computed is the measure that we are interested in.
+We average the results over 10 different instances of the point
+set and query point in order to see the trend clearer; thus
+each point on the plots in this section is the result of averaging of 100 runs
+of the code (10 instances, 10 random permutations per instance).
 
@@
     \label{fig:ann_const_ratio}
-    \includegraphics[width=0.7\textwidth]{log_dependency_constant_ratio.png}
+    \includegraphics[width=0.7\textwidth]{pics/log_dependency_constant_ratio.png}
     \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
     points.}
+\end{figure}
+
+\begin{figure}[ht]
+    \label{fig:ann_dimension_dependency}
+    \includegraphics[width=0.7\textwidth]{pics/ann_computed_distances_dimension_dep.png}
+    \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
 \end{figure}
@@
     \item Uniform. Points are sampled uniformly at random from the unit cube in $\R^d$.
-    \item Normal. Points are sampled from the standard normal distribution.
+    \item Normal. Points are sampled from the normal distribution.
 \end{enumerate}
@@
 the maximal number of points is $30,000$.
-In high dimensions the algorithm does not perform as well for very small epsilon,
-but for $\eps = 0.01$ even in dimension 20 just about 10 percent of the distances
-are computed for 7,000 points (while for 1,000 points more than 0.3 of the distances must be computed).
-We note, however, that the constants in asymptotics are very different for different combinations
-of distributions of query points and points in the point set.
 
-%The algorithm can be informally described as follows. Pick a random candidate $p_i$ and try to verify that
-%it is an approximate nearest neighbor by scanning all other points. Pick the next random point $p_j$
-%and assume that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$. This allows us to discard not just the point
-%$p_j$, but also the points that are very close to $p_j$ and very far from $p_j$ using triangle inequality.
-%Indeed, consider the ball of radius $ \dist(p_j, q) - \frac{ \dist(p_i, q) }{ 1+\eps} $ centered at $p_j$.
-%The triangle inequality implies that all points in this ball are at the distance at least $\frac{ \dist(p_i, q) }{ 1+\eps}$
-%from $q$. Now consider the ball of radius $\dist(p_j, q) + 2 \frac{\dist(p_i, q)}{1+ \eps}$ centered at $p_j$. 
-%Every point outside of this ball is a
+In order to empirically verify the upper bound $O(\log n)$,
+we plot the number of computed distances divided
+by the logarithm of the number of points in figure \ref{fig:ann_const_ratio} (for $d = 2$).
+We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
+the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
+really computes only a very small number of distances to the query point.
+As expected, in high dimensions the algorithm does not perform as well.
+In figure \ref{fig:ann_dimension_dependency} we plot the average number of computed distances
+for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
+the sublinearity of the growth becomes clear only when the number of points is relatively large,
+approaching 30000.
 
-%The algorithm that we analyze is based on the following simple observation.
-%Suppose that we want to to prove that a candidate for the approximate nearest neighbor is $p_1$,
-
-%The goal is to compute
-%as few $\dist(q, p_k)$ as possible.
+% We also looked at the dependency on the density. Note that our algorithm is not invariant under scaling: if we multiply the metric
+% by $\lambda$, we should also change $\eps$ to $\lambda \eps$, if we want
+% the same behavior. Therefore we could \textit{a priori} expect some
+% changes even if we change the parameters of the uniform distribution
+% without changing epsilon. However,
+% when we tested the algorithm on uniform data sampled from $[-a,a]^d$
+% for $a \in \Set{1, 10, 100, 1000, 10000}$ and on normal data
+% sampled from normal distribution with scales from the same set,
+% we did not observe any significant difference.
 
@@
 %was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
-%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
+%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been addedhttps://v2.overleaf.com/project/5bf6fa26032afa24d94383a5
 %to the spanner. By claim 1
@@
 
-\section{Conclusion}
+\section{Conclusion and Future Work}
 \label{sec:conclusion}
+We presented experimental evidence showing that in low
+doubling dimension we can avoid many distance computations,
+if we try to get as much information as possible from the triangle inequality
+applied to the previously computed distance. The \textsc{BlindGreedy} spanner
+performs especially well, being able to produce  sparse spanners even for small values of  $\eps$.
+For the approximate nearest neighbor algorithm we proved a logarithmic upper bound,
+and, even though the constant in the proof was very large, we demonstrated
+experimentally that this does not show in practice.
+
+The next step on the theoretical side could be proving some upper bound
+for blind spanners; it is also interesting to look at other metric space
+algorithms and data structures in the setting of expensive distance computation.
+On the more applied side, the further development of this approach would
+be to find finite spaces of low dimension in practically
+important cases, like spaces of persistence diagrams or images endowed with Wasserstein (a.k.a. Earth Movers)
+metric. We know that low-dimensional subspaces of functions do not necessarily produce
+low-dimensional spaces of persistence diagrams, but there may be some conditions
+which guarantee that dimension does not increase too much.
+% An easy example: consider a fixed function $f$ to which we apply
+% a finite-dimensional family of perturbations, $f(x) + g(x; \lambda_1, \dots, \lambda_n)$;
+% for instance, each $\lambda_i$ is associated with some $x_i$ and adds a Gaussian noise
+% scaled by $\lambda$.
+% If the perturbations are very small, then they will produce some small wiggles on the function,
+% and then the optimal matcih
 

### 2018-11-30 paper/draft.tex
Migrate from Overleaf

@@
 
+\newcommand{\myparagraph}[1]{\textbf{#1.}}
+
 %\newcommand{\remark}[1]{\textbf{[#1]}}
@@
 
-%\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
-%\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
+\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
+\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
 
@@
 \authorrunning{M. Kerber and A. Nigmetov}
+
+\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+\subjclass{dummy}
+% \category{dummy}
 
@@
 
-\subsection{Motivation.}
+Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
+in a metric space $(\metricspace,\dist)$, 
+consider the following standard operations:
+
+\begin{description}
+\item[Approximate Nearest Neighbor] Given $\eps>0$ and a point $q\in\metricspace$,
+find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
+\[\dist(q,p_i)\leq(1+\eps)\dist(q,p_j)\]
+
+\item[Spanner] Given $\eps>0$, compute a weighted graph $G$ with vertices in $\pointset$
+such that for any $u,v\in\pointset$, the shortest path distance between $u$ and $v$
+is at most $(1+\eps)\dist(u,v)$.
+\end{description}
+
+The performance of algorithms for these problems depends on the number of points,
+the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
+It is a common assumption to assume $\complexity$ to be a constant; 
+There are good reasons for that: the most common case of a metric space
+is $\metricspace=\R^d$ with $d$ some constant, in which case $\metric$ can be evaluated in $O(d)=O(1)$ time.
+Even if $d$ is considered non-constant,
+it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
+Another typical assumption is that all pairwise distances are part of the input
+in which case $\complexity$ is $O(1)$.
+
+However, we argue that in some situations, distance computations
+in $\metricspace$ can be costly and $\complexity$ might be incomparable
+with $n$. Our motivation comes from topological summaries
+such as persistence diagrams or Reeb graphs, which are of interest
+in the field of topological data analysis. A persistence diagram
+is a point set in $\R^2$, and the distance between two diagrams
+is determined by a min-cost matching between the point sets.
+If the diagrams have $N$ points, computing this matching requires
+polynomial time in $N$, and $N$ might well be larger than $n$, the number
+of diagrams considered. For the case of Reeb graphs, the situation is even
+worse: while several metrics on Reeb graphs have been proposed,
+not even an constant-factor approximation algorithm is known that runs
+in polynomial time in the size of the graphs.
+\michael{This paragraph needs references. As you know the literature better than me,
+I leave that part to you.}
+
+In such situations with expensive distance computations, 
+it makes sense to study a different cost model, where only the number of distance computations
+is taken into account. For instance, that means that quadratic time operations in terms of $n$
+are not counted towards the time complexity, as long as these operations do not query any distance
+in $\metricspace$. We also ignore the space complexity in our model.
+
+We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
+of $\metricspace$ is bounded by a constant. 
+In that situation, standard constructions from computational geometry provide partial answers:
+Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
+an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
+performing $O(\log n)$ distance computations per query point.
+However, for our relaxed cost model, we pose the question whether simpler constructions achieve
+comparable, or even fewer distance computations.
+
+We also propose a slight variant of our model: we assume that we also have access to an (efficient)
+$2$-approximation algorithm for the distance queries. Queries to this approximation algorithm
+are not counted in the model, hence we can assume that for each pair of points $(u,v)$, we
+know a number $a_{u,v}$ with $\dist(u,v)\leq a_{u,v}\leq 2\dist(u,v)$. This induces an approximate ordering
+of all distances in the metric space, and it is plausible to assume that such an ordering will simplify
+algorithmic tasks on metric spaces, at least in practice.
+
+\myparagraph{Contributions}
+%
+We propose simple algorithms for spanner construction and approximate nearest neighbor search
+and evaluate them theoretically and experimentally in the defined cost model.
+
+Our algorithms are based on the following simple idea: since distance computations are expensive
+and should be avoided, we try to obtain maximal information out of the distances that have been computed
+so far. 
+This information consists of lower and upper bounds for unknown distances, obtained from known distances
+by triangle inequality (see Figure~\ref{fig:1st_example}).
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=6cm]{intro_example.eps}
+\caption{The compute distances are shown as edges in a graph. Note that the exact distance
+of $p_1$ and $p_2$ is unknown. The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
+constitutes an upper bound on the distance by triangle inequality.
+However, we can also infer that $\metric(p_1,p_2)\geq 3$:
+otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
+would be shorter than the edge $(p_3,p_4)$, again contradicting
+triangle inequality.}
+\label{fig:1st_example}
+\end{figure}
+
+we see a weighted graph on $P$ whose edges are the distances that have
+been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
+The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
+constitutes an upper bound on the distance by triangle inequality.
+However, we can also infer that $\metric(p_1,p_2)\geq 3$:
+otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
+would be shorter than the edge $(p_3,p_4)$, again contradicting
+triangle inequality.
+Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
+for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$
+We describe our idea with an example. In Figure~\ref{fig:1st_example},
+we see a situation on $P$ whose edges are the distances that have
+been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
+The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
+constitutes an upper bound on the distance by triangle inequality.
+However, we can also infer that $\metric(p_1,p_2)\geq 3$:
+otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
+would be shorter than the edge $(p_3,p_4)$, again contradicting
+triangle inequality.
+Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
+for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$
 
@@
 
-The next option is to make an additional assumption: while
-computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a cheap approximation algorithm
-that gives, say, a $2$-approximation of $\dist$ much faster (the output of this approximation algorithm is guaranteed to
+The last option that we consider is based on an additional assumption: while
+computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
+cheap approximation algorithm
+that gives a $2$-approximation of $\dist$ much faster 
+(the output of this approximation algorithm is guaranteed to
 lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
 that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
-We now introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
-these intervals are not disjoint, but, by calling our algorithm we can assign
+We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
+these intervals are not disjoint, but by calling the approximation algorithm we can assign
 each pair $(p_i, p_j)$ to one of the buckets.
@@
 and we take $k = \lfloor \log v \rfloor - 1$.
-If we find a bucket for each pair $(p_i, p_j)$, we have some sort of a weak ordering: we cannot say anything
+Afer we found a bucket for each pair $(p_i, p_j)$, 
+we have some sort of a weak ordering: we cannot say anything
 about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
@@
 traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
-we travers the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
+we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
 we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
@@
 
-Note that after we added the edge $(p_i, p_j)$, the ratio of the upper and lower bound
-for this pair becomes 1, so a blind spanner algorithm always terminates (of course,
-we assume that the procedure \textsc{GetEdgeToAdd} picks an edge only once).
+% Note that after we added the edge $(p_i, p_j)$, the ratio of the upper and lower bound
+% for this pair becomes 1, so a blind spanner algorithm always terminates (of course,
+% we assume that the procedure \textsc{GetEdgeToAdd} picks an edge only once).
 

### 2018-11-30 cpp_version/include/cover_tree/wspd.h
read points, not distances - for larger files

@@ -62,6 +62,7 @@ namespace wasser_spanner {
         bool is_valid() const;
         size_t size() const;
         void make_spanner();
+        void make_spanner1();
         double get_spanner_distance(VertexDescriptor i, VertexDescriptor j);
         double get_max_relative_error();
     };

### 2018-11-30 cpp_version/include/dynamic_spanner.h

@@ -95,6 +95,17 @@ public:
         }
     }
 
+    void make_exact_distance_keeper()
+    {
+        for(int i = 0; i < m_num_points; ++i) {
+            for(int j = 0; j < m_num_points; ++j) {
+                m_matrix[i][j].lower_bound = m_matrix[i][j].upper_bound = m_matrix[i][j].distance;
+                m_matrix[i][j].distance_requested = false;
+                m_matrix[i][j].exact_distance_used = false;
+             }
+        }
+    }
+
     std::vector<std::pair<double, double>> get_buckets()
     {
         int n_buckets = ceil(log2(m_max_dist / m_min_dist));

### 2018-11-30 cpp_version/src/main.cpp

@@ -24,6 +24,8 @@ namespace spd = spdlog;
 
 using namespace wasser_spanner;
 
+
+
 int get_num_points(const std::string& fname, const char delimiter = ' ')
 {
     std::ifstream matr_file(fname);
@@ -48,6 +50,55 @@ int get_num_points(const std::string& fname, const char delimiter = ' ')
 
     matr_file.close();
     return result;
+}
+
+int get_dimension(std::string fname, const char delim = ' ')
+{
+    return get_num_points(fname, delim);
+}
+
+bool read_points_file(MatrixR& distance_matrix, double& max_distance, double& min_distance, std::string fname)
+{
+    std::ifstream points_file(fname);
+    if (not points_file.good()) {
+        std::cerr << "Cannot read matrix from file " << fname << std::endl;
+        return false;
+    }
+
+    std::vector<std::vector<double>> points;
+    int dim = get_dimension(fname);
+
+    std::string s;
+    while(std::getline(points_file, s)) {
+        points.emplace_back();
+        std::stringstream ss(s);
+        double a;
+        for (int i = 0; i < dim; ++i) {
+            ss >> a;
+            points.back().push_back(a);
+        }
+    }
+
+    int n_points = points.size();
+    min_distance = std::numeric_limits<double>::max();
+    max_distance = -1.0;
+
+    distance_matrix = MatrixR(n_points, std::vector<double>(n_points, 0.0));
+
+    for (size_t i = 0; i < n_points; ++i) {
+        for (size_t j = 0; j < i; ++j) {
+            double d = 0.0;
+            for (int k = 0; k < dim; ++k) {
+                d += (points[i][k] - points[j][k]) * (points[i][k] - points[j][k]);
+            }
+            distance_matrix[i][j] = distance_matrix[j][i] = sqrt(d);
+
+            max_distance = std::max(distance_matrix[i][j], max_distance);
+            if (i != j) {
+                min_distance = std::min(distance_matrix[i][j], min_distance);
+            }
+        }
+    }
 }
 
 bool read_distance_matrix(MatrixR& distance_matrix, double& max_distance, double& min_distance, const std::string& fname)
@@ -218,9 +269,9 @@ int main(int argc, char** argv)
     double max_dist = -1.0;
     double min_dist = std::numeric_limits<double>::max();
 
-    if (argc < 3) {
+    if (argc < 4) {
         std::cout << "Usage: " << argv[0]
[…]

### 2018-11-30 cpp_version/src/wspd.cpp

@@ -144,6 +144,23 @@ size_t WSPD::size() const
     return m_wspd.size();
 }
 
+void WSPD::make_spanner1()
+{
+    m_graph = std::unique_ptr<Graph>(new Graph(m_num_points));
+    initialize_incremental_components(*m_graph, m_conn_comps);
+    incremental_components(*m_graph, m_conn_comps);
+    for (const auto& wspd_node : m_wspd) {
+        VertexDescriptor i = wspd_node.first.get_point_idx();
+        VertexDescriptor j = wspd_node.second.get_point_idx();
+        if (!m_spanner->m_matrix[i][j].exact_distance_used) {
+            //std::cout << i << " " << j << " [ " << m_spanner->m_matrix[i][j].lower_bound << " , " << m_spanner->m_matrix[i][j].upper_bound << " ]" << std::endl;
+        }
+        auto d = m_spanner->get_distance_no_cache(i, j);
+        boost::add_edge(i, j, EdgeWeightProperty(d), *m_graph);
+        m_conn_comps.union_set(i, j);
+    }
+}
+
 void WSPD::make_spanner()
 {
     m_graph = std::unique_ptr<Graph>(new Graph(m_num_points));

### 2018-11-30 paper/draft.tex

@@
 
-\newcommand{\myparagraph}[1]{\textbf{#1.}}
-
 %\newcommand{\remark}[1]{\textbf{[#1]}}
@@
 \begin{abstract}
-
 In algorithms for finite metric spaces, it is common
@@
 There are good reasons for that: the most common case of a metric space
-is $\metricspace=\R^d$ with $d$ some constant, in which case $\metric$ can be evaluated in $O(d)=O(1)$ time.
+is $\metricspace=\R^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
 Even if $d$ is considered non-constant,
@@
 This information consists of lower and upper bounds for unknown distances, obtained from known distances
-by triangle inequality (see Figure~\ref{fig:1st_example}).
+by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
+operations whenever a new distance has been computed, turing the method useless in the standard computational model.
 
@@
 constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\metric(p_1,p_2)\geq 3$:
+However, we can also infer that $\dist(p_1,p_2)\geq 3$:
 otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
@@
 
-we see a weighted graph on $P$ whose edges are the distances that have
-been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
-The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\metric(p_1,p_2)\geq 3$:
-otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-would be shorter than the edge $(p_3,p_4)$, again contradicting
-triangle inequality.
-Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
-for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$
-We describe our idea with an example. In Figure~\ref{fig:1st_example},
-we see a situation on $P$ whose edges are the distances that have
-been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
-The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\metric(p_1,p_2)\geq 3$:
-otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-would be shorter than the edge $(p_3,p_4)$, again contradicting
-triangle inequality.
-Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
-for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$
+We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
+with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
+for each point pair, the set of all computed distances forms the spanner.
+The experimentally most succesful exploration strategy that we found is to
+repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
+We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
+\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds.\michael{Cite}
+Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
+\michael{Acutally, is that true? We could run BlindGreedy on the 2-approximations instead of on $[0,\infty]$,
+that might save something}
+We also compare with a spanner construction based on WSPDs. Our simple algorithms tend to give much smaller
+spanners on the tested example. Nevertheless, we leave the question open whether our construction
+yields a spanner of asymptotically linear size.
 
-In this paper we look at algorithms in metric spaces,
-in which computing a single distance between points can be very costly.
-The motivating example for us was Topological Data Analysis, where computation
-of bottleneck and Wasserstein distances between two persistence diagrams
-can be done in time polynomial in the number of points of the input diagrams.
-
-Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
-in a metric space $(\metricspace,\dist)$, 
-consider the following standard operations:
-
-\begin{description}
-\item[Nearest Neighbor] Given a point $q\in\metricspace$,
-find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
-\[d(q,p_i)\leq d(q,p_j)\]
-
-\item[Metric approximation] Given $\eps>0$, compute a metric $\approxmetric$
-such that for all $1\leq i,j\leq n$,
-%
-\[\dist(p_i,p_j)\leq\approxmetric(p_i,p_j)\leq (1+\eps)\dist(p_i,p_j).\]
-
-By ``computing'' $\approxmetric$, we mean to compute a $n\times n$-matrix $E$ 
-such that $E_{ij}=\approxmetric(p_i,p_j)$.
-\end{description}
-
-% We believe that the scenario when the input consists of a relatively small number of points, 
-% e.g., of order $10^2-10^3$, but the complexity of computing a distance is rather large,
-% is plausible (for the persistence diagrams, the size of 0-dimensional ones can be huge,
-% because in dimension 0 there are especially efficient algorithms that can handle much
-% larger inputs, so it is possible to produce diagrams with millions of  points in each).
-% In such a situation, the natural approach would be to avoid computing the
-% distance between two points if we do not need it,
-
-% We demonstrate that the number of distance computations can be further
-% reduced by a careful re-implementation of the underlying geometric algorithms.
-% Our improvement is based on the following simple observation: many geometric
-% constructions (such as cover tree and WSPD constructions) require distance
-% computations only in the form of the following predicate: 
-% ``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
-% We show that using the information of the already computed distances and
-% the triangle inequality, we can answer the above predicate in many cases
-% without computing the actual distance.
-
-% We describe our idea with an example. In Figure~\ref{fig:1st_example},
-% we see a weighted graph on $P$ whose edges are the distances that have
-% been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
-% The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-% constitutes an upper bound on the distance by triangle inequality.https://v2.overleaf.com/project/5bf6fa26032afa24d94383a5
-% However, we can also infer that $\dist(p_1,p_2)\geq 3$:
-% otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-% would be shorter than the edge $(p_3,p_4)$, again contradicting
-% triangle inequality.
-% Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
-% for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
-% distance computation is necessary. This approach is reminiscent to the usage
-% of \emph{numerical filters} in geometric computing in the CGAL library,
-% where geometric predicates in exact number types are first evaluated using 
-% certified floating-point arithmetic, and exact computation is only
-% invoked if the numerical evaluation did not arrive at a decisive result
-% (e.g.~\cite{bbp-interval,kerber-phd}).
-
-% Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
-% in a metric space $(\distspace,\dist)$, 
-% consider the following standard operations:
-
-% \begin{description}
-% \item[Nearest Neighbor] Given a point $q\in\distspace$,
-% find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
-% \[d(q,p_i)\leq d(q,p_j)\]
-
-% \item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
-% such that for all $1\leq i,j\leq n$,
-% %
-% \[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]
-
-% By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
-% such that $E_{ij}=\adist(p_i,p_j)$.
-% \end{description}
-
-% Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
-% for any two points in $\distspace$ in worst-case complexity $\complexity$.
-% Then trivially, the above problems can be solved (exactly) in
-% $O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
-% For low-dimensional\footnote{we will make this notion precise
-% later in the text} metric spaces, we can improve these bounds:
-% Using cover trees~\cite{cover-trees}, we can preprocess the point set
-% in $O(\complexity n\log n)$ time to answer nearest neighbor queries
-% in $O(\complexity \log n)$ time. Combining cover trees with well-separated
-% pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
-% induced by $\dist$ in $O(\complexity n\log n)$ time and
-% of linear size (assuming, for simplicity, that $\eps$ is a constant).
-% Out of this spanner,
-% we can construct an approximate metric
-% with an all-pair shortest path computation,
-% resulting in a total running time of
-% \[O(\complexity n\log n+ n^2\log n)\]
-% using Johnson's algorithm~\cite{johnson}.
-
-In the aforementioned techniques, $\complexity$ is usually considered
-as a constant-time operation and hence ignored in the analysis.
-There are good reasons for that: the most common case of a metric space
-is $\distspace=\R^d$ with $d$ some constant, in which case $\dist$ can be evaluated in $O(d)=O(1)$ time.
-Even if $d$ is considered non-constant,
-it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
-Another typical assumption is that all pairwise distances are part of the input
-in which case $\complexity$ is $O(1)$.
-
-However, we argue that in some situations, distance computations
-in $\distspace$ can be costly and $\complexity$ might be incomparable
-with $n$. Our motivating examples come from topological summaries
-such as persistence diagrams or Reeb graphs, which are of interest
-in the field of topological data analysis. A persistence diagram
-is a point set in $\R^2$, and the distance between two diagrams
-is determined by a min-cost matching between the point sets.
-If the diagrams have $N$ points, computing this matching requires
-polynomial time in $N$, and $N$ might well be larger than $n$, the number
-of diagrams considered. For the case of Reeb graphs, the situation is even
-worse: while several metrics on Reeb graphs have been proposed,
-not even an constant-factor approximation algorithm is known that runs
-in polynomial time in the size of the graphs.
-
-In the light of the above examples, we pose the question: 
-how can we practically reduce the number of distance computations 
-in algorithms for metric spaces?
-% Note that for low-dimensional spaces, the aforementioned techniques
-% already provide a partial solution: for approximating the metric, for instance,
-% we only compute $O(n\log n)$ distances instead of $O(n^2)$.
-
-\subsection{Contribution.}
-We consider the two well-known problems: finding a $1 + \eps$-spanner
-and finding an approximate nearest neighbor.
-%
-% We demonstrate that the number of distance computations can be further
-% reduced by a careful re-implementation of the underlying geometric algorithms.
-% Our improvement is based on the following simple observation: many geometric
-% constructions (such as cover tree and WSPD constructions) require distance
-% computations only in the form of the following predicate: 
-% ``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
-% We show that using the information of the already computed distances and
-% the triangle inequality, we can answer the above predicate in many cases
-% without computing the actual distance.
-
-% We describe our idea with an example. In Figure~\ref{fig:1st_example},
-% we see a weighted graph on $P$ whose edges are the distances that have
-% been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
-% The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-% constitutes an upper bound on the distance by triangle inequality.
-% However, we can also infer that $\dist(p_1,p_2)\geq 3$:
-% otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-% would be shorter than the edge $(p_3,p_4)$, again contradicting
-% triangle inequality.
-% Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
-% for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
-% distance computation is necessary. This approach is reminiscent to the usage
-% of \emph{numerical filters} in geometric computing in the CGAL library,
-% where geometric predicates in exact number types are first evaluated using 
-% certified floating-point arithmetic, and exact computation is only
-% invoked if the numerical evaluation did not arrive at a decisive result
-% (e.g.~\cite{bbp-interval,kerber-phd}).
-
-% \begin{figure}[h]
-
-% \centering
-% \includegraphics[width=6cm]{intro_example.eps}
-% \caption{An example of a partially revealed metric.}
-% \label{fig:1st_example}
-% \end{figure}
-
-% In our implementation, we maintain the best lower and upper bounds of all
-% pairwise distances in an $n\times n$ matrix of intervals. We show how to
-% update this matrix in $O(n^2)$ time when a new edge is added.
-% %We point out once more that a quadratic update time only pays off
-% %if distance computations are expensive, say $\complexity=\Omega(n^2)$.
-
-% We perform an extensive experimental evaluation of our idea.
-% We mostly concentrate on point clouds in Euclidean space~--
-% of course, the approach does not yield any performance benefits in that
-% situation because distance computations are cheap; however, the Euclidean
-% case provides a simple setup for experiments.
-% We investigate how the dimensionality, the clustering density, 
-% the approximation quality, and noise 
-% in the point cloud affect the number of distance computations.
-% Our result show that $\ldots$
-
-% We also provide some experimental results for the case of persistence diagrams
-% which motivated this line of research. We point out that the space of
-% persistence diagrams fails to be ``low-dimensional''. To be precise,
-% its doubling dimension is $\infty$, and a finite set of $n$ diagram
-% can have the worst-case doublng dimension $\log n$. This worst-case
-% behavior is also the expected one if the diagrams are chosen randomly
-% (in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
-% However, we demonstrate that
-% for natural collections of persistence diagrams, a substantial fraction
-% of distance computations can be avoided. In detail, $\ldots$
-
-%We also have derived two minor technical results necessary for 
-%our implementation. First of all, since our underlying distance computation
-%algorithm for persistence diagrams only returns approximate answers,
-%we had to adapt all algorithms to cope with approximate distances.
-%Moreover, we show how a WSPDs can be constructed
-%out of cover trees (we only found 
-%the theoretically more efficient version of
-%net trees used in the literature, but net trees are more difficult to implement
-%efficiently). Both results can be obtained using standard methods, 
-%but seem to be not present in prior work.
+For approximate nearest neighbor, we devise a simple randomized incremental algorithm and show that
+the number of distance queries to find an approximate nearest neighbor is $O(\log n)$ in expectation.
+Our proof is based on the well-known observation that the nearest neighbor changes $O(\log n)$ times
+in expectation when traversing the sequence of points, combined with a packing argument certifying that
+only a constant number of distances needs to be computed in-between two minima.
+We also experimentally evaluate our approach and observe that the approach follows 
+roughly the theoretical prediction.
 

### 2018-11-30 cpp_version/include/dynamic_spanner.h
interim code changes

@@ -239,9 +239,9 @@ public:
         if (dist_ij == -1.0)
             dist_ij = m_matrix[i][j].distance;
 
-#pragma omp parallel
+//#pragma omp parallel
         {
-#pragma omp for schedule(dynamic)
+//#pragma omp for schedule(dynamic)
             for (size_t x = 0; x < m_num_points; x++) {
                 for (size_t y = x + 1; y < m_num_points; y++) {
                     Pair_of_points_info& info_xy = m_matrix[x][y];
@@ -276,7 +276,7 @@ public:
             }
         }
 
-#pragma omp for schedule(dynamic)
+//#pragma omp for schedule(dynamic)
         for (size_t x = 0; x < m_num_points; x++) {
             for (size_t y = x + 1; y < m_num_points; y++) {
                 Pair_of_points_info& info_xy = m_matrix[x][y];
@@ -313,9 +313,9 @@ public:
         }
 
 #if 1
-#pragma omp parallel
+//#pragma omp parallel
         {
-#pragma omp for schedule(dynamic)
+//#pragma omp for schedule(dynamic)
             for (size_t x = 0; x < m_num_points; x++) {
                 for (size_t y = x + 1; y < m_num_points; y++) {
                     Pair_of_points_info& info_xy = m_matrix[x][y];

### 2018-11-30 cpp_version/src/main.cpp

@@ -305,7 +305,7 @@ int main(int argc, char** argv)
     DynamicSpannerR ct_spanner(dist_matrix);
     std::cout << "Spanner initialized" << std::endl;
 
-#if 1
+#if 0
 
     DynamicSpannerR copy(spanner);
 //

### 2018-11-30 python_version/save_in_text_files.py

@@ -6,13 +6,13 @@ from utils_euclidean import *
 import joblib as jl
 
 
-def save_to_files(point_method, dim, n_points, point_file_name=None, matrix_file_name=None, norm = None):
+def save_to_files(point_method, dim, n_points, n_instance, params = None, point_file_name=None, matrix_file_name=None, norm = None):
     if point_file_name == None:
-        point_file_name = "points_{}_{}_{}.txt".format(dim, n_points, point_method)
+        point_file_name = "points_{}_{}_{}_instance_{}.txt".format(dim, n_points, point_method, n_instance)
     if matrix_file_name == None:
         matrix_file_name = "dist_matrix_" + point_file_name
     pm_dic = {"unif": get_uniform_points,
-              "normal": get_points,
+              "normal": get_normal_points,
               "clustered": get_clustered_points,
               "exp": get_exponential_points,
               "beta_05_05": get_beta_05_05,
@@ -24,38 +24,49 @@ def save_to_files(point_method, dim, n_points, point_file_name=None, matrix_file
     if not point_method in pm_dic:
         print("Unknown point_method")
         exit(1)
-    points = pm_dic[point_method](n_points, dim)
+    if params == None:
+        points = pm_dic[point_method](n_points, dim)
+    else:
+        points = pm_dic[point_method](n_points, dim, **params)
     np.savetxt(point_file_name, points, "%15.7f")
     dist_matr = distance_matrix(points)
     np.savetxt(matrix_file_name, dist_matr, "%15.7f")
 
-def save_to_files(point_method, dim, n_points, norm = 2.0, point_file_name=None, matrix_file_name=None):
-    if point_file_name == None:
-        point_file_name = "points_{}_{}_{}.txt".format(dim, n_points, point_method)
-    if matrix_file_name == None:
-        matrix_file_name = "dist_matrix_norm_{}_{}".format(norm,  point_file_name)
-    pm_dic = {"unif": get_uniform_points,
-              "normal": get_points,
-              "clustered": get_clustered_points,
-              "exp": get_exponential_points,
-              "beta_05_05": get_beta_05_05,
-              "beta_5_1": get_beta_5_1,
-              "beta_1_3": get_beta_1_3,
-              "beta_2_2": get_beta_2_2,
-              "beta_2_5": get_beta_2_5,
-              }
[…]

### 2018-11-30 python_version/spanner_experiment_data/plots.py

@@ -65,22 +65,7 @@ if __name__ == "__main__":
     # sys.exit(0)
 
     ds_non_blind_greedy.drop(columns=['spanner_method'], inplace=True)
-    ds_blind_greedy.drop(columns=['spanner_met
-        x = ds_to_plot2.n_points
-        y = ds_to_plot2.dist_num
-
-
-        y = y / np.log(x)
-        # x = np.log(x)
-        print(x)
-        print(y)
-        fit = np.polyfit(x, y, 1)
-        fit_fn= np.poly1d(fit)
-        print(fit_fn(x))
-        plt.xlabel("#points")
-        plt.ylabel("#computed distances / log(#points)")
-        plt.plot(x, y)
-        # plt.plot(x, y, 'ro', x, fit_fn(x), '--k')hod'], inplace=True)
+    ds_blind_greedy.drop(columns=['spanner_method'], inplace=True)
     ds_blind_random.drop(columns=['spanner_method'], inplace=True)
     ds_quasi_greedy.drop(columns=['spanner_method'], inplace=True)
     ds_quasi_shaker.drop(columns=['spanner_method'], inplace=True)
@@ -89,7 +74,8 @@ if __name__ == "__main__":
     ds_blind_random_bad_ratio_cf.drop(columns=['spanner_method'], inplace=True)
     ds_blind_random_bad_ratio_cf_lbf.drop(columns=['spanner_method'], inplace=True)
 
-    ds_all_dims = ds_non_blind_greedy.join(ds_blind_greedy, lsuffix=' (non-blind greedy)', rsuffix=' (blind greedy)', how="outer")
+    # ds_all_dims = ds_non_blind_greedy.join(ds_blind_greedy, lsuffix=' (non-blind greedy)', rsuffix=' (blind greedy)', how="outer")
+    ds_all_dims = ds_non_blind_greedy.join(ds_blind_greedy, rsuffix=' (blind greedy)', how="outer")
     ds_all_dims = ds_all_dims.join(ds_blind_random, rsuffix=' (blind random)')
     ds_all_dims = ds_all_dims.join(ds_quasi_greedy, rsuffix=' (quasi-sorted greedy').join(ds_quasi_shaker, rsuffix=' (quasi-sorted shaker)')
     ds_all_dims = ds_all_dims.join(ds_blind_random_bad_ratio, rsuffix=' (blind random bad ratio)')

### 2018-11-30 python_version/utils_euclidean.py

@@ -10,6 +10,9 @@ def get_points(n_points, dim):
     return np.random.randn(n_points, dim)
 
 
+def get_normal_points(n_points, dim, scale = 1.0):
+    return np.random.normal(scale = scale, size = (n_points, dim))
+
 def get_uniform_points(n_points, dim, max_coord=100.0):
     return np.random.uniform(0.0, max_coord, (n_points, dim))
 

### 2018-11-30 paper/draft.tex

@@
 \newcommand{\doublingdimension}{\Delta}
+
+\newcommand{\ignore}[1]{}
 
@@
 are not counted in the model, hence we can assume that for each pair of points $(u,v)$, we
-know a number $a_{u,v}$ with $\dist(u,v)\leq a_{u,v}\leq 2\dist(u,v)$. This induces an approximate ordering
+know a number $A_{u,v}$ with $\dist(u,v)\leq A_{u,v}\leq 2\dist(u,v)$. This induces an approximate ordering
 of all distances in the metric space, and it is plausible to assume that such an ordering will simplify
@@
 
-\section{Notation and Preliminary Notions}
-Let $(X, \dist)$ be a finite metric space with $n$. Let $G$ be a graph with set of vertices $X$,
-its edges have natural weighting $w(x,y) = \dist(x, y)$ which induces the graph metric $\dist_G$ on $X$.
-Graph $X$ is called a $1 + \eps$-spanner of $P$, if $\dist_G$ is a good approximation of $\dist$:
-$\dist(x,y) \leq \dist_G(x, y) \leq (1+\eps) \dist(x, y)$. Obviously, $G$ must have $\Omega(n)$ edges,
-and we consider $G$ optimal, if it has $O(n)$ edges (for each edge of $G$ we must compute the corresponding
-distance, if we want to use the spanner instead of the original metric, which dictates this
-choice of optimality measure).
+\myparagraph{Outline}
 
+\section{Preliminary Notions}
+%
 A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
@@
 and $k$ is the smallest number having that property.
-\textit{Doubling dimension} of a doubling space is defined as $\log k$
+The \textit{doubling dimension} of a doubling space is defined as $\log k$
 (since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
@@
 is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).
-
-In this paper we consider some fixed metric space $(X, \dist)$, and, to get interesting
-results we must assume doubling property.
 
@@
 \end{proof}
+
+In the following, we will assume throughout that every considered metric space
+has a constant doubling dimension.
+\michael{I think that this sentence is enough. I would remove all what follows.}
 If the doubling dimension of $(X, \dist)$ is larger than $d$, then
@@
 
-\section{Metric Approximation}
+\section{Algorithms for spanner construction}
 
-It is known that optimal spanner can be constructed by the following greedy algorithm: 
-start with an empty graph on vertices $p_i$,
-sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
-until it becomes a $1+\eps$-spanner.
+\myparagraph{Spanners and known constructions}
+%
+Let $(X, \dist)$ be a finite metric space with $n$ vertices. 
+One way to encode the metric space is a complete weighted graph on $X$,
+where the weights correspond to the distances of the points.
+A subgraph $G$ of this graph is called an \emph{$(1+\eps)$-spanner} for $(X,\dist)$ 
+if for any pair of points $(u,v)$,
+the shortest path distance $d_{uv}$ of $u$ and $v$ in $G$ satisfies $d_{u,v}\leq (1+\eps)\dist(u,v)$.
+In other words, the shortest path metric of $G$ is a good approximation of the actual distance for every pair of points.
+Clearly, it is a necessary condition that $G$ is connected, hence every spanner must have at least $n-1$ edges.
 
- In our setting this is the most expensive approach, because it requires
- computing all pairwise distances first. We propose an alternative approach (or rather a family of algorithms), which we call
- \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
- before it decides to add the edge $(p_i, p_j)$.
- The algorithm also maintains lower and upper bounds
- for all pairwise distances, which are updated after a new edge is added.
- These bounds are used in the stopping criterion, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
- for the distance between $p_i$ and $p_j$, respectively, 
- then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
- for all $i \neq j$.
+The \emph{greedy spanner}\michael{Cite something} is a simple algorithm to compute linear-sized spanners:
+\begin{algorithmic}
+\label{alg:greedy_spanner}
+\Function{GreedySpanner}{$P, \eps$}
+    \State {$E\gets\emptyset$}
+    \State {Sort all pairwise distances of points in $P$}
+    \ForAll {pairs $(p_i,p_j)$ in increasing order}
+    \State {$d_{ij}\gets$ Shortest path distance in $(P,E)$}
+    \If{$d_{ij}>(1+\eps)\dist(p_i,p_j)$}
+    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
+    \EndIf
+    \EndFor
+    \Return {$(P,E)$}
+\EndFunction
+\end{algorithmic}
+
+\michael{Add references here whereever appropriate}
+The greedy spanner is guaranteed to return a spanner of size $O(n)$
+(for constant doubling dimension and fixed $\eps>0$. However, it clearly
+has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
+this means that in our cost model, the greedy spanner has the worst possible
+performance.
+
+On the other hand, spanner constructions based on WSPDs only compute
+$O(n\log n+n \eps^{-d})$ distances 
+to construct an $(1+\eps)$-spanner in doubling dimension $d$.
+The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
+this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
+computations. However, the algorithm is significantly more involved.
+
+\myparagraph{Blind spanners}
+%
+We introduce a new framework for constructing spanners
+which we call \emph{blind spanners}: the idea is to maintain,
+for every pair of points $(p_i,p_j)$,
+a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
+initially set to $[0,\infty)$. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
+we pick one of them, compute its distance and update the lower and upper bounds of
+all pairs with respect to the newly acquired information. Here is the pseudocode:
 
@@
     \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
-    \State {$(i,j) \gets $} \Call{GetEdgeToAdd}{}
+    \State {$(i,j) \gets $} \Call{GetNextEdgeToAdd()}{}
     \State {$v \gets \dist(p_i, pj)$}
@@
 and $\infty$ is larger than any real number,
-thus making the predicate in the while loop well-defined.  The procedure \textsc{UpdateBounds} is described in the next subsection,
-now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
-and the update is monotonic: no upper bound can increase, no lower bound can decrease, and, after the distance
-between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
-The most important part of this algorithm is the way of choosing the next edge to be added, and
-there are several possibilities, which give very different results.
+thus making the predicate in the while loop well-defined.
 
-The first possibility will be called \textsc{BlindGreedy}, where we choose a pair $(p_i, p_j)$ that maximizes
-the ratio $b_{i,j} / a_{i,j}$. If the number of maximizing pairs is more than 1, we choose one of them at random.
-Note that our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
+We give the details of the \textsc{UpdateBounds} procedure next.
+Suppose that
+Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
+First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
+of $p_i$ and $p_j$ is exactly $v$.
+To update the upper bound of some entry $b_{k,\ell}$,
+we observe that the shortest path from $p_k$ to $p_\ell$ might now
+go through the new edge. Hence, we update
+\[
+    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
+\]
+Repeating this for all $k,\ell$ yields the updated upper bounds.
+Note that this results in $O(n^2)$ arithmetic operations,
+but no distance computation.
+
+For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
+\[
+    v-b_{k,i}-b_{\ell,j}
+\]
+is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
+the triangle inequality
+%
+\[\dist(p_i,p_j)\leq \dist(p_i,p_k)+\dist(p_k,p_\ell)+\dist(p_\ell,p_j)\]
+by rearranging terms and plugging in the upper bounds for $\dist(p_i,p_k)$
+and $\dist(p_k,p_\ell)$. An analogue bound holds with $i$ and $j$ swapped.
+
+Moreover, the inequalities
+\begin{align*}
+  a_{j,\ell} - v - b_{k,i}&\leq \dist(p_k,p_\ell)\\
+  a_{j,k} - v - b_{j,i}&\leq \dist(p_k,p_\ell)
+\end{align*}
+hold by triangle inequality, and the same is true with $i$ and $j$ swapped.
+This yields $6$ lower bounds for $\dist(p_k,p_\ell)$, and $a_{k,\ell}$
+is updated to the maximum of these six lower bounds and its current value.
+
+\myparagraph{Heuristics}
+%
+The last missing ingredient of our algorithm is the procedure \textsc{GetNextEdgeToAdd},
+that is, how to select the next distance to be computed. We propose two natural choices
+
+\begin{description}
+\item[\textsc{BlindRandom}] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
+we pick one pair uniformly at random
+\item[\textsc{BlindGreedy}] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
+If the maximizing pair is not unique, choose among the maximizing pairs uniformly at random.
+\end{description}
+%
+The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
+in that way hoping to gather most additional information about the metric space.
+Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
 have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
@@
 
-The second possibility will be referred to as \textsc{BlindRandom}, where among all edges $(p_i, p_j)$
-for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach
-by insisting that we must first create a connected graph, or that we must first have some positive lower bound
-for all edges, and only after these requirements are satisfied, we consider other pairs.
+\michael{Perhaps talk about force lower bound here briefly (commented out for now}
+%The second possibility will be referred to as \textsc{BlindRandom}, where among all edges $(p_i, p_j)$
+%for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach
+%by insisting that we must first create a connected graph, or that we must first have some positive lower bound
+%for all edges, and only after these requirements are satisfied, we consider other pairs.
 
+\michael{I have shortened this part considerably. The buckets are nice, but perhaps not needed.
+I commented out the old version}
+
+The next two heuristics assume the existence of a $2$-approximation algorithm for distance
+computation. Denoting by $A_{i,j}$ the number satisfying $\dist(p_i,p_j)\leq A_{i,j}\leq 2 d(p_i,p_j)$,
+We sort all pairwise distances according to the values $A_{i,j}$.
+This yields a roughly sorted sequence of distance, because when $\dist(p_i,p_j)>2\dist(p_k,p_\ell)$,
+then $A_{i,j}>A_{k,\ell}$ is guaranteed.
+We propose two further heuristics that attempt to make use of this sorted sequence.
+\begin{description}
+\item [\textsc{BlindQuasiSortedGreedy}] Traverse the pairs in increasing order with respect to $A_{i,j}$.
+\item [\textsc{BlindQuasiSortedShaker}] Alternates between pairs with small and large $A_{i,j}$
+by traversing in increasing order of $A_{i,j}$ in odd iterations and in decreasing order
+in even iterations.
+\end{description}
+
+\textsc{BlindQuasiSortedGreedy} tries to mimick the blind greedy spanner and hence appears as a natural
+choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
+results. The reason is that no pair acquires useful lower bounds when only short distance are queried
+(the greedy spanner does not have this issue because it knows the distance and hence does not need
+lower bounds).
+Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
+useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
+which alternates between short and long distances.
+
+\ignore{
 The last option that we consider is based on an additional assumption: while
@@
 We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.
+}
 
- \subsection{Maintaining Lower and Upper Bounds}
-
- In this subsection we explain the update procedure that we use. Suppose that
- $a_{k,l}$ and $b_{k,l}$ are valid lower and upper bounds, so that
-\[
-     \forall 1 \leq k < l \leq n \quad a_{k,l} \leq \dist(p_k, p_l) \leq b_{k,l}.
-\]
-Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
-First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$. 
-To update the upper bound of some entry $b_{k,\ell}$,
-we observe that the shortest path from $p_k$ to $p_\ell$ might now
-go through the new edge. Hence, we update
-\[
-    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
-\]
-Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
-time per entry.
-
-For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
-\[
-    v-b_{k,i}-b_{\ell,j}
-\]
-is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
-the triangle inequality
-%
-\[\dist(p_i,p_j)\leq \dist(p_i,p_k)+\dist(p_k,p_\ell)+\dist(p_\ell,p_j)\]
-by rearranging terms and plugging in the upper bounds for $\dist(p_i,p_k)$
-and $\dist(p_k,p_\ell)$. An analogue bound holds with $i$ and $j$ swapped. Hence, we update
-%
-\[a_{k,\ell}\gets \max\{a_{k,\ell},v-b_{k,i}-b_{\ell,j},v-b_{k,j}-b_{\ell,i}\}\]
-%
-for all $k,\ell$.
-
-% Note that after we added the edge $(p_i, p_j)$, the ratio of the upper and lower bound
-% for this pair becomes 1, so a blind spanner algorithm always terminates (of course,
-% we assume that the procedure \textsc{GetEdgeToAdd} picks an edge only once).
-
-\subsection{Experimental Results}
+\section{Experiments on spanners}
 At the moment we cannot provide any theoretical guarantees for the heuristics that we propose to compute blind spanners,

### 2018-12-01 paper/draft.tex
Changes in the ANN proof

@@
  $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
- Then $P$ is finite and its cardinality is bounded by some $C$ that depends only on $d$
- and the ratio $r/R$.
+ Then 
+ \[|P|\leq \left(\frac{2R}{r}\right)^{2d}\]
 % Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $.
@@
 $B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
-contain at most one point from $P$, the claim follows.
+contain at most one point from $P$, 
+\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{2(\log \frac{R}{r/2})d}=\left(\frac{2R}{r}\right)^{2d}.\qedhere\]
 \end{proof}
@@
 
-\section{Approximate Nearest Neighbor}
-
-In this section we consider the standard problem of finding an approximate nearest neighbor: given
+\section{Approximate Nearest Neighbors}
+\label{sec:ann}
+We consider the standard problem of finding an approximate nearest neighbor: given
 $n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
@@
 We assume for simplicity
-that all exact pairwise distances $\dist(p_i, p_j)$ are already computed.
+that all exact pairwise distances $\dist(p_i, p_j)$ are already computed
+(a slight modification of the algorithm can also be applied if only a spanner is available).
 Our goal is to reduce the number of computed distances $\dist(p_i, q)$. We shall also assume w.l.o.g.
@@
 
+Our approach can be summarzied as follows. 
+Fix a random permutation of the points of $P$ 
+and consider the points in that order 
+Fix a random permutation of the points of $P$ (to simplify notation,
+we re-index them, so the order is again $p_1, \dots, p_n$).
+During the loop, we maintain lower bounds of each $p_i$
+to the query point $q$, which are initially all set to $0$.
+We also remember the closest neighbor $c$ that we have seen so
+far and its distance $v$ to $q$.
+We refer to this point as the \emph{candidate}.
+We maintain the invariant that $c$ is an approximate nearest neighbor
+to $q$ for the points $\{p_1,\ldots,p_i\}$.
+When reaching the point $p_i$, we check whether the lower
+bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
+If so, $c$ remains an approximate nearest neighbor and
+we do not query the distance of $p_i$ to $q$.
+Otherwise, we compute $\dist(p_i,q)$ and update the lower bounds
+of all points according to the newly computed distance.
+If $p_i$ is closer to $q$ than $c$, we update $c$ and $v$
+accordingly. At the end of the loop, $c$ is an approximate nearest
+neighbor. The pseudocode of the procedure follows.
+
 \begin{algorithmic}
 \label{alg:ann_blind}
-\Procedure{UpdateBounds}{$p_i, r_i$}
-    \For{$k = i + 1, \dots, n$}
-        \State {$a_k \gets \max(a_k, |\dist(p_i, p_k) - r_i|)$}
-    \EndFor
-\EndProcedure
 
@@
     \State {$[p_1, \dots, p_n] \gets \mbox{random permutation of }P$}
-    \State { $a_i \gets 0$}
+    \State { $a_i \gets 0$ for $i=1,\ldots,n$}
     \Comment {$a_i$ is lower bound for $\dist(p_i, q)$}
@@
 
-The algorithm that we analyze can be described as follows. 
-Fix a random permutation of the points of $P$ (to simplify notation,
-we re-index them, so the order is again $p_1, \dots, p_n$). Pick $p_1$
-as the current candidate, denote $v = r_1$ and start traversing the list of $p_i$'s.
-At each element $p_i$ look at the lower bound for $r_i$,
-if it is larger than $v / (1+\eps)$, then we can safely discard $p_i$.
-Otherwise we must compute the distance $r_i$, and we immediately
- use it to update the lower bounds for all the next points $p_k$'s. 
-If, after computing $r_i$, we see that $p_i$ is a better candidate (we compare $r_ir$ with $v$, not with $v/(1+\eps)$),
-we update the current candidate $c$ and the distance to the query point $v$ accordingly, 
-otherwise $c$ and $v$ remain unchanged.
+We remark that we obtain an exact nearest neighbor algorithm
+when setting $\eps$ to $0$, which means to 
+replacing the condition in the if-statement of the loop by
+with $a_i\geq v$.
+
+The procedure to maintain the lower bounds $a_i$
+is very simple and follows directly
+from triangle inequality.
+
+\begin{algorithmic}
+\Procedure{UpdateBounds}{$p_i, r_i$}
+    \For{$k = i + 1, \dots, n$}
+        \State {$a_k \gets \max(a_k, |\dist(p_i, p_k) - r_i|)$}
+    \EndFor
+\EndProcedure
+\end{algorithmic}
 
 \begin{theorem}
+\label{thm:ann_bound}
     If $(X, \dist)$ is a doubling space, then, for any fixed $\eps > 0$ the
-    algorithm \ref{alg:ann_blind} computes $O(\log n)$ distances $\dist(p_i, q)$ in expectation.
+    algorithm computes $O(\log n)$ distances $\dist(p_i, q)$ in expectation.
 \end{theorem}
 
-Before we start the proof, let us make some remarks. First, we want
-to rephrase what discarding a point means. 
-Let $v$ be the distance from the current candidate $c$ to $q$;
-suppose that we have computed $r_i = \dist(p_i, q)$ which is larger than $v$,
-and want to use it to estimate $r_j = \dist(p_j, q)$ for some other $p_j$. If we assume 
-that $\dist(p_i, p_j) < r_i - v/(1+\eps)$,
-and substitute this in the triangle inequality $\dist(p_j, q) \geq \dist(p_i, q) - \dist(p_i, p_j)$, then we obtain $\dist(p_j, q) \geq v / (1 + \eps)$.
-Thus, if $p_i$ fails to be closer to $q$ than the current candidate, then all points in the ball of radius $r_i - v / (1 + \eps)$,
-even those who are closer to $q$ than $c$, can be discarded, because we are looking for an approximate nearest neighbor.
-Figure \ref{fig:ann_discard_circles} illustrates this.
-Note that with the triangle inequality we can also discard points that are too far from $p_i$, that is the shaded exterior of the bigger balls in the figure.
+Towards the proof, we will use the following geometric lemma which
+can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
+further distance computations of points very close to $p_i$ or very far from $p_i$
+will be avoided.
 
-Secondly,
-we could modify the algorithm to search for the exact neighbor, just by setting $\eps = 0$. In this case, however, 
-the figure \ref{fig:bad_exact_nn_example},
-shows a set of points in the plane for which, regardless of the order we choose them,
-at each iteration of our algorithm we are able to discard only one point (the shaded balls do not contain any other point from the point set), 
-therefore we may not hope to avoid computation of $O(n)$ distances, if we
-are trying to solve the exact nearest neighbor problem.
-The necessity of the doubling property assumption is also easily justified: if all $d(p_i, p_j)$
-and $d(p_i, q)$ are approximately the same, at each iteration of our algorithm we will be able to discard only one point.
+\begin{lemma}
+\label{lem:bound_lemma}
+Assume $r_i=\dist(p_i,q)$ is computed in the algorithm, and let $j>i$.
+\begin{enumerate}
+\item If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps}) r_i$, the algorithm will not compute the distance
+of $p_j$ to $q$.
+\item If $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$, the algorithm will not compute
+the distance of $p_j$ to $q$.
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+The algorithm computes $r_i$ by assumption and updates
+all lower bounds. For $p_j$, it sets $a_j\gets \max (a_j,|\dist(p_i,p_j)-r_i|)$.
+If $|\dist(p_i,p_j)-r_i|\geq (1+\frac{1}{1+\eps})r_i$,
+so 
+\[a_j\geq (1+\frac{1}{1+\eps}r_i - r_i = \frac{r_i}{1+\eps}.\]
+Likewise, if $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$,
+\[a_j\geq r_i-\dist(p_i,p_j) \geq r_i - \frac{\eps}{1+\eps}r_i = \frac{r_i}{1+\eps}.\]
+In both cases, after the point $p_i$ is handled,
+$v\leq r_i$ clearly holds. Since $v$ is only decreasing
+and $a_j$ is only increasing in the algorithm,
+it follows that $a_j\geq \frac{v}{1+\eps}$ when $p_j$ is handled,
+so the algorithm proceeds without a distance computation.
+\end{proof}
 
+With the previous lemma, we can derive an upper bound for the number
+of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
+as follows. In what follows, we let $c_1,\ldots,c_n$ denote the candidates
+at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
+their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
+
+\begin{lemma}
+Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
+\[\left(\frac{2(2+\eps) v_k}{\eps v_\ell}\right)^{2d}\]
+distances to $q$.
+\end{lemma}
 \begin{proof}
+By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
+whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_i=\frac{2+\eps}{1+\eps} v_k$
+around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
+is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
+Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
+have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
+\end{proof}
+
+A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
+the number of computed distances is a constant (since $v_k=v_\ell$). This means that to prove our statement, it would suffice to show
+that the candidate changes only a logarithmic number of times in expectation.
+While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
+that argument.
+
+\begin{proof} (of Theorem~\ref{thm:ann_bound})
+\michael{I need to rewrite this...to be continued asap}
 Let $i_k$ be the {$k$-th} position where the sequence  $m_i := \min \Set{m_1, \dots, m_i}$

### 2018-12-01 paper/draft.tex
Finished Section 5+Conclusion

@@
 \title{Metric Spaces with Expensive Distances}
-\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
+\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
 \author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
@@
 in polynomial time in the size of the graphs.
+Another instance is a collection of high-resolution images
+endowed with the Wasserstein (or Earth Movers) metric.
 \michael{This paragraph needs references. As you know the literature better than me,
@@
 
-\section{Approximate Nearest Neighbors}
+\section{Approximate nearest neighbors}
 \label{sec:ann}
@@
 (a slight modification of the algorithm can also be applied if only a spanner is available).
-Our goal is to reduce the number of computed distances $\dist(p_i, q)$. We shall also assume w.l.o.g.
-that all $\dist(p_i, q)$ are distinct.
+Our goal is to reduce the number of computed distances $\dist(p_i, q)$.
 
@@
 and consider the points in that order 
-Fix a random permutation of the points of $P$ (to simplify notation,
+(to simplify notation,
 we re-index them, so the order is again $p_1, \dots, p_n$).
@@
 far and its distance $v$ to $q$.
-We refer to this point as the \emph{candidate}.
+We refer to the point $c$ as the \emph{candidate}.
 We maintain the invariant that $c$ is an approximate nearest neighbor
@@
 all lower bounds. For $p_j$, it sets $a_j\gets \max (a_j,|\dist(p_i,p_j)-r_i|)$.
-If $|\dist(p_i,p_j)-r_i|\geq (1+\frac{1}{1+\eps})r_i$,
-so 
-\[a_j\geq (1+\frac{1}{1+\eps}r_i - r_i = \frac{r_i}{1+\eps}.\]
+If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps})r_i$, it follows that 
+\[a_j\geq (1+\frac{1}{1+\eps})r_i - r_i = \frac{r_i}{1+\eps}.\]
 Likewise, if $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$,
@@
 
+In what follows, we let $c_1,\ldots,c_n$ denote the candidates
+at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
+their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
 With the previous lemma, we can derive an upper bound for the number
 of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
-as follows. In what follows, we let $c_1,\ldots,c_n$ denote the candidates
-at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
-their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
+as follows.
 
 \begin{lemma}
+\label{lem:sequence_lemma}
 Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
@@
 By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
-whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_i=\frac{2+\eps}{1+\eps} v_k$
+whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
 around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
@@
 A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
-the number of computed distances is a constant (since $v_k=v_\ell$). This means that to prove our statement, it would suffice to show
+the number of computed distances is a constant (since $v_k=v_\ell$). 
+This means that to prove Theorem~\ref{thm:ann_bound}, it would suffice to show
 that the candidate changes only a logarithmic number of times in expectation.
@@
 \begin{proof} (of Theorem~\ref{thm:ann_bound})
-\michael{I need to rewrite this...to be continued asap}
-Let $i_k$ be the {$k$-th} position where the sequence  $m_i := \min \Set{m_1, \dots, m_i}$
-changes, so that
+%
+In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
+such that $r_i<r_k$ for all $1\leq i\leq k-1$. We call an element
+of this form a \emph{minimum} of the sequence. 
+A standard backwards analysis argument~\cite{seidel-backwards} shows that the probability
+of $p_k$ being a minimum is at most $1/k$, so that the number of minima
+in the sequence is $O(\log n)$ in expectation.
+
+Note that for $\eps>0$,
+a minimum $p_k$ is not necessarily the candidate $c_k$ because a previous point
+in the sequence close to $p_k$ might have caused the lower bound $a_k$ to be
+in the interval $[\frac{v_k}{1+\eps},v_k]$, which leads to not
+computing the distance $r_k$. However, it is true that
+$v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
+nearest neighbor of $\{p_1,\ldots,p_k\}$.
+
+Now, let $p_k$, $p_\ell$ be two consecutive minima in the sequence
+(we also allow that $\ell=n+1$ if $k$ is the last minimum in the sequence).
+Note that $v_{\ell-1}\geq r_k$ because each $v_j$ is equal to $r_i$ for
+some $i\leq j$, and in the sequence $r_1,\ldots,r_{\ell-1}$, $r_k$
+is minimal by construction. Using Lemma~\ref{lem:sequence_lemma},
+the number of distance computations among the points
+$p_k,\ldots,p_{\ell-1}$ is at most
 \[
-m_1 = m_2 = \dots = m_{i_1 - 1} > m_{i_1} = m_{i_1 + 1} = \dots = m_{i_2 - 1} > m_{i_2} = \dots. 
-\]
-In other words, if we traversed the list $p_1, \dots, p_n$ and used the bruteforce algorithm
-for exact nearest neighbor, then $i_k$ would the indices where the current candidate
-changes, because we found a point that is closer to $q$ than all the previously seen points.
-We shall refer to $i_k$ as the \textit{change position}.
-Standard backwards analysis argument (e.g., \cite{dubch}, Sect. 4.4) shows that $k=O(log n)$.
-
-    Consider two consecutive change positions $s:=i_j$ and $t:=i_{j+1}$. We claim that in the sequence
-$p_s,...,p_{t-1}$
-only a constant number of distances are computed by our algorithm, irrespective of its length. This claim immediately implies the theorem.
-
-To see that it true, let $c_1,...,c_n$ be the candidates after the $i$-th step of
-the for loop of the algorithm. We have no guarantee that $c_s=p_s$, because $p_s$ it is possible
-    that $p_s$ was discarded without computing $r_s$ (see the picture). However, we know that
-$d(c_s,q)<=(1+ \eps) r_s$.
-Indeed, suppose that $c_s \neq p_s$ and $r_s < d(c_s,q)/(1+\eps)$. In this
-    case $p_s$ cannot be discarded 
-by its lower bound, and hence the distance to $p_s$ would
-have been queried, but this distance is a minimum among $r_1, \dots, r_s$,
-    and then we must reset $c_s$ to be $p_s$, a contradiction.
-
-It follows that any point in the sequence $p_s,...,p_{t-1}$ for which a
-distance is queried must lie in the ball around $c_s$ of radius at most $2(1+\eps)r_s$.
-
-Moreover, we know that for any index $j=s,\dots, t-1$ it holds
-$d(c_j,q)>=r_s$.
-
-Indeed, assume the opposite. $c_j$ is equal to $p_k$ for some $k<t$. If $k<s$,
-that is, the candidate came from before, than $p_s$ would not be in the
-change position. If $s<k<t$, then at least one of the
-elements in $p_{s+1},\dots,p_k$ is a change position, contradicting
-the assumption that $p_s$ and $p_t$ were chosen to be consecutive change positions.
-
-Hence, it follows that whenever a distance is queried, we can exclude a
-ball of radius $\eps r_s$ around the query point. All queried points must
-hence have pairwise distance of at least $ \eps r_s / 2$. 
-Now we can apply the packing lemma \ref{lem:packing_lemma} to bound the number of
-queried points by some constant that depends only on $\eps$ and the dimension. Indeed,
-they all are contained in a ball of radius $R = 2(1+\eps)r_s$, and, if we divide this by $\eps r_s$, the ratio depends only on $\eps$.
+\left(\frac{2(2+\eps) v_k}{\eps v_{\ell-1}}\right)^{2d}\leq \left(\frac{2(2+\eps)(1+\eps)r_k}{\eps r_k}\right)^{2d}=\left(\frac{2(2+\eps)(1+\eps)}{\eps}\right)^{2d},\]
+which is a constant depending only of $\eps$ and $d$, irrespective of the length of the sequence.
+Since $p_1,\ldots,p_n$ decomposes into $O(\log n)$ such sequences in expectation, the result follows.
 \end{proof}
 
-We should note that the constant factor that is obtained in this proof can be very pessimistic,
-because in practice the algorithm can discard much more points.
+We point out that the proof fails for $\eps=0$ because in that case, we cannot exlude an $\eps$-ball
+of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
+Indeed, as the example in Figure~\ref{} shows, there are point sets where the expected number
+of distance computations for exact nearest neighbor is linear.
 
-%\begin{proof}
-%    We start with the lemma of purely combinatorial (or probabilistic) nature.
-%\begin{lemma}
-%    Let $z$ be a random permutation of $n$ distinct real numbers,
-%    $z = X_{1}, \dots, X_{n}$, and denote $x_t = \min \Set{X_1, \dots, X_t}$.
-%    The expected number of positions $t$ where the sequence $X_t$ changes is $O(n)$.
-%\end{lemma}
-%    \begin{proof}
-%        The proof is easily obtained with the standard backwards analysis technique and can be found in the appendix.
-%        %Define random variable $Y_i$ to be 1, if $X_{i-1} = X_{i}$, and 0 otherwise.
-%        %We must estimate $Y = \sum_{i = 2}^n Y_i$, and 
-%        %\[
-%        %    E[Y] = \sum_{i = 2}^n E[Y_i] = \sum_{i = 2}^n P(Y_i = 1).
-%        %\]
-%        %It remains to observe than $P(Y_i) = 1 / (n - i)$, so the expectation
-%        %of $Y$ is the sum of the 
-%    \end{proof}
+Finally, we remark that a fast $2$-approximation algorithm for $\dist$ would lead to a straight-forward
+optimization: compute a $2$-approximation of $\dist(p_i,q)$ for all $1\leq i\leq n$
+and let $m$ denote the minimal approximate distance encountered. Then, we can discard all points
+whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.
 
-%\begin{lemma}
-%    There exists a constant $C$ that depends only on $\eps$ and the doubling dimension
-%    such that after computing at most $C$ distances $\dist(p_k, q)$
-%    either the algorithm verifies that the current candidate $p_i$ is a correct answer and terminates
-%    or the algorithm finds another candidate $p_j$ such that $r_j < r_i$.
-%\end{lemma}
-%\begin{proof}
-%    First, all points that are not contained in the ball $B(p_i, 2r_i)$
-%    are discarded without computing the distance to them.
-%    Now we have to examine all points $p_j$ inside this ball. 
-%    Suppose that we cannot discard the current point $p_j$ based
-%    on the lower bound for $r_j$, then we must compute the distance $r_j$.
-%    There are two possibilities: either $r_j \leq r_i$ or $r_j > r_i$.
-%    The latter case means that we discard the point $p_j$,
-%    but we also discard all points of $P$ that are contained in some ball around it.
-%    The triangle inequality tells us that we can take the ball of radius $r_j - r_i / (1 + \eps)$,
-%    but in this proof it is sufficient to consider the ball of radius $r_i \eps / (1 + \eps)$.
-%    In the figure this radius corresponds to the width of the annulus surrounding $q$,
-%    and it is  geometrically clear that this ball is indeed smaller than the one we could have discarded.
-%    The advantage of choosing this small radius is that, if we divide it by $2r_i$, we 
-%    get the expression $\eps / (2(1 + \eps))$, that depends only on $\eps$.
-%    By lemma \ref{lem:packing_lemma_dd} we know that there is an upper bound $T$ for the number of points
-%    $p_i$ inside the ball $B(p_i, 2r)$ that are at the distance at least $r_i / 2 ( 1 + \eps)$
-%    from each other, and the constant $T$ depends only on $\eps$ and the doubling dimension.
-%    This means that even in the worst case after we have computed $T$ distances, we have discarded all
-%    points of $P$ that lie outside of the ball $B(q, r_i)$, because all other points that lie in $B(p_i, 2r_i) \setminus B(q, r_i)$ 
-%    do not require distance computation. Indeed, each of these points lies in
-%    one of the balls $B(p_j, r_i \eps / (1 + \eps)$ for whose center we computed the distance already,
-%    and is therefore discarded, Now, either we have exhausted the set $P$, and in this
-%    case we terminate with an approximate nearest neighbor, or the next available $p_j$ that requires
-%    distance computation necessarily lies in $B(q, r_i)$, thus we found a closer neighbor. This means that $C = T+1$.
-%    Note that the algorithm still computes an approximate nearest neighbor, not the exact one, 
-%    because some of the points that are inside the annulus can be discarded, 
-%    if there is a point from $P$ close to them but outside of the annulus. 
-%\end{proof}
-
-%Now the theorem follows from the lemma by standard argument
-%of the backwards analysis.
-
-%\end{proof}
-
-\subsection{Experimental Results}
+\subsection{Experiments on approximate nearest neighbors}
 
@@
 
-\section{Conclusion and Future Work}
+\section{Conclusion and future work}
 \label{sec:conclusion}
+%
+We have introduced a new cost model for the analysis of algorithms
+for metric spaces that fits the situation that computing an individual distance
+is more costly than other types of primitive operations.
+Our theoretical and experimental results are under the usual assumption
+that the metric space has a low doubling dimension.
+However, in our motivating example of collections of persistence diagrams
+or Reeb graphs, this assumption does not hold. For instance,
+the space of persistence diagrams has an infinite doubling dimension.
+Nevertheless, realisitic data sets are usually not just a random sample
+in that infinite-dimensional space, but have structures
+(e.g. clusters of close-by diagrams) which should be favorable for our approach
+We plan to consider the quality of our algorithms for persistence diagrams
+as future work.
+
+On the theoretical side, the obvious next question is whether our strategy
+for blind spanners yields a linear spanner in expectation. 
+Our experiments are not conclusive enough in this respect to make this conjecture yet.
+However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
+that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
+which is the sum of the cardinalities of all pairs in a WSPD.
+The weight of a WSPD can be quadratic, but premilinary
+experimental evaluation on worst-case examples do not show such a quadratic
+behavior. Therefore, we postpone the theortical analysis of the spanner construction
+to an extended version of this article.
+
+The existence of a $2$-approximation algorithm did not help us to significantly
+reduce the number of exact distance computations, although it seems obvious
+that knowing the all approximate distances is useful.
+We pose the question what heuristic could make more use of this feature.
+
+\ignore{
 We presented experimental evidence showing that in low
@@
 % and then the optimal matcih
+}
 

### 2018-12-03 cpp_version/include/quadtree.h
quadtree wspd

@@ -1,0 +1,8 @@
+//
+// Created by narn on 03.12.18.
+//
+
+#ifndef SPANNER_WASSERSTEIN_QUADTREE_H
+#define SPANNER_WASSERSTEIN_QUADTREE_H
+
+#endif //SPANNER_WASSERSTEIN_QUADTREE_H

### 2018-12-03 python_version/quadtree_wspd.py

@@ -1,0 +1,215 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import utils_euclidean as ue
+
+dist_matrix = np.zeros(shape=(0, 0))
+dist_calls_matrix = np.zeros(shape=(0, 0), dtype=np.bool)
+
+# euclidean_dist_calls = 0
+# euclidean_dist_dict = {}
+
+class Point:
+    def __init__(self, x, y):
+        self.x = x
+        self.y = y
+
+def euclidean_dist_with_cache(x, y):
+    assert(isinstance(x, int))
+    assert(isinstance(y, int))
+    # global euclidean_dist_calls
+    # global euclidean_dist_dict
+    global dist_matrix
+    global dist_calls_matrix
+    if x == y:
+        return 0.0
+    dist_calls_matrix[x][y] = dist_calls_matrix[y][x] = 1
+    return dist_matrix[x][y]
+
+    # key1 = (tuple(x), tuple(y))
+    # if key1 in euclidean_dist_dict:
+    #     return euclidean_dist_dict[key1]
+    # else:
+    #     key2 = (tuple(y), tuple(x))
+    #     result = np.sqrt(np.sum((x - y) * (x - y)))
+    #     euclidean_dist_calls += 1
+    #     euclidean_dist_dict[key1] = result
+    #     euclidean_dist_dict[key2] = result
+    #     return result
+
+
+def euclidean_dist_no_count(x, y):
+    return np.sqrt(np.sum((x - y) * (x - y)))
+
+
+class QuadtreeNode:
+
+    def __init__(self, min_x, min_y, max_x, max_y, level, points):
+        global global_points
+        assert (len(points) >= 1)
+        self.min_x = min_x
+        self.min_y = min_y
+        self.max_x = max_x
+        self.max_y = max_y
+        self.points = points
+
+        mid_x = 0.5 * (min_x + max_x)
+        mid_y = 0.5 * (min_y + max_y)
+
+        if len(points) == 1:
+            return
+
+        p00 = []
+        p01 = []
+        p10 = []
+        p11 = []
+
+        for p in points:
+            p_x = global_points[p][0]
+            p_y = global_points[p][1]
+            if p_x < mid_x:
+                if p_y < mid_y:
+                    p00.append(p)
+                else:
+                    p01.append(p)
+            else:
+                if p_y < mid_y:
+                    p10.append(p)
+                else:
+                    p11.append(p)
+
[…]

### 2018-12-04 paper/draft.tex
fix references

@@
 \usepackage{url}
-%\usepackage{floatflt}
-%\usepackage{mathptmx}
+\usepackage{floatflt}
+\usepackage{mathptmx}
 \usepackage{epsfig}
@@
 of diagrams considered. For the case of Reeb graphs, the situation is even
-worse: while several metrics on Reeb graphs have been proposed,
+worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
+\cite{di2016edit}),
 not even an constant-factor approximation algorithm is known that runs
@@
 Another instance is a collection of high-resolution images
-endowed with the Wasserstein (or Earth Movers) metric.
+endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).
 \michael{This paragraph needs references. As you know the literature better than me,
@@
 We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
-\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds.\michael{Cite}
+\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
@@
 has a constant doubling dimension.
-\michael{I think that this sentence is enough. I would remove all what follows.}
-If the doubling dimension of $(X, \dist)$ is larger than $d$, then
-there exists a ball of radius $R$ such that there are at least $2^d$ points with pairwise distance at least $R$. On the other hand,
-the maximal distance between any two of these points is at most $2R$, and we have a lot of points with approximately the same distance
-between them. If we want a high-quality spanner on these points, then we can expect that its size will be quadratic, not linear;
-so the space of high doubling dimension cannot have optimal spanners. Thus the question of approximating
-the non-doubling metric without computing many distances cannot be solved positively, and we should
-restrict our attention to the case of doubling spaces.
+%\michael{I think that this sentence is enough. I would remove all what follows.}
+%If the doubling dimension of $(X, \dist)$ is larger than $d$, then
+%there exists a ball of radius $R$ such that there are at least $2^d$ points with pairwise distance at least $R$. On the other hand,
+%the maximal distance between any two of these points is at most $2R$, and we have a lot of points with approximately the same distance
+%between them. If we want a high-quality spanner on these points, then we can expect that its size will be quadratic, not linear;
+%so the space of high doubling dimension cannot have optimal spanners. Thus the question of approximating
+%the non-doubling metric without computing many distances cannot be solved positively, and we should
+%restrict our attention to the case of doubling spaces.
 
@@
 
-The \emph{greedy spanner}\michael{Cite something} is a simple algorithm to compute linear-sized spanners:
+The \emph{greedy spanner}(\cite{althofer1993sparse}) is a simple algorithm to compute linear-sized spanners:
 \begin{algorithmic}
@@
 \michael{Add references here whereever appropriate}
-The greedy spanner is guaranteed to return a spanner of size $O(n)$
-(for constant doubling dimension and fixed $\eps>0$. However, it clearly
+The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
+(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
+it was also shown to return the sparsest graph. However, it clearly
 has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
@@
 have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
-lower bounds for every edge before it will start adding any other edges.
+lower bounds for every edge before it will start adding any other edges. Based on this observation,
+we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
+first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
+then the algorith can only choose one of the corresponding edges, etc).
 
@@
 \begin{figure}[ht]
-    \label{fig:spanner-sparseness}
     \includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
@@
     in dimension 2, $\eps = 0.1$.}
+    \label{fig:spanner_sparseness}
 \end{figure}
@@
 \begin{figure}[ht]
-    \label{fig:blind-greedy-only}
     \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
     \caption{}
+    \label{fig:blind_greedy_only}
 \end{figure}
@@
 \begin{figure}[ht]
-    \label{fig:blind-rbr-variants}
     \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
     \caption{Comparison of the four variants of \textsc{BlindRandomBadRatio} algorithm.}
+    \label{fig:blind_rbr_variants}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
+    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\epsilon$. Data is for 400 normally distributed points in $\R^2$.}
+    \label{fig:spanner_eps_dependence}
 \end{figure}
@@
 
-The plot \ref{fig:spanner-sparseness}.
+The plot \ref{fig:spanner_sparseness}.
 shows the number of edges of the spanner.
@@
 \begin{table}[]
-\label{tbl:regr_coeff_spanner}
 \begin{tabular}{|l|l|l|}
@@
 on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
+\label{tbl:regr_coeff_spanner}
 \end{table}
 
-As for different variants of the \textsc{BlindRandomBadRatio} algorithm,
+As for different variants of the \textsc{BlindRandom} algorithm,
 we note that their performance is almost the same, and the algorithm works
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the \textsc{BlindRandomBadRatio} (see \ref{fig:blind-rbr-variants}).
+the other two variants of the \textsc{BlindRandom} (see \ref{fig:blind_rbr_variants}).
 
@@
 disappointing. We show the results for the best algorithm, \textsc{BlindGreedy},
-in dimensions 2, 3 and 4 in the plot \ref{fig:blind-greedy-only}.
+in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_only}.
 We can see that already in dimension 4 the best available algorithm
@@
 
-The plot \ref{fig:spanner-eps-dependence}
-compares the \textsc{BlindGreedy} and \textsc{BlindRandomRatioForceLowerBound} algorithms
-on \textbf{uniform} point sets for different choices of $\eps$.
+The plot \ref{fig:spanner_eps_dependence}
+compares the \textsc{BlindGreedy} and \textsc{Greedy} algorithms
+on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
+is approximately the same for both algorithms. Since it is not cleary seen from the picture, we also note 
+that the ratio of the number of edges decreases for
+smaller values of $\eps$: 
+for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
+while for $\eps = 1/32$ the ratio is 2.6
 
@@
 \begin{figure}[ht]
-    \label{fig:ann_const_ratio}
     \includegraphics[width=0.7\textwidth]{pics/log_dependency_constant_ratio.png}
@@
     points.}
+    \label{fig:ann_const_ratio}
 \end{figure}
@@
 \begin{figure}[ht]
-    \label{fig:ann_dimension_dependency}
     \includegraphics[width=0.7\textwidth]{pics/ann_computed_distances_dimension_dep.png}
     \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
+    \label{fig:ann_dimension_dependency}
 \end{figure}

### 2018-12-04 cpp_version/include/cover_tree/wspd.h
interim

@@ -42,6 +42,7 @@ namespace wasser_spanner {
 
         CoverTree* m_cover_tree;
         int m_num_points;
+        int m_num_spanner_edges;
         double m_epsilon;
         double m_tau;
         DynamicSpannerR* m_spanner;

### 2018-12-04 cpp_version/include/dynamic_spanner.h

@@ -239,9 +239,9 @@ public:
         if (dist_ij == -1.0)
             dist_ij = m_matrix[i][j].distance;
 
-//#pragma omp parallel
+#pragma omp parallel
         {
-//#pragma omp for schedule(dynamic)
+#pragma omp for schedule(dynamic)
             for (size_t x = 0; x < m_num_points; x++) {
                 for (size_t y = x + 1; y < m_num_points; y++) {
                     Pair_of_points_info& info_xy = m_matrix[x][y];
@@ -276,7 +276,7 @@ public:
             }
         }
 
-//#pragma omp for schedule(dynamic)
+#pragma omp for schedule(dynamic)
         for (size_t x = 0; x < m_num_points; x++) {
             for (size_t y = x + 1; y < m_num_points; y++) {
                 Pair_of_points_info& info_xy = m_matrix[x][y];
@@ -313,9 +313,9 @@ public:
         }
 
 #if 1
-//#pragma omp parallel
+#pragma omp parallel
         {
-//#pragma omp for schedule(dynamic)
+#pragma omp for schedule(dynamic)
             for (size_t x = 0; x < m_num_points; x++) {
                 for (size_t y = x + 1; y < m_num_points; y++) {
                     Pair_of_points_info& info_xy = m_matrix[x][y];

### 2018-12-04 cpp_version/include/quadtree.h

@@ -5,4 +5,24 @@
 #ifndef SPANNER_WASSERSTEIN_QUADTREE_H
 #define SPANNER_WASSERSTEIN_QUADTREE_H
 
+#include <vector>
+
+namespace quadtree {
+
+using Matrix = std::vector<std::vector<double>>;
+
+struct QuadtreeNode {
+  std::vector<int> m_children;
+  int m_level;
+};
+
+
+struct Quadtree {
+  Quadtree(const Matrix& distance_matrix, Matrix& distance_computed_matrix);
+};
+
+
+
+}
+
 #endif //SPANNER_WASSERSTEIN_QUADTREE_H

### 2018-12-04 cpp_version/src/main.cpp

@@ -305,7 +305,7 @@ int main(int argc, char** argv)
     DynamicSpannerR ct_spanner(dist_matrix);
     std::cout << "Spanner initialized" << std::endl;
 
-#if 0
+#if 1
 
     DynamicSpannerR copy(spanner);
 //
@@ -322,7 +322,13 @@ int main(int argc, char** argv)
     // TODO very ugly
     DynamicSpannerR* ds = &ct.m_dspanner;
 
-    std::cout << "Cover tree built. Requested distance : " << ds->get_fraction_of_requested_distances() << ", " << ds->get_number_of_requested_distances() << " out of " << ds->m_num_points * (ds->m_num_points  -1 )/ 2 << std::endl;
+    console->info("Cover tree built. Requested distance: {}, fraction = {} / {} = {}",
+            ds->get_number_of_requested_distances(),
+            ds->get_number_of_requested_distances(),
+            ds->m_num_points * (ds->m_num_points  -1 )/ 2,
+            ds->get_fraction_of_requested_distances());
+
+//    std::cout << "Cover tree built. Requested distance : " << ds->get_fraction_of_requested_distances() << ", " << ds->get_number_of_requested_distances() << " out of " <<  << std::endl;
 //              << ", computed distances : " << ds->get_fraction_of_computed_distances() << std::endl << std::endl;
 
     copy = *ds;
@@ -335,7 +341,8 @@ int main(int argc, char** argv)
     WspdNode::dspanner = &ct.m_dspanner;
     WSPD wspd(ct, eps);
 
-    std::cout << "WSPD built. Requested distance : " << ds->get_fraction_of_requested_distances() << ", " << ds->get_number_of_requested_distances() << std::endl;
+    console->info("WSPD built. Requested distance: {}, fraction = {}", ds->get_number_of_requested_distances(), ds->get_fraction_of_requested_distances());
+//    std::cout << "WSPD built. Requested distance : " << ds->get_fraction_of_requested_distances() << ", " << ds->get_number_of_requested_distances() << std::endl;
 //              << ", computed distances : " << ds->get_fraction_of_computed_distances() << std::endl << std::endl;
 
 //    copy = *ds;
@@ -347,10 +354,13 @@ int main(int argc, char** argv)
 
     //ds->print_ratios();
     wspd.make_spanner1();
[…]

### 2018-12-04 cpp_version/src/wspd.cpp

@@ -146,6 +146,7 @@ size_t WSPD::size() const
 
 void WSPD::make_spanner1()
 {
+    m_num_spanner_edges = 0;
     m_graph = std::unique_ptr<Graph>(new Graph(m_num_points));
     initialize_incremental_components(*m_graph, m_conn_comps);
     incremental_components(*m_graph, m_conn_comps);
@@ -157,6 +158,7 @@ void WSPD::make_spanner1()
         }
         auto d = m_spanner->get_distance_no_cache(i, j);
         boost::add_edge(i, j, EdgeWeightProperty(d), *m_graph);
+        m_num_spanner_edges++;
         m_conn_comps.union_set(i, j);
     }
 }

### 2018-12-04 python_version/quadtree_wspd.py

@@ -201,7 +201,7 @@ def make_spanner(eps):
 
 
 if __name__ == "__main__":
-    n_points = 4000
+    n_points = 8000
     n_pairs = int(n_points * (n_points - 1 ) / 2)
     np.random.seed(1)
     global_points = ue.get_uniform_points(n_points, dim=2, max_coord=1.0)

### 2018-12-04 python_version/spanner_experiment_data/plots.py

@@ -32,6 +32,18 @@ import matplotlib.pyplot as plt
 
 if __name__ == "__main__":
 
+    if True:
+        ds = pd.read_csv("eps_dep.csv", header = 0, sep = ";")
+        ds.set_index(["dim"], drop=True)
+        ds  = ds.loc[(ds['dim'] == 2)].set_index(["Epsilon"], drop = True)
+        # ds.plot(y=["Edges (greedy)", "Edges (blind greedy)"], logx = True, xticks = [0.0625, 0.09375, 0.125, 0.25, 0.5, 1, 2])
+        ds.plot(y=["Edges (greedy)", "Edges (blind greedy)"], logx = True)
+        plt.xticks([0.0625, 0.09375, 0.125, 0.25, 0.5, 1, 2], ['1/32', '1/16', '1/8', '1/4', '1/2', '1','2'])
+        plt.show()
+
+
+    sys.exit(0)
+
     # ds = pd.read_csv("n_edges_blind_greedy_dimensions.csv", header  = 0, sep = ';')
     # ds = ds.set_index(["# points"], drop=True)
     # ds = ds.sort_index(ascending=True)

### 2018-12-04 paper/draft.tex
Change constant in packing lemma; add bad example for exact NN

@@
 \keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-\subjclass{dummy}
+%\subjclass{\ccsdesc[500]{Theory of computation~Nearest neighbor algorithms}}
+\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
 % \category{dummy}
@@
 
-\myparagraph{Outline}
-
 \section{Preliminary Notions}
@@
  Then 
- \[|P|\leq \left(\frac{2R}{r}\right)^{2d}\]
+ \[|P|\leq \left(\frac{4R}{r}\right)^{d}\]
 % Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $.
@@
 contain at most one point from $P$, 
-\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{2(\log \frac{R}{r/2})d}=\left(\frac{2R}{r}\right)^{2d}.\qedhere\]
+\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
 \end{proof}
@@
 where the weights correspond to the distances of the points.
-A subgraph $G$ of this graph is called an \emph{$(1+\eps)$-spanner} for $(X,\dist)$ 
+A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(X,\dist)$ 
 if for any pair of points $(u,v)$,
@@
 
-\michael{Perhaps talk about force lower bound here briefly (commented out for now}
+%\michael{Perhaps talk about force lower bound here briefly (commented out for now}
 %The second possibility will be referred to as \textsc{BlindRandom}, where among all edges $(p_i, p_j)$
@@
 
-\michael{I have shortened this part considerably. The buckets are nice, but perhaps not needed.
-I commented out the old version}
+%\michael{I have shortened this part considerably. The buckets are nice, but perhaps not needed.
+%I commented out the old version}
 
@@
     \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
-    \caption{}
+    \caption{Results of blind greedy spanner for different dimensions.}
     \label{fig:blind_greedy_only}
@@
 which is explained by the fact that the number of points
-on which we computed is not large enough to clearly see
-the linear dependency.
+on which we computed spanners is not large enough to clearly see
+the linear dependence.
 
@@
 \end{tabular}
-\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependency of the number of edges
+\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
 on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
@@
 
-Our approach can be summarzied as follows. 
+Our approach can be summarized as follows. 
 Fix a random permutation of the points of $P$ 
@@
 Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
-\[\left(\frac{2(2+\eps) v_k}{\eps v_\ell}\right)^{2d}\]
+\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
 distances to $q$.
@@
 \[
-\left(\frac{2(2+\eps) v_k}{\eps v_{\ell-1}}\right)^{2d}\leq \left(\frac{2(2+\eps)(1+\eps)r_k}{\eps r_k}\right)^{2d}=\left(\frac{2(2+\eps)(1+\eps)}{\eps}\right)^{2d},\]
+\left(\frac{4(2+\eps) v_k}{\eps v_{\ell-1}}\right)^{d}\leq \left(\frac{4(2+\eps)(1+\eps)r_k}{\eps r_k}\right)^{d}=\left(\frac{4(2+\eps)(1+\eps)}{\eps}\right)^{d},\]
 which is a constant depending only of $\eps$ and $d$, irrespective of the length of the sequence.
@@
 of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
-Indeed, as the example in Figure~\ref{} shows, there are point sets where the expected number
+Indeed, as the example in Figure~\ref{fig:exact_bad_example} shows, there are point sets where the expected number
 of distance computations for exact nearest neighbor is linear.
+
+\begin{figure}[ht]
+    \includegraphics[width=0.7\textwidth]{pics/exact_bad_example.png}
+    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
+    Next $p_i$ is placed in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$.}
+    \label{fig:exact_bad_example}
+\end{figure}
 
@@
     \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
-    \label{fig:ann_dimension_dependency}
+    \label{fig:ann_dimension_dependence}
 \end{figure}
@@
 As expected, in high dimensions the algorithm does not perform as well.
-In figure \ref{fig:ann_dimension_dependency} we plot the average number of computed distances
+In figure \ref{fig:ann_dimension_dependence} we plot the average number of computed distances
 for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
@@
 
-% We also looked at the dependency on the density. Note that our algorithm is not invariant under scaling: if we multiply the metric
+% We also looked at the dependence on the density. Note that our algorithm is not invariant under scaling: if we multiply the metric
 % by $\lambda$, we should also change $\eps$ to $\lambda \eps$, if we want

### 2018-12-05 paper/draft.tex
pgfplots

@@
 \usepackage{braket}
+\usepackage{pgfplots}
 
@@
 We consider two standard operations on metric spaces,
-namely the construction of an $\epsilon$-spanner
+namely the construction of a $1+\eps$-spanner
 and the computation of an approximate nearest
@@
 performing $O(\log n)$ distance computations per query point.
+Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
+and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylog time, with polynomial preprocessing
+and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
 However, for our relaxed cost model, we pose the question whether simpler constructions achieve
@@
 \begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind quasi-sorted greedy}
+
+    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind quasi-sorted shaker}
+
+    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{WSPD}
+
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, lower bound first}
+
+    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind greedy}
+
+    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Greedy}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
     \caption{Number of edges in blind spanners generated by different variants of the blind
-    algorithm. Greedy non-blind algorithm is included for comparison. The plot is for normally distributed points
+    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for normally distributed points
     in dimension 2, $\eps = 0.1$.}
@@
 \begin{figure}[ht]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
+        legend pos=outer north east]
+
+    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    %\addlegendentry{Blind quasi-sorted greedy}
+
+    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    %\addlegendentry{Blind quasi-sorted shaker}
+
+    %\addplot+ table [x=n_points, y=ratio_wspd, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    %\addlegendentry{WSPD}
+
+    %\addplot+ table [x=n_points, y=ratio_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    %\addlegendentry{Blind random}
+
+    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addlegendentry{Blind random, lower bound first}
+
+    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addlegendentry{Blind greedy}
+
+    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addlegendentry{Greedy}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
+    \caption{Ratio \# edges / \# points for different variants of spanner 
+    algorithms. The plot is for normally distributed points
+    in dimension 2, $\eps = 0.1$.}
+    \label{fig:spanner_ratio}
+\end{figure}
+
+\begin{figure}[ht!]
     \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
@@
 
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
-    \caption{Comparison of the four variants of \textsc{BlindRandomBadRatio} algorithm.}
+\begin{figure}[ht!]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, connect first}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, lower bound first}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, connect first, lower bound first}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Comparison of the four variants of \textsc{BlindRandom} algorithm.}
     \label{fig:blind_rbr_variants}
@@
 
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
-    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\epsilon$. Data is for 400 normally distributed points in $\R^2$.}
+%\begin{figure}[ht]
+%    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
+%\end{figure}
+
+\begin{figure}[ht!]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
+        legend pos=outer north east]
+
+    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {eps_dep.csv};
+    \addlegendentry{Greedy, dim = 2}
+
+    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {eps_dep.csv};
+    \addlegendentry{Blind greedy, dim = 2}
+
+    %\addplot+ table [x=Epsilon, y=edges_greedy_dim_3, col sep=comma] {eps_dep.csv};
+    %\addlegendentry{Greedy, dim = 3}
+
+    %\addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_3, col sep=comma] {eps_dep.csv};
+    %\addlegendentry{Blind greedy, dim = 3}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\R^2$ and $\R^3$.}
     \label{fig:spanner_eps_dependence}
 \end{figure}
+
+%\begin{figure}[ht]
+%    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
+%\end{figure}
 
@@
 
-The plot \ref{fig:spanner_sparseness}.
+Figure \ref{fig:spanner_sparseness}.
 shows the number of edges of the spanner.
@@
 perform significantly better than both variants of \textsc{BlindQuasiSorted}.
-it is important to remember that for all blind spanners the number of edges
+It is important to remember that for all blind spanners the number of edges
 is equal to the number of distances that were actually computed, while
 for non-blind greedy spanner the latter number is always $n \choose 2$.
-The second plot shows the ratio of the number of edges to the number of points.
+Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
 The ideal behavior is demonstrated by the non-blind greedy spanner,
@@
 We remark that we obtain an exact nearest neighbor algorithm
-when setting $\eps$ to $0$, which means to 
-replacing the condition in the if-statement of the loop by
+when setting $\eps$ to $0$, which means 
+replacing the condition in the if-statement of the loop
 with $a_i\geq v$.
@@
 
-In what follows, we let $c_1,\ldots,c_n$ denote the candidates
-at the end of the $i$-th iteration of the loop, and $v_1,\ldots,v_n$
-their distances to $q$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
+In what follows, we let $c_i$ denote the candidate
+at the end of the $i$-th iteration of the loop, and $v_i$
+the distance to $\dist(c_i, q)$, $i = 1, \dots, n$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
 With the previous lemma, we can derive an upper bound for the number
@@
     \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
-    Next $p_i$ is placed in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$.}
+    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
+    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
+    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
+    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
     \label{fig:exact_bad_example}

### 2018-12-05 paper/draft.tex
interim

@@
 
-The plot \ref{fig:spanner_eps_dependence}
+The plot in Figure \ref{fig:spanner_eps_dependence}
 compares the \textsc{BlindGreedy} and \textsc{Greedy} algorithms
@@
 As expected, in high dimensions the algorithm does not perform as well.
-In figure \ref{fig:ann_dimension_dependence} we plot the average number of computed distances
+In Figure \ref{fig:ann_dimension_dependence} we plot the average number of computed distances
 for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$

### 2018-12-05 paper/draft.tex
Spelling

@@
 Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
-and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylog time, with polynomial preprocessing
+and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
 and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
@@
 by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
-operations whenever a new distance has been computed, turing the method useless in the standard computational model.
+operations whenever a new distance has been computed, turning the method useless in the standard computational model.
 
@@
 for each point pair, the set of all computed distances forms the spanner.
-The experimentally most succesful exploration strategy that we found is to
+The experimentally most successful exploration strategy that we found is to
 repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
@@
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
-\michael{Acutally, is that true? We could run BlindGreedy on the 2-approximations instead of on $[0,\infty]$,
-that might save something}
-We also compare with a spanner construction based on WSPDs. Our simple algorithms tend to give much smaller
+We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
 spanners on the tested example. Nevertheless, we leave the question open whether our construction
@@
 
-\michael{Add references here whereever appropriate}
 The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
@@
 
-On the other hand, spanner constructions based on WSPDs only compute
+On the other hand, spanner constructions based on WSPD only compute
 $O(n\log n+n \eps^{-d})$ distances 
@@
 first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
-then the algorith can only choose one of the corresponding edges, etc).
+then the algorithm can only choose one of the corresponding edges, etc).
 
@@
 
-\textsc{BlindQuasiSortedGreedy} tries to mimick the blind greedy spanner and hence appears as a natural
+\textsc{BlindQuasiSortedGreedy} tries to mimic the blind greedy spanner and hence appears as a natural
 choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
@@
 
-We point out that the proof fails for $\eps=0$ because in that case, we cannot exlude an $\eps$-ball
+We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
 of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
@@
 %%
-%Exansion constant, doubling dimension.
+%Expansion constant, doubling dimension.
 
@@
 %was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
-%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been addedhttps://v2.overleaf.com/project/5bf6fa26032afa24d94383a5
+%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
 %to the spanner. By claim 1
@@
 the space of persistence diagrams has an infinite doubling dimension.
-Nevertheless, realisitic data sets are usually not just a random sample
+Nevertheless, realistic data sets are usually not just a random sample
 in that infinite-dimensional space, but have structures
@@
 which is the sum of the  cardinalities of all pairs in a WSPD.
-The weight of a WSPD can be quadratic, but premilinary
+The weight of a WSPD can be quadratic, but preliminary
 experimental evaluation on worst-case examples do not show such a quadratic
-behavior. Therefore, we postpone the theortical analysis of the spanner construction
+behavior. Therefore, we postpone the theoretical analysis of the spanner construction
 to an extended version of this article.
@@
 
-\ignore{
-We presented experimental evidence showing that in low
-doubling dimension we can avoid many distance computations,
-if we try to get as much information as possible from the triangle inequality
-applied to the previously computed distance. The \textsc{BlindGreedy} spanner
-performs especially well, being able to produce  sparse spanners even for small values of  $\eps$.
-For the approximate nearest neighbor algorithm we proved a logarithmic upper bound,
-and, even though the constant in the proof was very large, we demonstrated
-experimentally that this does not show in practice.
-
-The next step on the theoretical side could be proving some upper bound
-for blind spanners; it is also interesting to look at other metric space
-algorithms and data structures in the setting of expensive distance computation.
-On the more applied side, the further development of this approach would
-be to find finite spaces of low dimension in practically
-important cases, like spaces of persistence diagrams or images endowed with Wasserstein (a.k.a. Earth Movers)
-metric. We know that low-dimensional subspaces of functions do not necessarily produce
-low-dimensional spaces of persistence diagrams, but there may be some conditions
-which guarantee that dimension does not increase too much.
-% An easy example: consider a fixed function $f$ to which we apply
-% a finite-dimensional family of perturbations, $f(x) + g(x; \lambda_1, \dots, \lambda_n)$;
-% for instance, each $\lambda_i$ is associated with some $x_i$ and adds a Gaussian noise
-% scaled by $\lambda$.
-% If the perturbations are very small, then they will produce some small wiggles on the function,
-% and then the optimal matcih
-}
-
 \bibliography{bib}

### 2018-12-05 paper/draft.tex
interim

@@
 \section{Experiments on spanners}
-At the moment we cannot provide any theoretical guarantees for the heuristics that we propose to compute blind spanners,
-and it can happen that it will yield the complete graph as a spanner.
 We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
@@
 
-\subsection{Experiments on approximate nearest neighbors}
+\section{Experiments on approximate nearest neighbors}
 

### 2018-12-06 paper/draft.tex
Submitted version.

@@
 \usepackage{pgfplots}
+\usepackage{pgf,tikz}
+\usetikzlibrary{arrows}
 
 % \usepackage{ccaption}
+
+\pgfplotsset{compat=1.15}
 
@@
 polynomial time in $N$, and $N$ might well be larger than $n$, the number
-of diagrams considered. For the case of Reeb graphs, the situation is even
+of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
 worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
@@
 endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).
-\michael{This paragraph needs references. As you know the literature better than me,
-I leave that part to you.}
 
@@
 
-\section{Preliminary Notions}
+\section{Background and Definitions}
 %
+\myparagraph{Doubling dimension}
 A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
@@
 \label{lem:packing_lemma}
- Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
- $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
+ Let $(\metricspace,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
+ $B(x,R)$ in $\metricspace$ such that the distance between any two distinct points of $P$ is at least $r$.
  Then 
@@
 %there exists a ball of radius $R$ such that there are at least $2^d$ points with pairwise distance at least $R$. On the other hand,
-%the maximal distance between any two of these points is at most $2R$, and we have a lot of points with approximately the same distance
+%the maximal distance between any two of these points is at most $2R$, and we ave a lot of points with approximately the same distance
 %between them. If we want a high-quality spanner on these points, then we can expect that its size will be quadratic, not linear;
@@
 
+\myparagraph{Well-separated pair decomposition}
+Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
+if 
+\[
+\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
+\]
+A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
+The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
+Har-Peled and Mendel \cite{hm-fast} 
+introduced the notion of net-trees and
+generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
+\begin{enumerate}
+    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n)$ expected time.
+    \item If $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
+are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
+    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
+The algorithm uses the net-tree structure.
+\end{enumerate}
+The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
+et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
+than net-trees. We mention in passing that cover trees can also be used for building
+a spanner (this can be proven with the same methods), and we use cover trees for
+building WSPD spanners in one of our implementations.
+
 \section{Algorithms for spanner construction}
@@
 %
-Let $(X, \dist)$ be a finite metric space with $n$ vertices. 
-One way to encode the metric space is a complete weighted graph on $X$,
+Let $(\metricspace, \dist)$ be a finite metric space with $n$ points. 
+One way to encode the metric space is a complete weighted graph on $\metricspace$,
 where the weights correspond to the distances of the points.
-A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(X,\dist)$ 
+A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(\metricspace,\dist)$ 
 if for any pair of points $(u,v)$,
@@
 Suppose that
-Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
+$\dist(p_i,p_j)=v\in\R$ has been computed.
 First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
@@
 computation. Denoting by $A_{i,j}$ the number satisfying $\dist(p_i,p_j)\leq A_{i,j}\leq 2 d(p_i,p_j)$,
-We sort all pairwise distances according to the values $A_{i,j}$.
+we sort all pairwise distances according to the values $A_{i,j}$.
 This yields a roughly sorted sequence of distance, because when $\dist(p_i,p_j)>2\dist(p_k,p_\ell)$,
@@
 this space is well-suited for controlled experiments.
+In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
 
@@
 
-Figure \ref{fig:spanner_sparseness}.
+Figure \ref{fig:spanner_sparseness}
 shows the number of edges of the spanner.
@@
 is equal to the number of distances that were actually computed, while
-for non-blind greedy spanner the latter number is always $n \choose 2$.
+for non-blind greedy spanner the latter number is always $\binom{n}{2}$.
 Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the \textsc{BlindRandom} (see \ref{fig:blind_rbr_variants}).
+the other two variants of the \textsc{BlindRandom} (see Figure \ref{fig:blind_rbr_variants}).
+
+We must also comment on WSPD spanners. In our experiments they performed poorly
+on non-clustered data, while the plots in the extensive experimental study \cite{farshi2009experimental}
+show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
+We implemented two versions of  WSPD: one for the Euclidean case,
+using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
+spaces with cover trees, using the base 1.3, which is known to give much better practical
+performance than the base 2. They both give similar results, and we can only conclude
+that the advantage of WSPD shows up on larger point sets than the ones we deal with.
+The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
+and our blind algorithms, which have at least cubic complexity in the number of points,
+are infeasible for such $n$.
 
@@
 of the conjecture: the blind greedy algorithm computes
-a spanner with $O(n)$ edges. The disappointing fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
-(\textsc{BlindQuasiSortedGreedy} is much worse, requiring all the edges). It would seem plausible that, if we have
+a spanner with $O(n)$ edges. The noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
+(\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
 access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
@@
 \label{thm:ann_bound}
-    If $(X, \dist)$ is a doubling space, then, for any fixed $\eps > 0$ the
+    If $(\metricspace, \dist)$ is a doubling space, then, for any fixed $\eps > 0$ the
     algorithm computes $O(\log n)$ distances $\dist(p_i, q)$ in expectation.
@@
 \begin{proof}
+
+\definecolor{uququq}{rgb}{0.25,0.25,0.25}
+\definecolor{xdxdff}{rgb}{0.49,0.49,1}
+\definecolor{qqqqff}{rgb}{0,0,1}
+\begin{figure}[ht!]
+    \centering
+\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
+%\draw[->,color=black] (-1.5,0) -- (17,0);
+%\foreach \x in {,2,4,6,8,10,12,14,16}
+%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
+%\draw[->,color=black] (0,-7.5) -- (0,9.2);
+%\foreach \y in {-6,-4,-2,2,4,6,8}
+%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
+%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
+\clip(-1.5,-7.5) rectangle (17,9.2);
+\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
+\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
+\draw [fill=black,fill opacity=0.95] (5.84,3.09) circle (0.21cm);
+\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
+\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
+\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
+\draw (3.51,0.63)-- (0.62,1.33);
+\draw (3.51,0.63)-- (6.12,3.39);
+\draw (8.52,0.98)-- (3.51,0.63);
+\draw (8.52,0.98)-- (14.89,-3.85);
+\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps$};
+\begin{scriptsize}
+\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
+\draw[color=qqqqff] (3.71,0.94) node {$q$};
+\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
+\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
+\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
+\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
+\fill [color=uququq] (6.48,0.83) circle (1.5pt);
+\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
+\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
+\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
+\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
+\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
+\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
+\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
+\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
+\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
+\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
+\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
+\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
+\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
+\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
+\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
+\fill [color=uququq] (5.84,3.09) circle (1.5pt);
+\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
+\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
+\end{scriptsize}
+\end{tikzpicture}
+\caption{First two steps of the ANN algorithm. First $p_1$ is 
+chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
+distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them.
+Note that the point $p_5$, which
+is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$.
+This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis.
+The small black ball between the dashed circle and the solid circle has diameter $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
+\label{fig:ann_illustration}
+\end{figure}
+
+\begin{figure}[hb!]
+    \includegraphics[width=0.7\textwidth]{pics/exact_bad_example.png}
+    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
+    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
+    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
+    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
+    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
+    \label{fig:exact_bad_example}
+\end{figure}
+
 The algorithm computes $r_i$ by assumption and updates
@@
 
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{pics/exact_bad_example.png}
-    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
-    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
-    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
-    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
-    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
-    \label{fig:exact_bad_example}
-\end{figure}
-
 Finally, we remark that a fast $2$-approximation algorithm for $\dist$ would lead to a straight-forward
@@
 
-\section{Experiments on approximate nearest neighbors}
+\subsection{Experiments on approximate nearest neighbors}
 

### 2018-12-06 cpp_version/include/quadtree.h

@@ -5,23 +5,62 @@
 #ifndef SPANNER_WASSERSTEIN_QUADTREE_H
 #define SPANNER_WASSERSTEIN_QUADTREE_H
 
+#include <cmath>
 #include <vector>
+#include <memory>
 
 namespace quadtree {
 
 using Matrix = std::vector<std::vector<double>>;
 
-struct QuadtreeNode {
-  std::vector<int> m_children;
-  int m_level;
+struct Point2D {
+    double x;
+    double y;
 };
 
+
+double dist_2d(const Point2D& p, const Point2D& q)
+{
+    return std::sqrt( (p.x - q.x) * (p.x - q.x) + (p.y - q.y) * (p.y - q.y) );
+}
+
+struct QuadtreeNode {
+
+    using Point = Point2D;
+
+    Point m_min_point;
+    Point m_max_point;
+
+    std::vector<Point>& m_all_points;
+    std::vector<int>& m_node_point_indices;
+
+    std::unique_ptr<QuadtreeNode> m_child_1 { nullptr };
+    std::unique_ptr<QuadtreeNode> m_child_2 { nullptr };
+    std::unique_ptr<QuadtreeNode> m_child_3 { nullptr };
+    std::unique_ptr<QuadtreeNode> m_child_4 { nullptr };
+
+    std::vector<int> m_children;
+    int m_level;
+
+    QuadtreeNode(std::vector<Point>& _points,
+                 const std::vector<int>& _point_indices);
+};
+
+
+QuadtreeNode::QuadtreeNode(std::vector<quadtree::QuadtreeNode::Point>& _points,
+        const std::vector<int>& _point_indices,
+        Point2D p_min,
+        Point2D p_max) :
+        m_min_point(p_min),
+        m_max_point(p_max),
+
+{
+
+}
 
 struct Quadtree {
   Quadtree(const Matrix& distance_matrix, Matrix& distance_computed_matrix);
 };
-
-
 
 }
 

### 2018-12-06 python_version/quadtree_wspd.py

@@ -52,6 +52,7 @@ class QuadtreeNode:
         self.max_x = max_x
         self.max_y = max_y
         self.points = points
+        self.level = level
 
         mid_x = 0.5 * (min_x + max_x)
         mid_y = 0.5 * (min_y + max_y)
@@ -79,7 +80,7 @@ class QuadtreeNode:
                     p11.append(p)
 
         self.children = []
-
+   
         if p00:
             child00 = QuadtreeNode(min_x, min_y, mid_x, mid_y, level + 1, p00)
             self.children.append(child00)
@@ -122,9 +123,20 @@ class QuadtreeNode:
         return min([np.linalg.norm(p - q) for p in self.corner_points() for q in other_node.corner_points])
 
 
+    def size(self):
+        return len(self.points)
+
+
+    def print(self):
+        print(f"level = {self.level},  ({self.min_x}, {self.min_y}) - ({self.max_x}, {self.max_y}), {len(self.points)}")
+        for c in self.children:
+            c.print()
+
+
 class Quadtree:
     def __init__(self, points):
         self.root = QuadtreeNode(0.0, 0.0, 1.0, 1.0, 1, points)
+
 
 
 class WspdFromQuadtree:
@@ -157,6 +169,12 @@ class WspdFromQuadtree:
     def _dist_between_sets(self, node_1, node_2):
         global global_points
         return min([euclidean_dist_no_count(global_points[p],global_points[q]) for p in node_1.points for q in node_2.points])
+
+    def print_size(self):
+        for wspd_mem in self.wspd:
+            node_1 = wspd_mem[0]
+            node_2 = wspd_mem[1]
+            print("sizes {} {}".format(node_1.size(), node_2.size()))
 
     def check(self):
         # all pairs must be covered
@@ -196,12 +214,13 @@ def make_spanner(eps):
     global dist_matrix
     global dist_calls_matrix
     qt = Quadtree(range(len(global_points)))
+    qt.root.print()
     wspd = WspdFromQuadtree(eps, qt)
     return wspd
 
 
 if __name__ == "__main__":
-    n_points = 8000
+    n_points = 100
     n_pairs = int(n_points * (n_points - 1 ) / 2)
     np.random.seed(1)
     global_points = ue.get_uniform_points(n_points, dim=2, max_coord=1.0)
@@ -210,6 +229,7 @@ if __name__ == "__main__":
[…]

### 2018-12-06 python_version/save_in_text_files.py

@@ -76,7 +76,7 @@ def save_normal_files(dim, n_points, scale, n_instance):
 if __name__ == "__main__":
 
 
-    if False:
+    if True:
         results = jl.Parallel(n_jobs=-1)(
             jl.delayed(save_unif_files)(dim, n_points, max_coord)
             for dim in [2]
@@ -87,7 +87,7 @@ if __name__ == "__main__":
             )
         sys.exit(0)
 
-    if True:
+    if False:
         results = jl.Parallel(n_jobs=-1)(
             jl.delayed(save_normal_files)(dim, n_points, scale, n_instance)
             for dim in [2, 3, 4, 5]

### 2018-12-06 paper/draft.tex
Last touch

@@
 
-\pgfplotsset{compat=1.15}
+\pgfplotsset{compat=1.14}
 
@@
 with $n$. Our motivation comes from topological summaries
-such as persistence diagrams or Reeb graphs, which are of interest
+such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
 in the field of topological data analysis. A persistence diagram
@@
 \begin{enumerate}
-    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n)$ expected time.
+    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
     \item If $\{ \{A_1, B_1 \}, 
@@
 
-\textsc{BlindQuasiSortedGreedy} tries to mimic the blind greedy spanner and hence appears as a natural
+\textsc{BlindQuasiSortedGreedy} tries to mimic the greedy spanner and hence appears as a natural
 choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
@@
 We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
-experimentally the performance of these heuristics. Clearly, this has no practical meaning for $\R^d$, but
-this space is well-suited for controlled experiments.
+experimentally the performance of these heuristics. Clearly, for this metric space, our
+cost model is not meaningful since distance comparisons are cheap;
+but we picked this environment for controlled experiments.
 In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
@@
 Figure \ref{fig:spanner_sparseness}
-shows the number of edges of the spanner.
+shows the number of edges of the spanner for various variants of 
+blind and non-blind spanner constructions.
+Note that for all blind spanner variants, the number of computed distances
+is equal to the spanner size, while for the non-blind greedy spanner,
+this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
+by the size of the spanner.
 We can see that, even though none of the blind spanners can produce
@@
 perform significantly better than both variants of \textsc{BlindQuasiSorted}.
-It is important to remember that for all blind spanners the number of edges
-is equal to the number of distances that were actually computed, while
-for non-blind greedy spanner the latter number is always $\binom{n}{2}$.
 Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
@@
 
-We must also comment on WSPD spanners. In our experiments they performed poorly
+WSPD spanners performed poorly in our experiments
 on non-clustered data, while the plots in the extensive experimental study \cite{farshi2009experimental}
@@
 using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
-spaces with cover trees, using the base 1.3, which is known to give much better practical
-performance than the base 2. They both give similar results, and we can only conclude
+spaces with cover trees (using the base $\tau=1.3$). 
+They both give similar results, and we can only conclude
 that the advantage of WSPD shows up on larger point sets than the ones we deal with.
@@
 
-We also look at higher dimensions, but the situation there is more
-disappointing. We show the results for the best algorithm, \textsc{BlindGreedy},
+We also tested higher dimensions and show the results for the best algorithm, \textsc{BlindGreedy},
 in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_only}.
-We can see that already in dimension 4 the best available algorithm
-for the uniformly distributed point sets the blind greedy variant
-produces almost a complete graph up to 700 points.
+We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
+edges for $700$ points, which clearly shows some degrading for higher dimensions. Still,
+we remark that the WSPD spanner remains worse also in the higher-dimensional setup.
 
@@
 algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
-first. The data that we observed can serve as weak evidence
-of the conjecture: the blind greedy algorithm computes
-a spanner with $O(n)$ edges. The noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
+first. 
+If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
+Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
+tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
+Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
 (\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
 access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
-find a working heuristics.
+find a working heuristic.
 

### 2019-01-24 paper/draft.tex
Correct ANN proof illustration (diameter->radius, centre).

@@
 \documentclass[a4paper,USenglish]{socg-lipics-v2018}
+%\documentclass[a4paper,USenglish]{article}
+
+%\usepackage{arxiv}
+
 \usepackage{amsthm}
@@
 \draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
-\draw [fill=black,fill opacity=0.95] (5.84,3.09) circle (0.21cm);
+\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
 \draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
@@
 This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis.
-The small black ball between the dashed circle and the solid circle has diameter $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
+The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
 \label{fig:ann_illustration}

### 2019-01-24 paper/arxiv_version.tex
Add arxiv version.

--- 2019-01-24 paper/draft.tex
@@
-\documentclass[a4paper,USenglish]{socg-lipics-v2018}
-%\documentclass[a4paper,USenglish]{article}
-
-%\usepackage{arxiv}
-
+%\documentclass[a4paper,USenglish]{socg-lipics-v2018}
+\documentclass[a4paper,USenglish]{article}
+
+% uncomment for Arxiv
+\usepackage{arxiv}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+% END uncomment for Arxiv
+
+\usepackage{booktabs}       % professional-quality tables
 \usepackage{amsthm}
@@
 
-%\newtheorem{theorem}{Theorem}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{proposition}[theorem]{Proposition}
-%\newtheorem{corollary}[theorem]{Corollary}
-%\newtheorem{definition}[theorem]{Definition}
-%\newtheorem{remark}[theorem]{Remark}
-%\newtheorem{note}[theorem]{Note}
+% uncomment for Arxiv
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem{note}[theorem]{Note}
+% END uncomment for Arxiv
 
@@
 \title{Metric Spaces with Expensive Distances}
-\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
-\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
-\authorrunning{M. Kerber and A. Nigmetov}
-
-\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-%\subjclass{\ccsdesc[500]{Theory of computation~Nearest neighbor algorithms}}
-\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
-% \category{dummy}
-
-\Copyright{Michael Kerber and Arnur Nigmetov}
+% uncomment for LIPICS 
+%\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
+%\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
+%\authorrunning{M. Kerber and A. Nigmetov}
+
+%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+%%\subjclass{\ccsdesc[500]{Theory of computation~Nearest neighbor algorithms}}
+%\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
+%% \category{dummy}
+
+%\Copyright{Michael Kerber and Arnur Nigmetov}
+% END uncomment for LIPICS
+
+% uncomment for Arxiv
+\author{Michael Kerber\\
+Graz University of Technology \\
+Institut für Geometrie \\
+Kopernikusgasse 24, 8010 Graz, Austria \\
+\And
+Arnur Nigmetov\\
+Graz University of Technology \\
+Institut für Geometrie \\
+Kopernikusgasse 24, 8010 Graz, Austria \\
+}
+%\\
+%kerber@tugraz.at\\
+%Supported by Austrian Science Fund (FWF) grant number P 29984-N35
+%\And
+
+%Arnur Nigmetov
+%Graz University of Technology\\
+%Institut fuer Geometrie\\
+%Kopernikusgasse 24\\
+%8010 Graz\\
+%Austria\\
+%nigmetov@tugraz.at}
+
+%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+
+% END uncomment for arxiv
 
@@
 \end{abstract}
+
+% uncomment for Arxiv
+\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+% END uncomment for Arxiv
 

### 2019-03-30 paper/draft_esa.tex
Interim for ESA

--- 2019-01-24 paper/draft.tex
@@
-\documentclass[a4paper,USenglish]{socg-lipics-v2018}
+\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
 %\documentclass[a4paper,USenglish]{article}
 
-%\usepackage{arxiv}
+\bibliographystyle{plainurl}% the mandatory bibstyle
 
@@
 \title{Metric Spaces with Expensive Distances}
+\titlerunning{Metric Spaces with Expensive Distances}
+% uncomment for LIPICS 
 \author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
@@
 \keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-%\subjclass{\ccsdesc[500]{Theory of computation~Nearest neighbor algorithms}}
 \subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
@@
 \Copyright{Michael Kerber and Arnur Nigmetov}
+% END uncomment for LIPICS
+
+% uncomment for Arxiv
+%\author{Michael Kerber\\
+%Graz University of Technology\\
+%Institut fuer Geometrie\\
+%Kopernikusgasse 24\\
+%8010 Graz\\
+%Austria\\
+%kerber@tugraz.at\\
+%Supported by Austrian Science Fund (FWF) grant number P 29984-N35
+%\And
+
+%Arnur Nigmetov
+%Graz University of Technology\\
+%Institut fuer Geometrie\\
+%Kopernikusgasse 24\\
+%8010 Graz\\
+%Austria\\
+%nigmetov@tugraz.at}
+
+%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+
+% END uncomment for arxiv
 
@@
 \end{abstract}
+
+% uncomment for Arxiv
+%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+% END uncomment for Arxiv
 
@@
 
+\begin{figure}[ht]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
+        legend pos=outer north east]
+
+    %\addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    %\addlegendentry{Greedy, $\eps = 0.1$}
+
+    %\addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    %\addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
+    \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
+    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_ratio_mcgill_perturbed}
+\end{figure}
+
+\begin{figure}[ht]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
+    \caption{Ratio \# edges / \# all distances for blind greedy spanner 
+    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_ratio}
+\end{figure}
+
+\begin{figure}[ht!]
+    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
+    \caption{Results of blind greedy spanner for different dimensions.}
+    \label{fig:blind_greedy_only}
+\end{figure}
+
 %\begin{figure}[ht]
@@
 
+We also ran experiments on a more realistic data set.
+
+We also generate real persistence diagrams of the shapes from  the McGill shape benchmark \cite{zhang2005mcgill}, 
+to which we add some random noise.
+
 \section{Approximate nearest neighbors}
@@
 \draw (8.52,0.98)-- (14.89,-3.85);
-\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps$};
+\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps)$};
 \begin{scriptsize}
@@
 \bibliography{bib}
-\bibliographystyle{plain}
 

### 2019-03-30 cpp_version/include/dynamic_spanner.h

@@ -8,6 +8,7 @@
 #include <algorithm>
 #include <math.h>
 #include <random>
+#include <ostream>
 
 #include <cassert>
 
@@ -17,16 +18,10 @@ namespace spd = spdlog;
 
 constexpr int INVALID_ID = -1;
 
-template<class Real_ = double>
-class DynamicSpanner {
 
-public:
 
-    using Real = Real_;
-
-protected:
-
-    struct Pair_of_points_info {
+template<class Real>
+struct Pair_of_points_info {
 
       Real distance;
       bool distance_requested;
@@ -43,12 +38,34 @@ protected:
           exact_distance_used = false;
           upper_bound = std::numeric_limits<Real>::max();
           lower_bound = 0.0;
+
+
       }
+
     };
+
+
+template<class R>
+std::ostream& operator<<(std::ostream& os, const Pair_of_points_info<R>& x)
+{
+  os << "(d = " << x.distance << ", exact_distance_used = " << x.exact_distance_used;
+  os << ", low = " << x.lower_bound << ", upp = " << x.upper_bound << ")";
+  return os;
+}
+
+template<class Real_ = double>
+class DynamicSpanner {
 
 public:
 
-    using Matrix = std::vector<std::vector<Pair_of_points_info>>;
+    using Real = Real_;
+
+//protected:
+
+
+public:
+
+    using Matrix = std::vector<std::vector<Pair_of_points_info<Real>>>;
 
     using MatrixReal = typename std::vector<std::vector<Real>>;
 
@@ -77,7 +94,7 @@ public:
             m_matrix[i].resize(m_num_points);
 
             for (size_t j = 0; j < m_num_points; j++) {
-                m_matrix[i][j] = Pair_of_points_info(distance_matrix[i][j]);
+                m_matrix[i][j] = Pair_of_points_info<Real>(distance_matrix[i][j]);
             }
         }
 
@@ -200,8 +217,8 @@ public:
             }
         }
 
-        Pair_of_points_info& info = m_matrix[i][j];
-        Pair_of_points_info& info_mirror = m_matrix[j][i];
+        Pair_of_points_info<Real>& info = m_matrix[i][j];
+        Pair_of_points_info<Real>& info_mirror = m_matrix[j][i];
 
         info.distance_requested = true;
         info_mirror.distance_requested = true;
@@ -239,13 +256,13 @@ public:
         if (dist_ij == -1.0)
[…]

### 2019-03-30 cpp_version/src/main.cpp

@@ -172,11 +172,11 @@ bool read_distance_matrix_mcgill(MatrixR& distance_matrix, double& max_distance,
 
     std::vector<size_t> indices(n_points);
     std::iota(indices.begin(), indices.end(), 0);
-    std::random_device rd;
-    std::mt19937_64 twister(rd());
-    std::shuffle(indices.begin(), indices.end(), twister);
+//    std::random_device rd;
+//    std::mt19937_64 twister(rd());
+//    std::shuffle(indices.begin(), indices.end(), twister);
 
-    indices.resize(n_samples);
+//    indices.resize(n_samples);
 
     min_distance = std::numeric_limits<double>::max();
     max_distance = -1.0;
@@ -463,52 +463,53 @@ int main(int argc, char** argv)
         exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
         console->flush();
         exp_logger->flush();
+        greedy_spanner.check_ratio(eps);
     }
 
-    {
-        DynamicSpannerR quasi_greedy_spanner(dist_matrix);
-        quasi_greedy_spanner.construct_blind_quasi_sorted_greedy_eps_spanner(eps);
-        console->info("{};{}", dist_name, quasi_greedy_spanner.get_statistics());
-        exp_logger->info("{};{}", dist_name, quasi_greedy_spanner.get_statistics());
-    }
-
-    {
-        DynamicSpannerR quasi_shaker_spanner(dist_matrix);
-        quasi_shaker_spanner.construct_blind_quasi_sorted_shaker_eps_spanner(eps);
-        exp_logger->info("{};{}", dist_name, quasi_shaker_spanner.get_statistics());
-    }
-
-    {
-        DynamicSpannerR blind_random_spanner_1_1(dist_matrix);
-        blind_random_spanner_1_1.construct_blind_random_ratio_eps_spanner(true, true, eps);
-        console->info("{};{}", dist_name, blind_random_spanner_1_1.get_statistics());
-        exp_logger->info("{};{}", dist_name, blind_random_spanner_1_1.get_statistics());
-        console->flush();
-        exp_logger->flush();
-    }
-
-    {
-        DynamicSpannerR blind_random_spanner_0_0(dist_matrix);
-        blind_random_spanner_0_0.construct_blind_random_ratio_eps_spanner(false, false, eps);
-        console->info("{};{}", dist_name, blind_random_spanner_0_0.get_statistics());
[…]

### 2019-03-31 paper/esa_draft.tex
Add McGill original table

--- 2019-01-24 paper/arxiv_version.tex
@@
 
-% uncomment for Arxiv
 \usepackage{arxiv}
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-% END uncomment for Arxiv
-
-\usepackage{booktabs}       % professional-quality tables
 \usepackage{amsthm}
@@
 
-% uncomment for Arxiv
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{remark}[theorem]{Remark}
-\newtheorem{note}[theorem]{Note}
-% END uncomment for Arxiv
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{proposition}[theorem]{Proposition}
+%\newtheorem{corollary}[theorem]{Corollary}
+%\newtheorem{definition}[theorem]{Definition}
+%\newtheorem{remark}[theorem]{Remark}
+%\newtheorem{note}[theorem]{Note}
 
@@
 \author{Michael Kerber\\
-Graz University of Technology \\
-Institut für Geometrie \\
-Kopernikusgasse 24, 8010 Graz, Austria \\
+Graz University of Technology\\
+Institut fuer Geometrie\\
+Kopernikusgasse 24\\
+8010 Graz\\
+Austria\\
+kerber@tugraz.at\\
+Supported by Austrian Science Fund (FWF) grant number P 29984-N35
 \And
-Arnur Nigmetov\\
-Graz University of Technology \\
-Institut für Geometrie \\
-Kopernikusgasse 24, 8010 Graz, Austria \\
-}
-%\\
-%kerber@tugraz.at\\
-%Supported by Austrian Science Fund (FWF) grant number P 29984-N35
-%\And
-
-%Arnur Nigmetov
-%Graz University of Technology\\
-%Institut fuer Geometrie\\
-%Kopernikusgasse 24\\
-%8010 Graz\\
-%Austria\\
-%nigmetov@tugraz.at}
+
+Arnur Nigmetov
+Graz University of Technology\\
+Institut fuer Geometrie\\
+Kopernikusgasse 24\\
+8010 Graz\\
+Austria\\
+nigmetov@tugraz.at}
 

### 2019-04-11 paper/draft_esa.tex
Minor fixes.

@@
 \documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
-%\documentclass[a4paper,USenglish]{article}
 
@@
 \usepackage{pgfplots}
+\usepackage{booktabs}
 \usepackage{pgf,tikz}
@@
 
-% \usepackage{ccaption}
-
 \pgfplotsset{compat=1.14}
 
-\newcommand{\N}{\mathbb{N}}
-\newcommand{\Q}{\mathbb{Q}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\Z}{\mathbb{Z}}
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\ZZ}{\mathbb{Z}}
 \renewcommand{\S}{\mathbb{S}}
 \newcommand{\dappr}{\tilde{d}}
-%\newcommand{\Wlog}{W.l.o.g.\ }
-%\newcommand{\wlog}{w.l.o.g.\ }
 \newcommand{\eps}{\varepsilon}
@@
 
-%\newcommand{\remark}[1]{\textbf{[#1]}}
-
-%\newtheorem{theorem}{Theorem}
-%\newdef{definition}{Definition}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{corollary}[theorem]{Corollary}
-
 \newcommand{\myparagraph}[1]{\textbf{#1.}}
-
-%\newtheorem{theorem}{Theorem}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{proposition}[theorem]{Proposition}
-%\newtheorem{corollary}[theorem]{Corollary}
-%\newtheorem{definition}[theorem]{Definition}
-%\newtheorem{remark}[theorem]{Remark}
-%\newtheorem{note}[theorem]{Note}
-
-%\numberwithin{equation}{section}
-%\numberwithin{figure}{section}
 
@@
 % END uncomment for LIPICS
-
-% uncomment for Arxiv
-%\author{Michael Kerber\\
-%Graz University of Technology\\
-%Institut fuer Geometrie\\
-%Kopernikusgasse 24\\
-%8010 Graz\\
-%Austria\\
-%kerber@tugraz.at\\
-%Supported by Austrian Science Fund (FWF) grant number P 29984-N35
-%\And
-
-%Arnur Nigmetov
-%Graz University of Technology\\
-%Institut fuer Geometrie\\
-%Kopernikusgasse 24\\
-%8010 Graz\\
-%Austria\\
-%nigmetov@tugraz.at}
-
-%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-
-% END uncomment for arxiv
 
@@
 
-% uncomment for Arxiv
-%\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-% END uncomment for Arxiv
-
 \section{Introduction}
@@
 There are good reasons for that: the most common case of a metric space
-is $\metricspace=\R^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
+is $\metricspace=\RR^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
 Even if $d$ is considered non-constant,
@@
 in the field of topological data analysis. A persistence diagram
-is a point set in $\R^2$, and the distance between two diagrams
+is a point set in $\RR^2$, and the distance between two diagrams
 is determined by a min-cost matching between the point sets.
@@
  \[|P|\leq \left(\frac{4R}{r}\right)^{d}\]
-% Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $.
-
-% Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
-% Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $. 
+
 \end{lemma}
@@
 has a constant doubling dimension.
-%\michael{I think that this sentence is enough. I would remove all what follows.}
-%If the doubling dimension of $(X, \dist)$ is larger than $d$, then
-%there exists a ball of radius $R$ such that there are at least $2^d$ points with pairwise distance at least $R$. On the other hand,
-%the maximal distance between any two of these points is at most $2R$, and we ave a lot of points with approximately the same distance
-%between them. If we want a high-quality spanner on these points, then we can expect that its size will be quadratic, not linear;
-%so the space of high doubling dimension cannot have optimal spanners. Thus the question of approximating
-%the non-doubling metric without computing many distances cannot be solved positively, and we should
-%restrict our attention to the case of doubling spaces.
 
@@
 Suppose that
-$\dist(p_i,p_j)=v\in\R$ has been computed.
+$\dist(p_i,p_j)=v\in\RR$ has been computed.
 First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
@@
 then the algorithm can only choose one of the corresponding edges, etc).
-
-%\michael{Perhaps talk about force lower bound here briefly (commented out for now}
-%The second possibility will be referred to as \textsc{BlindRandom}, where among all edges $(p_i, p_j)$
-%for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach
-%by insisting that we must first create a connected graph, or that we must first have some positive lower bound
-%for all edges, and only after these requirements are satisfied, we consider other pairs.
-
-%\michael{I have shortened this part considerably. The buckets are nice, but perhaps not needed.
-%I commented out the old version}
 
@@
 \end{centering}
-    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
     \caption{Number of edges in blind spanners generated by different variants of the blind
@@
 
-    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{Blind quasi-sorted greedy}
-
-    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{Blind quasi-sorted shaker}
-
-    %\addplot+ table [x=n_points, y=ratio_wspd, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{WSPD}
-
-    %\addplot+ table [x=n_points, y=ratio_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{Blind random}
-
     \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
@@
 \end{centering}
-    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
     \caption{Ratio \# edges / \# points for different variants of spanner 
@@
     \caption{Results of blind greedy spanner for different dimensions.}
-    \label{fig:blind_greedy_only}
+    \label{fig:blind_greedy_dimensions}
 \end{figure}
@@
 
-%\begin{figure}[ht]
-%    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
-%\end{figure}
-
 \begin{figure}[ht!]
@@
 
-    %\addplot+ table [x=Epsilon, y=edges_greedy_dim_3, col sep=comma] {eps_dep.csv};
-    %\addlegendentry{Greedy, dim = 3}
-
-    %\addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_3, col sep=comma] {eps_dep.csv};
-    %\addlegendentry{Blind greedy, dim = 3}
-
 \end{axis}
@@
 \end{centering}
-    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\R^2$ and $\R^3$.}
+    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
     \label{fig:spanner_eps_dependence}
@@
 
-    %\addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
-    %\addlegendentry{Greedy, $\eps = 0.1$}
-
-    %\addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
-    %\addlegendentry{Blind greedy, $\eps = 0.1$}
-
         \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
@@
 \end{centering}
-    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
     \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
@@
 \end{centering}
-    %\includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
     \caption{Ratio \# edges / \# all distances for blind greedy spanner 
@@
     $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_ratio}
+    \label{fig:spanner_sparseness_mcgill_perturbed}
 \end{figure}
 
-\begin{figure}[ht!]
-    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
-    \caption{Results of blind greedy spanner for different dimensions.}
-    \label{fig:blind_greedy_only}
-\end{figure}
-
-%\begin{figure}[ht]
-%    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
-%\end{figure}
+\begin{table}\centering
+    \ra{1.3}
+    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
+                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
+$\eps = 0.1$\\
+\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
+\% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
+$\eps = 0.2$\\
+\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
+\% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
+$\eps = 0.5$\\
+\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
+\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
+\bottomrule
+\end{tabular}
+\caption{Comparison of \textsc{BlindGreedy} and \textsc{Greedy} algorithms for 0-dimensional
+diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
+\label{tbl:mcgill_original_blind_greedy_spanner}
+\end{table}
 
@@
 We also tested higher dimensions and show the results for the best algorithm, \textsc{BlindGreedy},
-in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_only}.
+in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_dimensions}.
 We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
@@
 \begin{enumerate}
-    \item Uniform. Points are sampled uniformly at random from the unit cube in $\R^d$.
+    \item Uniform. Points are sampled uniformly at random from the unit cube in $\RR^d$.
     \item Normal. Points are sampled from the normal distribution.
@@
 approaching 30000.
-
-% We also looked at the dependence on the density. Note that our algorithm is not invariant under scaling: if we multiply the metric
-% by $\lambda$, we should also change $\eps$ to $\lambda \eps$, if we want
-% the same behavior. Therefore we could \textit{a priori} expect some
-% changes even if we change the parameters of the uniform distribution
-% without changing epsilon. However,
-% when we tested the algorithm on uniform data sampled from $[-a,a]^d$
-% for $a \in \Set{1, 10, 100, 1000, 10000}$ and on normal data
-% sampled from normal distribution with scales from the same set,
-% we did not observe any significant difference.
-
-%\section{Geometric concepts}
-
-%We revise the concepts that were designed for nearest neighbor queries
-%and metric approximation in metric spaces with low dimensionality.
-%We will also review the computation methods and how they depend
-%on the predicate ``$\dist(p_i,p_j)\leq t$?'' mentioned above.
-
-%\paragraph{Notion of Dimensionality.}
-%%
-%Expansion constant, doubling dimension.
-
-%\subsection{Cover trees}
-
-%\subsection{WSPD}
-%Given $s > 1$, two disjoint subsets $A, B$ of a metric space $(X, d)$ are called $s$-\textit{well-separated},
-%if 
-%\[
-%\forall a \in A \, \forall b \in B \, d(a, b) \geq s \max(\diam(A), \diam(B))
-%\]
-%A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
-%\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $X$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
-%The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd}.
-
-%\subsection{Spanners}
-
-%\subsection{Approximate Greedy Spannner}
-
-%Suppose that we have fast algorithms for computing $\lambda$-approximation of $d(x,y)$.
-%In other words, for every pair of points $x,y$ we can compute an interval $[c, \lambda c]$ that is guaranteed to contain 
-%the true distance $d(x,y)$.  Note that for every interval of this form there exists 
-%a unique $k \in \{1, \dots, N-1 \}$ such that $[c, \lambda c] \subset [\lambda^k, \lambda^{k+2}]$. 
-%Indeed, $k = \lceil \log_{\lambda}(c) \rceil$.
-%We also assume that we know the range of all possible values of $k$ and they are all non-negative,
-%$k = 1\dots N$. We call the intervals $[\lambda^k, \lambda^{k+2}]$ \textit{buckets}, even though they are overlapping.
-%Now we can formulate the algorithm for finding a spanner.
-
-%\begin{algorithmic}
-%\label{alg:local_init}
-%\Function{Approximate greedy spanner}{$rho$}
-%    \ForAll {pairs $x,y \in X$}
-%    \State {Place $x, y$ into corresponding bucket }
-%    \EndFor
-%\EndFunction
-%\end{algorithmic}
-
-%We can use essentially the same proof as in \cite{bose2010computing}.
-%Claim 1. If $s > \lambda^2 $ and $A, B$ are $s$-separated, then all pairs $(a, a')$, $(b, b')$ with $a, a' \in A$
-%and $b, b' \in B$ are processed before any of the pairs $(a, b)$.
-
-%Indeed, if $s > \lambda^2$, then for every points $a_1, a_2 \in A$ and $b \in B$
-%we have $\lambda^2 d(a_1, a_2) < d(a_1, b)$. If $\dappr(a_1, a_2)$ and $\dappr(a_1, b)$ denote the $\lambda$-approximations
-%of the corresponding distances, then the inequality $\frac{\dappr(a_1, b)}{\dappr(a_1, a_2)} > \lambda$
-%holds , hence $(a_1, a_2)$ will be placed in the bucket that precedes the bucket of $(a_1, b)$.
-
-%Claim 2. If $s > \max(\lambda^2, \frac{2t+2}{t-1})$ and $A, B$ are two $s$-well-separated subsets of $X$,
-%then in the approximate greedy spanner computed by \ref{alg:approx_greedy_spanner} there is at most one edge between $A$ and $B$.
-
-%Suppose that the pair $(a_1, b_1)$ with $a_1 \in A$, $b_1 \in B$ was processed and the edge $(a, b)$
-%was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
-%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
-%to the spanner. By claim 1
-%$d_G(a_2, a_1) \leq td(a_2, a_1)$ and $d_G(b_1, b_2) \leq t d(b_1, b_2)$.
-%By properties of WSPD, $d_G(a_1, b_1) = d(a_1, b_1) \leq (1 + \frac{2}{s})d(a_2, b_2)$
-%and $t d(a_2, a_1) \leq \frac{t}{s} d(a_2, b_2)$, $t d(b_1, b_2) \leq \frac{t}{s} d(a_2, b_2)$.
-
-%Combining these inequalities, we get
-
-%\begin{equation}
-%    \begin{split}
-%        d_G(a_2, b_2) & \leq  d_G(a_2, a_1) + d_G(a_1, b_1) + d_G(b_1, b_2)  \\
-%                      & \leq  \frac{t}{s} d(a_2, b_2) + (1 + \frac{2}{s})d(a_2, b_2) + \frac{t}{s} d(a_2, b_2) \\
-%                      & =     (1 + \frac{2t + 2}{s}) d(a_2, b_2)
-%    \end{split}
-%\end{equation}
-
-%Solving for $s$ the inequality $1 + \frac{2t + 2}{s} < t$ gives $s > \frac{2t+2}{t-1}$.
 

### 2019-04-14 paper/draft_esa.tex
interim

@@
 % uncomment for LIPICS 
-\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
-\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
-\authorrunning{M. Kerber and A. Nigmetov}
+\author{Anonymous Authors}{Anon}{}{}{}
+%\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
+%\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
+%\authorrunning{M. Kerber and A. Nigmetov}
+\authorrunning{Anonymous Authors}
 
@@
 
-\Copyright{Michael Kerber and Arnur Nigmetov}
+%\Copyright{Michael Kerber and Arnur Nigmetov}
+\Copyright{Anonymous Authors}
 % END uncomment for LIPICS
+
+\def \expDistFigPath {pics/}
+\def \expDistDataPath {./}
 
@@
 \centering
-\includegraphics[width=6cm]{intro_example.eps}
+\includegraphics[width=6cm]{\expDistFigPath  intro_example.eps}
 \caption{The compute distances are shown as edges in a graph. Note that the exact distance
@@
 
-    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind quasi-sorted greedy}
 
-    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind quasi-sorted shaker}
 
-    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{WSPD}
 
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind random}
 
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind random, lower bound first}
 
-    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind greedy}
 
-    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Greedy}
@@
 
-    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
     \addlegendentry{Blind random, lower bound first}
 
-    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
     \addlegendentry{Blind greedy}
 
-    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
     \addlegendentry{Greedy}
@@
 \begin{figure}[ht!]
-    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
     \caption{Results of blind greedy spanner for different dimensions.}
@@
 
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind random}
 
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind random, connect first}
 
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind random, lower bound first}
 
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1.txt};
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
     \addlegendentry{Blind random, connect first, lower bound first}
@@
 
-    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {eps_dep.csv};
+    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
     \addlegendentry{Greedy, dim = 2}
 
-    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {eps_dep.csv};
+    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
     \addlegendentry{Blind greedy, dim = 2}
@@
 
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
         \addlegendentry{Greedy, $\eps = 0.2$}
 
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
         \addlegendentry{Blind greedy, $\eps = 0.2$}
 
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
         \addlegendentry{Greedy, $\eps = 0.5$}
 
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
         \addlegendentry{Blind greedy, $\eps = 0.5$}
@@
 
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
     \addlegendentry{Greedy, $\eps = 0.1$}
 
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
     \addlegendentry{Blind greedy, $\eps = 0.1$}
 
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
     \addlegendentry{Greedy, $\eps = 0.2$}
 
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
     \addlegendentry{Blind greedy, $\eps = 0.2$}
 
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
     \addlegendentry{Greedy, $\eps = 0.5$}
 
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {mcgill_perturbed_blind_greedy.csv};
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
     \addlegendentry{Blind greedy, $\eps = 0.5$}
@@
 \begin{figure}[hb!]
-    \includegraphics[width=0.7\textwidth]{pics/exact_bad_example.png}
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
     \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
@@
 \begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{pics/log_dependency_constant_ratio.png}
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
     \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
@@
 \begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{pics/ann_computed_distances_dimension_dep.png}
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
     \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}

### 2019-04-14 paper/draft_esa.tex
Interim

@@
 
-\usepackage{amsthm}
-\usepackage{amsmath}
-\usepackage{mathbbol}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{algorithm}
-\usepackage{url}
-\usepackage{floatflt}
-\usepackage{mathptmx}
-\usepackage{epsfig}
-\usepackage{wrapfig}
-\usepackage{color}
-
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\usepackage{comment}
-\usepackage{braket}
-\usepackage{pgfplots}
-\usepackage{booktabs}
-\usepackage{pgf,tikz}
-\usetikzlibrary{arrows}
-
-\pgfplotsset{compat=1.14}
-
-\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
-
-\newcommand{\NN}{\mathbb{N}}
-\newcommand{\QQ}{\mathbb{Q}}
-\newcommand{\RR}{\mathbb{R}}
-\newcommand{\ZZ}{\mathbb{Z}}
-\renewcommand{\S}{\mathbb{S}}
-\newcommand{\dappr}{\tilde{d}}
-\newcommand{\eps}{\varepsilon}
-\newcommand{\orient}{O}
-\newcommand{\cpro}{\Pi}
-\newcommand{\fib}{\mathrm{Fib}}
-
-\newcommand{\metricspace}{\mathcal{M}}
-\newcommand{\approxmetric}{\gamma}
-\newcommand{\parent}{\mathrm{par}}
-\newcommand{\point}{p}
-\newcommand{\level}{\ell}
-\newcommand{\diam}{\mathrm{diam}}
-\newcommand{\pointset}{P}
-\newcommand{\distspace}{\mathcal{M}}
-\newcommand{\dist}{\delta}
-\newcommand{\adist}{\gamma}
-\newcommand{\complexity}{C_{\dist}}
-\newcommand{\doublingdimension}{\Delta}
-
-\newcommand{\ignore}[1]{}
-
-\newcommand{\myparagraph}[1]{\textbf{#1.}}
-
-\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
-\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
+\input{my_preamble.tex}
 
@@
 This information consists of lower and upper bounds for unknown distances, obtained from known distances
-by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
+by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
 operations whenever a new distance has been computed, turning the method useless in the standard computational model.
@@
 \begin{description}
-\item[\textsc{BlindRandom}] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
+\item[\brndm] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
 we pick one pair uniformly at random
-\item[\textsc{BlindGreedy}] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
+\item[\bgrdy] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
 If the maximizing pair is not unique, choose among the maximizing pairs uniformly at random.
@@
 %
-The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
+The idea behind \bgrdy is that we query an edge for which we know the least,
 in that way hoping to gather most additional information about the metric space.
-Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
+Also, our conventions imply that in \bgrdy the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
 have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
 lower bounds for every edge before it will start adding any other edges. Based on this observation,
-we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
+we also tested variations of the \brndm algorithm, where the algorithm
 first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
@@
 \begin{description}
-\item [\textsc{BlindQuasiSortedGreedy}] Traverse the pairs in increasing order with respect to $A_{i,j}$.
-\item [\textsc{BlindQuasiSortedShaker}] Alternates between pairs with small and large $A_{i,j}$
+\item [\bqsgrdy] Traverse the pairs in increasing order with respect to $A_{i,j}$.
+\item [\bqsshaker] Alternates between pairs with small and large $A_{i,j}$
 by traversing in increasing order of $A_{i,j}$ in odd iterations and in decreasing order
@@
 
-\textsc{BlindQuasiSortedGreedy} tries to mimic the greedy spanner and hence appears as a natural
+\bqsgrdy tries to mimic the greedy spanner and hence appears as a natural
 choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
@@
 Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
-useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
+useful for lower bounds. This motivates \bqsshaker
 which alternates between short and long distances.
@@
 in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
-We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.
+We refer to the former option as \bqsgrdy and to the latter one as \bqsshaker.
 }
@@
 \end{centering}
-    \caption{Comparison of the four variants of \textsc{BlindRandom} algorithm.}
+    \caption{Comparison of the four variants of \brndm algorithm.}
     \label{fig:blind_rbr_variants}
@@
 \end{tabular}
-\caption{Comparison of \textsc{BlindGreedy} and \textsc{Greedy} algorithms for 0-dimensional
+\caption{Comparison of \bgrdy and \textsc{Greedy} algorithms for 0-dimensional
 diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
@@
 
-Figure \ref{fig:spanner_sparseness}
+\cref{fig:spanner_sparseness}
 shows the number of edges of the spanner for various variants of 
@@
 spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
-\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
+\bgrdy and all variants of \brndm
 perform significantly better than both variants of \textsc{BlindQuasiSorted}.
-Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
+\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
 The ideal behavior is demonstrated by the non-blind greedy spanner,
@@
 is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
-linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
-the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
+linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
+the estimated exponents $\alpha$ for \bgrdy and standard
 greedy algorithms. Note that even for the greedy algorithm these estimated
@@
 
-As for different variants of the \textsc{BlindRandom} algorithm,
+As for different variants of the \brndm algorithm,
 we note that their performance is almost the same, and the algorithm works
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the \textsc{BlindRandom} (see Figure \ref{fig:blind_rbr_variants}).
+the other two variants of the \brndm (see \cref{fig:blind_rbr_variants}).
 
@@
 
-We also tested higher dimensions and show the results for the best algorithm, \textsc{BlindGreedy},
-in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_dimensions}.
+We also tested higher dimensions and show the results for the best algorithm, \bgrdy,
+in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
 We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
@@
 
-The plot in Figure \ref{fig:spanner_eps_dependence}
-compares the \textsc{BlindGreedy} and \textsc{Greedy} algorithms
+The plot in \cref{fig:spanner_eps_dependence}
+compares the \bgrdy and \textsc{Greedy} algorithms
 on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
@@
 Summing up, we can conclude from the experiments that 
- the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
+ the \bgrdy algorithm performs rather well, but also \brndm
 algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
@@
 Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
-(\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
+(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
 access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
@@
 
-We also ran experiments on a more realistic data set.
-
-We also generate real persistence diagrams of the shapes from  the McGill shape benchmark \cite{zhang2005mcgill}, 
-to which we add some random noise.
+We also ran experiments on two more realistic data sets,
+namely the persistence diagrams of the shapes 
+from  the McGill shape benchmark \cite{zhang2005mcgill}. 
+More precisely, we generated two sets of diagrams,
+\dtype{original} and \dtype{perturbed}. The \dtype{original}
+diagrams were computed on the points of the 450 original
+point clouds from the shape dataset; we only looked
+at dimension 0, since the diagrams in this dimension are the largest.
+We also added random noise to points of the original shapes,
+and computed persistence diagrams in dimension 0 of these modified
+shape; these diagrams form the \dtype{perturbed} dataset.
+Persistence diagrams were computed with \libraryname{DIPHA}
+library.
+
+Results of \bgrdy algorithm on the whole \dtype{original} dataset 
+for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner};
+\grdy algorithm is used as baseline. Our experiments in Euclidean setting
+demonstrate that other variants of blind algorithms are not competitive,
+so we do not present them. The first thing to note is that for higher values of $q$
+any advantage of our algorithm disappears. However, the cardinality of the \grdy
+spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
+then even the \grdy spanner is almost half of the complete graph.
+We can offer the following explanation. For higher values of $q$,
+long edges of an optimal matching play a more important role
+than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
+approaches bottleneck distance $\btdist$, which depends on the longest
+edge only. If all diagrams come from the same dataset,
+then there is a large fraction of pairs of diagrams whose optimal matchings
+have approximately the same length. This fact means that we are in the case
+of high doubling dimension, being close to a clique, and there is no hope
+of getting a small spanner. On the other hand, computation of $\btdist$ in practice
+requires significantly less time than computation of $\wsdist{q}$, and the case
+of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.
+
+We plot the sparseness of the spanners obtained by \grdy
+and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
+for different values of $\eps$, always using the $W_1$ distance.
+We can see
 
@@
 is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$.
-This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis.
+This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis.
 The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}

### 2019-04-14 paper/draft_esa.tex
Interim

@@
 of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.
-
-We plot the sparseness of the spanners obtained by \grdy
+For $wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
+computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, than we
+reduce the number of distances computed by a factor of 10.
+
+In order to see the growth of the spanner size, we use the \dtype{perturbed}
+dataset.  We plot the sparseness of the spanners obtained by \grdy
 and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
 for different values of $\eps$, always using the $W_1$ distance.
-We can see
+As expected, the ratio of computed distances decreases as the number of diagrams
+grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
+the growth of the spanner size relative to the number of diagrams. 
+For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
+linear growth) for larger values of $n$, while for $\eps = 0.2$
+this
+
+Linear
+growth of \grdy spanner can be seen distinctly for $\eps = 0.5$ (the plot
+is almost constant), while for $\eps = 0.2$ even the \grdy algorithm
+stabilizes
 

### 2019-04-15 paper/draft_esa.tex
Fix ESA draft Latex compilation

--- 2019-04-14 paper/draft_esa.tex
@@
 \keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
+\ccsdesc[500]{Theory of computation~Random search heuristics}
+\ccsdesc[500]{Theory of computation~Random search heuristics}
+%% \category{dummy}
+
 % \category{dummy}
@@
 
-\section{Introduction}
-
-Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
-in a metric space $(\metricspace,\dist)$, 
-consider the following standard operations:
-
-\begin{description}
-\item[Approximate Nearest Neighbor] Given $\eps>0$ and a point $q\in\metricspace$,
-find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
-\[\dist(q,p_i)\leq(1+\eps)\dist(q,p_j)\]
-
-\item[Spanner] Given $\eps>0$, compute a weighted graph $G$ with vertices in $\pointset$
-such that for any $u,v\in\pointset$, the shortest path distance between $u$ and $v$
-is at most $(1+\eps)\dist(u,v)$.
-\end{description}
-
-The performance of algorithms for these problems depends on the number of points,
-the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
-It is a common assumption to assume $\complexity$ to be a constant; 
-There are good reasons for that: the most common case of a metric space
-is $\metricspace=\RR^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
-Even if $d$ is considered non-constant,
-it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
-Another typical assumption is that all pairwise distances are part of the input
-in which case $\complexity$ is $O(1)$.
-
-However, we argue that in some situations, distance computations
-in $\metricspace$ can be costly and $\complexity$ might be incomparable
-with $n$. Our motivation comes from topological summaries
-such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
-in the field of topological data analysis. A persistence diagram
-is a point set in $\RR^2$, and the distance between two diagrams
-is determined by a min-cost matching between the point sets.
-If the diagrams have $N$ points, computing this matching requires
-polynomial time in $N$, and $N$ might well be larger than $n$, the number
-of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
-worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
-\cite{di2016edit}),
-not even an constant-factor approximation algorithm is known that runs
-in polynomial time in the size of the graphs.
-Another instance is a collection of high-resolution images
-endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).
-
-In such situations with expensive distance computations, 
-it makes sense to study a different cost model, where only the number of distance computations
-is taken into account. For instance, that means that quadratic time operations in terms of $n$
-are not counted towards the time complexity, as long as these operations do not query any distance
-in $\metricspace$. We also ignore the space complexity in our model.
-
-We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
-of $\metricspace$ is bounded by a constant. 
-In that situation, standard constructions from computational geometry provide partial answers:
-Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
-an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
-performing $O(\log n)$ distance computations per query point.
-Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
-and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
-and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
-However, for our relaxed cost model, we pose the question whether simpler constructions achieve
-comparable, or even fewer distance computations.
-
-We also propose a slight variant of our model: we assume that we also have access to an (efficient)
-$2$-approximation algorithm for the distance queries. Queries to this approximation algorithm
-are not counted in the model, hence we can assume that for each pair of points $(u,v)$, we
-know a number $A_{u,v}$ with $\dist(u,v)\leq A_{u,v}\leq 2\dist(u,v)$. This induces an approximate ordering
-of all distances in the metric space, and it is plausible to assume that such an ordering will simplify
-algorithmic tasks on metric spaces, at least in practice.
-
-\myparagraph{Contributions}
-%
-We propose simple algorithms for spanner construction and approximate nearest neighbor search
-and evaluate them theoretically and experimentally in the defined cost model.
-
-Our algorithms are based on the following simple idea: since distance computations are expensive
-and should be avoided, we try to obtain maximal information out of the distances that have been computed
-so far. 
-This information consists of lower and upper bounds for unknown distances, obtained from known distances
-by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
-operations whenever a new distance has been computed, turning the method useless in the standard computational model.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=6cm]{\expDistFigPath  intro_example.eps}
-\caption{The compute distances are shown as edges in a graph. Note that the exact distance
-of $p_1$ and $p_2$ is unknown. The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
-constitutes an upper bound on the distance by triangle inequality.
-However, we can also infer that $\dist(p_1,p_2)\geq 3$:
-otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
-would be shorter than the edge $(p_3,p_4)$, again contradicting
-triangle inequality.}
-\label{fig:1st_example}
-\end{figure}
-
-We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
-with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
-for each point pair, the set of all computed distances forms the spanner.
-The experimentally most successful exploration strategy that we found is to
-repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
-We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
-\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
-Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
-We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
-spanners on the tested example. Nevertheless, we leave the question open whether our construction
-yields a spanner of asymptotically linear size.
-
-For approximate nearest neighbor, we devise a simple randomized incremental algorithm and show that
-the number of distance queries to find an approximate nearest neighbor is $O(\log n)$ in expectation.
-Our proof is based on the well-known observation that the nearest neighbor changes $O(\log n)$ times
-in expectation when traversing the sequence of points, combined with a packing argument certifying that
-only a constant number of distances needs to be computed in-between two minima.
-We also experimentally evaluate our approach and observe that the approach follows 
-roughly the theoretical prediction.
-
-\section{Background and Definitions}
-%
-\myparagraph{Doubling dimension}
-A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
-if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
-and $k$ is the smallest number having that property.
-The \textit{doubling dimension} of a doubling space is defined as $\log k$
-(since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
-we always use $\log$ to denote the logarithm with base 2).
-It is easy to see that a subspace of a space with doubling dimension $d$ 
-is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).
-
-We shall need the following lemma, which is just a reformulation of the well-known
-packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).
-
-\begin{lemma}
-\label{lem:packing_lemma}
- Let $(\metricspace,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
- $B(x,R)$ in $\metricspace$ such that the distance between any two distinct points of $P$ is at least $r$.
- Then 
- \[|P|\leq \left(\frac{4R}{r}\right)^{d}\]
-
-\end{lemma}
-\begin{proof}
-We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
-balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
-we cover
-$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
-contain at most one point from $P$, 
-\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
-\end{proof}
-
-In the following, we will assume throughout that every considered metric space
-has a constant doubling dimension.
-
-\myparagraph{Well-separated pair decomposition}
-Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
-if 
-\[
-\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
-\]
-A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
-The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
-Har-Peled and Mendel \cite{hm-fast} 
-introduced the notion of net-trees and
-generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
-\begin{enumerate}
-    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
-    \item If $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
-are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
-    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
-The algorithm uses the net-tree structure.
-\end{enumerate}
-The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
-et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
-than net-trees. We mention in passing that cover trees can also be used for building
-a spanner (this can be proven with the same methods), and we use cover trees for
-building WSPD spanners in one of our implementations.
-
-\section{Algorithms for spanner construction}
-
-\myparagraph{Spanners and known constructions}
-%
-Let $(\metricspace, \dist)$ be a finite metric space with $n$ points. 
-One way to encode the metric space is a complete weighted graph on $\metricspace$,
-where the weights correspond to the distances of the points.
-A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(\metricspace,\dist)$ 
-if for any pair of points $(u,v)$,
-the shortest path distance $d_{uv}$ of $u$ and $v$ in $G$ satisfies $d_{u,v}\leq (1+\eps)\dist(u,v)$.
-In other words, the shortest path metric of $G$ is a good approximation of the actual distance for every pair of points.
-Clearly, it is a necessary condition that $G$ is connected, hence every spanner must have at least $n-1$ edges.
-
-The \emph{greedy spanner}(\cite{althofer1993sparse}) is a simple algorithm to compute linear-sized spanners:
-\begin{algorithmic}
-\label{alg:greedy_spanner}
-\Function{GreedySpanner}{$P, \eps$}
-    \State {$E\gets\emptyset$}
-    \State {Sort all pairwise distances of points in $P$}
-    \ForAll {pairs $(p_i,p_j)$ in increasing order}
-    \State {$d_{ij}\gets$ Shortest path distance in $(P,E)$}
-    \If{$d_{ij}>(1+\eps)\dist(p_i,p_j)$}
-    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
-    \EndIf
-    \EndFor
-    \Return {$(P,E)$}
-\EndFunction
-\end{algorithmic}
-
-The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
-(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
-it was also shown to return the sparsest graph. However, it clearly
-has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
-this means that in our cost model, the greedy spanner has the worst possible
-performance.
-
-On the other hand, spanner constructions based on WSPD only compute
-$O(n\log n+n \eps^{-d})$ distances 
-to construct an $(1+\eps)$-spanner in doubling dimension $d$.
-The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
-this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
-computations. However, the algorithm is significantly more involved.
-
-\myparagraph{Blind spanners}
-%
-We introduce a new framework for constructing spanners
-which we call \emph{blind spanners}: the idea is to maintain,
-for every pair of points $(p_i,p_j)$,
-a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
-initially set to $[0,\infty)$. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
-we pick one of them, compute its distance and update the lower and upper bounds of
-all pairs with respect to the newly acquired information. Here is the pseudocode:
-
-\begin{algorithmic}
-\label{alg:blind_spanner}
-\Function{BlindSpanner}{$P, \eps$}
-    \State {$E \gets \emptyset$}
-    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
-    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
-    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
-    \State {$(i,j) \gets $} \Call{GetNextEdgeToAdd()}{}
-    \State {$v \gets \dist(p_i, pj)$}
-    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
-    \State \Call{UpdateBounds}{$i, j, v$}
-    \EndWhile
-\EndFunction
-\end{algorithmic}
-
-In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
-and $\infty$ is larger than any real number,
-thus making the predicate in the while loop well-defined.
-
-We give the details of the \textsc{UpdateBounds} procedure next.
-Suppose that
-$\dist(p_i,p_j)=v\in\RR$ has been computed.
-First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
-of $p_i$ and $p_j$ is exactly $v$.
-To update the upper bound of some entry $b_{k,\ell}$,
-we observe that the shortest path from $p_k$ to $p_\ell$ might now
-go through the new edge. Hence, we update
-\[
-    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
-\]
-Repeating this for all $k,\ell$ yields the updated upper bounds.
-Note that this results in $O(n^2)$ arithmetic operations,
-but no distance computation.
-
-For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
-\[
-    v-b_{k,i}-b_{\ell,j}
-\]
-is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
-the triangle inequality
-%
-\[\dist(p_i,p_j)\leq \dist(p_i,p_k)+\dist(p_k,p_\ell)+\dist(p_\ell,p_j)\]
-by rearranging terms and plugging in the upper bounds for $\dist(p_i,p_k)$
-and $\dist(p_k,p_\ell)$. An analogue bound holds with $i$ and $j$ swapped.
-
-Moreover, the inequalities
-\begin{align*}
-  a_{j,\ell} - v - b_{k,i}&\leq \dist(p_k,p_\ell)\\
-  a_{j,k} - v - b_{j,i}&\leq \dist(p_k,p_\ell)
-\end{align*}
-hold by triangle inequality, and the same is true with $i$ and $j$ swapped.
-This yields $6$ lower bounds for $\dist(p_k,p_\ell)$, and $a_{k,\ell}$
-is updated to the maximum of these six lower bounds and its current value.
-
-\myparagraph{Heuristics}
-%
-The last missing ingredient of our algorithm is the procedure \textsc{GetNextEdgeToAdd},
-that is, how to select the next distance to be computed. We propose two natural choices
-
-\begin{description}
-\item[\textsc{BlindRandom}] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
-we pick one pair uniformly at random
-\item[\textsc{BlindGreedy}] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
-If the maximizing pair is not unique, choose among the maximizing pairs uniformly at random.
-\end{description}
-%
-The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
-in that way hoping to gather most additional information about the metric space.
-Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
-have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
-lower bounds for every edge before it will start adding any other edges. Based on this observation,
-we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
-first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
-then the algorithm can only choose one of the corresponding edges, etc).
-
-The next two heuristics assume the existence of a $2$-approximation algorithm for distance
-computation. Denoting by $A_{i,j}$ the number satisfying $\dist(p_i,p_j)\leq A_{i,j}\leq 2 d(p_i,p_j)$,
-we sort all pairwise distances according to the values $A_{i,j}$.
-This yields a roughly sorted sequence of distance, because when $\dist(p_i,p_j)>2\dist(p_k,p_\ell)$,
-then $A_{i,j}>A_{k,\ell}$ is guaranteed.
-We propose two further heuristics that attempt to make use of this sorted sequence.
-\begin{description}
-\item [\textsc{BlindQuasiSortedGreedy}] Traverse the pairs in increasing order with respect to $A_{i,j}$.
-\item [\textsc{BlindQuasiSortedShaker}] Alternates between pairs with small and large $A_{i,j}$
-by traversing in increasing order of $A_{i,j}$ in odd iterations and in decreasing order
-in even iterations.
-\end{description}
-
-\textsc{BlindQuasiSortedGreedy} tries to mimic the greedy spanner and hence appears as a natural
-choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
-results. The reason is that no pair acquires useful lower bounds when only short distance are queried
-(the greedy spanner does not have this issue because it knows the distance and hence does not need
-lower bounds).
-Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
-useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
-which alternates between short and long distances.
-
-\ignore{
-The last option that we consider is based on an additional assumption: while
-computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
-cheap approximation algorithm
-that gives a $2$-approximation of $\dist$ much faster 
-(the output of this approximation algorithm is guaranteed to
-lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
-that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
-We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
-these intervals are not disjoint, but by calling the approximation algorithm we can assign
-each pair $(p_i, p_j)$ to one of the buckets.
-Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
-then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
-that fully contains this interval, thus we need to find $k$ such that
-$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
-and we take $k = \lfloor \log v \rfloor - 1$.
-Afer we found a bucket for each pair $(p_i, p_j)$, 
-we have some sort of a weak ordering: we cannot say anything
-about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
-are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
-traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
-we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
-we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
-order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
-in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
-We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.
-}
-
-\section{Experiments on spanners}
-We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
-experimentally the performance of these heuristics. Clearly, for this metric space, our
-cost model is not meaningful since distance comparisons are cheap;
-but we picked this environment for controlled experiments.
-In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
-
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind quasi-sorted greedy}
-
-    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind quasi-sorted shaker}
-
-    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{WSPD}
-
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind greedy}
-
-    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Greedy}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Number of edges in blind spanners generated by different variants of the blind
-    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for normally distributed points
-    in dimension 2, $\eps = 0.1$.}
-    \label{fig:spanner_sparseness}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-    \addlegendentry{Blind greedy}
-
-    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-    \addlegendentry{Greedy}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges / \# points for different variants of spanner 
-    algorithms. The plot is for normally distributed points
-    in dimension 2, $\eps = 0.1$.}
-    \label{fig:spanner_ratio}
-\end{figure}
-
-\begin{figure}[ht!]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
-    \caption{Results of blind greedy spanner for different dimensions.}
-    \label{fig:blind_greedy_dimensions}
-\end{figure}
-
-\begin{figure}[ht!]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first, lower bound first}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Comparison of the four variants of \textsc{BlindRandom} algorithm.}
-    \label{fig:blind_rbr_variants}
-\end{figure}
-
-\begin{figure}[ht!]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-    \addlegendentry{Greedy, dim = 2}
-
-    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-    \addlegendentry{Blind greedy, dim = 2}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
-    \label{fig:spanner_eps_dependence}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
-        legend pos=outer north east]
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.5$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
-    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_ratio_mcgill_perturbed}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.5$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges / \# all distances for blind greedy spanner 
-    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_sparseness_mcgill_perturbed}
-\end{figure}
-
-\begin{table}\centering
-    \ra{1.3}
-    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
-                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
-                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
-                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
-$\eps = 0.1$\\
-\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
-\% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
-$\eps = 0.2$\\
-\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
-\% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
-$\eps = 0.5$\\
-\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
-\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
-\bottomrule
-\end{tabular}
-\caption{Comparison of \textsc{BlindGreedy} and \textsc{Greedy} algorithms for 0-dimensional
-diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
-\label{tbl:mcgill_original_blind_greedy_spanner}
-\end{table}
-
-We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
-\begin{enumerate}
-    \item In the \textbf{uniform} test set points are sampled uniformly at
-        random from the unit cube in $\mathbb{R}^d$.
-    \item In the \textbf{normal} test set points are sampled from the standard
-        normal distribution in $\mathbb{R}^d$.
-    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
-        at random from $[0,10000]^d$, and then we add normally distributed noise around
-        each of the centers. The number of clusters is chosen so that each cluster
-        contains 50 points.
-    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
-        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
-\end{enumerate}
-In all experiments the algorithms that we tested compared in the same way,
-so we only present results for the \textbf{uniform} point set in dimension 2.
-
-Figure \ref{fig:spanner_sparseness}
-shows the number of edges of the spanner for various variants of 
-blind and non-blind spanner constructions.
-Note that for all blind spanner variants, the number of computed distances
-is equal to the spanner size, while for the non-blind greedy spanner,
-this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
-by the size of the spanner.
-We can see that, even though none of the blind spanners can produce
-spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
-\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
-perform significantly better than both variants of \textsc{BlindQuasiSorted}.
-Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
-The ideal behavior is demonstrated by the non-blind greedy spanner,
-for which this ratio stays practically constant, confirming the linear growth.
-None of the blind algorithms seems to have this property, but among them
-the blind greedy spanner is the best one. If we assume that the number of edges
-is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
-linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
-the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
-greedy algorithms. Note that even for the greedy algorithm these estimated
-exponents can be significantly larger than 1,
-which is explained by the fact that the number of points
-on which we computed spanners is not large enough to clearly see
-the linear dependence.
-
-\begin{table}[]
-\begin{tabular}{|l|l|l|}
-\hline
-dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
-2         &         1.08                &  1.12                 \\ \hline
-3         &         1.24                &  1.41                \\ \hline
-4         &         1.42                &  1.77                 \\ \hline
-\end{tabular}
-\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
-on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
-\label{tbl:regr_coeff_spanner}
-\end{table}
-
-As for different variants of the \textsc{BlindRandom} algorithm,
-we note that their performance is almost the same, and the algorithm works
-significantly better than \textsc{QuasiSorted} variants, but obviously
-worse than the blind greedy variant. There is a consistent, though small,
-difference between the variants that do not force lower bounds first and
-the other two variants of the \textsc{BlindRandom} (see Figure \ref{fig:blind_rbr_variants}).
-
-WSPD spanners performed poorly in our experiments
-on non-clustered data, while the plots in the extensive experimental study \cite{farshi2009experimental}
-show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
-We implemented two versions of  WSPD: one for the Euclidean case,
-using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
-spaces with cover trees (using the base $\tau=1.3$). 
-They both give similar results, and we can only conclude
-that the advantage of WSPD shows up on larger point sets than the ones we deal with.
-The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
-and our blind algorithms, which have at least cubic complexity in the number of points,
-are infeasible for such $n$.
-
-We also tested higher dimensions and show the results for the best algorithm, \textsc{BlindGreedy},
-in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_dimensions}.
-We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
-edges for $700$ points, which clearly shows some degrading for higher dimensions. Still,
-we remark that the WSPD spanner remains worse also in the higher-dimensional setup.
-
-The plot in Figure \ref{fig:spanner_eps_dependence}
-compares the \textsc{BlindGreedy} and \textsc{Greedy} algorithms
-on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
-is approximately the same for both algorithms. Since it is not cleary seen from the picture, we also note 
-that the ratio of the number of edges decreases for
-smaller values of $\eps$: 
-for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
-while for $\eps = 1/32$ the ratio is 2.6
-
-Summing up, we can conclude from the experiments that 
- the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
-algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
-first. 
-If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
-Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
-tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
-Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
-(\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
-access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
-find a working heuristic.
-
-We also ran experiments on a more realistic data set.
-
-We also generate real persistence diagrams of the shapes from  the McGill shape benchmark \cite{zhang2005mcgill}, 
-to which we add some random noise.
-
-\section{Approximate nearest neighbors}
-\label{sec:ann}
-We consider the standard problem of finding an approximate nearest neighbor: given
-$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
-find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
-section, and we shall also use the shorthand notation
-\[
-    r_i := \dist(p_i, q).
-\]
-We assume for simplicity
-that all exact pairwise distances $\dist(p_i, p_j)$ are already computed
-(a slight modification of the algorithm can also be applied if only a spanner is available).
-Our goal is to reduce the number of computed distances $\dist(p_i, q)$.
-
-Our approach can be summarized as follows. 
-Fix a random permutation of the points of $P$ 
-and consider the points in that order 
-(to simplify notation,
-we re-index them, so the order is again $p_1, \dots, p_n$).
-During the loop, we maintain lower bounds of each $p_i$
-to the query point $q$, which are initially all set to $0$.
-We also remember the closest neighbor $c$ that we have seen so
-far and its distance $v$ to $q$.
-We refer to the point $c$ as the \emph{candidate}.
-We maintain the invariant that $c$ is an approximate nearest neighbor
-to $q$ for the points $\{p_1,\ldots,p_i\}$.
-When reaching the point $p_i$, we check whether the lower
-bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
-If so, $c$ remains an approximate nearest neighbor and
-we do not query the distance of $p_i$ to $q$.
-Otherwise, we compute $\dist(p_i,q)$ and update the lower bounds
-of all points according to the newly computed distance.
-If $p_i$ is closer to $q$ than $c$, we update $c$ and $v$
-accordingly. At the end of the loop, $c$ is an approximate nearest
-neighbor. The pseudocode of the procedure follows.
-
-\begin{algorithmic}
-\label{alg:ann_blind}
-
-\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
-    \State {$[p_1, \dots, p_n] \gets \mbox{random permutation of }P$}
-    \State { $a_i \gets 0$ for $i=1,\ldots,n$}
-    \Comment {$a_i$ is lower bound for $\dist(p_i, q)$}
-    \State {$ c \gets p_1, \quad v \gets \dist(p_1, q)$}
-    \Comment {$c$ keeps the current candidate}
-    \State \Call {UpdateBounds}{$p_1, v$}
-    \For{$i = 2\dots n$}
-        \If {$a_i \geq \frac{v}{1+\eps}$}
-            \State {\textbf{continue}}
-        \Else
-            \State {Compute $r_i = \dist(p_i, q)$}
-            \State \Call{UpdateBounds}{$p_i, r_i$}
-            \If {$r_i < v$}
-                \State {$c \gets p_i,\quad v\gets r_i$}
-            \EndIf
-        \EndIf
-    \EndFor
-    \State \Return {$c, v$}
-\EndFunction
-\end{algorithmic}
-
-We remark that we obtain an exact nearest neighbor algorithm
-when setting $\eps$ to $0$, which means 
-replacing the condition in the if-statement of the loop
-with $a_i\geq v$.
-
-The procedure to maintain the lower bounds $a_i$
-is very simple and follows directly
-from triangle inequality.
-
-\begin{algorithmic}
-\Procedure{UpdateBounds}{$p_i, r_i$}
-    \For{$k = i + 1, \dots, n$}
-        \State {$a_k \gets \max(a_k, |\dist(p_i, p_k) - r_i|)$}
-    \EndFor
-\EndProcedure
-\end{algorithmic}
-
-\begin{theorem}
-\label{thm:ann_bound}
-    If $(\metricspace, \dist)$ is a doubling space, then, for any fixed $\eps > 0$ the
-    algorithm computes $O(\log n)$ distances $\dist(p_i, q)$ in expectation.
-\end{theorem}
-
-Towards the proof, we will use the following geometric lemma which
-can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
-further distance computations of points very close to $p_i$ or very far from $p_i$
-will be avoided.
-
-\begin{lemma}
-\label{lem:bound_lemma}
-Assume $r_i=\dist(p_i,q)$ is computed in the algorithm, and let $j>i$.
-\begin{enumerate}
-\item If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps}) r_i$, the algorithm will not compute the distance
-of $p_j$ to $q$.
-\item If $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$, the algorithm will not compute
-the distance of $p_j$ to $q$.
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-
-\definecolor{uququq}{rgb}{0.25,0.25,0.25}
-\definecolor{xdxdff}{rgb}{0.49,0.49,1}
-\definecolor{qqqqff}{rgb}{0,0,1}
-\begin{figure}[ht!]
-    \centering
-\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.5cm,y=0.5cm]
-%\draw[->,color=black] (-1.5,0) -- (17,0);
-%\foreach \x in {,2,4,6,8,10,12,14,16}
-%\draw[shift={(\x,0)},color=black] (0pt,2pt) -- (0pt,-2pt) node[below] {\footnotesize $\x$};
-%\draw[->,color=black] (0,-7.5) -- (0,9.2);
-%\foreach \y in {-6,-4,-2,2,4,6,8}
-%\draw[shift={(0,\y)},color=black] (2pt,0pt) -- (-2pt,0pt) node[left] {\footnotesize $\y$};
-%\draw[color=black] (0pt,-10pt) node[right] {\footnotesize $0$};
-\clip(-1.5,-7.5) rectangle (17,9.2);
-\draw [fill=black,fill opacity=0.5] (8.52,0.98) circle (1.02cm);
-\draw [fill=black,fill opacity=0.15] (8.52,0.98) circle (4cm);
-\draw [fill=black,fill opacity=0.95] (6.12,3.39) circle (0.42cm);
-\draw [fill=black,fill opacity=0.5] (7.65,-3.14) circle (1.32cm);
-\draw [line width=2pt] (3.51,0.63) circle (1.9cm);
-\draw [dash pattern=on 2pt off 2pt] (3.51,0.63) circle (1.49cm);
-\draw (3.51,0.63)-- (0.62,1.33);
-\draw (3.51,0.63)-- (6.12,3.39);
-\draw (8.52,0.98)-- (3.51,0.63);
-\draw (8.52,0.98)-- (14.89,-3.85);
-\draw (11.88,-1) node[anchor=north west] {$r_2 +r_1 / ( 1 + \eps)$};
-\begin{scriptsize}
-\fill [color=qqqqff] (3.51,0.63) circle (1.5pt);
-\draw[color=qqqqff] (3.71,0.94) node {$q$};
-\fill [color=xdxdff] (0.62,1.33) circle (1.5pt);
-\draw[color=black] (3.24,1.41) node {$r_1/(1+\eps)$};
-\fill [color=qqqqff] (8.52,0.98) circle (1.5pt);
-\draw[color=qqqqff] (8.9,1.29) node {$p_2$};
-\fill [color=uququq] (6.48,0.83) circle (1.5pt);
-\fill [color=qqqqff] (7.83,2.21) circle (1.5pt);
-\draw[color=qqqqff] (8.21,2.52) node {$p_3$};
-\fill [color=qqqqff] (7.65,-3.14) circle (1.5pt);
-\draw[color=qqqqff] (8.01,-2.82) node {$p_4$};
-\fill [color=qqqqff] (6.62,0.6) circle (1.5pt);
-\draw[color=qqqqff] (6.98,0.92) node {$p_5$};
-\fill [color=qqqqff] (1.49,-2.39) circle (1.5pt);
-\draw[color=qqqqff] (1.87,-2.08) node {$p_6$};
-\fill [color=qqqqff] (1.95,-2.65) circle (1.5pt);
-\draw[color=qqqqff] (2.31,-2.33) node {$p_7$};
-\fill [color=qqqqff] (2.3,-2.75) circle (1.5pt);
-\draw[color=qqqqff] (2.68,-2.43) node {$p_8$};
-\fill [color=qqqqff] (2.79,-2.9) circle (1.5pt);
-\draw[color=qqqqff] (3.17,-2.57) node {$p_9$};
-\fill [color=xdxdff] (14.89,-3.85) circle (1.5pt);
-\fill [color=uququq] (5.84,3.09) circle (1.5pt);
-\fill [color=xdxdff] (4.52,4.29) circle (1.5pt);
-\draw[color=xdxdff] (4.89,4.61) node {$p_1$};
-\end{scriptsize}
-\end{tikzpicture}
-\caption{First two steps of the ANN algorithm. First $p_1$ is 
-chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
-distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them.
-Note that the point $p_5$, which
-is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$.
-This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis.
-The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
-\label{fig:ann_illustration}
-\end{figure}
-
-\begin{figure}[hb!]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
-    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
-    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
-    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
-    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
-    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
-    \label{fig:exact_bad_example}
-\end{figure}
-
-The algorithm computes $r_i$ by assumption and updates
-all lower bounds. For $p_j$, it sets $a_j\gets \max (a_j,|\dist(p_i,p_j)-r_i|)$.
-If $\dist(p_i,p_j)\geq (1+\frac{1}{1+\eps})r_i$, it follows that 
-\[a_j\geq (1+\frac{1}{1+\eps})r_i - r_i = \frac{r_i}{1+\eps}.\]
-Likewise, if $\dist(p_i,p_j)\leq\frac{\eps}{1+\eps}r_i$,
-\[a_j\geq r_i-\dist(p_i,p_j) \geq r_i - \frac{\eps}{1+\eps}r_i = \frac{r_i}{1+\eps}.\]
-In both cases, after the point $p_i$ is handled,
-$v\leq r_i$ clearly holds. Since $v$ is only decreasing
-and $a_j$ is only increasing in the algorithm,
-it follows that $a_j\geq \frac{v}{1+\eps}$ when $p_j$ is handled,
-so the algorithm proceeds without a distance computation.
-\end{proof}
-
-In what follows, we let $c_i$ denote the candidate
-at the end of the $i$-th iteration of the loop, and $v_i$
-the distance to $\dist(c_i, q)$, $i = 1, \dots, n$. Clearly, $v_1,\ldots,v_n$ is a decreasing sequence.
-With the previous lemma, we can derive an upper bound for the number
-of distance computations in an arbitrary subsequence of $p_1,\ldots,p_n$
-as follows.
-
-\begin{lemma}
-\label{lem:sequence_lemma}
-Among the points $p_k,\ldots,p_\ell$ with $1\leq k< \ell\leq n$, the algorithm computes at most
-\[\left(\frac{4(2+\eps) v_k}{\eps v_\ell}\right)^{d}\]
-distances to $q$.
-\end{lemma}
-\begin{proof}
-By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
-whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
-around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
-is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
-Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
-have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
-\end{proof}
-
-A consequence of the lemma is that as long as a candidate $c$ is fixed in the algorithm,
-the number of computed distances is a constant (since $v_k=v_\ell$). 
-This means that to prove Theorem~\ref{thm:ann_bound}, it would suffice to show
-that the candidate changes only a logarithmic number of times in expectation.
-While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
-that argument.
-
-\begin{proof} (of Theorem~\ref{thm:ann_bound})
-%
-In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
-such that $r_i<r_k$ for all $1\leq i\leq k-1$. We call an element
-of this form a \emph{minimum} of the sequence. 
-A standard backwards analysis argument~\cite{seidel-backwards} shows that the probability
-of $p_k$ being a minimum is at most $1/k$, so that the number of minima
-in the sequence is $O(\log n)$ in expectation.
-
-Note that for $\eps>0$,
-a minimum $p_k$ is not necessarily the candidate $c_k$ because a previous point
-in the sequence close to $p_k$ might have caused the lower bound $a_k$ to be
-in the interval $[\frac{v_k}{1+\eps},v_k]$, which leads to not
-computing the distance $r_k$. However, it is true that
-$v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
-nearest neighbor of $\{p_1,\ldots,p_k\}$.
-
-Now, let $p_k$, $p_\ell$ be two consecutive minima in the sequence
-(we also allow that $\ell=n+1$ if $k$ is the last minimum in the sequence).
-Note that $v_{\ell-1}\geq r_k$ because each $v_j$ is equal to $r_i$ for
-some $i\leq j$, and in the sequence $r_1,\ldots,r_{\ell-1}$, $r_k$
-is minimal by construction. Using Lemma~\ref{lem:sequence_lemma},
-the number of distance computations among the points
-$p_k,\ldots,p_{\ell-1}$ is at most
-\[
-\left(\frac{4(2+\eps) v_k}{\eps v_{\ell-1}}\right)^{d}\leq \left(\frac{4(2+\eps)(1+\eps)r_k}{\eps r_k}\right)^{d}=\left(\frac{4(2+\eps)(1+\eps)}{\eps}\right)^{d},\]
-which is a constant depending only of $\eps$ and $d$, irrespective of the length of the sequence.
-Since $p_1,\ldots,p_n$ decomposes into $O(\log n)$ such sequences in expectation, the result follows.
-\end{proof}
-
-We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
-of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
-Indeed, as the example in Figure~\ref{fig:exact_bad_example} shows, there are point sets where the expected number
-of distance computations for exact nearest neighbor is linear.
-
-Finally, we remark that a fast $2$-approximation algorithm for $\dist$ would lead to a straight-forward
-optimization: compute a $2$-approximation of $\dist(p_i,q)$ for all $1\leq i\leq n$
-and let $m$ denote the minimal approximate distance encountered. Then, we can discard all points
-whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.
-
-\section{Experiments on approximate nearest neighbors}
-
-In order to experimentally evaluate the performance of our algorithm,
-we generate random point sets and random query points, and for each query point
-run the algorithm 10 times. The average number of distances to the query point
-that were actually computed is the measure that we are interested in.
-We average the results over 10 different instances of the point
-set and query point in order to see the trend clearer; thus
-each point on the plots in this section is the result of averaging of 100 runs
-of the code (10 instances, 10 random permutations per instance).
-
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
-    \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
-    points.}
-    \label{fig:ann_const_ratio}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
-    \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
-    \label{fig:ann_dimension_dependence}
-\end{figure}
-
-We used the following methods of generating random points:
-\begin{enumerate}
-    \item Uniform. Points are sampled uniformly at random from the unit cube in $\RR^d$.
-    \item Normal. Points are sampled from the normal distribution.
-\end{enumerate}
-Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
-and from the normal distribution centered at the origin with scale 100,
-thus we get query points that are "inside" the point set and also "outside".
-We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
-the maximal number of points is $30,000$.
-
-In order to empirically verify the upper bound $O(\log n)$,
-we plot the number of computed distances divided
-by the logarithm of the number of points in figure \ref{fig:ann_const_ratio} (for $d = 2$).
-We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
-the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
-really computes only a very small number of distances to the query point.
-As expected, in high dimensions the algorithm does not perform as well.
-In Figure \ref{fig:ann_dimension_dependence} we plot the average number of computed distances
-for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
-the sublinearity of the growth becomes clear only when the number of points is relatively large,
-approaching 30000.
-
-\section{Conclusion and future work}
-\label{sec:conclusion}
-%
-We have introduced a new cost model for the analysis of algorithms
-for metric spaces that fits the situation that computing an individual distance
-is more costly than other types of primitive operations.
-Our theoretical and experimental results are under the usual assumption
-that the metric space has a low doubling dimension.
-However, in our motivating example of collections of persistence diagrams
-or Reeb graphs, this assumption does not hold. For instance,
-the space of persistence diagrams has an infinite doubling dimension.
-Nevertheless, realistic data sets are usually not just a random sample
-in that infinite-dimensional space, but have structures
-(e.g. clusters of close-by diagrams) which should be favorable for our approach
-We plan to consider the quality of our algorithms for persistence diagrams
-as future work.
-
-On the theoretical side, the obvious next question is whether our strategy
-for blind spanners yields a linear spanner in expectation. 
-Our experiments are not conclusive enough in this respect to make this conjecture yet.
-However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
-that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
-which is the sum of the  cardinalities of all pairs in a WSPD.
-The weight of a WSPD can be quadratic, but preliminary
-experimental evaluation on worst-case examples do not show such a quadratic
-behavior. Therefore, we postpone the theoretical analysis of the spanner construction
-to an extended version of this article.
-
-The existence of a $2$-approximation algorithm did not help us to significantly
-reduce the number of exact distance computations, although it seems obvious
-that knowing the all approximate distances is useful.
-We pose the question what heuristic could make more use of this feature.
+\input{expensive-distances-main.tex}
 

### 2019-04-15 paper/expensive-distances-main.tex
Fix ESA draft Latex compilation

--- 2019-01-24 paper/draft.tex
@@
-\documentclass[a4paper,USenglish]{socg-lipics-v2018}
-%\documentclass[a4paper,USenglish]{article}
-
-%\usepackage{arxiv}
-
-\usepackage{amsthm}
-\usepackage{amsmath}
-\usepackage{mathbbol}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{algorithm}
-\usepackage{url}
-\usepackage{floatflt}
-\usepackage{mathptmx}
-\usepackage{epsfig}
-\usepackage{wrapfig}
-\usepackage{color}
-
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\usepackage{comment}
-\usepackage{braket}
-\usepackage{pgfplots}
-\usepackage{pgf,tikz}
-\usetikzlibrary{arrows}
-
-% \usepackage{ccaption}
-
-\pgfplotsset{compat=1.14}
-
-\newcommand{\N}{\mathbb{N}}
-\newcommand{\Q}{\mathbb{Q}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\Z}{\mathbb{Z}}
-\renewcommand{\S}{\mathbb{S}}
-\newcommand{\dappr}{\tilde{d}}
-%\newcommand{\Wlog}{W.l.o.g.\ }
-%\newcommand{\wlog}{w.l.o.g.\ }
-\newcommand{\eps}{\varepsilon}
-\newcommand{\orient}{O}
-\newcommand{\cpro}{\Pi}
-\newcommand{\fib}{\mathrm{Fib}}
-
-\newcommand{\metricspace}{\mathcal{M}}
-\newcommand{\approxmetric}{\gamma}
-\newcommand{\parent}{\mathrm{par}}
-\newcommand{\point}{p}
-\newcommand{\level}{\ell}
-\newcommand{\diam}{\mathrm{diam}}
-\newcommand{\pointset}{P}
-\newcommand{\distspace}{\mathcal{M}}
-\newcommand{\dist}{\delta}
-\newcommand{\adist}{\gamma}
-\newcommand{\complexity}{C_{\dist}}
-\newcommand{\doublingdimension}{\Delta}
-
-\newcommand{\ignore}[1]{}
-
-%\newcommand{\remark}[1]{\textbf{[#1]}}
-
-%\newtheorem{theorem}{Theorem}
-%\newdef{definition}{Definition}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{corollary}[theorem]{Corollary}
-
-\newcommand{\myparagraph}[1]{\textbf{#1.}}
-
-%\newtheorem{theorem}{Theorem}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{proposition}[theorem]{Proposition}
-%\newtheorem{corollary}[theorem]{Corollary}
-%\newtheorem{definition}[theorem]{Definition}
-%\newtheorem{remark}[theorem]{Remark}
-%\newtheorem{note}[theorem]{Note}
-
-%\numberwithin{equation}{section}
-%\numberwithin{figure}{section}
-
-\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
-\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
-
-\title{Metric Spaces with Expensive Distances}
-\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
-\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
-\authorrunning{M. Kerber and A. Nigmetov}
-
-\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-%\subjclass{\ccsdesc[500]{Theory of computation~Nearest neighbor algorithms}}
-\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
-% \category{dummy}
-
-\Copyright{Michael Kerber and Arnur Nigmetov}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-In algorithms for finite metric spaces, it is common
-to assume that the distance between two points
-can be computed in constant time, and complexity
-bounds are expressed only in terms of the number
-of points of the metric space.
-We introduce a different model where we assume
-that the computation of a single distance is
-an expensive operation and consequently, the goal is
-to minimize the number of such distance queries.
-This model is motivated by metric spaces
-that appear in the context of topological data analysis.
-
-We consider two standard operations on metric spaces,
-namely the construction of a $1+\eps$-spanner
-and the computation of an approximate nearest
-neighbor for a given query point.
-In both cases, we partially explore the metric space
-through distance queries and infer lower and upper bounds
-for yet unexplored distances through triangle inequality.
-For spanners, we evaluate several exploration strategies
-through extensive experimental evaluation.
-For approximate nearest neighbors, we prove that our
-strategy returns an approximate nearest neighbor
-after a logarithmic number of distance queries.
-\end{abstract}
-
 \section{Introduction}
@@
 %
-The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
+The idea behind \bgrdy is that we query an edge for which we know the least,
 in that way hoping to gather most additional information about the metric space.
-Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
+Also, our conventions imply that in \bgrdy the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
 have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
 lower bounds for every edge before it will start adding any other edges. Based on this observation,
-we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
+we also tested variations of the \brndm algorithm, where the algorithm
 first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
 then the algorithm can only choose one of the corresponding edges, etc).
-
-%\michael{Perhaps talk about force lower bound here briefly (commented out for now}
-%The second possibility will be referred to as \textsc{BlindRandom}, where among all edges $(p_i, p_j)$
-%for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach
-%by insisting that we must first create a connected graph, or that we must first have some positive lower bound
-%for all edges, and only after these requirements are satisfied, we consider other pairs.
-
-%\michael{I have shortened this part considerably. The buckets are nice, but perhaps not needed.
-%I commented out the old version}
 
@@
 
-    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{Blind quasi-sorted greedy}
-
-    %\addplot+ table [x=n_points, y=ratio_quasi_sorted_shaker, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{Blind quasi-sorted shaker}
-
-    %\addplot+ table [x=n_points, y=ratio_wspd, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{WSPD}
-
-    %\addplot+ table [x=n_points, y=ratio_blind_random, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
-    %\addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
     \addlegendentry{Blind random, lower bound first}
 
-    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
     \addlegendentry{Blind greedy}
 
-    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {all_methods_dim_2_eps_0.1_ratio.csv};
+    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
     \addlegendentry{Greedy}
@@
 \begin{figure}[ht!]
-    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
     \caption{Results of blind greedy spanner for different dimensions.}
-    \label{fig:blind_greedy_only}
+    \label{fig:blind_greedy_dimensions}
 \end{figure}
@@
 \end{centering}
-    \caption{Comparison of the four variants of \textsc{BlindRandom} algorithm.}
+    \caption{Comparison of the four variants of \brndm algorithm.}
     \label{fig:blind_rbr_variants}
 \end{figure}
-
-%\begin{figure}[ht]
-%    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
-%\end{figure}
 
@@
 
-    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {eps_dep.csv};
+    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
     \addlegendentry{Greedy, dim = 2}
 
-    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {eps_dep.csv};
+    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
     \addlegendentry{Blind greedy, dim = 2}
-
-    %\addplot+ table [x=Epsilon, y=edges_greedy_dim_3, col sep=comma] {eps_dep.csv};
-    %\addlegendentry{Greedy, dim = 3}
-
-    %\addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_3, col sep=comma] {eps_dep.csv};
-    %\addlegendentry{Blind greedy, dim = 3}
 
@@
 
-%\begin{figure}[ht]
-%    \includegraphics[width=0.7\textwidth]{pics/epsilon_dependency_400_pts_normal_500_dim_2.png}
-%\end{figure}
+\begin{figure}[ht]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
+        legend pos=outer north east]
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.5$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
+    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_ratio_mcgill_perturbed}
+\end{figure}
+
+\begin{figure}[ht]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Ratio \# edges / \# all distances for blind greedy spanner 
+    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_sparseness_mcgill_perturbed}
+\end{figure}
+
+\begin{table}\centering
+    \ra{1.3}
+    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
+                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
+$\eps = 0.1$\\
+\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
+\% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
+$\eps = 0.2$\\
+\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
+\% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
+$\eps = 0.5$\\
+\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
+\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
+\bottomrule
+\end{tabular}
+\caption{Comparison of \bgrdy and \textsc{Greedy} algorithms for 0-dimensional
+diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
+\label{tbl:mcgill_original_blind_greedy_spanner}
+\end{table}
 
@@
 
-We also tested higher dimensions and show the results for the best algorithm, \textsc{BlindGreedy},
-in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_only}.
+We also tested higher dimensions and show the results for the best algorithm, \bgrdy,
+in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
 We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
@@
 Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
-(\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
+(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
 access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
 find a working heuristic.
+
+We also ran experiments on two more realistic data sets,
+namely the persistence diagrams of the shapes 
+from  the McGill shape benchmark \cite{zhang2005mcgill}. 
+More precisely, we generated two sets of diagrams,
+\dtype{original} and \dtype{perturbed}. The \dtype{original}
+diagrams were computed on the points of the 450 original
+point clouds from the shape dataset; we only looked
+at dimension 0, since the diagrams in this dimension are the largest.
+We also added random noise to points of the original shapes,
+and computed persistence diagrams in dimension 0 of these modified
+shape; these diagrams form the \dtype{perturbed} dataset.
+Persistence diagrams were computed with \libraryname{DIPHA}
+library.
+
+Results of \bgrdy algorithm on the whole \dtype{original} dataset 
+for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner};
+\grdy algorithm is used as baseline. Our experiments in Euclidean setting
+demonstrate that other variants of blind algorithms are not competitive,
+so we do not present them. The first thing to note is that for higher values of $q$
+any advantage of our algorithm disappears. However, the cardinality of the \grdy
+spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
+then even the \grdy spanner is almost half of the complete graph.
+We can offer the following explanation. For higher values of $q$,
+long edges of an optimal matching play a more important role
+than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
+approaches bottleneck distance $\btdist$, which depends on the longest
+edge only. If all diagrams come from the same dataset,
+then there is a large fraction of pairs of diagrams whose optimal matchings
+have approximately the same length. This fact means that we are in the case
+of high doubling dimension, being close to a clique, and there is no hope
+of getting a small spanner. On the other hand, computation of $\btdist$ in practice
+requires significantly less time than computation of $\wsdist{q}$, and the case
+of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.
+For $wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
+computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, than we
+reduce the number of distances computed by a factor of 10.
+
+In order to see the growth of the spanner size, we use the \dtype{perturbed}
+dataset.  We plot the sparseness of the spanners obtained by \grdy
+and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
+for different values of $\eps$, always using the $W_1$ distance.
+As expected, the ratio of computed distances decreases as the number of diagrams
+grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
+the growth of the spanner size relative to the number of diagrams. 
+For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
+linear growth) for larger values of $n$, while for $\eps = 0.2$
+this
+
+Linear
+growth of \grdy spanner can be seen distinctly for $\eps = 0.5$ (the plot
+is almost constant), while for $\eps = 0.2$ even the \grdy algorithm
+stabilizes
 
@@
 We pose the question what heuristic could make more use of this feature.
-
-\bibliography{bib}
-\bibliographystyle{plain}
-
-\end{document}

### 2019-04-15 paper/expensive-distances-main.tex
Moved proof to appendix, still over limit.

@@
 
+\ignore{
 \begin{figure}[ht!]
@@
 \end{figure}
+}
 
@@
 \end{centering}
-    \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
-    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
     $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
@@
 \end{centering}
-    \caption{Ratio \# edges / \# all distances for blind greedy spanner 
-    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
     $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
@@
 \begin{table}\centering
-    \ra{1.3}
+    \ra{1.2}
     \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
@@
 $\eps = 0.1$\\
-\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
+%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
 \% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
 $\eps = 0.2$\\
-\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
+%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
 \% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
 $\eps = 0.5$\\
-\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
+%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
 \% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the \brndm (see \cref{fig:blind_rbr_variants}).
+the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
 
@@
 
-Summing up, we can conclude from the experiments that 
- the \bgrdy algorithm performs rather well, but also \brndm
+\ignore{
+Summing up, the \bgrdy algorithm is the best algorithm for our model, but also \brndm
 algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
@@
 find a working heuristic.
-
-We also ran experiments on two more realistic data sets,
+}
+
+From these experiments we conclude that the algorithm with the highest chance
+of saving a significant amount of distance computations on real data, like large
+persistence diagrams, is \bgrdy. Even though \brdnm also substantially reduces
+the number of computed distances, its performance is significantly worse;
+other algorithms that we tested do not produce sparse spanners.
+While it would seem plausible that access to approximate value of the distance 
+could be exploited in the spanner construction, 
+we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
+dense spanners.
+
+Finally, we ran experiments on two realistic data sets,
 namely the persistence diagrams of the shapes
@@
 and computed persistence diagrams in dimension 0 of these modified
-shape; these diagrams form the \dtype{perturbed} dataset.
+shapes; these diagrams form the \dtype{perturbed} dataset.
 Persistence diagrams were computed with \libraryname{DIPHA}
@@
 for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner};
-\grdy algorithm is used as baseline. Our experiments in Euclidean setting
-demonstrate that other variants of blind algorithms are not competitive,
-so we do not present them. The first thing to note is that for higher values of $q$
+\grdy algorithm is used as baseline\footnote{Of course, in the columns of \grdy algorithm
+we give its sparseness, i.e. number of edges divided by the number of all distances;
+the percent of computed distances is always 100.}.  The first thing to note is that for higher values of $q$
 any advantage of our algorithm disappears. However, the cardinality of the \grdy
@@
 then even the \grdy spanner is almost half of the complete graph.
-We can offer the following explanation. For higher values of $q$,
-long edges of an optimal matching play a more important role
+We can offer the following explanation, assuming the reader is familiar
+with the definition of $\wsdist{q}$. For higher values of $q$,
+longer edges of an optimal matching play a more important role
 than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
@@
 edge only. If all diagrams come from the same dataset,
-then there is a large fraction of pairs of diagrams whose optimal matchings
-have approximately the same length. This fact means that we are in the case
+then there is a large number of pairs of diagrams whose optimal matchings
+have the longest edges of approximately the same length. This fact means that we are in the case
 of high doubling dimension, being close to a clique, and there is no hope
-of getting a small spanner. On the other hand, computation of $\btdist$ in practice
+of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
 requires significantly less time than computation of $\wsdist{q}$, and the case
-of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.
-For $wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
+of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
+For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
 computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, than we
@@
 and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
-for different values of $\eps$, always using the $W_1$ distance.
+for different values of $\eps$, always using the $\wsdist{1}$ distance.
 As expected, the ratio of computed distances decreases as the number of diagrams
@@
 For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
-linear growth) for larger values of $n$, while for $\eps = 0.2$
-this
-
-Linear
-growth of \grdy spanner can be seen distinctly for $\eps = 0.5$ (the plot
-is almost constant), while for $\eps = 0.2$ even the \grdy algorithm
-stabilizes
+linear growth) for larger values of $n$.
 
@@
 \end{theorem}
+
+The proof is given in the appendix.
+
+\section{Experiments on approximate nearest neighbors}
+
+In order to experimentally evaluate the performance of our algorithm,
+we generate random point sets and random query points, and for each query point
+run the algorithm 10 times. The average number of distances to the query point
+that were actually computed is the measure that we are interested in.
+We average the results over 10 different instances of the point
+set and query point in order to see the trend clearer; thus
+each point on the plots in this section is the result of averaging of 100 runs
+of the code (10 instances, 10 random permutations per instance).
+
+\begin{figure}[ht]
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
+    \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
+    points.}
+    \label{fig:ann_const_ratio}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=0.7\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
+    \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
+    \label{fig:ann_dimension_dependence}
+\end{figure}
+
+We used the following methods of generating random points:
+\begin{enumerate}
+    \item Uniform. Points are sampled uniformly at random from the unit cube in $\RR^d$.
+    \item Normal. Points are sampled from the normal distribution.
+\end{enumerate}
+Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
+and from the normal distribution centered at the origin with scale 100,
+thus we get query points that are "inside" the point set and also "outside".
+We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
+the maximal number of points is $30,000$.
+
+In order to empirically verify the upper bound $O(\log n)$,
+we plot the number of computed distances divided
+by the logarithm of the number of points in \cref{fig:ann_const_ratio} (for $d = 2$).
+We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
+the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
+really computes only a very small number of distances to the query point.
+As expected, in high dimensions the algorithm does not perform as well.
+In \cref{fig:ann_dimension_dependence} we plot the average number of computed distances
+for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
+the sublinearity of the growth becomes clear only when the number of points is relatively large,
+approaching 30000.
+
+\section{Conclusion and future work}
+\label{sec:conclusion}
+%
+We have introduced a new cost model for the analysis of algorithms
+for metric spaces that fits the situation that computing an individual distance
+is more costly than other types of primitive operations.
+Our theoretical and experimental results are under the usual assumption
+that the metric space has a low doubling dimension.
+However, in our motivating example of collections of persistence diagrams
+or Reeb graphs, this assumption does not hold. For instance,
+the space of persistence diagrams has an infinite doubling dimension.
+Nevertheless, realistic data sets are usually not just a random sample
+in that infinite-dimensional space, but have structures
+(e.g. clusters of close-by diagrams) which should be favorable for our approach.
+
+On the theoretical side, the obvious next question is whether our strategy
+for blind spanners yields a linear spanner in expectation. 
+Our experiments are not conclusive enough in this respect to make this conjecture yet.
+However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
+that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
+which is the sum of the  cardinalities of all pairs in a WSPD.
+The weight of a WSPD can be quadratic, but preliminary
+experimental evaluation on worst-case examples do not show such a quadratic
+behavior. Therefore, we postpone the theoretical analysis of the spanner construction
+to an extended version of this article.
+
+The existence of a $2$-approximation algorithm did not help us to significantly
+reduce the number of exact distance computations, although it seems obvious
+that knowing the all approximate distances is useful.
+We pose the question what heuristic could make more use of this feature.
+
+\appendix
+\section{Proof of \cref{thm:ann_bound}}
 
@@
 \begin{proof}
-By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
+By the first part of \cref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
 whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
 around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
-is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
+is computed, the second part of \cref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
 Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
-have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
+have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
 \end{proof}
@@
 the number of computed distances is a constant (since $v_k=v_\ell$). 
-This means that to prove Theorem~\ref{thm:ann_bound}, it would suffice to show
+This means that to prove \cref{thm:ann_bound}, it would suffice to show
 that the candidate changes only a logarithmic number of times in expectation.
@@
 
-\begin{proof} (of Theorem~\ref{thm:ann_bound})
+\begin{proof} (of \cref{thm:ann_bound})
 %
@@
 some $i\leq j$, and in the sequence $r_1,\ldots,r_{\ell-1}$, $r_k$
-is minimal by construction. Using Lemma~\ref{lem:sequence_lemma},
+is minimal by construction. Using \cref{lem:sequence_lemma},
 the number of distance computations among the points
@@
 We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
-of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
-Indeed, as the example in Figure~\ref{fig:exact_bad_example} shows, there are point sets where the expected number
+of close-by points as in the second part of \cref{lem:bound_lemma}, and the packing argument fails.
+Indeed, as the example in \cref{fig:exact_bad_example} shows, there are point sets where the expected number
 of distance computations for exact nearest neighbor is linear.
@@
 whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.
-
-\section{Experiments on approximate nearest neighbors}
-
-In order to experimentally evaluate the performance of our algorithm,
-we generate random point sets and random query points, and for each query point
-run the algorithm 10 times. The average number of distances to the query point
-that were actually computed is the measure that we are interested in.
-We average the results over 10 different instances of the point
-set and query point in order to see the trend clearer; thus
-each point on the plots in this section is the result of averaging of 100 runs
-of the code (10 instances, 10 random permutations per instance).
-
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
-    \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
-    points.}
-    \label{fig:ann_const_ratio}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
-    \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
-    \label{fig:ann_dimension_dependence}
-\end{figure}
-
-We used the following methods of generating random points:
-\begin{enumerate}
-    \item Uniform. Points are sampled uniformly at random from the unit cube in $\RR^d$.
-    \item Normal. Points are sampled from the normal distribution.
-\end{enumerate}
-Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
-and from the normal distribution centered at the origin with scale 100,
-thus we get query points that are "inside" the point set and also "outside".
-We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
-the maximal number of points is $30,000$.
-
-In order to empirically verify the upper bound $O(\log n)$,
-we plot the number of computed distances divided
-by the logarithm of the number of points in figure \ref{fig:ann_const_ratio} (for $d = 2$).
-We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
-the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
-really computes only a very small number of distances to the query point.
-As expected, in high dimensions the algorithm does not perform as well.
-In Figure \ref{fig:ann_dimension_dependence} we plot the average number of computed distances
-for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
-the sublinearity of the growth becomes clear only when the number of points is relatively large,
-approaching 30000.
-
-\section{Conclusion and future work}
-\label{sec:conclusion}
-%
-We have introduced a new cost model for the analysis of algorithms
-for metric spaces that fits the situation that computing an individual distance
-is more costly than other types of primitive operations.
-Our theoretical and experimental results are under the usual assumption
-that the metric space has a low doubling dimension.
-However, in our motivating example of collections of persistence diagrams
-or Reeb graphs, this assumption does not hold. For instance,
-the space of persistence diagrams has an infinite doubling dimension.
-Nevertheless, realistic data sets are usually not just a random sample
-in that infinite-dimensional space, but have structures
-(e.g. clusters of close-by diagrams) which should be favorable for our approach
-We plan to consider the quality of our algorithms for persistence diagrams
-as future work.
-
-On the theoretical side, the obvious next question is whether our strategy
-for blind spanners yields a linear spanner in expectation. 
-Our experiments are not conclusive enough in this respect to make this conjecture yet.
-However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
-that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
-which is the sum of the  cardinalities of all pairs in a WSPD.
-The weight of a WSPD can be quadratic, but preliminary
-experimental evaluation on worst-case examples do not show such a quadratic
-behavior. Therefore, we postpone the theoretical analysis of the spanner construction
-to an extended version of this article.
-
-The existence of a $2$-approximation algorithm did not help us to significantly
-reduce the number of exact distance computations, although it seems obvious
-that knowing the all approximate distances is useful.
-We pose the question what heuristic could make more use of this feature.

### 2019-04-16 python_version/mcgill/print_mcgill.py
Add data from server to repo

@@ -1,0 +1,107 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+
+import numpy as np
+import dionysus as d
+import joblib as jl
+
+wasser_dist_calls = 0
+wasser_dist_dict = {}
+q = 1
+
+id_to_name = {}
+
+class MyDiagram:
+    def __init__(self, dgm_data, id, label=None):
+        self.dgm_data = dgm_data
+        self.id = id
+        self.label = label
+
+
+def wasser_dist(dgm1, dgm2, q):
+    global wasser_dist_calls
+    global wasser_dist_dict
+    if dgm1.id == dgm2.id:
+        return 0.0
+    key1 = (dgm1.id, dgm2.id)
+    key2 = (dgm2.id, dgm1.id)
+    if key1 in wasser_dist_dict:
+        return wasser_dist_dict[key1]
+    else:
+        wasser_dist_calls += 1
+        # print("wasser_dist called, %d, %d, calling Hera" % (dgm1.id, dgm2.id))
+        result = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q,
+                                        delta=0.0001)
+        # print("Hera returned")
+        wasser_dist_dict[key1] = result
+        wasser_dist_dict[key2] = result
+        return result
+
+
+def read_mcgill_diagrams(diag_dir_root, dim):
+    global id_to_name
+    diag_dirs = [os.path.join(diag_dir_root, f) for f in os.listdir(diag_dir_root)
+                 if f.endswith('Ply')]
+    dim_ext = '.%d.txt' % dim
+    dgms = []
+    dgm_id = 0
+    for diag_dir in diag_dirs:
+        diag_files = [os.path.join(diag_dir, f) for f in os.listdir(diag_dir)
+                      if f.endswith(dim_ext)]
+        for dgm_f in diag_files:
+            pts = []
+            with open(dgm_f, 'r') as dgm_file:
+                for line in dgm_file:
+                    x, y = [float(x) for x in line.split()]
+                    pts.append((x, y))
+            dgms.append(MyDiagram(d.Diagram(pts[:]), dgm_id, diag_dir))
+            id_to_name[dgm_id] = dgm_f
+            dgm_id += 1
+            print("Read file %s, number of points %d" % (dgm_f, len(pts)))
[…]

### 2019-04-16 python_version/mcgill/test_mcgill.py

@@ -1,0 +1,103 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+
+import numpy as np
+import dionysus as d
+import joblib as jl
+
+wasser_dist_calls = 0
+wasser_dist_dict = {}
+q = 1
+
+
+class MyDiagram:
+    def __init__(self, dgm_data, id, label=None):
+        self.dgm_data = dgm_data
+        self.id = id
+        self.label = label
+
+
+def wasser_dist(dgm1, dgm2, q):
+    global wasser_dist_calls
+    global wasser_dist_dict
+    if dgm1.id == dgm2.id:
+        return 0.0
+    key1 = (dgm1.id, dgm2.id)
+    key2 = (dgm2.id, dgm1.id)
+    if key1 in wasser_dist_dict:
+        return wasser_dist_dict[key1]
+    else:
+        wasser_dist_calls += 1
+        # print("wasser_dist called, %d, %d, calling Hera" % (dgm1.id, dgm2.id))
+        result = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q,
+                                        delta=0.01)
+        # print("Hera returned")
+        wasser_dist_dict[key1] = result
+        wasser_dist_dict[key2] = result
+        return result
+
+
+def read_mcgill_diagrams(diag_dir_root, dim):
+    diag_dirs = [os.path.join(diag_dir_root, f) for f in os.listdir(diag_dir_root)
+                 if f.endswith('Ply')]
+    dim_ext = '.%d.txt' % dim
+    dgms = []
+    dgm_id = 0
+    for diag_dir in diag_dirs:
+        diag_files = [os.path.join(diag_dir, f) for f in os.listdir(diag_dir)
+                      if f.endswith(dim_ext)]
+        for dgm_f in diag_files:
+            pts = []
+            with open(dgm_f, 'r') as dgm_file:
+                for line in dgm_file:
+                    x, y = [float(x) for x in line.split()]
+                    pts.append((x, y))
+            dgms.append(MyDiagram(d.Diagram(pts[:]), dgm_id, dgm_f))
+            dgm_id += 1
+            print("Read file %s, number of points %d" % (dgm_f, len(pts)))
+    print("Read %d diagrams" % len(dgms))
+    return dgms
+
+
+def get_distance_tuple(i, j, dgm1, dgm2, q):
[…]

### 2019-04-16 python_version/mcgill/test_mcgill_no_parallel.py

@@ -1,0 +1,100 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+
+import numpy as np
+import dionysus as d
+
+wasser_dist_calls = 0
+wasser_dist_dict = {}
+q = 1
+
+
+class MyDiagram:
+    def __init__(self, dgm_data, id, label=None):
+        self.dgm_data = dgm_data
+        self.id = id
+        self.label = label
+
+
+def wasser_dist(dgm1, dgm2, q):
+    global wasser_dist_calls
+    global wasser_dist_dict
+    if dgm1.id == dgm2.id:
+        return 0.0
+    key1 = (dgm1.id, dgm2.id)
+    key2 = (dgm2.id, dgm1.id)
+    if key1 in wasser_dist_dict:
+        return wasser_dist_dict[key1]
+    else:
+        wasser_dist_calls += 1
+        # print("wasser_dist called, %d, %d, calling Hera" % (dgm1.id, dgm2.id))
+        result = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q,
+                                        delta=0.01)
+        # print("Hera returned")
+        wasser_dist_dict[key1] = result
+        wasser_dist_dict[key2] = result
+        return result
+
+
+def read_mcgill_diagrams(diag_dir_root, dim):
+    diag_dirs = [os.path.join(diag_dir_root, f) for f in os.listdir(diag_dir_root)
+                 if f.endswith('Ply')]
+    dim_ext = '.%d.txt' % dim
+    dgms = []
+    dgm_id = 0
+    for diag_dir in diag_dirs:
+        diag_files = [os.path.join(diag_dir, f) for f in os.listdir(diag_dir)
+                      if f.endswith(dim_ext)]
+        for dgm_f in diag_files:
+            pts = []
+            with open(dgm_f, 'r') as dgm_file:
+                for line in dgm_file:
+                    x, y = [float(x) for x in line.split()]
+                    pts.append((x, y))
+            dgms.append(MyDiagram(d.Diagram(pts[:]), dgm_id, dgm_f))
+            dgm_id += 1
+            print("Read file %s, number of points %d" % (dgm_f, len(pts)))
+    print("Read %d diagrams" % len(dgms))
+    return dgms
+
+
+def get_distance_tuple(i, j, dgm1, dgm2, q):
[…]

### 2019-04-16 python_version/mcgill/test_mcgill_time.py

@@ -1,0 +1,81 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+
+import numpy as np
+import dionysus as d
+import datetime as dt
+
+
+class MyDiagram:
+    def __init__(self, dgm_data, id, label=None):
+        self.dgm_data = dgm_data
+        self.id = id
+        self.label = label
+
+
+
+def read_mcgill_diagrams(diag_dir_root, dim):
+    diag_dirs = [os.path.join(diag_dir_root, f) for f in os.listdir(diag_dir_root)
+                 if f.endswith('Ply')]
+    dim_ext = '.%d.txt' % dim
+    dgms = []
+    dgm_id = 0
+    for diag_dir in diag_dirs:
+        diag_files = [os.path.join(diag_dir, f) for f in os.listdir(diag_dir)
+                      if f.endswith(dim_ext)]
+        for dgm_f in diag_files:
+            pts = []
+            with open(dgm_f, 'r') as dgm_file:
+                for line in dgm_file:
+                    x, y = [float(x) for x in line.split()]
+                    pts.append((x, y))
+            dgms.append(MyDiagram(d.Diagram(pts[:]), dgm_id, dgm_f))
+            dgm_id += 1
+            # print("Read file %s, number of points %d" % (dgm_f, len(pts)))
+    print("Read %d diagrams" % len(dgms))
+    return dgms
+
+
+def get_distance_tuple(i, j, dgm1, dgm2, q, delta):
+    # print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, started computation" % (i, j, q, dgm1.label, dgm2.label))
+    start_time = dt.datetime.now()
+    res = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q, delta=delta)
+    end_time = dt.datetime.now()
+    elapsed_time = (end_time - start_time).microseconds
+    # print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, d = %f, finished computation" % (i, j, q, dgm1.label, dgm2.label, res))
+    return (i, j, res, elapsed_time)
+    
+
+if __name__ == "__main__":
+    if len(sys.argv) < 2:
+        print("dim\n")
+        exit(0)
+    dim = int(sys.argv[1])
+    delta = 0.01
+    diag_dir_root = '/home/anigmetov/Downloads/mcgill_shape_benchmark_dgms_no_duplicates/'
[…]

### 2019-04-16 paper/expensive-distances-main.tex
interim

@@
 
-\begin{figure}[ht]
-    \begin{centering}
+\begin{figure}
+\captionsetup[subfigure]{font=footnotesize}
+\centering
+\subcaptionbox{Ratio \# edges / \# points for different variants of spanner 
+            algorithms. The plot is for normally distributed points
+            in dimension 2, $\eps = 0.1$.}[.5\textwidth]{%
 \begin{tikzpicture}
     \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
-        legend pos=outer north east]
+        legend pos=north east]
 
@@
     \end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges / \# points for different variants of spanner 
-    algorithms. The plot is for normally distributed points
-    in dimension 2, $\eps = 0.1$.}
             \label{fig:spanner_ratio}
+\end{tikzpicture}}%
+\subcaptionbox{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}[.5\textwidth]{
+    \begin{tikzpicture}
+            \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
+                legend pos=north east]
+
+            \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+            \addlegendentry{Greedy, dim = 2}
+
+            \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+            \addlegendentry{Blind greedy, dim = 2}
+
+        \end{axis}
+        \label{fig:spanner_eps_dependence}
+    \end{tikzpicture}}
 \end{figure}
 
-\begin{figure}[ht!]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
-    \caption{Results of blind greedy spanner for different dimensions.}
-    \label{fig:blind_greedy_dimensions}
-\end{figure}
+%\begin{figure}[ht]
+%    \begin{minipage}{0.45\textwidth}
+%        \begin{tikzpicture}
+%            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
+%                legend pos=north east]
+
+%                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+%                \addlegendentry{Blind random, lower bound first}
+
+%                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+%                \addlegendentry{Blind greedy}
+
+%                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+%                \addlegendentry{Greedy}
+
+%            \end{axis}
+%        \end{tikzpicture}
+%            \caption{Ratio \# edges / \# points for different variants of spanner 
+%            algorithms. The plot is for normally distributed points
+%            in dimension 2, $\eps = 0.1$.}
+%            \label{fig:spanner_ratio}
+%    \end{minipage}
+%    \begin{minipage}{0.45\textwidth}
+%        \begin{tikzpicture}
+%            \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
+%                legend pos=north east]
+
+%            \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+%            \addlegendentry{Greedy, dim = 2}
+
+%            \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+%            \addlegendentry{Blind greedy, dim = 2}
+
+%        \end{axis}
+%        \end{tikzpicture}
+%        \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
+%        \label{fig:spanner_eps_dependence}
+%   \end{minipage}
+%\end{figure}
 
@@
     \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-    \addlegendentry{Greedy, dim = 2}
-
-    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-    \addlegendentry{Blind greedy, dim = 2}
-
-\end{axis}
-\end{tikzpicture}
+    \includegraphics[width=0.8\textwidth]{\expDistFigPath blind-greedy-dims.png}
+    \caption{Results of \bgrdy spanner for different dimensions.}
+    \label{fig:blind_greedy_dimensions}
     \end{centering}
-    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
-    \label{fig:spanner_eps_dependence}
 \end{figure}
@@
 
-\begin{table}[]
+\begin{table}[b]
 \begin{tabular}{|l|l|l|}
@@
 of saving a significant amount of distance computations on real data, like large
-persistence diagrams, is \bgrdy. Even though \brdnm also substantially reduces
+persistence diagrams, is \bgrdy. Even though \brndm also substantially reduces
 the number of computed distances, its performance is significantly worse;
@@
 \begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
+    \begin{minipage}{0.45\textwidth}
+        \includegraphics[width=0.9\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
         \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
@@
         \label{fig:ann_const_ratio}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
+    \end{minipage}
+    \begin{minipage}{0.45\textwidth}
+        \includegraphics[width=0.9\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
         \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
         \label{fig:ann_dimension_dependence}
+    \end{minipage}
 \end{figure}

### 2019-04-16 python_version/nearest_neighbor_mcgill.py

@@ -78,7 +78,7 @@ class AnnFinder:
                 self.discard_2(alternative_cand_idx, cand_idx)
                 self.discard_1(cand_idx)
 
-        assert(self.check_nearest_neighbor(self.points[cand_idx], epsilon))
+        # assert(self.check_nearest_neighbor(self.points[cand_idx], epsilon))
 
         return self.points[cand_idx]
 
@@ -200,19 +200,15 @@ def run_experiment_ann(n_points, fname, epsilon, n_attempts):
 
 if __name__ == "__main__":
     np.random.seed(1)
-    n_pointses = [ i for i in range(100, 4000, 200)]
+    n_pointses = [ 420 ]
     dims = [0, 1, 2]
-    qs = [1, 2, 3, 4]
-    fnames = ["dist_matrix_q_{}_dim_{}.txt".format(q, dim) for q in qs for dim in dims]
+    qs = [1, 2, 3]
+
+    fnames = ["mcgill/dist_matrix_mcgill_original_no_parallel_q_{}_dim_{}.txt".format(q, dim) for q in qs for dim in dims]
     epsilons = [0.0001, 0.0005, 0.001, 0.005, 0.01, 0.1, 0.5]
     n_attempts = 20
-    # n_pointses = [ 100 ]
-    # dims = [0]
-    # qs = [1]
-    # epsilons = [0.1]
-    # n_attempts = 2
 
-    results_ann = jl.Parallel(n_jobs=8)(
+    results_ann = jl.Parallel(n_jobs=1)(
         jl.delayed(run_experiment_ann)(n_points, fname, eps, n_attempts)
         for n_points in n_pointses
         for fname in fnames

### 2019-04-16 python_version/mcgill/test_mcgill_save_time.py
save individual time

@@ -1,0 +1,81 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+
+import numpy as np
+import dionysus as d
+import datetime as dt
+
+
+class MyDiagram:
+    def __init__(self, dgm_data, id, label=None):
+        self.dgm_data = dgm_data
+        self.id = id
+        self.label = label
+
+
+
+def read_mcgill_diagrams(diag_dir_root, dim):
+    diag_dirs = [os.path.join(diag_dir_root, f) for f in os.listdir(diag_dir_root)
+                 if f.endswith('Ply')]
+    dim_ext = '.%d.txt' % dim
+    dgms = []
+    dgm_id = 0
+    for diag_dir in diag_dirs:
+        diag_files = [os.path.join(diag_dir, f) for f in os.listdir(diag_dir)
+                      if f.endswith(dim_ext)]
+        for dgm_f in diag_files:
+            pts = []
+            with open(dgm_f, 'r') as dgm_file:
+                for line in dgm_file:
+                    x, y = [float(x) for x in line.split()]
+                    pts.append((x, y))
+            dgms.append(MyDiagram(d.Diagram(pts[:]), dgm_id, dgm_f))
+            dgm_id += 1
+            # print("Read file %s, number of points %d" % (dgm_f, len(pts)))
+    print("Read %d diagrams" % len(dgms))
+    return dgms
+
+
+def get_distance_tuple(i, j, dgm1, dgm2, q, delta):
+    print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, started computation" % (i, j, q, dgm1.label, dgm2.label))
+    start_time = dt.datetime.now()
+    res = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q, delta=delta)
+    end_time = dt.datetime.now()
+    elapsed_time = (end_time - start_time).microseconds
+    print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, d = %f, finished computation" % (i, j, q, dgm1.label, dgm2.label, res))
+    return (i, j, elapsed_time, elapsed_time)
+    
+
+if __name__ == "__main__":
+    if len(sys.argv) < 2:
+        print("dim\n")
+        exit(0)
+    dim = int(sys.argv[1])
+    delta = 0.01
[…]

### 2019-04-16 cpp_version/include/dynamic_spanner.h
interim

@@ -749,9 +749,9 @@ public:
         std::uniform_real_distribution<double> dis(1.0, approximation_factor);
 
 
-#pragma omp parallel
+//#pragma omp parallel
         {
-#pragma omp for schedule(dynamic)
+//#pragma omp for schedule(dynamic)
             for (int i = 0; i < m_num_points; ++i) {
                 for (int j = i + 1; j < m_num_points; ++j) {
                     auto& info_ij = m_matrix[i][j];

### 2019-04-16 cpp_version/src/main.cpp

@@ -8,6 +8,7 @@
 #include <iostream>
 #include <algorithm>
 #include <random>
+#include <chrono>
 
 #include "cover_tree/cover_tree.h"
 #include "cover_tree/wspd.h"
@@ -458,7 +459,12 @@ int main(int argc, char** argv)
 
     {
         DynamicSpannerR greedy_spanner(dist_matrix);
+        auto begin_greedy = std::chrono::steady_clock::now();
         greedy_spanner.construct_greedy_eps_spanner(eps);
+        auto end_greedy = std::chrono::steady_clock::now();
+        auto dur_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_greedy - begin_greedy).count();
+        console->info("greedy_time = {} microseconds", dur_greedy);
+        exp_logger->info("greedy_time = {} microseconds", dur_greedy);
         console->info("{};{}", dist_name, greedy_spanner.get_statistics());
         exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
         console->flush();
@@ -513,12 +519,17 @@ int main(int argc, char** argv)
 
     {
         DynamicSpannerR blind_greedy_spanner(dist_matrix);
+        auto begin_blind_greedy = std::chrono::steady_clock::now();
         blind_greedy_spanner.construct_blind_greedy_eps_spanner(eps);
+        auto end_blind_greedy = std::chrono::steady_clock::now();
+        auto dur_blind_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_blind_greedy - begin_blind_greedy).count();
+        console->info("blind_greedy_time = {} microseconds", dur_blind_greedy);
+        exp_logger->info("blind_greedy_time = {} microseconds", dur_blind_greedy);
         console->info("{};{}", dist_name, blind_greedy_spanner.get_statistics());
         exp_logger->info("{};{}", dist_name, blind_greedy_spanner.get_statistics());
         exp_logger->flush();
         console->flush();
-        blind_greedy_spanner.check_ratio(eps);
+        //blind_greedy_spanner.check_ratio(eps);
     }
 
 //    {

### 2019-04-16 python_version/nearest_neighbor_mcgill.py

@@ -205,6 +205,12 @@ if __name__ == "__main__":
     qs = [1, 2, 3]
 
     fnames = ["mcgill/dist_matrix_mcgill_original_no_parallel_q_{}_dim_{}.txt".format(q, dim) for q in qs for dim in dims]
+    # np.random.seed()
+    # random.seed()
+    # n_pointses = [ i for i in range(100, 4000, 200)]
+    # dims = [0, 1, 2]
+    # qs = [1, 2]
+    # fnames = ["dist_matrix_q_{}_dim_{}.txt".format(q, dim) for q in qs for dim in dims]
     epsilons = [0.0001, 0.0005, 0.001, 0.005, 0.01, 0.1, 0.5]
     n_attempts = 20
 

### 2019-04-16 cpp_version/include/dynamic_spanner.h
Add distance timings from matrix, change call

@@ -73,16 +73,19 @@ public:
 
     size_t m_num_points;
     Matrix m_matrix;
+    MatrixReal m_distance_timings_matrix;
     double m_min_dist;
     double m_max_dist;
     std::mt19937 m_twister;
     double m_epsilon { 0.0 };
+    double m_time_elapsed { 0.0 };
     std::string m_strategy;
 
 public:
 
-    DynamicSpanner(const MatrixReal& distance_matrix)
+    DynamicSpanner(const MatrixReal& distance_matrix, const MatrixReal& distance_timings_matrix)
             :
+            m_distance_timings_matrix(distance_timings_matrix),
             m_min_dist(std::numeric_limits<double>::max()),
             m_max_dist(0.0),
             m_twister(std::random_device()())
@@ -442,6 +445,7 @@ public:
         info.upper_bound = info.lower_bound = info.distance;
         info_mirror.upper_bound = info_mirror.lower_bound = info.distance;
         update_bounds_using_distance(i, j);
+        m_time_elapsed += m_distance_timings_matrix[i][j];
         return info.distance;
     }
 
@@ -466,6 +470,11 @@ public:
             }
         }
         return result;
+    }
+
+    double get_microseconds_to_compute_distances() const
+    {
+        return m_time_elapsed;
     }
 
     size_t get_number_of_requested_distances() const

### 2019-04-16 cpp_version/src/main.cpp

@@ -139,7 +139,7 @@ bool read_distance_matrix(MatrixR& distance_matrix, double& max_distance, double
     return true;
 }
 
-bool read_distance_matrix_mcgill(MatrixR& distance_matrix, double& max_distance, double& min_distance, const std::string& fname, std::vector<int>::size_type n_samples)
+bool read_distance_matrix_mcgill_and_sample(MatrixR& distance_matrix, double& max_distance, double& min_distance, const std::string& fname, std::vector<int>::size_type n_samples)
 {
     std::ifstream matr_file(fname);
     if (not matr_file.good()) {
@@ -199,6 +199,53 @@ bool read_distance_matrix_mcgill(MatrixR& distance_matrix, double& max_distance,
 
     return true;
 }
+
+bool read_distance_matrix_mcgill(MatrixR& distance_matrix, double& max_distance, double& min_distance, const std::string& fname)
+{
+    std::ifstream matr_file(fname);
+    if (not matr_file.good()) {
+        std::cerr << "Cannot read matrix from file " << fname << std::endl;
+        return false;
+    }
+
+    int n_points;
+
+    matr_file >> n_points;
+
+    distance_matrix = MatrixR(n_points, std::vector<double>(n_points, 0.0));
+
+    for (size_t i = 0; i < n_points; ++i) {
+        for (size_t j = 0; j < n_points; ++j) {
+            matr_file >> distance_matrix[i][j];
+        }
+    }
+
+    for (size_t i = 0; i < n_points; ++i) {
+        for (size_t j = i; j < n_points; ++j) {
+            if (i == j)
+                assert(distance_matrix[i][j] == 0.0);
+            else
+                assert(distance_matrix[i][j] == distance_matrix[j][i]);
+        }
+    }
+
+    min_distance = std::numeric_limits<double>::max();
+    max_distance = -1.0;
+
+    for(size_t i = 0; i < n_points; ++i) {
+        for(size_t j = 0; j < n_points; ++j) {
+            max_distance = std::max(distance_matrix[i][j], max_distance);
+            if (i != j) {
+                min_distance = std::min(distance_matrix[i][j], min_distance);
+            }
+        }
+    }
+
+    assert(0.0 < min_distance and min_distance <= max_distance);
+
+    return true;
+}
+
 
 
[…]

### 2019-04-17 paper/expensive-distances-main.tex
add timings to paper

--- 2019-04-15 paper/expensive-distances-main.tex
@@
 In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
+By \textit{sparseness} of a graph with $n$
+vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.
 
@@
     \caption{Number of edges in blind spanners generated by different variants of the blind
-    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for normally distributed points
+    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for uniformly distributed points
     in dimension 2, $\eps = 0.1$.}
@@
 \begin{figure}[ht]
-    \begin{centering}
         \begin{tikzpicture}
@@
         \end{tikzpicture}
-\end{centering}
             \caption{Ratio \# edges / \# points for different variants of spanner 
@@
             \label{fig:spanner_ratio}
-\end{figure}
-
-\begin{figure}[ht!]
-    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
-    \caption{Results of blind greedy spanner for different dimensions.}
-    \label{fig:blind_greedy_dimensions}
-\end{figure}
-
-\ignore{
-\begin{figure}[ht!]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first, lower bound first}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Comparison of the four variants of \brndm algorithm.}
-    \label{fig:blind_rbr_variants}
-\end{figure}
-}
-
-\begin{figure}[ht!]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-    \addlegendentry{Greedy, dim = 2}
-
-    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-    \addlegendentry{Blind greedy, dim = 2}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
-    \label{fig:spanner_eps_dependence}
 \end{figure}
@@
 \end{centering}
-    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
+    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
     $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
@@
                           & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
-$\eps = 0.1$\\
 %\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
-\% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
-$\eps = 0.2$\\
+$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
 %\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
-\% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
-$\eps = 0.5$\\
+$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
 %\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
-\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
+%\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
+$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
 \bottomrule
 \end{tabular}
-\caption{Comparison of \bgrdy and \textsc{Greedy} algorithms for 0-dimensional
-diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
-\label{tbl:mcgill_original_blind_greedy_spanner}
+    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
+ with Wasserstein distance $W_q$.}
+\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
 \end{table}
 
-We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
+\begin{table}\centering
+    \ra{1.2}
+    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
+                          & Bruteforce & Blind greedy && Bruteforce & Blind greedy && Bruteforce & Blind greedy \\ \midrule
+
+$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
+$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
+$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
+\bottomrule
+\end{tabular}
+    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
+    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
+\label{tbl:mcgill_original_blind_greedy_spanner_timing}
+\end{table}
+
+We tested the algorithms for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
 \begin{enumerate}
@@
 by the size of the spanner.
-We can see that, even though none of the blind spanners can produce
-spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
+We can see that, even though none of the blind spanners produces
+spanners of the same quality as the standard greedy algorithm,
 \bgrdy and all variants of \brndm
 perform significantly better than both variants of \textsc{BlindQuasiSorted}.
-\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
-The ideal behavior is demonstrated by the non-blind greedy spanner,
-for which this ratio stays practically constant, confirming the linear growth.
-None of the blind algorithms seems to have this property, but among them
-the blind greedy spanner is the best one. If we assume that the number of edges
-is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
-linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
-the estimated exponents $\alpha$ for \bgrdy and standard
-greedy algorithms. Note that even for the greedy algorithm these estimated
-exponents can be significantly larger than 1,
-which is explained by the fact that the number of points
-on which we computed spanners is not large enough to clearly see
-the linear dependence.
-
-\begin{table}[]
-\begin{tabular}{|l|l|l|}
-\hline
-dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
-2         &         1.08                &  1.12                 \\ \hline
-3         &         1.24                &  1.41                \\ \hline
-4         &         1.42                &  1.77                 \\ \hline
-\end{tabular}
-\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
-on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
-\label{tbl:regr_coeff_spanner}
-\end{table}
-
+
+\ignore{
 As for different variants of the \brndm algorithm,
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
+the other two variants of the \brndm.  (see \cref{fig:blind_rbr_variants}). 
+}
 
@@
 
-We also tested higher dimensions and show the results for the best algorithm, \bgrdy,
-in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
-We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
-edges for $700$ points, which clearly shows some degrading for higher dimensions. Still,
-we remark that the WSPD spanner remains worse also in the higher-dimensional setup.
-
-The plot in \cref{fig:spanner_eps_dependence}
-compares the \bgrdy and \textsc{Greedy} algorithms
-on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
-is approximately the same for both algorithms. Since it is not cleary seen from the picture, we also note 
-that the ratio of the number of edges decreases for
-smaller values of $\eps$: 
-for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
-while for $\eps = 1/32$ the ratio is 2.6
-
-\ignore{
-Summing up, the \bgrdy algorithm is the best algorithm for our model, but also \brndm
-algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
-first. 
-If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
-Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
-tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
-Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
-(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
-access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
-find a working heuristic.
-}
+%\ignore{
+%Summing up, the \bgrdy algorithm is the best algorithm for our model, but also \brndm
+%algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
+%first. 
+%If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
+%Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
+%tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
+%Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
+%(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
+%access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
+%find a working heuristic.
+%}
 
@@
 Results of \bgrdy algorithm on the whole \dtype{original} dataset 
-for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner};
-\grdy algorithm is used as baseline\footnote{Of course, in the columns of \grdy algorithm
-we give its sparseness, i.e. number of edges divided by the number of all distances;
-the percent of computed distances is always 100.}.  The first thing to note is that for higher values of $q$
+for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness}
+and in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+\grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
 any advantage of our algorithm disappears. However, the cardinality of the \grdy
@@
 For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
-computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, than we
+computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, then we
 reduce the number of distances computed by a factor of 10.
@@
 
-The proof is given in the appendix.
+The proof is given in the appendix \cref{sec:proof}.
 
@@
 \appendix
+
+\section{Additional Experimental Results on Spanners}
+\label{sec:spanners_euclidean_addendum}
+
+\begin{figure}[ht!]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, connect first}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, lower bound first}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, connect first, lower bound first}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Comparison of the four variants of \brndm algorithm.}
+    \label{fig:blind_rbr_variants}
+\end{figure}
+
+\begin{figure}[ht!]
+    \begin{centering}
+    \includegraphics[width=0.8\textwidth]{\expDistFigPath blind-greedy-dims.png}
+    \caption{Results of \bgrdy spanner for different dimensions.}
+    \label{fig:blind_greedy_dimensions}
+    \end{centering}
+\end{figure}
+
+As for different variants of the \brndm algorithm,
+we note that their performance is almost the same, and the algorithm works
+significantly better than \textsc{QuasiSorted} variants, but obviously
+worse than the blind greedy variant. There is a consistent, though small,
+difference between the variants that do not force lower bounds first and
+the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
+
+\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
+The ideal behavior is demonstrated by the non-blind greedy spanner,
+for which this ratio stays practically constant, confirming the linear growth.
+None of the blind algorithms seems to have this property, but among them
+the blind greedy spanner is the best one. If we assume that the number of edges
+is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
+linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
+the estimated exponents $\alpha$ for \bgrdy and standard
+greedy algorithms. Note that even for the greedy algorithm these estimated
+exponents can be significantly larger than 1,
+which is explained by the fact that the number of points
+on which we computed spanners is not large enough to clearly see
+the linear dependence.
+
+\begin{table}[b]
+\begin{tabular}{|l|l|l|}
+\hline
+dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
+2         &         1.08                &  1.12                 \\ \hline
+3         &         1.24                &  1.41                \\ \hline
+4         &         1.42                &  1.77                 \\ \hline
+\end{tabular}
+\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
+on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
+\label{tbl:regr_coeff_spanner}
+\end{table}
+
+We also tested higher dimensions and show the results for the best algorithm, \bgrdy,
+in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
+We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
+edges for $700$ points, which clearly shows some degrading for higher dimensions. Still,
+we remark that the WSPD spanner remains worse also in the higher-dimensional setup.
+
+The plot in \cref{fig:spanner_eps_dependence}
+compares the \bgrdy and \textsc{Greedy} algorithms
+on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
+is approximately the same for both algorithms. Since it is not cleary seen from the picture, we also note 
+that the ratio of the number of edges decreases for
+smaller values of $\eps$: 
+for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
+while for $\eps = 1/32$ the ratio is 2.6
+
+\begin{figure}[ht]
+        \begin{tikzpicture}
+            \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
+                legend pos=north east]
+
+            \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+            \addlegendentry{Greedy, dim = 2}
+
+            \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+            \addlegendentry{Blind greedy, dim = 2}
+
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
+        \label{fig:spanner_eps_dependence}
+\end{figure}
+
 \section{Proof of \cref{thm:ann_bound}}
+\label{sec:proof_ann}
 

### 2019-04-17 paper/expensive-distances-main.tex
Some changes in paper

@@
 \section{Experiments on spanners}
-We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
-experimentally the performance of these heuristics. Clearly, for this metric space, our
-cost model is not meaningful since distance comparisons are cheap;
-but we picked this environment for controlled experiments.
-In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
+
+We performed two types of experiments: on points sampled from a low-dimensional Euclidean
+space and on persistence diagrams with Wasserstein distance.
+Clearly, our model is not meaningful in a Euclidean space, where
+distance computations are cheap.
+However, in this case we can easily check the performance
+of our heuristics on a variety of easily controllable
+data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
 By \textit{sparseness} of a graph with $n$
@@
 
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind quasi-sorted greedy}
-
-    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind quasi-sorted shaker}
-
-    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{WSPD}
-
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind greedy}
-
-    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Greedy}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Number of edges in blind spanners generated by different variants of the blind
-    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for uniformly distributed points
-    in dimension 2, $\eps = 0.1$.}
-    \label{fig:spanner_sparseness}
-\end{figure}
-
-\begin{figure}[ht]
-        \begin{tikzpicture}
-            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
-                legend pos=outer north east]
-
-                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind random, lower bound first}
-
-                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind greedy}
-
-                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Greedy}
-
-            \end{axis}
-        \end{tikzpicture}
-            \caption{Ratio \# edges / \# points for different variants of spanner 
-            algorithms. The plot is for normally distributed points
-            in dimension 2, $\eps = 0.1$.}
-            \label{fig:spanner_ratio}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
-        legend pos=outer north east]
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.5$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_ratio_mcgill_perturbed}
-\end{figure}
-
-\begin{figure}[ht]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.5$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_sparseness_mcgill_perturbed}
-\end{figure}
-
-\begin{table}\centering
-    \ra{1.2}
-    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
-                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
-                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
-                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
-%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
-$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
-%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
-$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
-%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
-%\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
-$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
-\bottomrule
-\end{tabular}
-    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
- with Wasserstein distance $W_q$.}
-\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
-\end{table}
-
-\begin{table}\centering
-    \ra{1.2}
-    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
-                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
-                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
-                          & Bruteforce & Blind greedy && Bruteforce & Blind greedy && Bruteforce & Blind greedy \\ \midrule
-
-$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
-$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
-$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
-\bottomrule
-\end{tabular}
-    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
-    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
-\label{tbl:mcgill_original_blind_greedy_spanner_timing}
-\end{table}
+%We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
+%experimentally the performance of these heuristics. Clearly, for this metric space, our
+%cost model is not meaningful since distance comparisons are cheap;
+%but we picked this environment for controlled experiments.
+%In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
+%By \textit{sparseness} of a graph with $n$
+%vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.
 
@@
 \end{enumerate}
-In all experiments the algorithms that we tested compared in the same way,
+In addition to the standard Euclidean norm on $\RR^n$, we used also $\ell_1$ and $\ell_\infty$ norms.
+In all these experiments the algorithms that we tested compared in the same way,
 so we only present results for the \textbf{uniform} point set in dimension 2.
@@
 
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind quasi-sorted greedy}
+
+    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind quasi-sorted shaker}
+
+    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{WSPD}
+
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, lower bound first}
+
+    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind greedy}
+
+    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Greedy}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Number of edges in blind spanners generated by different variants of the blind
+    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for uniformly distributed points
+    in dimension 2, $\eps = 0.1$.}
+    \label{fig:spanner_sparseness}
+\end{figure}
+
+\begin{figure}[!htbp]
+        \begin{tikzpicture}
+            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
+                legend pos=outer north east]
+
+                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind random, lower bound first}
+
+                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind greedy}
+
+                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Greedy}
+
+            \end{axis}
+        \end{tikzpicture}
+            \caption{Ratio \# edges / \# points for different variants of spanner 
+            algorithms. The plot is for normally distributed points
+            in dimension 2, $\eps = 0.1$.}
+            \label{fig:spanner_ratio}
+\end{figure}
+
 From these experiments we conclude that the algorithm with the highest chance
-of saving a significant amount of distance computations on real data, like large
-persistence diagrams, is \bgrdy. Even though \brndm also substantially reduces
+of saving a significant amount of distance computations on real data is \bgrdy. 
+Even though \brndm also substantially reduces
 the number of computed distances, its performance is significantly worse;
@@
 
-Finally, we ran experiments on two realistic data sets,
-namely the persistence diagrams of the shapes
-from  the McGill shape benchmark \cite{zhang2005mcgill}. 
+The persistence diagrams that we used as
+realistic data sets,
+come from the McGill shape benchmark \cite{zhang2005mcgill}.
 More precisely, we generated two sets of diagrams,
@@
 point clouds from the shape dataset; we only looked
-at dimension 0, since the diagrams in this dimension are the largest.
+at dimension 0, since the diagrams in this dimension are larger in size.
 We also added random noise to points of the original shapes,
@@
 
+We also compare the time that is needed to compute the matrix of all pairwise distances
+naively with the time that is needed to approximate it with \bgrdy spanner.
+The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+If one compares the running time with the sparseness results,
+it becomes evident that in this case the overhead of \bgrdy algorithm
+is rather small, compared to the time needed to compute the distances.
+Let us emphasize that the diagrams we have are still rather small, with a typical cardinality
+between 100 and 300 points. Nevertheless, even on this data set \bgrdy spanner
+provides a substantial improvement in terms of real running time.
+
 In order to see the growth of the spanner size, we use the \dtype{perturbed}
@@
 linear growth) for larger values of $n$.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
+        legend pos=outer north east]
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.5$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_ratio_mcgill_perturbed}
+\end{figure}
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_sparseness_mcgill_perturbed}
+\end{figure}
+
+\begin{table}\centering
+    \ra{1.2}
+    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
+                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
+%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
+$\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
+%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
+$\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
+%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
+%\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
+$\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
+\bottomrule
+\end{tabular}
+    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
+ with Wasserstein distance $W_q$.}
+\label{tbl:mcgill_original_blind_greedy_spanner_sparseness}
+\end{table}
+
+\begin{table}\centering
+    \ra{1.2}
+    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
+                          & Bruteforce & Blind greedy && Bruteforce & Blind greedy && Bruteforce & Blind greedy \\ \midrule
+
+$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
+$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
+$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
+\bottomrule
+\end{tabular}
+    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
+    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
+\label{tbl:mcgill_original_blind_greedy_spanner_timing}
+\end{table}
 
@@
 
-The proof is given in the appendix \cref{sec:proof}.
+The full proof is given in the appendix \cref{sec:proof_ann}.
 

### 2019-04-18 cpp_version/src/main.cpp
add distance time

@@ -517,6 +517,7 @@ int main(int argc, char** argv)
         greedy_spanner.construct_greedy_eps_spanner(eps);
         auto end_greedy = std::chrono::steady_clock::now();
         auto dur_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_greedy - begin_greedy).count();
+        dur_greedy += greedy_spanner.get_microseconds_to_compute_distances();
         console->info("greedy_time = {} microseconds", dur_greedy);
         exp_logger->info("greedy_time = {} microseconds", dur_greedy);
         console->info("{};{}", dist_name, greedy_spanner.get_statistics());
@@ -577,6 +578,7 @@ int main(int argc, char** argv)
         blind_greedy_spanner.construct_blind_greedy_eps_spanner(eps);
         auto end_blind_greedy = std::chrono::steady_clock::now();
         auto dur_blind_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_blind_greedy - begin_blind_greedy).count();
+        dur_blind_greedy += blind_greedy_spanner.get_microseconds_to_compute_distances();
         console->info("blind_greedy_time = {} microseconds", dur_blind_greedy);
         exp_logger->info("blind_greedy_time = {} microseconds", dur_blind_greedy);
         console->info("{};{}", dist_name, blind_greedy_spanner.get_statistics());

### 2019-04-18 cpp_version/src/main.cpp
interim

@@ -524,7 +524,7 @@ int main(int argc, char** argv)
         exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
         console->flush();
         exp_logger->flush();
-        greedy_spanner.check_ratio(eps);
+        //greedy_spanner.check_ratio(eps);
     }
 
 //    {

### 2019-04-18 paper/expensive-distances-main.tex
interim

@@
 Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
-an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
+a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
 performing $O(\log n)$ distance computations per query point.
@@
 only a constant number of distances needs to be computed in-between two minima.
-We also experimentally evaluate our approach and observe that the approach follows 
-roughly the theoretical prediction.
+Our experimental results match the theoretical predictions
+%We also experimentally evaluate our approach and observe that the approach follows 
+%roughly the theoretical prediction.
 
@@
 
-We shall need the following lemma, which is just a reformulation of the well-known
-packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).
-
-\begin{lemma}
-\label{lem:packing_lemma}
- Let $(\metricspace,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
- $B(x,R)$ in $\metricspace$ such that the distance between any two distinct points of $P$ is at least $r$.
- Then 
- \[|P|\leq \left(\frac{4R}{r}\right)^{d}\]
-
-\end{lemma}
-\begin{proof}
-We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
-balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
-we cover
-$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
-contain at most one point from $P$, 
-\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
-\end{proof}
-
 In the following, we will assume throughout that every considered metric space
 has a constant doubling dimension.
-
-\myparagraph{Well-separated pair decomposition}
-Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
-if 
-\[
-\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
-\]
-A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
-The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
-Har-Peled and Mendel \cite{hm-fast} 
-introduced the notion of net-trees and
-generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
-\begin{enumerate}
-    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
-    \item If $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
-are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
-    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
-The algorithm uses the net-tree structure.
-\end{enumerate}
-The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
-et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
-than net-trees. We mention in passing that cover trees can also be used for building
-a spanner (this can be proven with the same methods), and we use cover trees for
-building WSPD spanners in one of our implementations.
-
-\section{Algorithms for spanner construction}
 
@@
 
-On the other hand, spanner constructions based on WSPD only compute
+On the other hand, spanner constructions based on WSPD\footnote{Well-Separated Pair Decomposition; 
+for convenience we remind the definition in \cref{sec:appendix_background}} only compute
 $O(n\log n+n \eps^{-d})$ distances 
-to construct an $(1+\eps)$-spanner in doubling dimension $d$.
+to construct a \espanner in doubling dimension $d$.
 The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
@@
 
-\myparagraph{Blind spanners}
+\section{Blind spanners}
 %
@@
 \section{Experiments on spanners}
+\label{sec:experiment_spanners}
 
@@
         normal distribution in $\mathbb{R}^d$.
-    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
+    \item In the \textbf{clustered} test set we first sample cluster centres uniformly 
         at random from $[0,10000]^d$, and then we add normally distributed noise around
-        each of the centers. The number of clusters is chosen so that each cluster
+        each of the centres. The number of clusters is chosen so that each cluster
         contains 50 points.
@@
 diagrams were computed on the points of the 450 original
-point clouds from the shape dataset; we only looked
-at dimension 0, since the diagrams in this dimension are larger in size.
+point clouds from the shape dataset; we only used
+dimension 0, since the diagrams in this dimension are larger in size.
 We also added random noise to points of the original shapes,
@@
 
-Results of \bgrdy algorithm on the whole \dtype{original} dataset
-for different values of $\eps$ and $q$ are summarized in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness}
-and in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
+for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
 \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
@@
 The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
-If one compares the running time with the sparseness results,
-it becomes evident that in this case the overhead of \bgrdy algorithm
-is rather small, compared to the time needed to compute the distances.
-Let us emphasize that the diagrams we have are still rather small, with a typical cardinality
-between 100 and 300 points. Nevertheless, even on this data set \bgrdy spanner
-provides a substantial improvement in terms of real running time.
+First of all, we should explain why the timings in brute-force column
+depend on $\eps$. Wasserstein distance can either be computed exactly,
+by using the Hungarian algorithm, as done in \dionisus, or
+it can be computed with the user-given accuracy using the auction algorithm,
+as in \hera. Exact computation is very expensive (see \cite{kerber2017geometry}), and in our code for experiments
+with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
+In order to make the comparison absolutely fair, when we compute a \espanner,
+we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
+Let us emphasize that the diagrams we have are rather small, with a typical cardinality
+between 100 and 300 points. Nevertheless, on this data set \bgrdy spanner
+provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.
 
 In order to see the growth of the spanner size, we use the \dtype{perturbed}
-dataset.  We plot the sparseness of the spanners obtained by \grdy
+dataset, drawing from it random samples.
+We plot the sparseness of the spanners obtained by \grdy
 and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
@@
                           \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
-                          & Bruteforce & Blind greedy && Bruteforce & Blind greedy && Bruteforce & Blind greedy \\ \midrule
-
-$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
-$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
-$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
+                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
+%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
+%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
+%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
+$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
+$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  0  &  13557\\
+$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  0  &  9383\\
+$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  0  &  5146\\
 \bottomrule
@@
 \section{Experiments on approximate nearest neighbors}
+\label{sec:experiment_ann}
 
@@
         \includegraphics[width=0.9\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
-        \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
+        \caption{Ratio $\mbox{computed distances} / \log(n)$ for ANN algorithm. Data is for uniformly distributed
         points.}
@@
 approaching 30000.
+
+Similar experiments were performed on \dtype{perturbed} data set of persistence
+diagrams, which was described in \cref{sec:experiment_spanners}.
+For a given number $n$, we randomly select $n$ diagrams from the dataset
+and from the remaining diagrams we randomly choose 20 query points.
+In \cref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
+distances to $\log(n)$ for $\wsdist{1}$ and $\wsdist{2}$ metrics.
+The average ratio appears to fluctuate around 7 for $\wsdist{1}$ and around 8 for $\wsdist{2}$,
+which even makes the algorithm
+reasonable for practical application, if pairwise distances are already available.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Computed distances / log(\# Diagrams),
+        legend pos=outer north east]
+    \addplot+ table [x=n_points, y=avg_ratio_dim_0_q_1_eps_001, col sep=semicolon] {\expDistDataPath ann_mcgill_results.csv};
+        \addlegendentry{Average ratio, $\wsdist{1}$}
+
+    \addplot+ table [x=n_points, y=avg_ratio_dim_0_q_2_eps_001, col sep=semicolon] {\expDistDataPath ann_mcgill_results.csv};
+        \addlegendentry{Average ratio, $\wsdist{2}$}
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{ Ratio $\mbox{\# computed distances} / n$ for ANN algorithm. The plot is for \dtype{perturbed} persistence diagrams,
+    $\eps = 0.01$ with $\wsdist{1}$ and $\wsdist{2}$ metrics.}
+    \label{fig:ann_const_ratio_mcgill}
+\end{figure}
 
@@
 in that infinite-dimensional space, but have structures
-(e.g. clusters of close-by diagrams) which should be favorable for our approach.
+(e.g. clusters of close-by diagrams) which should be favourable for our approach.
+
+In particular, our experiments on real persistence diagrams
+show that the blind spanner can significantly accelerate
+computation of all pairwise distances. We must admit that,
+while in our theoretical model we completely ignore dependency on the number of points $n$,
+the $\Omega(n^3)$ complexity of our algorithms makes it practically
+non-competetive, when $n$ becomes large.
+On the other hand, this overhead is relatively easy to estimate
+in advance, just by knowing $n$, and make a decision,
+whether \bgrdy is worth trying. 
+The distances that \bgrdy computes would be computed anyway (and can be re-used),
+and the size of the \bgrdy spanner can itself be considered as a piece
+of information about the data set under consideration.
 
@@
 \label{sec:spanners_euclidean_addendum}
+In this section we give additional plots that give more information
+about the behaviour of \bgrdy spanner in Euclidean space.
 
@@
 
-As for different variants of the \brndm algorithm,
-we note that their performance is almost the same, and the algorithm works
-significantly better than \textsc{QuasiSorted} variants, but obviously
-worse than the blind greedy variant. There is a consistent, though small,
-difference between the variants that do not force lower bounds first and
-the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
+%As for different variants of the \brndm algorithm,
+%we note that their performance is almost the same, and the algorithm works
+%significantly better than \textsc{QuasiSorted} variants, but obviously
+%worse than the blind greedy variant. There is a consistent, though small,
+%difference between the variants that do not force lower bounds first and
+%the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
 
@@
 
-We also tested higher dimensions and show the results for the best algorithm, \bgrdy,
+We also tested higher dimensions and show the results
 in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
 We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
-edges for $700$ points, which clearly shows some degrading for higher dimensions. Still,
-we remark that the WSPD spanner remains worse also in the higher-dimensional setup.
+edges for $700$ points, which clearly shows some degrading for higher dimensions.
 
@@
 on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
-is approximately the same for both algorithms. Since it is not cleary seen from the picture, we also note 
+is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
 that the ratio of the number of edges decreases for
@@
 \label{sec:proof_ann}
+
+We shall need the following lemma, which is just a reformulation of the well-known
+packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).
+
+\begin{lemma}
+\label{lem:packing_lemma}
+ Let $(\metricspace,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
+ $B(x,R)$ in $\metricspace$ such that the distance between any two distinct points of $P$ is at least $r$.
+ Then 
+ \[|P|\leq \left(\frac{4R}{r}\right)^{d}\]
+
+\end{lemma}
+\begin{proof}
+We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
+balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
+we cover
+$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
+contain at most one point from $P$, 
+\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
+\end{proof}
 
@@
 is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$.
-This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the bruteforce algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis.
+This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis.
 The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
@@
 whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.
+
+\section{Well-Separated Pair Decomposition}
+\label{sec:appendix_background}
+Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
+if 
+\[
+\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
+\]
+A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
+of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
+The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
+Har-Peled and Mendel \cite{hm-fast} 
+introduced the notion of net-trees and
+generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
+\begin{enumerate}
+    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
+    \item If $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
+are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
+    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
+The algorithm uses the net-tree structure.
+\end{enumerate}
+The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
+et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
+than net-trees. We mention in passing that cover trees can also be used for building
+a spanner (this can be proven with the same methods), and we use cover trees for
+building WSPD spanners in one of our implementations.
+
+\section{Wasserstein Distance}

### 2019-04-19 cpp_version/include/dynamic_spanner.h
hera factor

@@ -80,7 +80,7 @@ public:
     double m_epsilon { 0.0 };
     double m_time_elapsed { 0.0 };
     std::string m_strategy;
-    double m_hera_factor {  1.0 / 1.01 };
+    double m_hera_factor {  1.0 / (1.01) };
 
 public:
 
@@ -89,7 +89,8 @@ public:
             m_distance_timings_matrix(distance_timings_matrix),
             m_min_dist(std::numeric_limits<double>::max()),
             m_max_dist(0.0),
-            m_twister(std::random_device()())
+            //m_twister(std::random_device()())
+            m_twister(1)
     {
         m_num_points = distance_matrix.size();
         m_matrix.resize(m_num_points);
@@ -291,10 +292,10 @@ public:
                     }
                     Real new_upper_bound = std::min(new_upper_bound_1, new_upper_bound_2);
 
-                    //if (new_upper_bound < m_hera_factor * info_xy.distance) {
-                    //    std::cerr << "ERROR HERE: " << "i = " << i << ", j = " << j << ", dist_ij = " << dist_ij << ", x = " << x << ", y = " << y << ", " << info_xy << ", new_ub_1  = " << new_upper_bound_1 << ", new ub2 = " << new_upper_bound_2 << std::endl;
-                    //    throw std::runtime_error("bad upper bound");
-                    //}
+                    if (new_upper_bound < m_hera_factor * info_xy.distance) {
+                        std::cerr << "ERROR HERE: " << "i = " << i << ", j = " << j << ", dist_ij = " << dist_ij << ", x = " << x << ", y = " << y << ", " << info_xy << ", new_ub_1  = " << new_upper_bound_1 << ", new ub2 = " << new_upper_bound_2 << std::endl;
+                        throw std::runtime_error("bad upper bound");
+                    }
 
                     info_xy.upper_bound = std::min(info_xy.upper_bound, new_upper_bound);
                     info_yx.upper_bound = info_xy.upper_bound;
@@ -324,7 +325,7 @@ public:
                 Real new_lower_bound_2;
                 Real new_lower_bound_2_1 = m_matrix[x][j].upper_bound;
                 Real new_lower_bound_2_2 = m_matrix[i][y].upper_bound;
-                //std::cout << "Upper bounds xj " <<  new_lower_bound_2_1 << " iy " << new_lower_bound_2_2 << std::endl;
[…]

### 2019-04-19 cpp_version/src/main.cpp

@@ -511,21 +511,21 @@ int main(int argc, char** argv)
 
 #else
 
-    {
-        DynamicSpannerR greedy_spanner(dist_matrix, dist_matrix_timings);
-        auto begin_greedy = std::chrono::steady_clock::now();
-        greedy_spanner.construct_greedy_eps_spanner(eps);
-        auto end_greedy = std::chrono::steady_clock::now();
-        auto dur_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_greedy - begin_greedy).count();
-        dur_greedy += greedy_spanner.get_microseconds_to_compute_distances();
-        console->info("greedy_time = {} microseconds", dur_greedy);
-        exp_logger->info("greedy_time = {} microseconds", dur_greedy);
-        console->info("{};{}", dist_name, greedy_spanner.get_statistics());
-        exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
-        console->flush();
-        exp_logger->flush();
-        //greedy_spanner.check_ratio(eps);
-    }
+    //{
+    //    DynamicSpannerR greedy_spanner(dist_matrix, dist_matrix_timings);
+    //    auto begin_greedy = std::chrono::steady_clock::now();
+    //    greedy_spanner.construct_greedy_eps_spanner(eps);
+    //    auto end_greedy = std::chrono::steady_clock::now();
+    //    auto dur_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_greedy - begin_greedy).count();
+    //    dur_greedy += greedy_spanner.get_microseconds_to_compute_distances();
+    //    console->info("greedy_time = {} microseconds", dur_greedy);
+    //    exp_logger->info("greedy_time = {} microseconds", dur_greedy);
+    //    console->info("{};{}", dist_name, greedy_spanner.get_statistics());
+    //    exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
+    //    console->flush();
+    //    exp_logger->flush();
+    //    //greedy_spanner.check_ratio(eps);
+    //}
 
 //    {
 //        DynamicSpannerR quasi_greedy_spanner(dist_matrix);

### 2019-04-19 python_version/table_printer.py
table_printer changed

@@ -1,0 +1,66 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import glob
+import sys
+import re
+from math import floor
+
+
+def sum_file(fname):
+    result = 0
+    with open(fname, 'r') as f:
+        first_line = True
+        for s in f:
+            if first_line:
+                first_line = False
+                continue
+            result += float(s)
+    return floor(result / 2e6)
+
+if __name__ == "__main__":
+    epss = ["0.01", "0.1", "0.2", "0.5"]
+    qs = [1,2,3]
+    dims = [0]
+
+    final_table = {}
+
+    for dim in dims:
+        for eps in epss:
+            for q in qs:
+                dist_fname = "mcgill_orig/dist_matrix_mcgill_original_no_parallel_q_{}_dim_{}.txt".format(q, dim)
+                dist_timing_fname = "mcgill_orig/dist_timing_matrix_mcgill_original_q_{}_dim_{}_delta_{}.txt".format(q, dim, eps)
+                fname_base = os.path.basename(dist_fname)
+                log_name="log_mcgill_timing_orig_correct_spanner_{}_eps_{}.txt".format(fname_base, eps)
+                log_exists = os.path.isfile(log_name)
+                dist_timing_matrix_exists = os.path.isfile(dist_timing_fname)
+                if not log_exists and not dist_timing_matrix_exists:
+                    continue
+                blind_greedy_time = 0
+                bruteforce_time = 0
+                if dist_timing_matrix_exists:
+                    bruteforce_time = sum_file(dist_timing_fname)
+                if log_exists:
+                    with open(log_name) as f:
+                        for s in f:
+                            sr = re.search("blind_greedy_time = ([0-9]+) microseconds", s)
+                            if sr:
+                                blind_greedy_time = floor( int(sr.group(1)) / 1e6 )
+
+                final_table[(dim, eps, q)] = (bruteforce_time, blind_greedy_time)
+
+
+    for dim in dims:
+        for eps in epss:
+            s = "$\eps = {}$    & ".format(eps)
+            for q in qs:
+                try:
+                    bt, bgt = final_table[(dim, eps, q)]
+                except KeyError:
+                    bt = bgt = 0
+                if q < 3:
+                    s += "  {}  &  {}  &&".format(bt, bgt)
+                elif q == 3:
+                    s += "  {}  &  {}\\\\".format(bt, bgt)
+            print(s)

### 2019-04-19 cpp_version/include/dynamic_spanner.h
disable checks for timings

@@ -292,10 +292,10 @@ public:
                     }
                     Real new_upper_bound = std::min(new_upper_bound_1, new_upper_bound_2);
 
-                    if (new_upper_bound < m_hera_factor * info_xy.distance) {
-                        std::cerr << "ERROR HERE: " << "i = " << i << ", j = " << j << ", dist_ij = " << dist_ij << ", x = " << x << ", y = " << y << ", " << info_xy << ", new_ub_1  = " << new_upper_bound_1 << ", new ub2 = " << new_upper_bound_2 << std::endl;
-                        throw std::runtime_error("bad upper bound");
-                    }
+                    //if (new_upper_bound < m_hera_factor * info_xy.distance) {
+                    //    std::cerr << "ERROR HERE: " << "i = " << i << ", j = " << j << ", dist_ij = " << dist_ij << ", x = " << x << ", y = " << y << ", " << info_xy << ", new_ub_1  = " << new_upper_bound_1 << ", new ub2 = " << new_upper_bound_2 << std::endl;
+                    //    throw std::runtime_error("bad upper bound");
+                    //}
 
                     info_xy.upper_bound = std::min(info_xy.upper_bound, new_upper_bound);
                     info_yx.upper_bound = info_xy.upper_bound;
@@ -334,13 +334,13 @@ public:
                 }
                 Real new_lower_bound = std::max(new_lower_bound_1, new_lower_bound_2);
 
-                if (new_lower_bound > info_xy.distance) {
-                    std::cerr << "ERROR HERE: " << "x = " << x << ", y = " << y << ", " << info_xy << ", new_lb_1  = " << new_lower_bound_1 << ", new_lb2 = " << new_lower_bound_2 << std::endl;
-                    std::cerr << "ERROR HERE: " << "i = " << i << ", j = " << j << ", dist_ij = " << dist_ij <<", with factor: "
-                        << m_hera_factor * dist_ij <<
-                        ", Upper bounds xj " <<  new_lower_bound_2_1 << " iy " << new_lower_bound_2_2 << std::endl;
-                    throw std::runtime_error("bad lower bound");
-                }
+                //if (new_lower_bound > info_xy.distance) {
+                //    std::cerr << "ERROR HERE: " << "x = " << x << ", y = " << y << ", " << info_xy << ", new_lb_1  = " << new_lower_bound_1 << ", new_lb2 = " << new_lower_bound_2 << std::endl;
[…]

### 2019-04-19 cpp_version/src/main.cpp

@@ -511,21 +511,21 @@ int main(int argc, char** argv)
 
 #else
 
-    //{
-    //    DynamicSpannerR greedy_spanner(dist_matrix, dist_matrix_timings);
-    //    auto begin_greedy = std::chrono::steady_clock::now();
-    //    greedy_spanner.construct_greedy_eps_spanner(eps);
-    //    auto end_greedy = std::chrono::steady_clock::now();
-    //    auto dur_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_greedy - begin_greedy).count();
-    //    dur_greedy += greedy_spanner.get_microseconds_to_compute_distances();
-    //    console->info("greedy_time = {} microseconds", dur_greedy);
-    //    exp_logger->info("greedy_time = {} microseconds", dur_greedy);
-    //    console->info("{};{}", dist_name, greedy_spanner.get_statistics());
-    //    exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
-    //    console->flush();
-    //    exp_logger->flush();
-    //    //greedy_spanner.check_ratio(eps);
-    //}
+    {
+        DynamicSpannerR greedy_spanner(dist_matrix, dist_matrix_timings);
+        auto begin_greedy = std::chrono::steady_clock::now();
+        greedy_spanner.construct_greedy_eps_spanner(eps);
+        auto end_greedy = std::chrono::steady_clock::now();
+        auto dur_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_greedy - begin_greedy).count();
+        dur_greedy += greedy_spanner.get_microseconds_to_compute_distances();
+        console->info("greedy_time = {} microseconds", dur_greedy);
+        exp_logger->info("greedy_time = {} microseconds", dur_greedy);
+        console->info("{};{}", dist_name, greedy_spanner.get_statistics());
+        exp_logger->info("{};{}", dist_name, greedy_spanner.get_statistics());
+        console->flush();
+        exp_logger->flush();
+        //greedy_spanner.check_ratio(eps);
+    }
 
 //    {
 //        DynamicSpannerR quasi_greedy_spanner(dist_matrix);

### 2019-04-21 paper/expensive-distances-main.tex
interim

@@
 Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
-a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
+a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
 performing $O(\log n)$ distance computations per query point.
@@
 %
-We propose simple algorithms for spanner construction and approximate nearest neighbor search
+We propose simple algorithms for spanner construction and approximate nearest neighbour search
 and evaluate them theoretically and experimentally in the defined cost model.
@@
 We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
-spanners on the tested example. Nevertheless, we leave the question open whether our construction
-yields a spanner of asymptotically linear size.
-
-For approximate nearest neighbor, we devise a simple randomized incremental algorithm and show that
-the number of distance queries to find an approximate nearest neighbor is $O(\log n)$ in expectation.
-Our proof is based on the well-known observation that the nearest neighbor changes $O(\log n)$ times
+spanners on the tested examples. Nevertheless, we leave the question open whether our construction
+yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
+outperforms naive computation of pairwise distances or runs slighly slower, depending on the approximation
+quality.
+
+For approximate nearest neighbour, we devise a simple randomized incremental algorithm and show that
+the number of distance queries to find an approximate nearest neighbour is $O(\log n)$ in expectation.
+Our proof is based on the well-known observation that the nearest neighbour changes $O(\log n)$ times
 in expectation when traversing the sequence of points, combined with a packing argument certifying that
 only a constant number of distances needs to be computed in-between two minima.
-Our experimental results match the theoretical predictions
+Our experimental results match the theoretical predictions.
 %We also experimentally evaluate our approach and observe that the approach follows 
@@
 $\dist(p_i,p_j)=v\in\RR$ has been computed.
-First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
+First, we reset $a_{i,j}$, $a_{j,i}$, $b_{i,j}$ and $b_{j,i}$ to $v$, since the distance
 of $p_i$ and $p_j$ is exactly $v$.
@@
 %$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
-$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
+%$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
 $\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  0  &  13557\\
@@
 
-\section{Approximate nearest neighbors}
+\section{Approximate nearest neighbours}
 \label{sec:ann}
-We consider the standard problem of finding an approximate nearest neighbor: given
+We consider the standard problem of finding an approximate nearest neighbour: given
 $n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
@@
 to the query point $q$, which are initially all set to $0$.
-We also remember the closest neighbor $c$ that we have seen so
+We also remember the closest neighbour $c$ that we have seen so
 far and its distance $v$ to $q$.
 We refer to the point $c$ as the \emph{candidate}.
-We maintain the invariant that $c$ is an approximate nearest neighbor
+We maintain the invariant that $c$ is an approximate nearest neighbour
 to $q$ for the points $\{p_1,\ldots,p_i\}$.
@@
 bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
-If so, $c$ remains an approximate nearest neighbor and
+If so, $c$ remains an approximate nearest neighbour and
 we do not query the distance of $p_i$ to $q$.
@@
 accordingly. At the end of the loop, $c$ is an approximate nearest
-neighbor. The pseudocode of the procedure follows.
+neighbour. The pseudocode of the procedure follows.
 
@@
 
-We remark that we obtain an exact nearest neighbor algorithm
+We remark that we obtain an exact nearest neighbour algorithm
 when setting $\eps$ to $0$, which means 
@@
 
-The full proof is given in the appendix \cref{sec:proof_ann}.
-
-\section{Experiments on approximate nearest neighbors}
+The proof idea is the following. First, we notice
+that current candidate $c$ allows us to exclude points
+$p_i$ that are too far from $c$. This means that if $p_i$ is
+outside of the ball of radius $2\dist(c, q)$ centered at $c$, then the lower
+bound for $p_i$ will be larger than $\dist(c, q)$,
+and discarding $p_i$ does not cost us a distance computation.
+On the other hand, once we computed a distance to $c$
+from a point $p_j$, which is inside this ball but happens to be farther from $q$ than $c$,
+we can use it to discard points which are very close to $p_j$, 
+namely, inside a small ball of radius approximately $\eps \dist(c, q)$, because we
+are solving approximate nearest neighbour problem.
+Then we use a standard packing argument for doubling spaces, applied to the large and small balls, to show
+that the number of distance computations that we pay per candidate is constant.
+It remains only to prove that the number of different candidates is $O(\log n)$
+in expectation; there is a technicality that prevents using the standard
+backwards analysis argument verbatim, but it can be adapted.
+The full proof is given in \cref{sec:proof_ann}.
+
+\section{Experiments on approximate nearest neighbours}
 \label{sec:experiment_ann}
@@
 \label{sec:spanners_euclidean_addendum}
-In this section we give additional plots that give more information
+In this section we include plots that give more information
 about the behaviour of \bgrdy spanner in Euclidean space.
@@
 
-We shall need the following lemma, which is just a reformulation of the well-known
+The following lemma is just a reformulation of the well-known
 packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).
@@
 
-Towards the proof, we will use the following geometric lemma which
+The following geometric lemma
 can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
@@
 \end{tikzpicture}
-\caption{First two steps of the ANN algorithm. First $p_1$ is 
-chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
-distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them.
-Note that the point $p_5$, which
-is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$.
-This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis.
+\caption{First two steps of the ANN algorithm. 
+%    First $p_1$ is 
+%chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
+%distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered 
+%at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
+%is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$
+%in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. 
+%This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the
+%brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
 The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
@@
     \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
-    \caption{Example of point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
-    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
+    \caption{Example of point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
+    The exact nearest neighbour is the point $p_1$, next point $p_i$ is placed 
     in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
-    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
+    is the true nearest neighbour cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
     $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
@@
 \end{proof}
+
+In \cref{fig:ann_illustration} we illustrate this. First, $p_1$ is
+chosen as the current candidate, and we must compute $\dist(p_2, q)$.
+After that the algorithm will not compute
+distance to any of the points inside the heavily shaded ball or outside
+the lightly shaded ball that are centered at $p_2$,
+because their lower bounds allow us to discard them.
 
@@
 have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
+See \cref{fig:ann_illustration}.
 \end{proof}
@@
 that the candidate changes only a logarithmic number of times in expectation.
-While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
-that argument.
+Let us consider \cref{fig:ann_illustration} again. Notice that the point $p_5$, which
+is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
+$p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are farther from 
+$q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the 
+candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
+thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis,
+and we need to modify it slightly.
 
@@
 $v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
-nearest neighbor of $\{p_1,\ldots,p_k\}$.
+nearest neighbour of $\{p_1,\ldots,p_k\}$.
 
@@
 Indeed, as the example in \cref{fig:exact_bad_example} shows, there are point sets where the expected number
-of distance computations for exact nearest neighbor is linear.
+of distance computations for exact nearest neighbour is linear.
 
@@
 
-\section{Well-Separated Pair Decomposition}
+\section{Additional Definitions}
+\myparagraph{Well-Separated Pair Decomposition}
 \label{sec:appendix_background}
@@
 
-\section{Wasserstein Distance}
+\myparagraph{Wasserstein Distance}
+A persistence diagram is a multiset of points $(x, y)$ above the diagonal
+$\Delta$ and all points of $\Delta$ with infinite cardinality.

### 2019-04-21 python_version/mcgill/test_mcgill_save_time.py
interim

@@ -40,42 +40,49 @@ def read_mcgill_diagrams(diag_dir_root, dim):
 
 
 def get_distance_tuple(i, j, dgm1, dgm2, q, delta):
-    print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, started computation" % (i, j, q, dgm1.label, dgm2.label))
+    # print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, started computation" % (i, j, q, dgm1.label, dgm2.label))
     start_time = dt.datetime.now()
     res = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q, delta=delta)
     end_time = dt.datetime.now()
     elapsed_time = (end_time - start_time).microseconds
-    print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, d = %f, finished computation" % (i, j, q, dgm1.label, dgm2.label, res))
+    # print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, d = %f, time = %d microsec, finished computation" % (i, j, q, dgm1.label, dgm2.label, res, elapsed_time))
     return (i, j, elapsed_time, elapsed_time)
     
 
 if __name__ == "__main__":
-    if len(sys.argv) < 2:
-        print("dim\n")
-        exit(0)
-    dim = int(sys.argv[1])
-    delta = 0.01
-    diag_dir_root = '/home/anigmetov/Downloads/mcgill_shape_benchmark_dgms_no_duplicates/'
-    dgms = read_mcgill_diagrams(diag_dir_root, dim)
-    n_pts = len(dgms)
-    print("n_pts = %d\n" % n_pts)
-    total_distances = float(n_pts * (n_pts - 1) / 2)
-    data = np.array(dgms[:])
-    times = {}
-    for q in [1,2,3]:
-        times[q] = 0
-        dist_matr_fname = "dist_timing_matrix_mcgill_original_q_{}_dim_{}.txt".format(q, dim)
-        dist_matrix = np.zeros((n_pts, n_pts))
-        for i in range(n_pts):
-            for j in range(i+1, n_pts):
-                (i, j, wd, et) = get_distance_tuple(i, j, dgms[i], dgms[j], q, delta)
-                dist_matrix[i][j] = wd
-                dist_matrix[j][i] = wd
-                times[q] += et
[…]

### 2019-04-21 paper/expensive-distances-main.tex

@@
 spanners on the tested examples. Nevertheless, we leave the question open whether our construction
-yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
-outperforms naive computation of pairwise distances or runs slighly slower, depending on the approximation
-quality.
+yields a spanner of asymptotically linear size.
 
@@
 $\dist(p_i,p_j)=v\in\RR$ has been computed.
-First, we reset $a_{i,j}$, $a_{j,i}$, $b_{i,j}$ and $b_{j,i}$ to $v$, since the distance
+First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
 of $p_i$ and $p_j$ is exactly $v$.
@@
 \label{sec:spanners_euclidean_addendum}
-In this section we include plots that give more information
+In this section we give additional plots that give more information
 about the behaviour of \bgrdy spanner in Euclidean space.
@@
 
-The following lemma is just a reformulation of the well-known
+We shall need the following lemma, which is just a reformulation of the well-known
 packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).
@@
 
-The following geometric lemma
+Towards the proof, we will use the following geometric lemma which
 can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
@@
 \end{tikzpicture}
-\caption{First two steps of the ANN algorithm. 
-%    First $p_1$ is 
-%chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
-%distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered 
-%at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
-%is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$
-%in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. 
-%This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the
-%brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
+\caption{First two steps of the ANN algorithm. First $p_1$ is 
+chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
+distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered at $p_2$, because their lower bounds allow us to discard them.
+Note that the point $p_5$, which
+is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$.
+This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis.
 The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
@@
 \end{proof}
-
-In \cref{fig:ann_illustration} we illustrate this. First, $p_1$ is
-chosen as the current candidate, and we must compute $\dist(p_2, q)$.
-After that the algorithm will not compute
-distance to any of the points inside the heavily shaded ball or outside
-the lightly shaded ball that are centered at $p_2$,
-because their lower bounds allow us to discard them.
 
@@
 have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
-See \cref{fig:ann_illustration}.
 \end{proof}
@@
 that the candidate changes only a logarithmic number of times in expectation.
-Let us consider \cref{fig:ann_illustration} again. Notice that the point $p_5$, which
-is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
-$p_6,p_7,p_8,p_9$ in the annulus between the dashed and solid circle, which are farther from 
-$q$ than $p_5$, will be chosen as $c$. This shows that in our algorithm the distance from the 
-candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
-thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis,
-and we need to modify it slightly.
+While we have not found a simple proof for this claim, we can prove the statement with a slight variant of
+that argument.
 
@@
 
-\section{Additional Definitions}
-\myparagraph{Well-Separated Pair Decomposition}
+\section{Well-Separated Pair Decomposition}
 \label{sec:appendix_background}
@@
 
-\myparagraph{Wasserstein Distance}
-A persistence diagram is a multiset of points $(x, y)$ above the diagonal
-$\Delta$ and all points of $\Delta$ with infinite cardinality.
+\section{Wasserstein Distance}

### 2019-04-21 python_version/mcgill/test_mcgill_save_time_attempts.py
attempt

@@ -1,0 +1,88 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+
+import numpy as np
+import dionysus as d
+import datetime as dt
+
+
+class MyDiagram:
+    def __init__(self, dgm_data, id, label=None):
+        self.dgm_data = dgm_data
+        self.id = id
+        self.label = label
+
+
+
+def read_mcgill_diagrams(diag_dir_root, dim):
+    diag_dirs = [os.path.join(diag_dir_root, f) for f in os.listdir(diag_dir_root)
+                 if f.endswith('Ply')]
+    dim_ext = '.%d.txt' % dim
+    dgms = []
+    dgm_id = 0
+    for diag_dir in diag_dirs:
+        diag_files = [os.path.join(diag_dir, f) for f in os.listdir(diag_dir)
+                      if f.endswith(dim_ext)]
+        for dgm_f in diag_files:
+            pts = []
+            with open(dgm_f, 'r') as dgm_file:
+                for line in dgm_file:
+                    x, y = [float(x) for x in line.split()]
+                    pts.append((x, y))
+            dgms.append(MyDiagram(d.Diagram(pts[:]), dgm_id, dgm_f))
+            dgm_id += 1
+            # print("Read file %s, number of points %d" % (dgm_f, len(pts)))
+    print("Read %d diagrams" % len(dgms))
+    return dgms
+
+
+def get_distance_tuple(i, j, dgm1, dgm2, q, delta):
+    # print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, started computation" % (i, j, q, dgm1.label, dgm2.label))
+    start_time = dt.datetime.now()
+    res = d.wasserstein_distance(dgm1.dgm_data, dgm2.dgm_data, q=q, delta=delta)
+    end_time = dt.datetime.now()
+    elapsed_time = (end_time - start_time).microseconds
+    # print("i = %d, j = %d, q = %d, fname1 = %s, fname2 = %s, d = %f, time = %d microsec, finished computation" % (i, j, q, dgm1.label, dgm2.label, res, elapsed_time))
+    return (i, j, elapsed_time, elapsed_time)
+    
+
+if __name__ == "__main__":
+    # if len(sys.argv) < 2:
+    #     print("dim\n")
+    #     exit(0)
+    # for dim in [2, 1, 0]:
+    for dim in [2,0]:
[…]

### 2019-04-21 paper/draft_esa.tex
interim
Fits in page limit

@@
 % uncomment for LIPICS 
-\author{Anonymous Authors}{Anon}{}{}{}
+\author{Anonymous Authors}{Anonymous}{}{}{}
 %\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
@@
 \keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-\ccsdesc[500]{Theory of computation~Random search heuristics}
 \ccsdesc[500]{Theory of computation~Random search heuristics}
@@
 of points of the metric space.
-We introduce a different model where we assume
+We introduce a different model, where we assume
 that the computation of a single distance is

### 2019-04-21 paper/expensive-distances-main.tex
Add timings
Fits in page limit

--- 2019-04-21 paper/expensive-distances-main.tex
@@
 the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
-It is a common assumption to assume $\complexity$ to be a constant; 
+It is a common assumption to assume $\complexity$ to be a constant.
 There are good reasons for that: the most common case of a metric space
@@
 yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
-outperforms naive computation of pairwise distances or runs slighly slower, depending on the approximation
+outperforms naive computation of all pairwise distances or runs slighly slower, depending on the approximation
 quality.
@@
 
-\begin{figure}[!htbp]
-        \begin{tikzpicture}
-            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
-                legend pos=outer north east]
-
-                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind random, lower bound first}
-
-                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind greedy}
-
-                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Greedy}
-
-            \end{axis}
-        \end{tikzpicture}
-            \caption{Ratio \# edges / \# points for different variants of spanner 
-            algorithms. The plot is for normally distributed points
-            in dimension 2, $\eps = 0.1$.}
-            \label{fig:spanner_ratio}
-\end{figure}
-
 From these experiments we conclude that the algorithm with the highest chance
@@
 we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
-Let us emphasize that the diagrams we have are rather small, with a typical cardinality
-between 100 and 300 points. Nevertheless, on this data set \bgrdy spanner
+Let us emphasize that the diagrams we have are rather small, with average cardinality
+350 points. Nevertheless, on this data set \bgrdy spanner
 provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.
@@
 %$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
-$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  0  &  13557\\
-$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  0  &  9383\\
-$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  0  &  5146\\
+%$\eps = 0.01$    &   4093  &  0  &&  5331  &  0  &&  16852  &  0\\
+$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  11781  &  13557\\
+$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  10541   &  9383\\
+$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  9418    &  5146\\
 \bottomrule
@@
 
+\begin{figure}[!htbp]
+        \begin{tikzpicture}
+            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
+                legend pos=outer north east]
+
+                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind random, lower bound first}
+
+                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind greedy}
+
+                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Greedy}
+
+            \end{axis}
+        \end{tikzpicture}
+            \caption{Ratio \# edges / \# points for different variants of spanner 
+            algorithms. The plot is for normally distributed points
+            in dimension 2, $\eps = 0.1$.}
+            \label{fig:spanner_ratio}
+\end{figure}
+
 \cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
@@
 None of the blind algorithms seems to have this property, but among them
-the blind greedy spanner is the best one. If we assume that the number of edges
+the \bgrdy spanner is the best one. If we assume that the number of edges
 is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
@@
 
-We also tested higher dimensions and show the results
-in dimensions 2, 3 and 4 in \cref{fig:blind_greedy_dimensions}.
-We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
+The plot in \cref{fig:blind_greedy_dimensions}
+compares performance of \bgrdy in different dimensions.
+We see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
 edges for $700$ points, which clearly shows some degrading for higher dimensions.
@@
 compares the \bgrdy and \textsc{Greedy} algorithms
-on \textbf{uniform} point sets for different choices of $\eps$. We can see that dependence on $\eps$
+on \textbf{uniform} point sets for different choices of $\eps$. Dependence on $\eps$
 is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
@@
 \end{proof}
-
-The following geometric lemma
-can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
-further distance computations of points very close to $p_i$ or very far from $p_i$
-will be avoided.
 
@@
 \end{lemma}
+
+\cref{lem:bound_lemma}
+can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
+further distance computations of points very close to $p_i$ or very far from $p_i$
+will be avoided.
+
 \begin{proof}

### 2019-04-23 python_version/mcgill/print_mcgill_avg_attempts.py
interim

@@ -1,0 +1,32 @@
+#!/usr/bin/env python3
+
+import os
+import os.path
+import sys
+import re
+
+import numpy as np
+
+def get_time_from_last_line(fname):
+    with open(dist_matr_fname, 'r') as dmf:
+        last_line = ""
+        for s in dmf:
+            last_line = s
+        reg_res = re.match("(.*) seconds")
+        if reg_res:
+            return float(reg_res.group(1))
+        else:
+            return 0.0
+
+
+if __name__ == "__main__":
+    for dim in [2,0]:
+        for delta in [0.01, 0.5, 0.2, 0.1]:
+            for q in [1,2,3]:
+                total_time = 0.0
+                n_attempts = 5
+                for attempt in range(n_attempts):
+                    dist_matr_fname = "dist_timing_matrix_mcgill_original_q_{}_dim_{}_delta_{}_attempt_{}.txt".format(q, dim, delta, attempt)
+                    time_in_file = get_time_from_last_line(fname)
+                    total_time += time_in_file
+                avg_time = total_time / n_attempts

### 2019-04-23 python_version/mcgill/test_mcgill_save_time_attempts.py

@@ -85,4 +85,4 @@ if __name__ == "__main__":
                         for i in range(n_pts):
                             for j in range(0, n_pts):
                                 dmf.write("%f\n" % dist_matrix[i][j])
-                        dmf.write("\n%f seconds\n" % times[q] / 1e6)
+                        dmf.write("\n%f seconds\n" % (times[q] / 1e6))

### 2019-04-23 paper/expensive-distances-main.tex
almost final

@@
 
-\section{Background and Definitions}
+\section{Background}
 %
@@
 
-On the other hand, spanner constructions based on WSPD\footnote{Well-Separated Pair Decomposition; 
-for convenience we remind the definition in \cref{sec:appendix_background}} only compute
+On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
+for convenience, we remind the definition of WSPD in \cref{sec:appendix_background}} only compute
 $O(n\log n+n \eps^{-d})$ distances 
@@
 Even though \brndm also substantially reduces
-the number of computed distances, its performance is significantly worse;
+the number of computed distances, its performance is worse;
 other algorithms that we tested do not produce sparse spanners.
@@
 we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
-dense spanners.
-
-The persistence diagrams that we used as
-realistic data sets,
-come from the McGill shape benchmark \cite{zhang2005mcgill}.
+dense spanners. More data from experiments in $\RR^n$ can be found in
+\cref{sec:spanners_euclidean_addendum}.
+
+The realistic data sets that we used
+are the persistence diagrams
+computed from the McGill shape benchmark \cite{zhang2005mcgill}.
 More precisely, we generated two sets of diagrams,
@@
 Persistence diagrams were computed with \libraryname{DIPHA}
-library.
+library \cite{bauer2014distributed}.
 
@@
 then even the \grdy spanner is almost half of the complete graph.
-We can offer the following explanation, assuming the reader is familiar
-with the definition of $\wsdist{q}$. For higher values of $q$,
+We can offer the following explanation\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}. For higher values of $q$,
+\label{}
 longer edges of an optimal matching play a more important role
@@
 The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
-First of all, we should explain why the timings in brute-force column
+First of all, we should explain why the timings in the brute-force columns
 depend on $\eps$. Wasserstein distance can either be computed exactly,
-by using the Hungarian algorithm, as done in \dionisus, or
-it can be computed with the user-given accuracy using the auction algorithm,
-as in \hera. Exact computation is very expensive (see \cite{kerber2017geometry}), and in our code for experiments
+by using the Hungarian algorithm, or
+it can be computed with the user-given accuracy with the auction algorithm,
+as in \hera. Exact computation is very expensive \cite{kerber2017geometry}, and in our code for experiments
 with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
-In order to make the comparison absolutely fair, when we compute a \espanner,
+To make the comparison absolutely fair, when we compute a \espanner,
 we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
@@
 It remains only to prove that the number of different candidates is $O(\log n)$
-in expectation; there is a technicality that prevents using the standard
-backwards analysis argument verbatim, but it can be adapted.
+in expectation; the standard
+backwards analysis argument cannot be applied verbatim, because of a technicality,
+but it can be adapted.
 The full proof is given in \cref{sec:proof_ann}.
@@
     \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
-    \caption{Example of point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
+    \caption{Example of a point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
     The exact nearest neighbour is the point $p_1$, next point $p_i$ is placed 
@@
 \myparagraph{Wasserstein Distance}
-A persistence diagram is a multiset of points $(x, y)$ above the diagonal
+Let $\diag$ denote the line in the plane with the equation $y = x$; we call it the diagonal.
+A persistence diagram is a finite multiset of points $(x, y)$ with $x > y$
 $\Delta$ and all points of $\Delta$ with infinite cardinality.
+The points with $y > x$ are called \textit{off-diagonal} points,
+and points on $\Delta$ are called \textit{diagonal} points.
+The \emph{$q$-Wasserstein distance} is defined as
+\[
+    \wsdist{q}(X,Y) = \left[ \inf_{\eta:X\rightarrow Y} \sum_{x\in X} \|x-\eta(x)\|_\infty^q\right]^{1/q},
+\]
+where $\eta$ ranges over all bijections between $X$ and $Y$.
+Equivalent definition, which is better suited for actual computations,
+is as follows. Let $\pi\colon \RR^2 \to \Delta$ be defined
+as $(x,y) \mapsto (\frac{x+y}{2}, \frac{x+y}{2})$; we refer to $\pi$ as \textit{diagonal
+projection}. Let $\tX$ be the set $(X\setminus \Delta) \cup \pi(Y \setminus \Delta)$
+and $\tY$ be the set $(Y\setminus \Delta) \cup \pi(X \setminus \Delta)$; in plain words,
+to get $\tX$, we add to the off-diagonal points of $X$
+the diagonal projections of the off-diagonal points of $Y$,
+and perform the same operation to get $\tY$. Notice that $\tX$ and $\tY$
+are finite sets of the same cardinality. Let $G$ be a complete bipartite
+graph with vertices $\tX \coprod \tY$, where each vertex from $\tX$
+is connected with each vertex of from $\tY$. If edge $(x,y)$ of $G$
+has at least one of the vertices $x$ or $y$ not on the diagonal,
+then the edge is assigned weight $\ell_{\infty}(x, y)^q$, otherwise
+the edge is assigned weight 0. Wasserstein distance is defined as the $q$-th
+root of the minimal cost of a perfect matching on $G$; in other words,
+in order to find $\wsdist{q}$, one must solve an \textit{assignment problem}.

### 2019-04-24 paper/expensive-distances-main.tex
some Michaels' changes

@@
 yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
-outperforms naive computation of all pairwise distances or runs slighly slower, depending on the approximation
+outperforms naive computation of all pairwise distances or runs slightly slower, depending on the approximation
 quality.
@@
         contains 50 points.
-    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
+    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{\xi_d})$,
         where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
@@
 shapes; these diagrams form the \dtype{perturbed} dataset.
-Persistence diagrams were computed with \libraryname{DIPHA}
+Persistence diagrams were computed with the \libraryname{DIPHA}
 library \cite{bauer2014distributed}.
@@
 for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
-\grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
-any advantage of our algorithm disappears. However, the cardinality of the \grdy
+the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
+\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}} 
+any advantage of our algorithm disappears. A plausible explanation 
+is that for higher values of $q$, the doubling dimension becomes larger;
+this is confirmed by the increase in the size of the greedy spanner.
+\ignore{
+However, the cardinality of the \grdy
 spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
 then even the \grdy spanner is almost half of the complete graph.
-We can offer the following explanation\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}. For higher values of $q$,
-\label{}
 longer edges of an optimal matching play a more important role
@@
 of high doubling dimension, being close to a clique, and there is no hope
-of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
+of getting a small spanner. On the other hand, computation of $\btdist$ in practice
 requires significantly less time than computation of $\wsdist{q}$, and the case
@@
 For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
-computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, then we
+computations for $\eps = 0.1$; for a 1.5 approximation, we
 reduce the number of distances computed by a factor of 10.
@@
 350 points. Nevertheless, on this data set \bgrdy spanner
-provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.
+provides a substantial improvement in terms of real running time, and we expect this advantage
+to grow for diagrams of larger cardinality.
 

### 2019-04-24 paper/expensive-distances-main.tex
interim

@@
 polynomial time in $N$, and $N$ might well be larger than $n$, the number
-of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
+of diagrams considered \cite{cohen2007stability}. For the case of Reeb graphs, the situation is even
 worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
@@
 Another instance is a collection of high-resolution images
-endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).
+endowed with the Wasserstein (or Earth Movers) metric \cite{rubner2000earth}.
 
@@
 We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
-\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
+\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
@@
 yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
-outperforms naive computation of all pairwise distances or runs slightly slower, depending on the approximation
+runs faster than naive computation of all pairwise distances or runs slightly slower, depending on the approximation
 quality.
@@
 
-The \emph{greedy spanner}(\cite{althofer1993sparse}) is a simple algorithm to compute linear-sized spanners:
+The \emph{greedy spanner} \cite{althofer1993sparse} is a simple algorithm to compute linear-sized spanners:
 \begin{algorithmic}
@@
 
-The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
+The greedy spanner is guaranteed \cite{althofer1993sparse} to return a spanner of size $O(n)$
 (for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
-it was also shown to return the sparsest graph.  However, it clearly
+it was also shown to return the sparsest graph.  However, it
 has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
@@
 We performed two types of experiments: on points sampled from a low-dimensional Euclidean
-space and on persistence diagrams with Wasserstein distance.
+space and on persistence diagrams
+with Wasserstein distance\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}.
 Clearly, our model is not meaningful in a Euclidean space, where
@@
 \cref{sec:spanners_euclidean_addendum}.
-
-The realistic data sets that we used
-are the persistence diagrams
-computed from the McGill shape benchmark \cite{zhang2005mcgill}.
-More precisely, we generated two sets of diagrams,
-\dtype{original} and \dtype{perturbed}. The \dtype{original}
-diagrams were computed on the points of the 450 original
-point clouds from the shape dataset; we only used
-dimension 0, since the diagrams in this dimension are larger in size.
-We also added random noise to points of the original shapes,
-and computed persistence diagrams in dimension 0 of these modified
-shapes; these diagrams form the \dtype{perturbed} dataset.
-Persistence diagrams were computed with the \libraryname{DIPHA}
-library \cite{bauer2014distributed}.
-
-Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
-for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
-the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
-\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}} 
-any advantage of our algorithm disappears. A plausible explanation 
-is that for higher values of $q$, the doubling dimension becomes larger;
-this is confirmed by the increase in the size of the greedy spanner.
-\ignore{
-However, the cardinality of the \grdy
-spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
-then even the \grdy spanner is almost half of the complete graph.
-longer edges of an optimal matching play a more important role
-than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
-approaches bottleneck distance $\btdist$, which depends on the longest
-edge only. If all diagrams come from the same dataset,
-then there is a large number of pairs of diagrams whose optimal matchings
-have the longest edges of approximately the same length. This fact means that we are in the case
-of high doubling dimension, being close to a clique, and there is no hope
-of getting a small spanner. On the other hand, computation of $\btdist$ in practice
-requires significantly less time than computation of $\wsdist{q}$, and the case
-of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
-For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
-computations for $\eps = 0.1$; for a 1.5 approximation, we
-reduce the number of distances computed by a factor of 10.
-
-We also compare the time that is needed to compute the matrix of all pairwise distances
-naively with the time that is needed to approximate it with \bgrdy spanner.
-The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
-First of all, we should explain why the timings in the brute-force columns
-depend on $\eps$. Wasserstein distance can either be computed exactly,
-by using the Hungarian algorithm, or
-it can be computed with the user-given accuracy with the auction algorithm,
-as in \hera. Exact computation is very expensive \cite{kerber2017geometry}, and in our code for experiments
-with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
-To make the comparison absolutely fair, when we compute a \espanner,
-we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
-Let us emphasize that the diagrams we have are rather small, with average cardinality
-350 points. Nevertheless, on this data set \bgrdy spanner
-provides a substantial improvement in terms of real running time, and we expect this advantage
-to grow for diagrams of larger cardinality.
-
-In order to see the growth of the spanner size, we use the \dtype{perturbed}
-dataset, drawing from it random samples.
-We plot the sparseness of the spanners obtained by \grdy
-and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
-for different values of $\eps$, always using the $\wsdist{1}$ distance.
-As expected, the ratio of computed distances decreases as the number of diagrams
-grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
-the growth of the spanner size relative to the number of diagrams.
-For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
-linear growth) for larger values of $n$.
-
-\begin{figure}[!htbp]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
-        legend pos=outer north east]
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.5$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_ratio_mcgill_perturbed}
-\end{figure}
-
-\begin{figure}[!htbp]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.5$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_sparseness_mcgill_perturbed}
-\end{figure}
 
@@
 \end{table}
+
+The realistic data sets that we used
+are the persistence diagrams
+computed from the McGill shape benchmark \cite{zhang2005mcgill}.
+More precisely, we generated two sets of diagrams,
+\dtype{original} and \dtype{perturbed}. The \dtype{original}
+diagrams were computed on the points of the 450 original
+point clouds from the shape dataset; we only used
+dimension 0, since the diagrams in this dimension are larger in size.
+We also added random noise to points of the original shapes,
+and computed persistence diagrams in dimension 0 of these modified
+shapes; these diagrams form the \dtype{perturbed} dataset.
+Persistence diagrams were computed with the \libraryname{DIPHA}
+library \cite{bauer2014distributed}. 
+Wasserstein distance between diagrams can either be computed exactly,
+by using the Hungarian algorithm, or
+it can be computed with the user-given accuracy with the auction algorithm.
+Exact computation is very expensive \cite{kerber2017geometry}, 
+and in our experiments we always compute $\wsdist{q}$ using \hera with relative error 0.01.
+
+Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
+for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
+the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
+any advantage of our algorithm disappears. A plausible explanation
+is that for higher values of $q$, the doubling dimension becomes larger;
+this is confirmed by the increase in size of the greedy spanner.
+For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
+computations for $\eps = 0.1$; for a 1.5 approximation, we
+reduce the number of distances computed by a factor of 10.
+
+We also compare the time that is needed to compute the matrix of all pairwise distances
+naively with the time that is needed to approximate it with \bgrdy spanner.
+The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+We compared our approach for a fixed
+$\eps$ with the time to compute all pairwise-distance between the
+persistence diagrams using \hera, using the same approximation parameter
+$\eps$. Let us emphasize that the diagrams we have are rather small,
+with average cardinality
+350 points. Nevertheless, on this data set \bgrdy spanner
+provides a substantial improvement in terms of real running time,
+and we expect this advantage to grow for diagrams of larger cardinality.
+
+In order to see the growth of the spanner size, we use the \dtype{perturbed}
+dataset, drawing from it random samples.
+We plot the sparseness of the spanners obtained by \grdy
+and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
+for different values of $\eps$, always using the $\wsdist{1}$ distance.
+As expected, the ratio of computed distances decreases as the number of diagrams
+grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
+the growth of the spanner size relative to the number of diagrams.
+For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
+linear growth) for larger values of $n$.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_sparseness_mcgill_perturbed}
+\end{figure}
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
+        legend pos=outer north east]
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.5$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_ratio_mcgill_perturbed}
+\end{figure}
 
@@
 
+%As for different variants of the \brndm algorithm,
+%we note that their performance is almost the same, and the algorithm works
+%significantly better than \textsc{QuasiSorted} variants, but obviously
+%worse than the blind greedy variant. There is a consistent, though small,
+%difference between the variants that do not force lower bounds first and
+%the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
+
+\begin{figure}[!htbp]
+        \begin{tikzpicture}
+            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
+                legend pos=outer north east]
+
+                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind random, lower bound first}
+
+                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind greedy}
+
+                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Greedy}
+
+            \end{axis}
+        \end{tikzpicture}
+            \caption{Ratio \# edges / \# points for different variants of spanner 
+            algorithms. The plot is for normally distributed points
+            in dimension 2, $\eps = 0.1$.}
+            \label{fig:spanner_ratio}
+\end{figure}
+
 \begin{figure}[ht!]
@@
     \end{centering}
-\end{figure}
-
-%As for different variants of the \brndm algorithm,
-%we note that their performance is almost the same, and the algorithm works
-%significantly better than \textsc{QuasiSorted} variants, but obviously
-%worse than the blind greedy variant. There is a consistent, though small,
-%difference between the variants that do not force lower bounds first and
-%the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
-
-\begin{figure}[!htbp]
-        \begin{tikzpicture}
-            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
-                legend pos=outer north east]
-
-                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind random, lower bound first}
-
-                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind greedy}
-
-                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Greedy}
-
-            \end{axis}
-        \end{tikzpicture}
-            \caption{Ratio \# edges / \# points for different variants of spanner 
-            algorithms. The plot is for normally distributed points
-            in dimension 2, $\eps = 0.1$.}
-            \label{fig:spanner_ratio}
 \end{figure}

### 2019-04-24 paper/expensive-distances-main.tex
interim

@@
 yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
-runs faster than naive computation of all pairwise distances or runs slightly slower, depending on the approximation
-quality.
+runs either faster or slightly slower than naive computation of all pairwise distances, depending on the approximation quality.
 

### 2019-04-25 paper/draft_esa.tex
Page breaks
Changes in abstract and intro

@@
 We consider two standard operations on metric spaces,
-namely the construction of a $1+\eps$-spanner
+namely the construction of a $(1+\eps)$-spanner
 and the computation of an approximate nearest

### 2019-04-25 paper/expensive-distances-main.tex
Page breaks
Rescale images
Changes in abstract and intro

--- 2019-04-24 paper/expensive-distances-main.tex
@@
 Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
-a {$1+\eps$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
+a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
 performing $O(\log n)$ distance computations per query point.
-Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
+Krauthgamer and Lee \cite{krauthgamer2005black} investigated the \textit{black box model},
 and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
@@
 Our algorithms are based on the following simple idea: since distance computations are expensive
-and should be avoided, we try to obtain maximal information out of the distances that have been computed
+and should be avoided, we try to obtain maximal information out of the distances computed
 so far. 
@@
 \emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
+We demonstrate experimentally that on a collection
+of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
+computing all $\binom{n}{2}$ distances.
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
-We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
-spanners on the tested examples. Nevertheless, we leave the question open whether our construction
-yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
-runs faster than naive computation of all pairwise distances or runs slightly slower, depending on the approximation
-quality.
+We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
+spanners on the tested examples. 
+Nevertheless, we leave the question open whether our construction
+yields a spanner of asymptotically linear size.
 
@@
 We pose the question what heuristic could make more use of this feature.
+
+\newpage
 

### 2019-04-25 paper/expensive-distances-main.tex
changes

@@
 \centering
-\includegraphics[width=6cm]{\expDistFigPath  intro_example.eps}
+\includegraphics[width=3.5cm]{\expDistFigPath  intro_example.eps}
 \caption{The compute distances are shown as edges in a graph. Note that the exact distance
@@
 For approximate nearest neighbour, we devise a simple randomized incremental algorithm and show that
-the number of distance queries to find an approximate nearest neighbour is $O(\log n)$ in expectation.
+the number of distance queries needed is $O(\log n)$ in expectation.
 Our proof is based on the well-known observation that the nearest neighbour changes $O(\log n)$ times
@@
 The \textit{doubling dimension} of a doubling space is defined as $\log k$
-(since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
+(%since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
 we always use $\log$ to denote the logarithm with base 2).
-It is easy to see that a subspace of a space with doubling dimension $d$ 
-is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).
+A subspace of a space with doubling dimension $d$ 
+is always of doubling dimension $\leq 2d$.
 
@@
 The greedy spanner is guaranteed \cite{althofer1993sparse} to return a spanner of size $O(n)$
-(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
+(for constant doubling dimension and fixed $\eps>0$); in an experimental study \cite{farshi2009experimental}
 it was also shown to return the sparsest graph.  However, it
-has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
+has to compute all $\binom{n}{2}$ pairwise distances for sorting;
 this means that in our cost model, the greedy spanner has the worst possible
@@
 On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
-for convenience, we remind the definition of WSPD in \cref{sec:appendix_background}} only compute
+for convenience, we repeat the definition of WSPD in \cref{sec:appendix_background}}~\cite{cal-kos-wspd,hm-fast} only compute
 $O(n\log n+n \eps^{-d})$ distances 
@@
 a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
-initially set to $[0,\infty)$. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
+initially set to $0$ and $\infty$, respectively. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
 we pick one of them, compute its distance and update the lower and upper bounds of
@@
 hold by triangle inequality, and the same is true with $i$ and $j$ swapped.
-This yields $6$ lower bounds for $\dist(p_k,p_\ell)$, and $a_{k,\ell}$
+This yields six lower bounds for $\dist(p_k,p_\ell)$, and $a_{k,\ell}$
 is updated to the maximum of these six lower bounds and its current value.
@@
 that is, how to select the next distance to be computed. We propose two natural choices
-
+%
 \begin{description}
 \item[\brndm] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
-we pick one pair uniformly at random
+we pick one uniformly at random
 \item[\bgrdy] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
@@
 in that way hoping to gather most additional information about the metric space.
-Also, our conventions imply that in \bgrdy the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
+Also, our conventions imply that in \bgrdy the edges with $a_{i,j} = 0$ or $b_{i,j} = \infty$
 have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
@@
 space and on persistence diagrams
-with Wasserstein distance\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}.
-Clearly, our model is not meaningful in a Euclidean space, where
+with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
+Clearly, our model is not meaningful in Euclidean spaces because
 distance computations are cheap.
@@
         normal distribution in $\mathbb{R}^d$.
-    \item In the \textbf{clustered} test set we first sample cluster centres uniformly 
+    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
         at random from $[0,10000]^d$, and then we add normally distributed noise around
@@
 In addition to the standard Euclidean norm on $\RR^n$, we used also $\ell_1$ and $\ell_\infty$ norms.
-In all these experiments the algorithms that we tested compared in the same way,
+In all these experiments, the algorithms that we tested compared in the same way,
 so we only present results for the \textbf{uniform} point set in dimension 2.
@@
 WSPD spanners performed poorly in our experiments
-on non-clustered data, while the plots in the extensive experimental study \cite{farshi2009experimental}
+on non-clustered data, while the plots in a previous experimental study \cite{farshi2009experimental}
 show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
@@
 The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
-and our blind algorithms, which have at least cubic complexity in the number of points,
+and our blind algorithms, which have cubic complexity in the number of points,
 are infeasible for such $n$.
@@
 
-From these experiments we conclude that the algorithm with the highest chance
+From our experiments, we conclude that the algorithm with the highest chance
 of saving a significant amount of distance computations on real data is \bgrdy. 
@@
 
-The realistic data sets that we used
+The other data sets that we used
 are the persistence diagrams
@@
 point clouds from the shape dataset; we only used
-dimension 0, since the diagrams in this dimension are larger in size.
-We also added random noise to points of the original shapes,
+persitence in dimension 0, since the diagrams in this dimension are larger in size.
+For the \dtype{perturbed} dataset,
+we added random noise to points of the original shapes,
 and computed persistence diagrams in dimension 0 of these modified
-shapes; these diagrams form the \dtype{perturbed} dataset.
+shapes.
 Persistence diagrams were computed with the \libraryname{DIPHA}
 library \cite{bauer2014distributed}. 
-Wasserstein distance between diagrams can either be computed exactly,
-by using the Hungarian algorithm, or
-it can be computed with the user-given accuracy with the auction algorithm.
-Exact computation is very expensive \cite{kerber2017geometry}, 
-and in our experiments we always compute $\wsdist{q}$ using \hera with relative error 0.01.
+Wasserstein distance were computed approximately
+with the software library \hera using a relative error of $0.01$.
+We restricted to approximate computation because computing the
+Wasserstein distance exactly (using the Hungarian algorithm)
+has been reported to be very expensice~\cite{kerber2017geometry}.
 
@@
 for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
-the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
-any advantage of our algorithm disappears. A plausible explanation
+the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
+the advantage of our algorithm fades away. A plausible explanation
 is that for higher values of $q$, the doubling dimension becomes larger;
 this is confirmed by the increase in size of the greedy spanner.
-For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
-computations for $\eps = 0.1$; for a 1.5 approximation, we
+For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
+computations for $\eps = 0.1$; for a $1.5$-approximation, we
 reduce the number of distances computed by a factor of 10.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}[scale=0.85]
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_sparseness_mcgill_perturbed}
+\end{figure}
 
@@
     \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.5$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_sparseness_mcgill_perturbed}
-\end{figure}
-
-\begin{figure}[!htbp]
-    \begin{centering}
-\begin{tikzpicture}
+\begin{tikzpicture}[scale=0.8]
     \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
@@
 $n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
-find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. This notation will be fixed throughout this 
-section, and we shall also use the shorthand notation
+find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. We also use the shorthand notation
 \[
@@
 (a slight modification of the algorithm can also be applied if only a spanner is available).
-Our goal is to reduce the number of computed distances $\dist(p_i, q)$.
+Our goal is to reduce the number of computed distances $\dist(p_i, q)$ involving the query point $q$.
 
@@
 accordingly. At the end of the loop, $c$ is an approximate nearest
-neighbour. The pseudocode of the procedure follows.
+neighbour. The pseudocode follows.
 
@@
 
-The proof idea is the following. First, we notice
-that current candidate $c$ allows us to exclude points
-$p_i$ that are too far from $c$. This means that if $p_i$ is
-outside of the ball of radius $2\dist(c, q)$ centered at $c$, then the lower
-bound for $p_i$ will be larger than $\dist(c, q)$,
-and discarding $p_i$ does not cost us a distance computation.
-On the other hand, once we computed a distance to $c$
-from a point $p_j$, which is inside this ball but happens to be farther from $q$ than $c$,
-we can use it to discard points which are very close to $p_j$, 
-namely, inside a small ball of radius approximately $\eps \dist(c, q)$, because we
-are solving approximate nearest neighbour problem.
-Then we use a standard packing argument for doubling spaces, applied to the large and small balls, to show
-that the number of distance computations that we pay per candidate is constant.
-It remains only to prove that the number of different candidates is $O(\log n)$
-in expectation; the standard
-backwards analysis argument cannot be applied verbatim, because of a technicality,
-but it can be adapted.
-The full proof is given in \cref{sec:proof_ann}.
+The full proof is given in \cref{sec:proof_ann},
+we summarize the main ideas:
+first, we notice that points ``far away'' from the current candidate $c$ 
+can be disregarded in the algorithm. More precisely, if $p_i$
+is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
+the lower bound for $p_i$ will be so large that
+the distance to of $p_i$ to $q$ does not have to be queried.
+Second, whenever a distance of $q$ to some $p_j$ is computed,
+we can disregard all points ``very close'' to $p_j$.
+Precisely, points in a ball of radius $\eps\dist(p_j,q)$
+will not invalidate the current candidate 
+to be an approximate nearest neighbor and hence,
+no distance query for such points is needed.
+
+Now, traversing the points $P$ in random order,
+we call a point a \emph{minimum} if it is closer to $q$
+than all previous points. Note that minima do not necessarily
+correspond to the candidates in the algorithm because
+we look for an approximate neareast neighbor only.
+Standard backwards analysis~\cite{seidel-backwards}
+shows that the expected number of minima encountered is $O(\log n)$.
+The first two ideas together with a packing argument show
+that between two consecutive minima, 
+the algorithm queries only a constant number of distances,
+and the bound follows.
+
+The proof strategy fails for $\eps=0$ because the $\eps$-ball
+around $p_j$ as described above degenerates, and the packing argument fails.
+Indeed, as the example in \cref{fig:exact_bad_example} (in~\cref{sec:proof_ann}) 
+shows, there are point sets where the expected number
+of distance computations for exact nearest neighbour is linear.
+
+Finally, a fast $2$-approximation algorithm for $\dist$ leads to a straight-forward
+optimization: compute a $2$-approximation of $\dist(p_i,q)$ for all $1\leq i\leq n$
+and let $m$ denote the minimal approximate distance encountered. Then, we can discard all points
+whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.
 
@@
     \begin{centering}
-\begin{tikzpicture}
+\begin{tikzpicture}[scale=0.8]
     \begin{axis}[xlabel=\# Diagrams,ylabel=\# Computed distances / log(\# Diagrams),
@@
 is more costly than other types of primitive operations.
-Our theoretical and experimental results are under the usual assumption
+Our theoretical results assume
 that the metric space has a low doubling dimension.
@@
 (e.g. clusters of close-by diagrams) which should be favourable for our approach.
+Moreover, the size of the \bgrdy spanner can itself be considered as a piece
+of information about the dimensionality of the data set under consideration.
 
@@
 whether \bgrdy is worth trying. 
-The distances that \bgrdy computes would be computed anyway (and can be re-used),
-and the size of the \bgrdy spanner can itself be considered as a piece
-of information about the data set under consideration.
+%The distances that \bgrdy computes would be computed anyway (and can be re-used),
+%and
 
@@
 for blind spanners yields a linear spanner in expectation. 
-Our experiments are not conclusive enough in this respect to make this conjecture yet.
-However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
-that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
+%Our experiments are not conclusive enough in this respect to make this conjecture yet.
+It has been brought to our attention\footnote{Yusu Wang, personal communication}
+that the size of the blind spanner is upper bounded by the \emph{weight} of the WSPD
 which is the sum of the  cardinalities of all pairs in a WSPD.
@@
 
-We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
-of close-by points as in the second part of \cref{lem:bound_lemma}, and the packing argument fails.
-Indeed, as the example in \cref{fig:exact_bad_example} shows, there are point sets where the expected number
-of distance computations for exact nearest neighbour is linear.
-
-Finally, we remark that a fast $2$-approximation algorithm for $\dist$ would lead to a straight-forward
-optimization: compute a $2$-approximation of $\dist(p_i,q)$ for all $1\leq i\leq n$
-and let $m$ denote the minimal approximate distance encountered. Then, we can discard all points
-whose approximate distance is larger than $2m$, and run the above algorithm on the remaining points.
-
 \section{Additional Definitions}

### 2019-04-25 paper/expensive-distances-main.tex
explain perturbed

@@
 diagrams were computed on the points of the 450 original
-point clouds from the shape dataset; we only used
+point clouds from the shape data set; we only used
 persitence in dimension 0, since the diagrams in this dimension are larger in size.
-For the \dtype{perturbed} dataset,
+For the \dtype{perturbed} data set,
 we added random noise to points of the original shapes,
-and computed persistence diagrams in dimension 0 of these modified
+creating 10 modified versions of each of them;
+then we computed persistence diagrams in dimension 0 of these modified
 shapes.
@@
 
-Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
+Sparseness values of \bgrdy spanner on the whole \dtype{original} data set
 for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
@@
 In order to see the growth of the spanner size, we use the \dtype{perturbed}
-dataset, drawing from it random samples.
+data set, drawing from it random samples.
 We plot the sparseness of the spanners obtained by \grdy
@@
 diagrams, which was described in \cref{sec:experiment_spanners}.
-For a given number $n$, we randomly select $n$ diagrams from the dataset
+For a given number $n$, we randomly select $n$ diagrams from the data set
 and from the remaining diagrams we randomly choose 20 query points.

### 2019-04-25 paper/draft_esa.tex
changes
Minor changes

--- 2019-04-21 paper/draft_esa.tex
@@
 We consider two standard operations on metric spaces,
-namely the construction of a $1+\eps$-spanner
+namely the construction of a $(1+\eps)$-spanner
 and the computation of an approximate nearest
-neighbor for a given query point.
+neighbour for a given query point.
 In both cases, we partially explore the metric space
@@
 through extensive experimental evaluation.
-For approximate nearest neighbors, we prove that our
-strategy returns an approximate nearest neighbor
+For approximate nearest neighbours, we prove that our
+strategy returns an approximate nearest neighbour
 after a logarithmic number of distance queries.
 \end{abstract}
+
+\newpage
 

### 2019-04-25 paper/expensive-distances-main.tex
Minor changes

@@
 \centering
-\includegraphics[width=3.5cm]{\expDistFigPath  intro_example.eps}
+\includegraphics[width=4cm]{\expDistFigPath  intro_example.eps}
 \caption{The compute distances are shown as edges in a graph. Note that the exact distance
@@
     \begin{centering}
-\begin{tikzpicture}[scale=0.85]
+\begin{tikzpicture}[scale=0.89]
     \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
@@
     \begin{centering}
-\begin{tikzpicture}[scale=0.8]
+\begin{tikzpicture}[scale=0.83]
     \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
@@
 and $\tY$ be the set $(Y\setminus \Delta) \cup \pi(X \setminus \Delta)$; in plain words,
-to get $\tX$, we add to the off-diagonal points of $X$
+to obtain $\tX$, we add to the off-diagonal points of $X$
 the diagonal projections of the off-diagonal points of $Y$,
-and perform the same operation to get $\tY$. Notice that $\tX$ and $\tY$
+and perform the same operation to obtain $\tY$. Notice that $\tX$ and $\tY$
 are finite sets of the same cardinality. Let $G$ be a complete bipartite
 graph with vertices $\tX \coprod \tY$, where each vertex from $\tX$
-is connected with each vertex of from $\tY$. If edge $(x,y)$ of $G$
+is connected with each vertex from $\tY$. If edge $(x,y)$ of $G$
 has at least one of the vertices $x$ or $y$ not on the diagonal,

### 2019-08-15 paper/sosa_acm.tex
Add SOSA submission, reformat to ACM

--- 2019-04-25 paper/expensive-distances-main.tex
@@
+\documentclass[natbib=false]{acmart}
+
+\usepackage{amsmath}%
+\usepackage{amssymb}%
+\usepackage{amsfonts}
+\usepackage{amsthm}
+\usepackage{enumerate}
+\usepackage{booktabs}
+\usepackage{nccmath}
+\usepackage{diagbox}
+\usepackage{url}
+\usepackage{xspace}
+\usepackage{mathbbol}
+\usepackage{comment}
+\usepackage{braket}
+
+% graphics
+\usepackage{tikz}
+\usepackage{pgf}
+\usepackage{graphicx}
+\usepackage{pgfplots}
+\pgfplotsset{compat=1.14}
+\usepackage{pgfplotstable}
+\usetikzlibrary{patterns}
+\usetikzlibrary{external}
+\usepackage{epsfig}
+\usepackage{wrapfig}
+\usepackage{color}
+\usepackage{mwe}
+\usepackage{floatflt}
+
+\usetikzlibrary{arrows}
+
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+
+\newcommand{\RR}{\mathbb{R}}
+
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\renewcommand{\epsilon}{\varepsilon}
+\renewcommand{\phi}{\varphi}
+\newcommand{\eps}{\varepsilon}
+
+\newcommand{\ignore}[1]{}
+\newcommand{\yaxis}{$y$\nobreakdash-axis\xspace}
+\newcommand{\xaxis}{$x$\nobreakdash-axis\xspace}
+
+\newcommand{\algname}[1]{{\textsc{#1}}\xspace}
+
+\newcommand{\dtype}[1]{{\textit{\small #1}}}
+
+\newcommand{\libraryname}[1]{{\textsc{#1}}\xspace}
+
+\newcommand{\hera}{\libraryname{Hera}}
+\newcommand{\Hera}{\libraryname{Hera}}
+\newcommand{\dionisus}{\libraryname{Dionysus}}
+\newcommand{\Dionisus}{\libraryname{Dionysus}}
+
+\pgfplotsset
+{
+    tick label style={font=\footnotesize},
+    legend style    ={font=\footnotesize},
+    label style     ={font=\footnotesize},
+    tickwidth       =2pt,
+    ylabel near ticks,
+    xlabel near ticks,
+    legend style     ={draw=none},
+    legend cell align=right,
+    legend plot pos  =right,
+    /pgfplots/narrow area legend/.style=
+    {
+        legend image code/.code={ \draw[#1] (0cm,-0.06cm) rectangle (0.6cm,0.06cm); }
+    },
+    /pgfplots/short line legend/.style=
+    {
+        legend image code/.code={ \draw[#1] (0cm,0cm) rectangle (0.6cm,0cm); }
+    },
+}
+
+\pgfplotscreateplotcyclelist{filled-color}
+{
+    every mark/.append style={solid,fill=blue!80!white},    mark=*          \\%
+    every mark/.append style={solid,fill=red!80!white},     mark=square*    \\%
+    every mark/.append style={solid,fill=orange!80!white},  mark=pentagon*  \\%
+    every mark/.append style={solid,fill=green},            mark=diamond*   \\%
+    every mark/.append style={solid,fill=black},            mark=triangle*  \\%
+}
+
+\newcommand{\addaveraged}[2][]
+           {\addplot+[#1,error bars/.cd, y dir=both,y explicit] table[x index=0, y index=1, y error index=2] {#2};}
+
+\pgfplotsset{
+    geommark/.style={black,mark options={solid,fill=red!80!white},mark=square*},
+    nongeommark/.style={black,mark options={solid,fill=blue!80!white},mark=*},
+    dionmark/.style={black,mark options={solid,fill=orange!80!white},mark=pentagon*},
+    greentrianglemark/.style={black,mark options={solid,fill=green!80!white},mark=triangle*},
+    browndiamondmark/.style={black,mark options={solid,fill=brown!80!white},mark=diamond*}
+}
+
+% all distances
+% interleaving
+\newcommand{\intdist}{{D_I}}
+% bottleneck
+\newcommand{\btdist}{{W_\infty}}
+% Wasserstein of order q (parameter)
+\newcommand{\wsdist}[1]{{W_{#1}}}
+
+%% \category{dummy}
+
+% \category{dummy}
+
+%\Copyright{Michael Kerber and Arnur Nigmetov}
+%\Copyright{Anonymous Authors}
+% END uncomment for LIPICS
+
+\def \expDistFigPath {data/expensive_distances/pics/}
+\def \expDistDataPath {data/expensive_distances/}
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+
+\renewcommand{\S}{\mathbb{S}}
+\newcommand{\dappr}{\tilde{d}}
+\newcommand{\orient}{O}
+\newcommand{\cpro}{\Pi}
+
+\newcommand{\metricspace}{\mathcal{M}}
+\newcommand{\approxmetric}{\gamma}
+\newcommand{\parent}{\mathrm{par}}
+\newcommand{\point}{p}
+\newcommand{\level}{\ell}
+\newcommand{\diam}{\mathrm{diam}}
+\newcommand{\diag}{{\Delta}}
+\newcommand{\pointset}{P}
+\newcommand{\distspace}{\mathcal{M}}
+\newcommand{\dist}{\delta}
+\newcommand{\adist}{\gamma}
+\newcommand{\complexity}{C_{\dist}}
+\newcommand{\doublingdimension}{\Delta}
+
+\newcommand{\espanner}{{$(1+\eps)$-spanner}}
+
+\newcommand{\grdy}{\algname{Greedy}}
+\newcommand{\bgrdy}{\algname{BlindGreedy}}
+\newcommand{\brndm}{\algname{BlindRandom}}
+\newcommand{\bqsgrdy}{\algname{BlindQuasiSortedGreedy}}
+\newcommand{\bqsshaker}{\algname{BlindQuasiSortedShaker}}
+
+\newcommand{\tX}{{\tilde{X}}}
+\newcommand{\tY}{{\tilde{Y}}}
+
+\newcommand{\myparagraph}[1]{\textbf{#1.}}
+
+\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
+\def\michael#1{\textcolor{red}{\textsc{Michael says: }{\marrow\sf #1}}}
+
+\def \expDistFigPath {pics/}
+\def \expDistDataPath {./}
+
+\usepackage[style=ACM-Reference-Format,backend=bibtex,sorting=none]{biblatex}
+\addbibresource{bib.bib}
+
+\title{Metric Spaces with Expensive Distances}
+
+\author{Michael Kerber}
+\email{kerber@tugraz.at}
+\affiliation{%
+  \institution{Graz University of Technology}
+  \streetaddress{Kopernikusgasse 24}
+  \city{Graz}
+  \postcode{8010}
+  \country{Austria}
+}
+
+\author{Arnur Nigmetov}
+\email{nigmetov@tugraz.at}
+\affiliation{%
+  \institution{Graz University of Technology}
+  \streetaddress{Kopernikusgasse 24}
+  \city{Graz}
+  \postcode{8010}
+  \country{Austria}
+}
+
+\begin{abstract}
+In algorithms for finite metric spaces, it is common
+to assume that the distance between two points
+can be computed in constant time, and complexity
+bounds are expressed only in terms of the number
+of points of the metric space.
+We introduce a different model, where we assume
+that the computation of a single distance is
+an expensive operation and consequently, the goal is
+to minimize the number of such distance queries.
+This model is motivated by metric spaces
+that appear in the context of topological data analysis.
+
+We consider two standard operations on metric spaces,
+namely the construction of a $(1+\epsilon)$-spanner
+and the computation of an approximate nearest
+neighbor for a given query point.
+In both cases, we partially explore the metric space
+through distance queries and infer lower and upper bounds
+for yet unexplored distances through triangle inequality.
+For spanners, we evaluate several exploration strategies
+through extensive experimental evaluation.
+For approximate nearest neighbors, we prove that our
+strategy returns an approximate nearest neighbor
+after a logarithmic number of distance queries.
+\end{abstract}
+
+\begin{document}
+
+\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+\ccsdesc[500]{Theory of computation~Random search heuristics}
+
+\copyrightyear{2019}
+\acmYear{2020}
+\acmConference[SOSA '20]{SIAM Symposium on Simplicity in Algorithms}{January 06--07, 2020}{Salt Lake City, UT, USA}
+\acmBooktitle{SIAM Symposium on Simplicity in Algorithms (SOSA '20), January 06--07, 2020, Salt Lake City, UT, USA}
+%\acmPrice{15.00}
+%\acmDOI{10.1145/3295500.3356188}
+%\acmISBN{978-1-4503-6229-0/19/11}
+
+\maketitle
+
 \section{Introduction}
@@
 \cite{di2016edit}),
-not even an constant-factor approximation algorithm is known that runs
+not even a constant-factor approximation algorithm is known that runs
 in polynomial time in the size of the graphs.
@@
 This information consists of lower and upper bounds for unknown distances, obtained from known distances
-by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
+by triangle inequality (see Figure \ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
 operations whenever a new distance has been computed, turning the method useless in the standard computational model.
@@
 \myparagraph{Spanners and known constructions}
-%
+
 Let $(\metricspace, \dist)$ be a finite metric space with $n$ points. 
@@
 
-On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
-for convenience, we repeat the definition of WSPD in \cref{sec:appendix_background}}~\cite{cal-kos-wspd,hm-fast} only compute
+\myparagraph{Well-Separated Pair Decomposition}
+An alternative algorithm for computing a spanner is based  on well-separated pair decomposition (WSPD).
+Let us recall the definition.  Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
+if 
+\[
+\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
+\]
+A well-separated pair decomposition is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
+of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
+The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
+Har-Peled and Mendel \cite{hm-fast} 
+introduced the notion of net-trees and
+generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
+\begin{enumerate}
+    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
+    \item If $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
+are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
+    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
+The algorithm uses the net-tree structure.
+\end{enumerate}
+The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
+et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
+than net-trees. We mention in passing that cover trees can also be used for building
+a spanner (this can be proven with the same methods), and we use cover trees for
+building WSPD spanners in one of our implementations.
+
+On the other hand, spanner constructions based on WSPD~\cite{cal-kos-wspd,hm-fast} only compute
 $O(n\log n+n \eps^{-d})$ distances 
@@
 space and on persistence diagrams
-with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
+with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See Section \ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
 Clearly, our model is not meaningful in Euclidean spaces because
@@
 
-\cref{fig:spanner_sparseness}
+Figure \ref{fig:spanner_sparseness}
 shows the number of edges of the spanner for various variants of 
@@
 difference between the variants that do not force lower bounds first and
-the other two variants of the \brndm.  (see \cref{fig:blind_rbr_variants}). 
+the other two variants of the \brndm.  (see Figure \ref{fig:blind_rbr_variants}). 
 }
@@
 dense spanners. More data from experiments in $\RR^n$ can be found in
-\cref{sec:spanners_euclidean_addendum}.
+Section \ref{sec:spanners_euclidean_addendum}.
 
@@
 point clouds from the shape data set; we only used
-persitence in dimension 0, since the diagrams in this dimension are larger in size.
+persistence in dimension 0, since the diagrams in this dimension are larger in size.
 For the \dtype{perturbed} data set,
@@
 Sparseness values of \bgrdy spanner on the whole \dtype{original} data set
-for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
+for different values of $\eps$ and $q$ are presented in Table \ref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
 the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
@@
 naively with the time that is needed to approximate it with \bgrdy spanner.
-The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+The timings are given in Table \ref{tbl:mcgill_original_blind_greedy_spanner_timing}.
 We compared our approach for a fixed
@@
 with average cardinality
-350 points. Nevertheless, on this data set \bgrdy spanner
+of 350 points. Nevertheless, on this data set \bgrdy spanner
 provides a substantial improvement in terms of real running time,
@@
 We plot the sparseness of the spanners obtained by \grdy
-and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
+and \bgrdy algorithms in Figure \ref{fig:spanner_sparseness_mcgill_perturbed}
 for different values of $\eps$, always using the $\wsdist{1}$ distance.
 As expected, the ratio of computed distances decreases as the number of diagrams
-grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
+grows. In Figure \ref{fig:spanner_ratio_mcgill_perturbed} we show
 the growth of the spanner size relative to the number of diagrams.
@@
 
-The full proof is given in \cref{sec:proof_ann},
+The full proof is given in the next section,
 we summarize the main ideas:
@@
 around $p_j$ as described above degenerates, and the packing argument fails.
-Indeed, as the example in \cref{fig:exact_bad_example} (in~\cref{sec:proof_ann}) 
+Indeed, as the example in Figure \ref{fig:exact_bad_example} (in Section~\ref{sec:proof_ann}) 
 shows, there are point sets where the expected number
@@
 
-\section{Experiments on approximate nearest neighbours}
-\label{sec:experiment_ann}
-
-In order to experimentally evaluate the performance of our algorithm,
-we generate random point sets and random query points, and for each query point
-run the algorithm 10 times. The average number of distances to the query point
-that were actually computed is the measure that we are interested in.
-We average the results over 10 different instances of the point
-set and query point in order to see the trend clearer; thus
-each point on the plots in this section is the result of averaging of 100 runs
-of the code (10 instances, 10 random permutations per instance).
-
-\begin{figure}[ht]
-    \begin{minipage}{0.45\textwidth}
-        \includegraphics[width=0.9\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
-        \caption{Ratio $\mbox{computed distances} / \log(n)$ for ANN algorithm. Data is for uniformly distributed
-        points.}
-        \label{fig:ann_const_ratio}
-    \end{minipage}
-    \begin{minipage}{0.45\textwidth}
-        \includegraphics[width=0.9\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
-        \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
-        \label{fig:ann_dimension_dependence}
-    \end{minipage}
-\end{figure}
-
-We used the following methods of generating random points:
-\begin{enumerate}
-    \item Uniform. Points are sampled uniformly at random from the unit cube in $\RR^d$.
-    \item Normal. Points are sampled from the normal distribution.
-\end{enumerate}
-Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
-and from the normal distribution centered at the origin with scale 100,
-thus we get query points that are "inside" the point set and also "outside".
-We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
-the maximal number of points is $30,000$.
-
-In order to empirically verify the upper bound $O(\log n)$,
-we plot the number of computed distances divided
-by the logarithm of the number of points in \cref{fig:ann_const_ratio} (for $d = 2$).
-We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
-the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
-really computes only a very small number of distances to the query point.
-As expected, in high dimensions the algorithm does not perform as well.
-In \cref{fig:ann_dimension_dependence} we plot the average number of computed distances
-for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
-the sublinearity of the growth becomes clear only when the number of points is relatively large,
-approaching 30000.
-
-Similar experiments were performed on \dtype{perturbed} data set of persistence
-diagrams, which was described in \cref{sec:experiment_spanners}.
-For a given number $n$, we randomly select $n$ diagrams from the data set
-and from the remaining diagrams we randomly choose 20 query points.
-In \cref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
-distances to $\log(n)$ for $\wsdist{1}$ and $\wsdist{2}$ metrics.
-The average ratio appears to fluctuate around 7 for $\wsdist{1}$ and around 8 for $\wsdist{2}$,
-which even makes the algorithm
-reasonable for practical application, if pairwise distances are already available.
-
-\begin{figure}[!htbp]
-    \begin{centering}
-\begin{tikzpicture}[scale=0.8]
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Computed distances / log(\# Diagrams),
-        legend pos=outer north east]
-    \addplot+ table [x=n_points, y=avg_ratio_dim_0_q_1_eps_001, col sep=semicolon] {\expDistDataPath ann_mcgill_results.csv};
-        \addlegendentry{Average ratio, $\wsdist{1}$}
-
-    \addplot+ table [x=n_points, y=avg_ratio_dim_0_q_2_eps_001, col sep=semicolon] {\expDistDataPath ann_mcgill_results.csv};
-        \addlegendentry{Average ratio, $\wsdist{2}$}
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{ Ratio $\mbox{\# computed distances} / n$ for ANN algorithm. The plot is for \dtype{perturbed} persistence diagrams,
-    $\eps = 0.01$ with $\wsdist{1}$ and $\wsdist{2}$ metrics.}
-    \label{fig:ann_const_ratio_mcgill}
-\end{figure}
-
-\section{Conclusion and future work}
-\label{sec:conclusion}
-%
-We have introduced a new cost model for the analysis of algorithms
-for metric spaces that fits the situation that computing an individual distance
-is more costly than other types of primitive operations.
-Our theoretical results assume
-that the metric space has a low doubling dimension.
-However, in our motivating example of collections of persistence diagrams
-or Reeb graphs, this assumption does not hold. For instance,
-the space of persistence diagrams has an infinite doubling dimension.
-Nevertheless, realistic data sets are usually not just a random sample
-in that infinite-dimensional space, but have structures
-(e.g. clusters of close-by diagrams) which should be favourable for our approach.
-Moreover, the size of the \bgrdy spanner can itself be considered as a piece
-of information about the dimensionality of the data set under consideration.
-
-In particular, our experiments on real persistence diagrams
-show that the blind spanner can significantly accelerate
-computation of all pairwise distances. We must admit that,
-while in our theoretical model we completely ignore dependency on the number of points $n$,
-the $\Omega(n^3)$ complexity of our algorithms makes it practically
-non-competetive, when $n$ becomes large.
-On the other hand, this overhead is relatively easy to estimate
-in advance, just by knowing $n$, and make a decision,
-whether \bgrdy is worth trying. 
-%The distances that \bgrdy computes would be computed anyway (and can be re-used),
-%and
-
-On the theoretical side, the obvious next question is whether our strategy
-for blind spanners yields a linear spanner in expectation. 
-%Our experiments are not conclusive enough in this respect to make this conjecture yet.
-It has been brought to our attention\footnote{Yusu Wang, personal communication}
-that the size of the blind spanner is upper bounded by the \emph{weight} of the WSPD
-which is the sum of the  cardinalities of all pairs in a WSPD.
-The weight of a WSPD can be quadratic, but preliminary
-experimental evaluation on worst-case examples do not show such a quadratic
-behavior. Therefore, we postpone the theoretical analysis of the spanner construction
-to an extended version of this article.
-
-The existence of a $2$-approximation algorithm did not help us to significantly
-reduce the number of exact distance computations, although it seems obvious
-that knowing the all approximate distances is useful.
-We pose the question what heuristic could make more use of this feature.
-
-\newpage
-
-\appendix
-
-\section{Additional Experimental Results on Spanners}
-\label{sec:spanners_euclidean_addendum}
-In this section we include plots that give more information
-about the behaviour of \bgrdy spanner in Euclidean space.
-
-\ignore{
-\begin{figure}[ht!]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first, lower bound first}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Comparison of the four variants of \brndm algorithm.}
-    \label{fig:blind_rbr_variants}
-\end{figure}
-}
-
-%As for different variants of the \brndm algorithm,
-%we note that their performance is almost the same, and the algorithm works
-%significantly better than \textsc{QuasiSorted} variants, but obviously
-%worse than the blind greedy variant. There is a consistent, though small,
-%difference between the variants that do not force lower bounds first and
-%the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
-
-\begin{figure}[!htbp]
-        \begin{tikzpicture}
-            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
-                legend pos=outer north east]
-
-                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind random, lower bound first}
-
-                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Blind greedy}
-
-                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
-                \addlegendentry{Greedy}
-
-            \end{axis}
-        \end{tikzpicture}
-            \caption{Ratio \# edges / \# points for different variants of spanner 
-            algorithms. The plot is for normally distributed points
-            in dimension 2, $\eps = 0.1$.}
-            \label{fig:spanner_ratio}
-\end{figure}
-
-\begin{figure}[ht!]
-    \begin{centering}
-    \includegraphics[width=0.8\textwidth]{\expDistFigPath blind-greedy-dims.png}
-    \caption{Results of \bgrdy spanner for different dimensions.}
-    \label{fig:blind_greedy_dimensions}
-    \end{centering}
-\end{figure}
-
-\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
-The ideal behavior is demonstrated by the non-blind greedy spanner,
-for which this ratio stays practically constant, confirming the linear growth.
-None of the blind algorithms seems to have this property, but among them
-the \bgrdy spanner is the best one. If we assume that the number of edges
-is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
-linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
-the estimated exponents $\alpha$ for \bgrdy and standard
-greedy algorithms. Note that even for the greedy algorithm these estimated
-exponents can be significantly larger than 1,
-which is explained by the fact that the number of points
-on which we computed spanners is not large enough to clearly see
-the linear dependence.
-
-\begin{table}[b]
-\begin{tabular}{|l|l|l|}
-\hline
-dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
-2         &         1.08                &  1.12                 \\ \hline
-3         &         1.24                &  1.41                \\ \hline
-4         &         1.42                &  1.77                 \\ \hline
-\end{tabular}
-\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
-on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
-\label{tbl:regr_coeff_spanner}
-\end{table}
-
-The plot in \cref{fig:blind_greedy_dimensions}
-compares performance of \bgrdy in different dimensions.
-We see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
-edges for $700$ points, which clearly shows some degrading for higher dimensions.
-
-The plot in \cref{fig:spanner_eps_dependence}
-compares the \bgrdy and \textsc{Greedy} algorithms
-on \textbf{uniform} point sets for different choices of $\eps$. Dependence on $\eps$
-is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
-that the ratio of the number of edges decreases for
-smaller values of $\eps$: 
-for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
-while for $\eps = 1/32$ the ratio is 2.6
-
-\begin{figure}[ht]
-        \begin{tikzpicture}
-            \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
-                legend pos=north east]
-
-            \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-            \addlegendentry{Greedy, dim = 2}
-
-            \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
-            \addlegendentry{Blind greedy, dim = 2}
-
-        \end{axis}
-        \end{tikzpicture}
-        \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
-        \label{fig:spanner_eps_dependence}
-\end{figure}
-
-\section{Proof of \cref{thm:ann_bound}}
+\section{Proof of Theorem \ref{thm:ann_bound}}
 \label{sec:proof_ann}
@@
 
-\cref{lem:bound_lemma}
+Lemma \ref{lem:bound_lemma}
 can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
@@
 %This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the
-%brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
+%brute-force algorithm, thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
 The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
@@
 
-In \cref{fig:ann_illustration} we illustrate this. First, $p_1$ is
+In Figure \ref{fig:ann_illustration} we illustrate this. First, $p_1$ is
 chosen as the current candidate, and we must compute $\dist(p_2, q)$.
@@
 \begin{proof}
-By the first part of \cref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
+By the first part of Lemma \ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
 whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
 around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
-is computed, the second part of \cref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
+is computed, the second part of Lemma \ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
 Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
-have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
-See \cref{fig:ann_illustration}.
+have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma \ref{lem:packing_lemma}.
+See Figure \ref{fig:ann_illustration}.
 \end{proof}
@@
 the number of computed distances is a constant (since $v_k=v_\ell$). 
-This means that to prove \cref{thm:ann_bound}, it would suffice to show
+This means that to prove Theorem \ref{thm:ann_bound}, it would suffice to show
 that the candidate changes only a logarithmic number of times in expectation.
-Let us consider \cref{fig:ann_illustration} again. Notice that the point $p_5$, which
+Let us consider Figure \ref{fig:ann_illustration} again. Notice that the point $p_5$, which
 is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
@@
 candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
-thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis,
+thus Theorem \ref{thm:ann_bound} does not immediately follow from standard backwards analysis,
 and we need to modify it slightly.
 
-\begin{proof} (of \cref{thm:ann_bound})
+\begin{proof} (of Theorem \ref{thm:ann_bound})
 %
@@
 some $i\leq j$, and in the sequence $r_1,\ldots,r_{\ell-1}$, $r_k$
-is minimal by construction. Using \cref{lem:sequence_lemma},
+is minimal by construction. Using Lemma \ref{lem:sequence_lemma},
 the number of distance computations among the points
@@
 
-\section{Additional Definitions}
-\myparagraph{Well-Separated Pair Decomposition}
+\section{Experiments on approximate nearest neighbours}
+\label{sec:experiment_ann}
+
+In order to experimentally evaluate the performance of our algorithm,
+we generate random point sets and random query points, and for each query point
+run the algorithm 10 times. The average number of distances to the query point
+that were actually computed is the measure that we are interested in.
+We average the results over 10 different instances of the point
+set and query point in order to see the trend clearer; thus
+each point on the plots in this section is the result of averaging of 100 runs
+of the code (10 instances, 10 random permutations per instance).
+
+\begin{figure}[ht]
+    \begin{minipage}{0.45\textwidth}
+        \includegraphics[width=0.9\textwidth]{\expDistFigPath log_dependency_constant_ratio.png}
+        \caption{Ratio $\mbox{computed distances} / \log(n)$ for ANN algorithm. Data is for uniformly distributed
+        points.}
+        \label{fig:ann_const_ratio}
+    \end{minipage}
+    \begin{minipage}{0.45\textwidth}
+        \includegraphics[width=0.9\textwidth]{\expDistFigPath ann_computed_distances_dimension_dep.png}
+        \caption{Number of computed distances for different dimensions. Points are chosen uniformly, $\eps = 0.01$.}
+        \label{fig:ann_dimension_dependence}
+    \end{minipage}
+\end{figure}
+
+We used the following methods of generating random points:
+\begin{enumerate}
+    \item Uniform. Points are sampled uniformly at random from the unit cube in $\RR^d$.
+    \item Normal. Points are sampled from the normal distribution.
+\end{enumerate}
+Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
+and from the normal distribution centered at the origin with scale 100,
+thus we get query points that are "inside" the point set and also "outside".
+We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
+the maximal number of points is $30,000$.
+
+In order to empirically verify the upper bound $O(\log n)$,
+we plot the number of computed distances divided
+by the logarithm of the number of points in Figure \ref{fig:ann_const_ratio} (for $d = 2$).
+We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
+the theoretical upper bound, but also shows that the algorithm in the low-dimensional case
+really computes only a very small number of distances to the query point.
+As expected, in high dimensions the algorithm does not perform as well.
+In Figure \ref{fig:ann_dimension_dependence} we plot the average number of computed distances
+for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
+the sublinearity of the growth becomes clear only when the number of points is relatively large,
+approaching 30000.
+
+Similar experiments were performed on \dtype{perturbed} data set of persistence
+diagrams, which was described in Section \ref{sec:experiment_spanners}.
+For a given number $n$, we randomly select $n$ diagrams from the data set
+and from the remaining diagrams we randomly choose 20 query points.
+In Figure \ref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
+distances to $\log(n)$ for $\wsdist{1}$ and $\wsdist{2}$ metrics.
+The average ratio appears to fluctuate around 7 for $\wsdist{1}$ and around 8 for $\wsdist{2}$,
+which even makes the algorithm
+reasonable for practical application, if pairwise distances are already available.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}[scale=0.8]
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Computed distances / log(\# Diagrams),
+        legend pos=outer north east]
+    \addplot+ table [x=n_points, y=avg_ratio_dim_0_q_1_eps_001, col sep=semicolon] {\expDistDataPath ann_mcgill_results.csv};
+        \addlegendentry{Average ratio, $\wsdist{1}$}
+
+    \addplot+ table [x=n_points, y=avg_ratio_dim_0_q_2_eps_001, col sep=semicolon] {\expDistDataPath ann_mcgill_results.csv};
+        \addlegendentry{Average ratio, $\wsdist{2}$}
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{ Ratio $\mbox{\# computed distances} / n$ for ANN algorithm. The plot is for \dtype{perturbed} persistence diagrams,
+    $\eps = 0.01$ with $\wsdist{1}$ and $\wsdist{2}$ metrics.}
+    \label{fig:ann_const_ratio_mcgill}
+\end{figure}
+
+\section{Conclusion and future work}
+\label{sec:conclusion}
+%
+We have introduced a new cost model for the analysis of algorithms
+for metric spaces that fits the situation that computing an individual distance
+is more costly than other types of primitive operations.
+Our theoretical results assume
+that the metric space has a low doubling dimension.
+However, in our motivating example of collections of persistence diagrams
+or Reeb graphs, this assumption does not hold. For instance,
+the space of persistence diagrams has an infinite doubling dimension.
+Nevertheless, realistic data sets are usually not just a random sample
+in that infinite-dimensional space, but have structures
+(e.g. clusters of close-by diagrams) which should be favourable for our approach.
+Moreover, the size of the \bgrdy spanner can itself be considered as a piece
+of information about the dimensionality of the data set under consideration.
+
+In particular, our experiments on real persistence diagrams
+show that the blind spanner can significantly accelerate
+computation of all pairwise distances. We must admit that,
+while in our theoretical model we completely ignore dependency on the number of points $n$,
+the $\Omega(n^3)$ complexity of our algorithms makes it practically
+non-competetive, when $n$ becomes large.
+On the other hand, this overhead is relatively easy to estimate
+in advance, just by knowing $n$, and make a decision,
+whether \bgrdy is worth trying. 
+%The distances that \bgrdy computes would be computed anyway (and can be re-used),
+%and
+
+On the theoretical side, the obvious next question is whether our strategy
+for blind spanners yields a linear spanner in expectation. 
+%Our experiments are not conclusive enough in this respect to make this conjecture yet.
+It has been brought to our attention\footnote{Yusu Wang, personal communication}
+that the size of the blind spanner is upper bounded by the \emph{weight} of the WSPD
+which is the sum of the  cardinalities of all pairs in a WSPD.
+The weight of a WSPD can be quadratic, but preliminary
+experimental evaluation on worst-case examples do not show such a quadratic
+behavior. Therefore, we postpone the theoretical analysis of the spanner construction
+to an extended version of this article.
+
+The existence of a $2$-approximation algorithm did not help us to significantly
+reduce the number of exact distance computations, although it seems obvious
+that knowing the all approximate distances is useful.
+We pose the question what heuristic could make more use of this feature.
+
+\newpage
+
+\appendix
+
+\section{Additional Experimental Results on Spanners}
+\label{sec:spanners_euclidean_addendum}
+In this section we include plots that give more information
+about the behaviour of \bgrdy spanner in Euclidean space.
+
+\ignore{
+\begin{figure}[ht!]
+    \begin{centering}
+\begin{tikzpicture}
+    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, connect first}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, lower bound first}
+
+    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
+    \addlegendentry{Blind random, connect first, lower bound first}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Comparison of the four variants of \brndm algorithm.}
+    \label{fig:blind_rbr_variants}
+\end{figure}
+}
+
+%As for different variants of the \brndm algorithm,
+%we note that their performance is almost the same, and the algorithm works
+%significantly better than \textsc{QuasiSorted} variants, but obviously
+%worse than the blind greedy variant. There is a consistent, though small,
+%difference between the variants that do not force lower bounds first and
+%the other two variants of the \brndm. \ignore{ (see Figure \ref{fig:blind_rbr_variants}). }
+
+\begin{figure}[!htbp]
+        \begin{tikzpicture}
+            \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
+                legend pos=outer north east]
+
+                \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind random, lower bound first}
+
+                \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Blind greedy}
+
+                \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
+                \addlegendentry{Greedy}
+
+            \end{axis}
+        \end{tikzpicture}
+            \caption{Ratio \# edges / \# points for different variants of spanner 
+            algorithms. The plot is for normally distributed points
+            in dimension 2, $\eps = 0.1$.}
+            \label{fig:spanner_ratio}
+\end{figure}
+
+\begin{figure}[ht!]
+    \begin{centering}
+    \includegraphics[width=0.5\textwidth]{\expDistFigPath blind-greedy-dims.png}
+    \caption{Results of \bgrdy spanner for different dimensions.}
+    \label{fig:blind_greedy_dimensions}
+    \end{centering}
+\end{figure}
+
+Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
+The ideal behavior is demonstrated by the non-blind greedy spanner,
+for which this ratio stays practically constant, confirming the linear growth.
+None of the blind algorithms seems to have this property, but among them
+the \bgrdy spanner is the best one. If we assume that the number of edges
+is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
+linear regression (after taking $\log$).  We give in Table \ref{tbl:regr_coeff_spanner}
+the estimated exponents $\alpha$ for \bgrdy and standard
+greedy algorithms. Note that even for the greedy algorithm these estimated
+exponents can be significantly larger than 1,
+which is explained by the fact that the number of points
+on which we computed spanners is not large enough to clearly see
+the linear dependence.
+
+\begin{table}[b]
+\begin{tabular}{|l|l|l|}
+\hline
+dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
+2         &         1.08                &  1.12                 \\ \hline
+3         &         1.24                &  1.41                \\ \hline
+4         &         1.42                &  1.77                 \\ \hline
+\end{tabular}
+\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
+on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
+\label{tbl:regr_coeff_spanner}
+\end{table}
+
+The plot in Figure \ref{fig:blind_greedy_dimensions}
+compares performance of \bgrdy in different dimensions.
+We see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
+edges for $700$ points, which clearly shows some degrading for higher dimensions.
+
+The plot in Figure \ref{fig:spanner_eps_dependence}
+compares the \bgrdy and \textsc{Greedy} algorithms
+on \textbf{uniform} point sets for different choices of $\eps$. Dependence on $\eps$
+is approximately the same for both algorithms. Since it is not clearly seen from the picture, we also note
+that the ratio of the number of edges decreases for
+smaller values of $\eps$: 
+for $\eps = 2$ the blind greedy spanner contains almost 6 times more edges than the greedy spanner,
+while for $\eps = 1/32$ the ratio is 2.6
+
+\begin{figure}[ht]
+        \begin{tikzpicture}
+            \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
+                legend pos=north east]
+
+            \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+            \addlegendentry{Greedy, dim = 2}
+
+            \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
+            \addlegendentry{Blind greedy, dim = 2}
+
+        \end{axis}
+        \end{tikzpicture}
+        \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
+        \label{fig:spanner_eps_dependence}
+\end{figure}
+
+\section{Definition of Wasserstein Distance}
 \label{sec:appendix_background}
-Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
-if 
-\[
-\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
-\]
-A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
-of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
-The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
-Har-Peled and Mendel \cite{hm-fast} 
-introduced the notion of net-trees and
-generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
-\begin{enumerate}
-    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
-    \item If $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
-are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
-    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
-The algorithm uses the net-tree structure.
-\end{enumerate}
-The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
-et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
-than net-trees. We mention in passing that cover trees can also be used for building
-a spanner (this can be proven with the same methods), and we use cover trees for
-building WSPD spanners in one of our implementations.
-
-\myparagraph{Wasserstein Distance}
 Let $\diag$ denote the line in the plane with the equation $y = x$; we call it the diagonal.
@@
 in order to find $\wsdist{q}$, one must solve an \textit{assignment problem}.
+
+\printbibliography
+
+\end{document}

### 2019-11-22 cpp_version/include/dynamic_spanner.h
EMD underway

@@ -519,6 +519,15 @@ public:
 
     }
 
+    double get_bruteforce_timing() const
+    {
+        double brute_force_timing = 0.0;
+        for(int i = 0; i < m_num_points; ++i)
+            for(int j = i + 1; j < m_num_points; ++j)
+                brute_force_timing += m_distance_timings_matrix[i][j];
+        return brute_force_timing;
+    }
+
     std::string get_statistics() const
     {
         double n_pairs = m_num_points * (m_num_points - 1) / 2;

### 2019-11-22 cpp_version/src/main.cpp

@@ -579,8 +579,10 @@ int main(int argc, char** argv)
         auto end_blind_greedy = std::chrono::steady_clock::now();
         auto dur_blind_greedy = std::chrono::duration_cast<std::chrono::microseconds>(end_blind_greedy - begin_blind_greedy).count();
         dur_blind_greedy += blind_greedy_spanner.get_microseconds_to_compute_distances();
-        console->info("blind_greedy_time = {} microseconds", dur_blind_greedy);
-        exp_logger->info("blind_greedy_time = {} microseconds", dur_blind_greedy);
+        double bf_time = blind_greedy_spanner.get_bruteforce_timing();
+        double savings_time_ratio = dur_blind_greedy / bf_time;
+        console->info("blind_greedy_time = {} microseconds out of {}, ratio  {} ", dur_blind_greedy, bf_time, savings_time_ratio);
+        exp_logger->info("blind_greedy_time = {} microseconds out of {}, ratio  {} ", dur_blind_greedy, bf_time, savings_time_ratio);
         console->info("{};{}", dist_name, blind_greedy_spanner.get_statistics());
         exp_logger->info("{};{}", dist_name, blind_greedy_spanner.get_statistics());
         exp_logger->flush();

### 2019-11-22 emd/emd_assemble_matrix.py

@@ -2,17 +2,7 @@
 
 import numpy as np
 
-def save_matrix(fname, a):
-    n_rows = a.shape[0]
-    n_cols = a.shape[1]
-    n_entries = n_rows * n_cols
-    with open(fname, 'w') as f:
-        f.write("%d\n" % n_rows)
-        for i in range(n_rows):
-            for j in range(n_cols):
-                f.write("%f\n" % a[i][j])
-
-
+from emd_utils import *
 
 if __name__ == "__main__":
     flist = [ "matrix-try" ]

### 2019-11-22 emd/emd_assemble_matrix_gtsrb.py

@@ -1,0 +1,116 @@
+#!/usr/bin/env python
+
+import random
+import numpy as np
+
+from emd_utils import *
+
+if __name__ == "__main__":
+    flist = [ "matrix-try" ]
+
+    n_procs = 10
+    n_images = 300
+    n_pairs = n_images * (n_images - 1) / 2
+
+    flist = [ "gtsrb_pair_file_i_%d_np_%d.%d.matrix.%d" % (n_images, n_procs, i, i) for i in range(n_procs) ]
+    flist_1 = [ "gtsrb_missing_.%d.matrix.%d" % (i, i) for i in range(n_procs) ]
+    flist.extend(flist_1)
+
+
+    d_matr = np.zeros((n_images, n_images), dtype = np.float32)
+    t_matr = np.zeros((n_images, n_images), dtype = np.float32)
+
+    image_fnames = [ "" for i in range(n_images) ]
+
+    for fname in flist:
+        print(fname)
+        with open(fname, 'r') as f:
+            for line in f:
+                a = line.split('\t')
+                i = int(a[0])
+                j = int(a[1])
+                image_fnames[i] = a[2]
+                image_fnames[j] = a[3]
+                d_matr[i][j] = d_matr[j][i] = float(a[4])
+                t_matr[i][j] = t_matr[j][i] = 1000000 * float(a[5])
+
+    computed_distances = np.sum(d_matr > 0.0, 1) + 1
+
+    complete_indices = np.where(computed_distances == n_images)
+    tot_incomplete_indices = np.where(computed_distances == 1)
+    partial_indices = np.where((computed_distances > 1) & (computed_distances < n_images))
+
+    print("complete: ", complete_indices[0].size)
+
+    partial_image_fnames = [ (image_fnames[i], i) for i in partial_indices[0] ]
+
+    # # print("partial: ", partial_image_fnames)
+
+    small_partial_image_fnames = []
+    max_size = 900
+    for pif, pif_idx in partial_image_fnames:
+        w, p = load_picture(pif)
+        if p.shape[0] > max_size:
+            print(pif, " is too large: ", p.shape)
+        else:
+            small_partial_image_fnames.append((pif, pif_idx))
+
+    small_partial_indices = [ pif_idx for (pif, pif_idx) in small_partial_image_fnames ]
+    aaa = [ x for x in complete_indices[0] ]
[…]

### 2019-11-22 emd/emd_gtsrb_check_size.py

@@ -1,0 +1,29 @@
+#!/usr/bin/env python
+
+from timeit import default_timer as timer
+import sys
+
+import glob
+import random
+
+
+from PIL import Image
+import numpy as np
+from emd import emd
+
+from emd_utils import *
+
+if __name__ == "__main__":
+    random.seed(42)
+    n_images = 300
+    im_files = read_gtsrb_file_list(n_images)
+    a = []
+    for fname in im_files:
+        r = load_picture(fname)
+        a.append(r)
+
+    a.sort()
+
+    for t in a:
+        print(t)
+

### 2019-11-22 emd/emd_prepare_gtsrb_files.py

@@ -3,31 +3,13 @@
 import glob
 import random
 
-def read_gtsrb_file_list(n_images):
-    im_files_all = glob.glob("traffic_signs/*.ppm")
-    im_files = random.sample(im_files_all, n_images)
-    return im_files
-
-def split(a, n):
-    k, m = divmod(len(a), n)
-    return (a[i * k + min(i, m):(i + 1) * k + min(i + 1, m)] for i in xrange(n))
-
-
-def prepare_file_lists_for_parallel(im_files, n_procs, file_start):
-    all_pairs = [ (i, j) for i in range(len(im_files)) for j in range(0, i) ]
-    all_pairs_split = split(all_pairs, n_procs)
-    for i, part in enumerate(all_pairs_split):
-        fname = file_start + "." + str(i)
-        with open(fname, 'w') as f:
-            for x, y in part:
-                f.write("%s %s %d %d\n" % (im_files[x], im_files[y], x, y))
-
+from emd_utils import *
 
 if __name__ == "__main__":
 
     random.seed(42)
 
-    n_images = 400
+    n_images = 300
 
     n_procs = 10
 

### 2019-11-22 emd/emd_process_one_file.py

@@ -16,21 +16,6 @@ def pic_positions(shape):
     result = np.asarray(result, dtype=np.float64)
     return result
 
-def load_picture(fname, pic_size = None):
-    a = Image.open(fname)
-    if pic_size:
-        a = a.resize((pic_size,pic_size))
-    # get grayvalues
-    a_weights = np.asarray(a.convert("L"), dtype=np.float64)
-    # normalize
-    a_weights = a_weights / np.sum(a_weights)
-    # flatten
-    n_pixels = a_weights.shape[0] * a_weights.shape[1]
-    a_weights = a_weights.reshape((n_pixels, 1))
-
-    return a_weights
-
-
 
 def compute_distance(fname_a, fname_b, pic_size):
     pic_shape = (pic_size, pic_size)

### 2019-11-22 emd/emd_process_one_file_gtsrb.py

@@ -7,50 +7,63 @@ from PIL import Image
 import numpy as np
 from emd import emd
 
-# return flattened list of pixel coordinates (support of measure)
-def pic_positions(shape):
-    result = []
-    for i in range(shape[0]):
-        for j in range(shape[1]):
-            result.append([float(i),float(j)])
-    result = np.asarray(result, dtype=np.float64)
-    return result
+from emd_utils import *
 
-def load_picture(fname):
-    a = Image.open(fname)
-    # get grayvalues
-    a_weights = np.asarray(a.convert("L"), dtype=np.float64)
-    # normalize
-    a_weights = a_weights / np.sum(a_weights)
+# # return flattened list of pixel coordinates (support of measure)
+# def pic_positions(shape):
+#     result = []
+#     for i in range(shape[0]):
+#         for j in range(shape[1]):
+#             result.append([float(i),float(j)])
+#     result = np.asarray(result, dtype=np.float64)
+#     return result
 
-    # get image support
-    a_positions = pic_positions(a_weights.shape)
-
-    # flatten
-    n_pixels = a_weights.shape[0] * a_weights.shape[1]
-    a_weights = a_weights.reshape((n_pixels, 1))
-
-    return (a_weights, a_positions)
+# def resize_pic(pic, coeff):
+#     new_width = pic.width / coeff
+#     new_height = pic.height / coeff
+#     return pic.resize((new_width, new_height), Image.BICUBIC)
 
 
-def compute_distance(fname_a, fname_b):
-    weights_a, positions_a = load_picture(fname_a)
-    weights_b, positions_b = load_picture(fname_b)
-    start = timer()
-    dval = emd(positions_a, positions_b, weights_a, weights_b)
-    end = timer()
-    elapsed = end - start
-    return (dval, elapsed)
 
 
-def compute_from_file(fname_in, fname_out):
-    with open(fname_in, 'r') as pair_file, open(fname_out, 'w') as matrix_file:
-        for line in pair_file:
-            fname_a, fname_b, i, j = line.split(' ')
-            i = int(i)
-            j = int(j)
-            dval, elapsed = compute_distance(fname_a, fname_b)
-            matrix_file.write("%d\t%d\t%s\t%s\t%f\t%f\n" % (i, j, fname_a, fname_b, dval, elapsed))
+# def load_picture(fname):
+#     a = Image.open(fname)
+
[…]

### 2019-11-22 emd/emd_utils.py

@@ -1,0 +1,107 @@
+#!/usr/bin/env python
+
+from timeit import default_timer as timer
+import sys
+
+import glob
+import random
+
+
+from PIL import Image
+import numpy as np
+from emd import emd
+
+
+def read_gtsrb_file_list(n_images):
+    im_files_all = glob.glob("traffic_signs/*.ppm")
+    im_files = random.sample(im_files_all, n_images)
+    return im_files
+
+
+def resize_pic(pic, coeff):
+    new_width = pic.width / coeff
+    new_height = pic.height / coeff
+    return pic.resize((new_width, new_height), Image.BICUBIC)
+
+
+
+def split(a, n):
+    k, m = divmod(len(a), n)
+    return (a[i * k + min(i, m):(i + 1) * k + min(i + 1, m)] for i in xrange(n))
+
+
+def prepare_file_lists_for_parallel(im_files, n_procs, file_start):
+    all_pairs = [ (i, j) for i in range(len(im_files)) for j in range(0, i) ]
+    all_pairs_split = split(all_pairs, n_procs)
+    for i, part in enumerate(all_pairs_split):
+        fname = file_start + "." + str(i)
+        with open(fname, 'w') as f:
+            for x, y in part:
+                f.write("%s %s %d %d\n" % (im_files[x], im_files[y], x, y))
+
+
+def save_matrix(fname, a):
+    n_rows = a.shape[0]
+    n_cols = a.shape[1]
+    n_entries = n_rows * n_cols
+    with open(fname, 'w') as f:
+        f.write("%d\n" % n_rows)
+        for i in range(n_rows):
+            for j in range(n_cols):
+                f.write("%f\n" % a[i][j])
+
+# return flattened list of pixel coordinates (support of measure)
+def pic_positions(shape):
+    result = []
+    for i in range(shape[0]):
+        for j in range(shape[1]):
+            result.append([float(i),float(j)])
+    result = np.asarray(result, dtype=np.float64)
+    return result
+
+
+def resize_pic(pic, coeff):
+    new_width = pic.width / coeff
+    new_height = pic.height / coeff
+    return pic.resize((new_width, new_height), Image.BICUBIC)
+
+
+def load_picture(fname):
+    a = Image.open(fname)
+
+    a = resize_pic(a, 2)
+
+    # get grayvalues
+    a_weights = np.asarray(a.convert("L"), dtype=np.float64)
+    # normalize
[…]

### 2019-11-22 emd/example.py

@@ -4,39 +4,28 @@ from PIL import Image
 import numpy as np
 from emd import emd
 
-def weights(pic_size):
+def get_pic_support(pic_shape):
     result = []
-    for i in range(pic_size):
-        for j in range(pic_size):
+    for i in range(pic_shape[0]):
+        for j in range(pic_shape[1]):
             result.append([float(i),float(j)])
     result = np.asarray(result, dtype=np.float64)
     return result
 
-pic_size = 64
-n_pixels = pic_size * pic_size
+def read_image(fname):
+    pic = Image.open(fname)
+    pic_weights = np.asarray(pic.convert("L"), dtype=np.float64)
+    pic_weights = pic_weights / np.sum(pic_weights)
+    pic_shape = pic_weights.shape
+    n_pixels = pic_shape[0] * pic_shape[1]
+    pic_weights = pic_weights.reshape((n_pixels, 1))
+    pic_support = get_pic_support(pic_shape)
+    return (pic_weights, pic_support)
 
-a = Image.open("bismarck.jpg")
-a = a.resize((pic_size,pic_size))
-a_weights = np.asarray(a.convert("L"), dtype=np.float64)
-a_weights = a_weights / np.sum(a_weights)
 
-b = Image.open("kipling.jpg")
-b = b.resize((pic_size,pic_size))
-b_weights = np.asarray(b.convert("L"), dtype=np.float64)
-b_weights = b_weights / np.sum(b_weights)
-
-a_weights = a_weights.reshape((n_pixels, 1))
-b_weights = b_weights.reshape((n_pixels, 1))
-
-print("read images")
-
-A = weights(pic_size)
-B = weights(pic_size)
-
-# print(A)
-# print(B)
-# print(a_weights)
-# print(b_weights)
-
-res = emd(A, B, a_weights, b_weights)
-print("distance = ", res)
+if __name__ == "__main__":
+    a_weights, A = read_image("00001.ppm")
+    b_weights, B = read_image("00002.ppm")
+    print("read images")
+    res = emd(A, B, a_weights, b_weights)
+    print("distance = ", res)

### 2019-11-22 paper/draft_socg.tex
Add SoCG version

--- 2019-04-25 paper/expensive-distances-main.tex
@@
+\documentclass[a4paper,UKenglish,cleveref, autoref]{socg-lipics-v2019}
+
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\input{my_preamble.tex}
+
+\title{Metric Spaces with Expensive Distances}
+\titlerunning{Metric Spaces with Expensive Distances}
+% uncomment for LIPICS 
+%\author{Anonymous Authors}{Anonymous}{}{}{}
+\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
+\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
+\authorrunning{M. Kerber and A. Nigmetov}
+%\authorrunning{Anonymous Authors}
+
+\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
+\ccsdesc[500]{Theory of computation~Random search heuristics}
+%% \category{dummy}
+
+% \category{dummy}
+
+\Copyright{Michael Kerber and Arnur Nigmetov}
+%\Copyright{Anonymous Authors}
+% END uncomment for LIPICS
+
+\def \expDistFigPath {pics/}
+\def \expDistDataPath {./}
+
+\begin{document}
+\maketitle
+
+\begin{abstract}
+In algorithms for finite metric spaces, it is common
+to assume that the distance between two points
+can be computed in constant time, and complexity
+bounds are expressed only in terms of the number
+of points of the metric space.
+We introduce a different model, where we assume
+that the computation of a single distance is
+an expensive operation and consequently, the goal is
+to minimize the number of such distance queries.
+This model is motivated by metric spaces
+that appear in the context of topological data analysis.
+
+We consider two standard operations on metric spaces,
+namely the construction of a $(1+\eps)$-spanner
+and the computation of an approximate nearest
+neighbour for a given query point.
+In both cases, we partially explore the metric space
+through distance queries and infer lower and upper bounds
+for yet unexplored distances through triangle inequality.
+For spanners, we evaluate several exploration strategies
+through extensive experimental evaluation.
+For approximate nearest neighbours, we prove that our
+strategy returns an approximate nearest neighbour
+after a logarithmic number of distance queries.
+\end{abstract}
+
 \section{Introduction}
@@
 
-\newpage
+\bibliography{bib}
 
@@
 in order to find $\wsdist{q}$, one must solve an \textit{assignment problem}.
+
+\end{document}

### 2019-11-22 paper/draft_socg.tex
Add  EMD results.

@@
 and the computation of an approximate nearest
-neighbour for a given query point.
+neighbor for a given query point.
 In both cases, we partially explore the metric space
@@
 through extensive experimental evaluation.
-For approximate nearest neighbours, we prove that our
-strategy returns an approximate nearest neighbour
+For approximate nearest neighbors, we prove that our
+strategy returns an approximate nearest neighbor
 after a logarithmic number of distance queries.
@@
 Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
-a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbours
+a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
 performing $O(\log n)$ distance computations per query point.
@@
 %
-We propose simple algorithms for spanner construction and approximate nearest neighbour search
+We propose simple algorithms for spanner construction and approximate nearest neighbor search
 and evaluate them theoretically and experimentally in the defined cost model.
@@
 
-For approximate nearest neighbour, we devise a simple randomized incremental algorithm and show that
+For approximate nearest neighbor, we devise a simple randomized incremental algorithm and show that
 the number of distance queries needed is $O(\log n)$ in expectation.
-Our proof is based on the well-known observation that the nearest neighbour changes $O(\log n)$ times
+Our proof is based on the well-known observation that the nearest neighbor changes $O(\log n)$ times
 in expectation when traversing the sequence of points, combined with a packing argument certifying that
@@
         at random from $[0,10000]^d$, and then we add normally distributed noise around
-        each of the centres. The number of clusters is chosen so that each cluster
+        each of the centers. The number of clusters is chosen so that each cluster
         contains 50 points.
@@
     \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
-                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{c} & \multicolumn{2}{c}{$q=2$} & \phantom{c} & \multicolumn{2}{c}{$q=3$}\\
                           \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
@@
 \label{tbl:mcgill_original_blind_greedy_spanner_timing}
+\end{table}
+
+\begin{table}\centering
+    \ra{1.2}
+    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
+        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
+        \cmidrule{2-3} \cmidrule{5-7} 
+                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
+$\eps = 0.1$    &   $0.19$  &  $0.42$    &&  $618,587$   &  $262,770$    & $0.42$   \\
+$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $618,587$   & $196,309$    & $0.31$   \\
+$\eps = 0.5$    &   $0.04$  &  $0.18$    &&  $618,587$   & $112,995$    & $0.18$   \\
+\bottomrule
+\end{tabular}
+    \caption{Results of \bgrdy algorithm on the dataset of 250 images
+    with Earth Mover Distance, time in seconds.}
+\label{tbl:emd_gtsrb}
 \end{table}
@@
 
-\section{Approximate nearest neighbours}
+\section{Approximate nearest neighbors}
 \label{sec:ann}
-We consider the standard problem of finding an approximate nearest neighbour: given
+We consider the standard problem of finding an approximate nearest neighbor: given
 $n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
@@
 to the query point $q$, which are initially all set to $0$.
-We also remember the closest neighbour $c$ that we have seen so
+We also remember the closest neighbor $c$ that we have seen so
 far and its distance $v$ to $q$.
 We refer to the point $c$ as the \emph{candidate}.
-We maintain the invariant that $c$ is an approximate nearest neighbour
+We maintain the invariant that $c$ is an approximate nearest neighbor
 to $q$ for the points $\{p_1,\ldots,p_i\}$.
@@
 bound $a_i$ satisfies $a_i\geq \frac{v}{1+\eps}$.
-If so, $c$ remains an approximate nearest neighbour and
+If so, $c$ remains an approximate nearest neighbor and
 we do not query the distance of $p_i$ to $q$.
@@
 accordingly. At the end of the loop, $c$ is an approximate nearest
-neighbour. The pseudocode follows.
+neighbor. The pseudocode follows.
 
@@
 
-We remark that we obtain an exact nearest neighbour algorithm
+We remark that we obtain an exact nearest neighbor algorithm
 when setting $\eps$ to $0$, which means 
@@
 shows, there are point sets where the expected number
-of distance computations for exact nearest neighbour is linear.
+of distance computations for exact nearest neighbor is linear.
 
@@
 
-\section{Experiments on approximate nearest neighbours}
+\section{Experiments on approximate nearest neighbors}
 \label{sec:experiment_ann}
@@
 the $\Omega(n^3)$ complexity of our algorithms makes it practically
-non-competetive, when $n$ becomes large.
+non-competitive, when $n$ becomes large.
 On the other hand, this overhead is relatively easy to estimate
@@
     \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
-    \caption{Example of a point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
-    The exact nearest neighbour is the point $p_1$, next point $p_i$ is placed 
+    \caption{Example of a point set where exact nearest neighbor search cannot be accelerated by maintaining bounds.
+    The exact nearest neighbor is the point $p_1$, next point $p_i$ is placed 
     in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
-    is the true nearest neighbour cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
+    is the true nearest neighbor cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
     $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
@@
 $v_k\leq (1+\eps)r_k$, because otherwise, $c_k$ would not be an approximate
-nearest neighbour of $\{p_1,\ldots,p_k\}$.
+nearest neighbor of $\{p_1,\ldots,p_k\}$.
 

### 2019-11-24 paper/draft_socg.tex
Add some text about EMD.

@@
 \emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
-We demonstrate experimentally that on a collection
-of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
-computing all $\binom{n}{2}$ distances.
+Our experiments with Earth Mover distance and persistence
+diagrams of 3-D shapes show that blind greedy spanner can speed up computations
+by a factor of 3, saving hundreds of hours of machine time,
+even though the images and diagrams in our data sets are of moderate size.
+%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
+%computing all $\binom{n}{2}$ distances.
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
@@
 
-We performed two types of experiments: on points sampled from a low-dimensional Euclidean
-space and on persistence diagrams
+We performed three types of experiments: on points sampled from a low-dimensional Euclidean
+space, on images with Earth Mover distance, and on persistence diagrams
 with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
@@
 \end{figure}
+
+For experiments with images, we used the German traffic sign
+dataset and the code for computing the Earth Mover distance
+by Gary Dornan.
+We randomly selected 250 images from the benchmark.
+The first two columns of \cref{tbl:emd_gtsrb}
+compare the sparseness of \bgrdy and greedy spanners.
+For smaller values of $\eps$,
+the \bgrdy spanner contains approximately 3 times more
+edges than the greedy spanner, so the ratio is roughly
+the same as for persistence diagrams.
+The last 3 columns show the advantage
+of \bgrdy in terms of the running time.
+For $10 \%$ approximation, we can reduce
+the running time by more than a factor of $2$,
+saving $98$ hours of computation time.
+For $20 \%$ approximation,
+we speed up the computations by a factor of 3.
 
@@
 
+\myparagraph{Earth Mover distance}
+Let $\mu_P$ and $\mu_Q$
+be two probability measures supported
+on finite sets $P = \{p_1, \dots, p_n \}$
+and $Q = \{ q_1, \dots, q_m \}$.
+An $m \times n$ matrix $C$
+is called a \textit{transportation plan},
+if it satisfies the following three conditions:
+\begin{enumerate}
+    \item $c_{i,j} \geq 0$ for all entries $c_{i, j}$.
+    \item $\sum_{i = 1}^m c_{i, j} = p_j$ for all $j = 1, \dots, n$.
+    \item $\sum_{j = 1}^n c_{i, j} = q_i$ for all $i = 1, \dots, m$.
+\end{enumerate}
+%By definition, the \textit{cost} of transportation plan $C$
+Let $\dist(\cdot, \cdot)$ denote the Euclidean
+distance in plane.
+The earth mover distance between $\mu_P$ and $\mu_Q$
+is defined as
+\[
+    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j)$,
+\]
+where $\inf$ is taken over all transportation
+plans $C$. Intuitively, $c_{i,j}$ is the
+mass concentrated at the point $p_i$ that
+is sent to the point $q_j$, and the earth mover
+distance is the smallest total amount of work
+needed to transform the measure $\mu_P$
+into the measure $\mu_Q$.
+
+Let $I_1$, $I_2$ be two greyvalue images.
+We divide each greyvalue of $I_{1,2}$
+by the sum of all greyvalues in the corresponding image.
+Thus we can interpret $I_i$
+as a probability measure supported on a rectangle
+that has one vertex at the origin and
+fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
+The earth mover distance between images $I_1$ and $I_2$
+is, by definition, the earth mover distance between
+these probability measures.
+
 \end{document}

### 2019-11-26 paper/draft_socg.tex
add citations

@@
 in polynomial time in the size of the graphs.
-Another instance is a collection of high-resolution images
-endowed with the Wasserstein (or Earth Movers) metric \cite{rubner2000earth}.
+Another instance is a collection of images
+equipped with the Wasserstein (or the Earth Mover's) metric \cite{rubner2000earth}.
 
@@
 \emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
-Our experiments with Earth Mover distance and persistence
+Our experiments with the Earth Mover's distance and persistence
 diagrams of 3-D shapes show that blind greedy spanner can speed up computations
-by a factor of 3, saving hundreds of hours of machine time,
-even though the images and diagrams in our data sets are of moderate size.
+by a factor of 2 to 5, saving hundreds of hours of machine time,
+even though the inputs in our data sets are of moderate size.
 %of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
@@
 For experiments with images, we used the German traffic sign
-dataset and the code for computing the Earth Mover distance
-by Gary Dornan.
+dataset \cite{gtsrb} and the code for computing the Earth Mover distance
+by Doran \cite{doran_pyemd}.
 We randomly selected 250 images from the benchmark.
@@
 \[
-    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j)$,
+    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j),
 \]

### 2019-12-04 paper/draft_socg.tex
Typo
add more data

@@
 In that situation, standard constructions from computational geometry provide partial answers:
-Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
+Using net-trees~\cite{hm-fast}, we can construct an $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
 a {$(1+\eps)$-spanner} immediately. Net-trees can also be used to compute approximate nearest neighbors
@@
 The experimentally most successful exploration strategy that we found is to
-repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
+repeatedly query the distance between a pair with the worst ratio of upper and lower bound.
 We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
 \emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
+
 Our experiments with the Earth Mover's distance and persistence
@@
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
-We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
+We also compare with a spanner construction based on WSPD. Our simple algorithm gives much smaller
 spanners on the tested examples. 
@@
 
+\myparagraph{Outline}
+We review the properties of spanner algorithms in \cref{sec:back}.
+In \cref{sec:blind_spanners}, we introduce different
+variants of the blind spanner algorithm. Experimental
+results on spanners are presented in \cref{sec:experiment_spanners}.
+Then we move to the ANN problem, giving
+an outline of the proof of the $O(\log n)$ bound in \cref{sec:ann}.
+The full proof can be found in the appendix.
+We report our experimental results for ANN in \cref{sec:experiment_ann}.
+
 \section{Background}
+\label{sec:back}
 %
@@
 has a constant doubling dimension.
+
+\myparagraph{Well-Separated Pair Decomposition}
+Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
+if 
+\[
+\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
+\]
+A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
+of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
+The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
+Har-Peled and Mendel \cite{hm-fast} 
+introduced the notion of net-trees and
+generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
+\begin{enumerate}
+    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
+    \item If $\{ \{A_1, B_1 \}, 
+\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
+are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
+    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
+The algorithm uses the net-tree structure.
+\end{enumerate}
+The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
+et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
+than net-trees. We mention in passing that cover trees can also be used for building
+a spanner (this can be proven with the same methods), and we use cover trees for
+building WSPD spanners in one of our implementations.
 
@@
 
-On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
-for convenience, we repeat the definition of WSPD in \cref{sec:appendix_background}}~\cite{cal-kos-wspd,hm-fast} only compute
+On the other hand, spanner constructions based on Well-Separated Pair Decomposition~\cite{cal-kos-wspd,hm-fast} only compute
 $O(n\log n+n \eps^{-d})$ distances 
@@
 \section{Blind spanners}
+\label{sec:blind_spanners}
 %
@@
 We performed three types of experiments: on points sampled from a low-dimensional Euclidean
-space, on images with Earth Mover distance, and on persistence diagrams
+space, on images with Earth Mover's distance, and on persistence diagrams
 with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
@@
                           & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
-%\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
 $\eps = 0.1$                & 0.14  & 0.28         && 0.32  & 0.72      &&   0.41  & 0.78 \\
-%\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
 $\eps = 0.2$                & 0.06 & 0.19          && 0.13 & 0.44       &&   0.20  & 0.52 \\
-%\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
-%\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
+$\eps = 0.4$               & 0.03 &  0.13          && 0.05 & 0.26        &&   0.06  & 0.34 \\
 $\eps = 0.5$               & 0.02 &  0.10          && 0.03 & 0.20        &&   0.05  & 0.27 \\
+$\eps = 0.8$               & 0.01 & 0.08           && 0.02  & 0.13        &&   0.02  & 0.17 \\
+$\eps = 1.0$               & 0.01 &  0.06  && 0.01 & 0.10        &&   0.02  & 0.14 \\
 \bottomrule
@@
                           & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
-%$\eps = 0.1$    &   4093  &  1266  &&  5331  &  4085  &&  16852  &  13557\\
-%$\eps = 0.2$    &   4093  &  878  &&  5331  &  2574  &&  16852  &  9383\\
-%$\eps = 0.5$    &   4093  &  483  &&  5331  &  1233  &&  16852  &  5146\\
-%$\eps = 0.01$    &   4093 &  - &&  5331 &  - &&  16852  &  - \\
-%$\eps = 0.01$    &   4093  &  0  &&  5331  &  0  &&  16852  &  0\\
-$\eps = 0.1$    &   2605  &  1266  &&  3719  &  4085  &&  11781  &  13557\\
-$\eps = 0.2$    &   2209  &  878  &&  3338  &  2574  &&  10541   &  9383\\
-$\eps = 0.5$    &   1744  &  483  &&  2835  &  1233  &&  9418    &  5146\\
+$\eps = 0.1$    &   $2605$  &  $1266$  &&  $3719$  &  $4085$  &&  $11$\,$781$  &  $13$\,$557$\\
+$\eps = 0.2$    &   $2209$  &  $878$  &&  $3338$  &  $2574$  &&  $10$\,$541$  &  $9383$ \\
+$\eps = 0.4$    &   $1988$  &  $586$  &&  $3088$  &  $1590$  &&  $-$    &  $6383$ \\
+$\eps = 0.5$    &   $1744$  &  $483$  &&  $2835$  &  $1233$  &&  $9418$    &  $5146$ \\
+$\eps = 0.8$    &   $1691$  &  $368$  &&  $2809$  &  $814$  &&  $-$    &  $3419$ \\
+$\eps = 1.0$    &   $1645$  &  $324$  &&  $2780$  &  $672$  &&  $-$    &  $-$ \\
 \bottomrule
@@
                 & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
-$\eps = 0.1$    &   $0.19$  &  $0.42$    &&  $618,587$   &  $262,770$    & $0.42$   \\
-$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $618,587$   & $196,309$    & $0.31$   \\
-$\eps = 0.5$    &   $0.04$  &  $0.18$    &&  $618,587$   & $112,995$    & $0.18$   \\
+$\eps = 0.1$    &   $0.20$  &  $0.43$    &&  $853\,682$   &  $248\,192$    & $0.29$   \\
+$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $853\,682$   & $192\,302$    & $0.22$   \\
+$\eps = 0.3$    &   $0.07$  &  $0.24$    &&  $853\,682$   & $161\,007$    & $0.19$   \\
+$\eps = 0.4$    &   $0.05$  &  $0.20$    &&  $853\,682$   & $140\,470$    & $0.16$   \\
+$\eps = 0.5$    &   $0.04$  &  $0.17$    &&  $853\,682$   & $129\,489$    & $0.15$   \\
+$\eps = 0.8$    &   $0.02$  &  $0.13$    &&  $853\,682$   & $104\,851$    & $0.12$   \\
+$\eps = 1.0$    &   $0.02$  &  $0.11$    &&  $853\,682$   & $92\,462$    & $0.11$   \\
 \bottomrule
@@
     \caption{Results of \bgrdy algorithm on the dataset of 250 images
-    with Earth Mover Distance, time in seconds.}
+    with Earth Mover's Distance, time in seconds.}
+\label{tbl:emd_gtsrb}
+\end{table}
+
+\begin{table}\centering
+    \ra{1.2}
+    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
+        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
+        \cmidrule{2-3} \cmidrule{5-7} 
+                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
+$\eps = 0.2$    &   $0.34$  &  $0.81$    &&  $185\,863$   & $133\,194$    & $0.72$   \\
+$\eps = 0.3$    &   $0.20$  &  $0.69$    &&  $185\,863$   & $105\,723$    & $0.56$   \\
+$\eps = 0.4$    &   $0.13$  &  $0.58$    &&  $185\,863$   & $86\,283$    & $0.46$   \\
+$\eps = 0.5$    &   $0.09$  &  $0.50$    &&  $185\,863$   & $72\,447$    & $0.38$   \\
+$\eps = 0.8$    &   $0.05$  &  $0.35$    &&  $185\,863$   & $48\,870$    & $0.26$   \\
+$\eps = 1.0$    &   $0.04$  &  $0.30$    &&  $185\,863$   & $40\,955$    & $0.22$   \\
+\bottomrule
+\end{tabular}
+    \caption{Results of \bgrdy algorithm on the dataset of 200 images
+    with Earth Mover's Distance on \textit{Lab} histograms, time in seconds.}
 \label{tbl:emd_gtsrb}
@@
 For experiments with images, we used the German traffic sign
-dataset \cite{gtsrb} and the code for computing the Earth Mover distance
+dataset GTSRB \cite{gtsrb} and the code for computing the Earth Mover's distance
 by Doran \cite{doran_pyemd}.
 We randomly selected 250 images from the benchmark.
-The first two columns of \cref{tbl:emd_gtsrb}
-compare the sparseness of \bgrdy and greedy spanners.
+The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
+The first two columns compare the sparseness of \bgrdy and greedy spanners.
 For smaller values of $\eps$,
@@
 of \bgrdy in terms of the running time.
-For $10 \%$ approximation, we can reduce
+If a $1.5$ approximation is acceptable,
+then the \bgrdy algorithm speeds up
+the computation by a factor of more than $5$,
+and even for a $10 \%$ approximation, it reduces
 the running time by more than a factor of $2$,
 saving $98$ hours of computation time.
-For $20 \%$ approximation,
-we speed up the computations by a factor of 3.
 
@@
 \section{Additional Definitions}
-\myparagraph{Well-Separated Pair Decomposition}
 \label{sec:appendix_background}
-Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
-if 
-\[
-\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
-\]
-A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
-of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
-The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
-Har-Peled and Mendel \cite{hm-fast} 
-introduced the notion of net-trees and
-generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
-\begin{enumerate}
-    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
-    \item If $\{ \{A_1, B_1 \}, 
-\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
-are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
-    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time.
-The algorithm uses the net-tree structure.
-\end{enumerate}
-The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
-et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
-than net-trees. We mention in passing that cover trees can also be used for building
-a spanner (this can be proven with the same methods), and we use cover trees for
-building WSPD spanners in one of our implementations.
 
@@
 
-\myparagraph{Earth Mover distance}
+\myparagraph{Earth Mover's distance}
 Let $\mu_P$ and $\mu_Q$
@@
 distance in plane.
-The earth mover distance between $\mu_P$ and $\mu_Q$
+The Earth Mover's distance between $\mu_P$ and $\mu_Q$
 is defined as
@@
 mass concentrated at the point $p_i$ that
-is sent to the point $q_j$, and the earth mover
+is sent to the point $q_j$, and the Earth Mover's
 distance is the smallest total amount of work
@@
 fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
-The earth mover distance between images $I_1$ and $I_2$
-is, by definition, the earth mover distance between
+The Earth Mover's distance between images $I_1$ and $I_2$
+is, by definition, the Earth Mover's distance between
 these probability measures.

### 2019-12-04 paper/draft_socg.tex
Add Lab Earth Mover data, move grey value to appendix

@@
 in polynomial time in the size of the graphs.
-Another instance is a collection of images
-equipped with the Wasserstein (or the Earth Mover's) metric \cite{rubner2000earth}.
+
+Another example is the Earth Mover's distance (EMD, \cite{rubner2000earth}), which
+is often used for images.
+To compute it, one has to solve the optimal
+transportation problem, which has cubical complexity. There exist many papers addressing this issue,
+e.g., \cite{wemd}, \cite{grounded_emd}, \cite{entropy_emd}.
+However, usually these solutions either provide no guarantees of the approximation
+quality at all, as in entropy-regularization based approaches,
+or give only a loose bound on the relative error (e.g., the wavelet EMD \cite{wemd}
+theoretically can differ from the EMD by a factor of 7).
 
@@
 diagrams of 3-D shapes show that blind greedy spanner can speed up computations
-by a factor of 2 to 5, saving hundreds of hours of machine time,
+by a factor of up to 5, saving hundreds of hours of machine time,
 even though the inputs in our data sets are of moderate size.
@@
 and we take $k = \lfloor \log v \rfloor - 1$.
-Afer we found a bucket for each pair $(p_i, p_j)$, 
+After we found a bucket for each pair $(p_i, p_j)$, 
 we have some sort of a weak ordering: we cannot say anything
@@
 We implemented two versions of  WSPD: one for the Euclidean case,
-using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
+using quad trees and the algorithm from \cite{hp-book}, and WSPD for general metric
 spaces with cover trees (using the base $\tau=1.3$). 
@@
 \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
-                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
+                          & \multicolumn{2}{c}{$q=1$} & \phantom{a} & \multicolumn{2}{c}{$q=2$} & \phantom{a} & \multicolumn{2}{c}{$q=3$}\\
                           \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
@@
 \end{tabular}
-    \caption{Sparseness of \bgrdy and \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
+    \caption{Sparseness of the \bgrdy and the \textsc{Greedy} spanners for 450 \dtype{original} persistence diagrams
  with Wasserstein distance $W_q$.}
@@
                           \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
-                          & Brute-force & Blind greedy && Brute-force & Blind greedy && Brute-force & Blind greedy \\ \midrule
-$\eps = 0.1$    &   $2605$  &  $1266$  &&  $3719$  &  $4085$  &&  $11$\,$781$  &  $13$\,$557$\\
-$\eps = 0.2$    &   $2209$  &  $878$  &&  $3338$  &  $2574$  &&  $10$\,$541$  &  $9383$ \\
-$\eps = 0.4$    &   $1988$  &  $586$  &&  $3088$  &  $1590$  &&  $-$    &  $6383$ \\
+                          & BF & BG && BF & BG && BF & BG \\ \midrule
+$\eps = 0.1$    &   $2605$  &  $1266$  &&  $3719$  &  $4085$  &&  $11\,781$  &  $13\,557$\\
+$\eps = 0.2$    &   $2209$  &  $878$  &&  $3338$  &  $2574$  &&  $10\,541$  &  $9383$ \\
+$\eps = 0.4$    &   $1988$  &  $586$  &&  $3088$  &  $1590$  &&  $10\,128$    &  $6383$ \\
 $\eps = 0.5$    &   $1744$  &  $483$  &&  $2835$  &  $1233$  &&  $9418$    &  $5146$ \\
-$\eps = 0.8$    &   $1691$  &  $368$  &&  $2809$  &  $814$  &&  $-$    &  $3419$ \\
-$\eps = 1.0$    &   $1645$  &  $324$  &&  $2780$  &  $672$  &&  $-$    &  $-$ \\
+$\eps = 0.8$    &   $1691$  &  $368$  &&  $2809$  &  $814$  &&  $9578$    &  $3419$ \\
+$\eps = 1.0$    &   $1645$  &  $324$  &&  $2780$  &  $672$  &&  $9539$    &  $2743$ \\
 \bottomrule
 \end{tabular}
-    \caption{Time required to approximate all pairwise distances with \bgrdy algorithm.
-    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.}
+    \caption{Time required to approximate all pairwise distances with the \bgrdy algorithm.
+    Data is for 450 \dtype{original} persistence diagrams with Wasserstein distance $W_q$, time in seconds.
+    BF stands for brute-force, BG for \bgrdy.}
 \label{tbl:mcgill_original_blind_greedy_spanner_timing}
@@
 
-\begin{table}\centering
-    \ra{1.2}
-    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
-        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
-        \cmidrule{2-3} \cmidrule{5-7} 
-                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
-$\eps = 0.1$    &   $0.20$  &  $0.43$    &&  $853\,682$   &  $248\,192$    & $0.29$   \\
-$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $853\,682$   & $192\,302$    & $0.22$   \\
-$\eps = 0.3$    &   $0.07$  &  $0.24$    &&  $853\,682$   & $161\,007$    & $0.19$   \\
-$\eps = 0.4$    &   $0.05$  &  $0.20$    &&  $853\,682$   & $140\,470$    & $0.16$   \\
-$\eps = 0.5$    &   $0.04$  &  $0.17$    &&  $853\,682$   & $129\,489$    & $0.15$   \\
-$\eps = 0.8$    &   $0.02$  &  $0.13$    &&  $853\,682$   & $104\,851$    & $0.12$   \\
-$\eps = 1.0$    &   $0.02$  &  $0.11$    &&  $853\,682$   & $92\,462$    & $0.11$   \\
-\bottomrule
-\end{tabular}
-    \caption{Results of \bgrdy algorithm on the dataset of 250 images
-    with Earth Mover's Distance, time in seconds.}
-\label{tbl:emd_gtsrb}
-\end{table}
+\myparagraph{Experiments with Persistence Diagrams}
+The other data sets that we used
+are the persistence diagrams
+computed from the McGill shape benchmark \cite{zhang2005mcgill}.
+More precisely, we generated two sets of diagrams,
+\dtype{original} and \dtype{perturbed}. The \dtype{original}
+diagrams were computed on the points of the 450 original
+point clouds from the shape data set; we only used
+persistence in dimension 0, since the diagrams in this dimension are larger in size.
+For the \dtype{perturbed} data set,
+we added random noise to points of the original shapes,
+creating 10 modified versions of each of them;
+then we computed persistence diagrams in dimension 0 of these modified
+shapes.
+Persistence diagrams were computed with the \libraryname{DIPHA}
+library \cite{bauer2014distributed}. 
+Wasserstein distance were computed approximately
+with the software library \hera using a relative error of $0.01$.
+We restricted to approximate computation, because computing the
+Wasserstein distance exactly (using the Hungarian algorithm)
+has been reported to be very expensive~\cite{kerber2017geometry}.
+
+Sparseness values of the \bgrdy spanner on the whole \dtype{original} data set
+for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
+the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
+the advantage of our algorithm fades away. A plausible explanation
+is that for higher values of $q$, the doubling dimension becomes larger;
+this is confirmed by the increase in size of the greedy spanner.
+For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
+computations for even $\eps = 0.1$; for a $1.5$-approximation, we
+reduce the number of distances computed by a factor of 10.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}[scale=0.89]
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
+        legend pos=outer north east]
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.1$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Greedy, $\eps = 0.5$}
+
+    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+    \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Sparseness of the \bgrdy spanner. The plot is for \dtype{perturbed} persistence diagrams,
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_sparseness_mcgill_perturbed}
+\end{figure}
+
+We also compare the time that is needed to compute the matrix of all pairwise distances
+naively with the time that is needed to approximate it with the \bgrdy spanner.
+The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+We compare our approach with the time to compute all pairwise-distance between the
+persistence diagrams with \hera, using the same approximation parameter
+$\eps$ (that is the reason why the BF column contains different values)\footnote{Let us be completely clear. The time in the BF column is the total running time for computing 
+$\wsdist{q}$ with the corresponding $\eps$, so \textit{all} distances in the BF case are approximate. The time in the BG column
+is the total time of the \bgrdy algorithm, which internally computes $\wsdist{q}$ with $\eps = 0.01$. In particular,
+the sparseness column in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness} shows the fraction of distances that are almost exact.}.
+Let us emphasize that the diagrams we have are rather small,
+with average cardinality of
+350 points. Nevertheless, on this data set the \bgrdy spanner
+provides a substantial improvement in terms of real running time,
+and we expect this advantage to grow for diagrams of larger cardinality.
+
+In order to see the growth of the spanner size, we use the \dtype{perturbed}
+data set, drawing from it random samples.
+We plot the sparseness of the spanners obtained by the \grdy
+and the \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
+for different values of $\eps$, always using the $\wsdist{1}$ distance.
+As expected, the ratio of computed distances decreases as the number of diagrams
+grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
+the growth of the spanner size relative to the number of diagrams.
+For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
+linear growth) for larger values of $n$.
+
+\begin{figure}[!htbp]
+    \begin{centering}
+\begin{tikzpicture}[scale=0.83]
+    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
+        legend pos=outer north east]
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.2$}
+
+        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Greedy, $\eps = 0.5$}
+
+        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
+        \addlegendentry{Blind greedy, $\eps = 0.5$}
+
+\end{axis}
+\end{tikzpicture}
+\end{centering}
+    \caption{Ratio \# edges /  \# diagrams for the \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
+    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
+    \label{fig:spanner_ratio_mcgill_perturbed}
+\end{figure}
 
@@
 \end{tabular}
-    \caption{Results of \bgrdy algorithm on the dataset of 200 images
+    \caption{Results of the \bgrdy algorithm on the dataset of 200 images
     with Earth Mover's Distance on \textit{Lab} histograms, time in seconds.}
-\label{tbl:emd_gtsrb}
+\label{tbl:emd_lab_caltech}
 \end{table}
 
-The other data sets that we used
-are the persistence diagrams
-computed from the McGill shape benchmark \cite{zhang2005mcgill}.
-More precisely, we generated two sets of diagrams,
-\dtype{original} and \dtype{perturbed}. The \dtype{original}
-diagrams were computed on the points of the 450 original
-point clouds from the shape data set; we only used
-persitence in dimension 0, since the diagrams in this dimension are larger in size.
-For the \dtype{perturbed} data set,
-we added random noise to points of the original shapes,
-creating 10 modified versions of each of them;
-then we computed persistence diagrams in dimension 0 of these modified
-shapes.
-Persistence diagrams were computed with the \libraryname{DIPHA}
-library \cite{bauer2014distributed}. 
-Wasserstein distance were computed approximately
-with the software library \hera using a relative error of $0.01$.
-We restricted to approximate computation because computing the
-Wasserstein distance exactly (using the Hungarian algorithm)
-has been reported to be very expensice~\cite{kerber2017geometry}.
-
-Sparseness values of \bgrdy spanner on the whole \dtype{original} data set
-for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
-the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
-the advantage of our algorithm fades away. A plausible explanation
-is that for higher values of $q$, the doubling dimension becomes larger;
-this is confirmed by the increase in size of the greedy spanner.
-For $\wsdist{1}$ we observe more encouraging figures, avoiding 70 \% of distance
-computations for $\eps = 0.1$; for a $1.5$-approximation, we
-reduce the number of distances computed by a factor of 10.
-
-\begin{figure}[!htbp]
-    \begin{centering}
-\begin{tikzpicture}[scale=0.89]
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.1$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Greedy, $\eps = 0.5$}
-
-    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-    \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Sparseness of \bgrdy spanner algorithm. The plot is for \dtype{perturbed} persistence diagrams,
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_sparseness_mcgill_perturbed}
-\end{figure}
-
-We also compare the time that is needed to compute the matrix of all pairwise distances
-naively with the time that is needed to approximate it with \bgrdy spanner.
-The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
-We compared our approach for a fixed
-$\eps$ with the time to compute all pairwise-distance between the
-persistence diagrams using \hera, using the same approximation parameter
-$\eps$. Let us emphasize that the diagrams we have are rather small,
-with average cardinality
-350 points. Nevertheless, on this data set \bgrdy spanner
-provides a substantial improvement in terms of real running time,
-and we expect this advantage to grow for diagrams of larger cardinality.
-
-In order to see the growth of the spanner size, we use the \dtype{perturbed}
-data set, drawing from it random samples.
-We plot the sparseness of the spanners obtained by \grdy
-and \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
-for different values of $\eps$, always using the $\wsdist{1}$ distance.
-As expected, the ratio of computed distances decreases as the number of diagrams
-grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
-the growth of the spanner size relative to the number of diagrams.
-For $\eps = 0.5$ the \bgrdy spanner starts to stabilize (i.e., demonstrates
-linear growth) for larger values of $n$.
-
-\begin{figure}[!htbp]
-    \begin{centering}
-\begin{tikzpicture}[scale=0.83]
-    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
-        legend pos=outer north east]
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.2$}
-
-        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Greedy, $\eps = 0.5$}
-
-        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
-        \addlegendentry{Blind greedy, $\eps = 0.5$}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Ratio \# edges /  \# diagrams for  \bgrdy spanner. The plot is for 0-dimensional diagrams of perturbed McGill data
-    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
-    \label{fig:spanner_ratio_mcgill_perturbed}
-\end{figure}
-
-For experiments with images, we used the German traffic sign
-dataset GTSRB \cite{gtsrb} and the code for computing the Earth Mover's distance
-by Doran \cite{doran_pyemd}.
-We randomly selected 250 images from the benchmark.
-The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
-The first two columns compare the sparseness of \bgrdy and greedy spanners.
-For smaller values of $\eps$,
-the \bgrdy spanner contains approximately 3 times more
-edges than the greedy spanner, so the ratio is roughly
-the same as for persistence diagrams.
+\myparagraph{Earth Mover's Distance} 
+We took the standard approach \cite{rubner2000earth} of transforming a colored image
+to the \textit{Lab} color space and creating a $16\times 16\times 16$
+distribution histogram. After that we compute the distance between normalized
+distributions with \cite{doran_pyemd}.
+We used the Caltech 101 dataset \cite{caltech101},
+randomly selecting 200 images from different (randomly chosen)
+categories.  We measured only the time needed to compute
+the Earth Mover's distance, without preprocessing (color space transformation
+and binning). The results are presented in \cref{tbl:emd_lab_caltech}.
+The first two columns of the table compare the sparseness of the \bgrdy and the greedy spanners.
 The last 3 columns show the advantage
 of \bgrdy in terms of the running time.
-If a $1.5$ approximation is acceptable,
+Even for the small relative error $\eps = 0.2$,
+the \bgrdy spanner saves about $28\%$ of time,
+and, if {$2$-approximation} is acceptable,
 then the \bgrdy algorithm speeds up
-the computation by a factor of more than $5$,
-and even for a $10 \%$ approximation, it reduces
-the running time by more than a factor of $2$,
-saving $98$ hours of computation time.
+the computation by a factor of almost $5$.
+
+Alternatively, one could also compute the Earth Mover's distance
+on gray value images by considering each image as a distribution on pixels.
+The results of this experiment are in \cref{sec:spanners_euclidean_addendum}, since this variant
+does not seem to be used in applications.
 
@@
 correspond to the candidates in the algorithm because
-we look for an approximate neareast neighbor only.
+we look for an approximate nearest neighbor only.
 Standard backwards analysis~\cite{seidel-backwards}
@@
 in that infinite-dimensional space, but have structures
-(e.g. clusters of close-by diagrams) which should be favourable for our approach.
+(e.g. clusters of close-by diagrams) which should be favorable for our approach.
 Moreover, the size of the \bgrdy spanner can itself be considered as a piece
@@
 In this section we include plots that give more information
-about the behaviour of \bgrdy spanner in Euclidean space.
+about the behavior of \bgrdy spanner in Euclidean space.
 
@@
 \end{figure}
+
+\begin{table}\centering
+    \ra{1.2}
+    \begin{tabular}{@{}rrrcrrrc@{}}\toprule
+        & \multicolumn{2}{c}{Sparseness} & \phantom{abc} & \multicolumn{3}{c}{Running time} \\
+        \cmidrule{2-3} \cmidrule{5-7} 
+                & Greedy  & Blind greedy && Brute-force & Blind greedy & Ratio \\ \midrule
+$\eps = 0.1$    &   $0.20$  &  $0.43$    &&  $853\,682$   &  $248\,192$    & $0.29$   \\
+$\eps = 0.2$    &   $0.11$  &  $0.30$    &&  $853\,682$   & $192\,302$    & $0.22$   \\
+$\eps = 0.3$    &   $0.07$  &  $0.24$    &&  $853\,682$   & $161\,007$    & $0.19$   \\
+$\eps = 0.4$    &   $0.05$  &  $0.20$    &&  $853\,682$   & $140\,470$    & $0.16$   \\
+$\eps = 0.5$    &   $0.04$  &  $0.17$    &&  $853\,682$   & $129\,489$    & $0.15$   \\
+$\eps = 0.8$    &   $0.02$  &  $0.13$    &&  $853\,682$   & $104\,851$    & $0.12$   \\
+$\eps = 1.0$    &   $0.02$  &  $0.11$    &&  $853\,682$   & $92\,462$    & $0.11$   \\
+\bottomrule
+\end{tabular}
+    \caption{Results of \bgrdy algorithm on the dataset of 250 images
+    with Earth Mover's Distance, time in seconds.}
+\label{tbl:emd_gtsrb}
+\end{table}
+
+\myparagraph{Earth Mover's Distance on Images}
+Let $I_1$, $I_2$ be two gray value images.
+If we divide each gray value of $I_{i}$
+by the total sum, we can interpret $I_i$
+as a probability measure supported on a rectangle
+that has one vertex at the origin and
+fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
+The Earth Mover's distance between images $I_1$ and $I_2$
+is, by definition, the Earth Mover's distance between
+these probability measures.
+Note that this defines a true metric on the space of images,
+while the standard approach with color histograms only defines
+a pseudometric, since any permutaion of pixels does not
+change the distribution in the color space.
+However, this metric is very hard to compute:
+for images with resolution $40\times 40$,
+it takes about 20 minutes. For this reason,
+we cannot test it on the Caltech 101 dataset.
+Instead, we used the German traffic sign
+dataset GTSRB \cite{gtsrb}. This
+dataset contains small images, and usually
+one traffic sign occupies most of the image area.
+It is interesting to see how
+the \bgrdy spanner would behave in this situation.
+
+We randomly selected 250 images from the dataset.
+The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
+As in \cref{sec:experiment_spanners}, 
+the first two columns give the sparseness.
+For smaller values of $\eps$,
+the \bgrdy spanner contains approximately 3 times more
+edges than the greedy spanner, so the ratio is roughly
+the same as for persistence diagrams.
+The last 3 columns compare the running time.
+Note that even for $\eps=0.1$,
+the \bgrdy spanner saves $70\%$ of the computation time.
+If we allow large relative error,
+we see the speed by almost a factor of $10$.
 
@@
 
-Let $I_1$, $I_2$ be two greyvalue images.
-We divide each greyvalue of $I_{1,2}$
-by the sum of all greyvalues in the corresponding image.
-Thus we can interpret $I_i$
-as a probability measure supported on a rectangle
-that has one vertex at the origin and
-fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
-The Earth Mover's distance between images $I_1$ and $I_2$
-is, by definition, the Earth Mover's distance between
-these probability measures.
-
 \end{document}

### 2019-12-04 paper/draft_socg.tex
Minor changes in paper

@@
 
-Alternatively, one could also compute the Earth Mover's distance
-on gray value images by considering each image as a distribution on pixels.
-The results of this experiment are in \cref{sec:spanners_euclidean_addendum}, since this variant
-does not seem to be used in applications.
+Alternatively, we also experimented with the Earth Mover’s distance on
+gray value images by considering each image as a distribution on pixels.
+For this (non-standard) variant, we obtain even larger speed-ups~-- see
+\cref{sec:spanners_euclidean_addendum} for details.
 
@@
 If we allow large relative error,
-we see the speed by almost a factor of $10$.
+we see the speed up by almost a factor of $10$.
 

### 2020-09-20 paper/draft_ijcga.tex
ijcga draft

--- 2019-12-04 paper/draft_socg.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref]{socg-lipics-v2019}
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
-
-\input{my_preamble.tex}
-
+\documentclass{ws-ijcga}
+
+\bibliographystyle{ws-ijcga}
+
+\usepackage{amsmath}%
+\usepackage{amssymb}%
+\usepackage{amsfonts}
+%\usepackage{amsthm}
+
+\usepackage{enumerate}
+\usepackage{booktabs}
+\usepackage{nccmath}
+\usepackage{diagbox}
+\usepackage{url}
+\usepackage{xspace}
+\usepackage{mathbbol}
+\usepackage{comment}
+\usepackage{braket}
+%\usepackage{mathptmx}
+
+% graphics
+\usepackage{tikz}
+\usepackage{pgf}
+\usepackage{graphicx}
+\usepackage{pgfplots}
+\pgfplotsset{compat=1.14}
+\usepackage{pgfplotstable}
+\usetikzlibrary{patterns}
+\usetikzlibrary{external}
+\usepackage{epsfig}
+\usepackage{wrapfig}
+\usepackage{color}
+\usepackage{mwe}
+\usepackage{floatflt}
+
+\usetikzlibrary{arrows}
+
+% copied from hera paper - check
+\usepackage{subcaption}
+
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{proposition}[theorem]{Proposition}
+%\newtheorem{corollary}[theorem]{Corollary}
+%\newtheorem{definition}[theorem]{Definition}
+
+% copied from expensive distances paper
+%
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
+
+% disable ugly latex defaults
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\renewcommand{\epsilon}{\varepsilon}
+\renewcommand{\phi}{\varphi}
+\newcommand{\eps}{\varepsilon}
+
+% hyphen in math mode
+\mathchardef\mhyphen="2D
+
+% common rings/fields/etc
+\newcommand{\NN}{\mathbb{N}}
+\newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\ZnZ}{\mathbb{Z}/n\mathbb{Z}}
+\newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\CC}{\mathbb{C}}
+\newcommand{\HH}{\mathbb{H}}
+\newcommand{\Fp}{\mathbb{F}_p}
+
+\newcommand{\hence}{\implies}
+\newcommand{\ergo}{\implies}
+
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\tr}{tr}
+\DeclareMathOperator*{\rk}{rank}
+\DeclareMathOperator*{\im}{im}
+\DeclareMathOperator*{\epi}{epi}
+% \char command is already defined
+\DeclareMathOperator*{\fchar}{char}
+
+%\theoremstyle{definition}
+%\newtheorem{mythm}{Theorem}
+%\newtheorem{mylem}{Lemma}
+%\newtheorem{mycor}{Corollary}
+%\newtheorem{myprop}{Proposition}
+%\newtheorem{mydef}{Definition}
+
+\newcommand{\yaxis}{$y$\nobreakdash-axis\xspace}
+\newcommand{\xaxis}{$x$\nobreakdash-axis\xspace}
+
+\newcommand{\algname}[1]{{\textsc{#1}}\xspace}
+
+% Hera stuff
+
+\newcommand{\ignore}[1]{}
+
+\newcommand{\dtype}[1]{{\textit{\small #1}}}
+
+\newcommand{\libraryname}[1]{{\textsc{#1}}\xspace}
+
+\newcommand{\hera}{\libraryname{Hera}}
+\newcommand{\Hera}{\libraryname{Hera}}
+\newcommand{\dionisus}{\libraryname{Dionysus}}
+\newcommand{\Dionisus}{\libraryname{Dionysus}}
+
+\pgfplotsset
+{
+    tick label style={font=\footnotesize},
+    legend style    ={font=\footnotesize},
+    label style     ={font=\footnotesize},
+    tickwidth       =2pt,
+    ylabel near ticks,
+    xlabel near ticks,
+    legend style     ={draw=none},
+    legend cell align=right,
+    legend plot pos  =right,
+    /pgfplots/narrow area legend/.style=
+    {
+        legend image code/.code={ \draw[#1] (0cm,-0.06cm) rectangle (0.6cm,0.06cm); }
+    },
+    /pgfplots/short line legend/.style=
+    {
+        legend image code/.code={ \draw[#1] (0cm,0cm) rectangle (0.6cm,0cm); }
+    },
+}
+
+\pgfplotscreateplotcyclelist{filled-color}
+{
+    every mark/.append style={solid,fill=blue!80!white},    mark=*          \\%
+    every mark/.append style={solid,fill=red!80!white},     mark=square*    \\%
+    every mark/.append style={solid,fill=orange!80!white},  mark=pentagon*  \\%
+    every mark/.append style={solid,fill=green},            mark=diamond*   \\%
+    every mark/.append style={solid,fill=black},            mark=triangle*  \\%
+}
+
+\newcommand{\addaveraged}[2][]
+           {\addplot+[#1,error bars/.cd, y dir=both,y explicit] table[x index=0, y index=1, y error index=2] {#2};}
+
+\pgfplotsset{
+    geommark/.style={black,mark options={solid,fill=red!80!white},mark=square*},
+    nongeommark/.style={black,mark options={solid,fill=blue!80!white},mark=*},
+    dionmark/.style={black,mark options={solid,fill=orange!80!white},mark=pentagon*},
+    greentrianglemark/.style={black,mark options={solid,fill=green!80!white},mark=triangle*},
+    browndiamondmark/.style={black,mark options={solid,fill=brown!80!white},mark=diamond*}
+}
+
+% all distances
+% interleaving
+\newcommand{\intdist}{{D_I}}
+% bottleneck
+\newcommand{\btdist}{{W_\infty}}
+% Wasserstein of order q (parameter)
+\newcommand{\wsdist}[1]{{W_{#1}}}
+
+% expensive distances stuff
+\input{my_preamble_exp_dist.tex}
 \title{Metric Spaces with Expensive Distances}
-\titlerunning{Metric Spaces with Expensive Distances}
-% uncomment for LIPICS 
-%\author{Anonymous Authors}{Anonymous}{}{}{}
-\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{https://orcid.org/0000-0002-8030-9299}{Supported by Austrian Science Fund (FWF) grant number P 29984-N35}
-\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
-\authorrunning{M. Kerber and A. Nigmetov}
-%\authorrunning{Anonymous Authors}
-
-\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
-\ccsdesc[500]{Theory of computation~Random search heuristics}
-%% \category{dummy}
-
-% \category{dummy}
-
-\Copyright{Michael Kerber and Arnur Nigmetov}
-%\Copyright{Anonymous Authors}
-% END uncomment for LIPICS
+
+\author{Michael Kerber}
+
+%\address{University Department, University Name, Address\\
+%City, State ZIP/Zone,Country\,\footnote{State completely without
+%abbreviations, the affiliation and mailing address,
+%including country. Typeset in 8 pt italic.}\\
+%e-mail address
+%}
+
+\author{Arnur Nigmetov}
+
+%\address{Group, Laboratory, Address\\
+%City, State ZIP/Zone, Country\\
+%e-mail address
+%}
 
@@
 This information consists of lower and upper bounds for unknown distances, obtained from known distances
-by triangle inequality (see \cref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
+by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
 operations whenever a new distance has been computed, turning the method useless in the standard computational model.
@@
 \myparagraph{Outline}
-We review the properties of spanner algorithms in \cref{sec:back}.
-In \cref{sec:blind_spanners}, we introduce different
+We review the properties of spanner algorithms in Section~\ref{sec:back}.
+In Section~\ref{sec:blind_spanners}, we introduce different
 variants of the blind spanner algorithm. Experimental
-results on spanners are presented in \cref{sec:experiment_spanners}.
+results on spanners are presented in Section~\ref{sec:experiment_spanners}.
 Then we move to the ANN problem, giving
-an outline of the proof of the $O(\log n)$ bound in \cref{sec:ann}.
+an outline of the proof of the $O(\log n)$ bound in Section~\ref{sec:ann}.
 The full proof can be found in the appendix.
-We report our experimental results for ANN in \cref{sec:experiment_ann}.
+We report our experimental results for ANN in Section~\ref{sec:experiment_ann}.
 
@@
 space, on images with Earth Mover's distance, and on persistence diagrams
-with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
+with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See Section~\ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
 Clearly, our model is not meaningful in Euclidean spaces because
@@
 
-\cref{fig:spanner_sparseness}
+Figure~\ref{fig:spanner_sparseness}
 shows the number of edges of the spanner for various variants of 
@@
 perform significantly better than both variants of \textsc{BlindQuasiSorted}.
-
-\ignore{
-As for different variants of the \brndm algorithm,
-we note that their performance is almost the same, and the algorithm works
-significantly better than \textsc{QuasiSorted} variants, but obviously
-worse than the blind greedy variant. There is a consistent, though small,
-difference between the variants that do not force lower bounds first and
-the other two variants of the \brndm.  (see \cref{fig:blind_rbr_variants}). 
-}
 
@@
 dense spanners. More data from experiments in $\RR^n$ can be found in
-\cref{sec:spanners_euclidean_addendum}.
+Section~\ref{sec:spanners_euclidean_addendum}.
 
@@
 Sparseness values of the \bgrdy spanner on the whole \dtype{original} data set
-for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
+for different values of $\eps$ and $q$ are presented in Table~\ref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
 the \grdy algorithm is used as baseline.  We observe that for higher values of $q$
@@
 naively with the time that is needed to approximate it with the \bgrdy spanner.
-The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
+The timings are given in Table~\ref{tbl:mcgill_original_blind_greedy_spanner_timing}.
 We compare our approach with the time to compute all pairwise-distance between the
@@
 is the total time of the \bgrdy algorithm, which internally computes $\wsdist{q}$ with $\eps = 0.01$. In particular,
-the sparseness column in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness} shows the fraction of distances that are almost exact.}.
+the sparseness column in Table~\ref{tbl:mcgill_original_blind_greedy_spanner_sparseness} shows the fraction of distances that are almost exact.}.
 Let us emphasize that the diagrams we have are rather small,
@@
 We plot the sparseness of the spanners obtained by the \grdy
-and the \bgrdy algorithms in \cref{fig:spanner_sparseness_mcgill_perturbed}
+and the \bgrdy algorithms in Figure~\ref{fig:spanner_sparseness_mcgill_perturbed}
 for different values of $\eps$, always using the $\wsdist{1}$ distance.
 As expected, the ratio of computed distances decreases as the number of diagrams
-grows. In \cref{fig:spanner_ratio_mcgill_perturbed} we show
+grows. In Figure~\ref{fig:spanner_ratio_mcgill_perturbed} we show
 the growth of the spanner size relative to the number of diagrams.
@@
 the Earth Mover's distance, without preprocessing (color space transformation
-and binning). The results are presented in \cref{tbl:emd_lab_caltech}.
+and binning). The results are presented in Table~\ref{tbl:emd_lab_caltech}.
 The first two columns of the table compare the sparseness of the \bgrdy and the greedy spanners.
@@
 For this (non-standard) variant, we obtain even larger speed-ups~-- see
-\cref{sec:spanners_euclidean_addendum} for details.
+Section~\ref{sec:spanners_euclidean_addendum} for details.
 
@@
 
-The full proof is given in \cref{sec:proof_ann},
+The full proof is given in Section~\ref{sec:proof_ann},
 we summarize the main ideas:
@@
 around $p_j$ as described above degenerates, and the packing argument fails.
-Indeed, as the example in \cref{fig:exact_bad_example} (in~\cref{sec:proof_ann}) 
+Indeed, as the example in Figure~\ref{fig:exact_bad_example} (in Section~\ref{sec:proof_ann}) 
 shows, there are point sets where the expected number
@@
 we plot the number of computed distances divided
-by the logarithm of the number of points in \cref{fig:ann_const_ratio} (for $d = 2$).
+by the logarithm of the number of points in Figure~\ref{fig:ann_const_ratio} (for $d = 2$).
 We see that this ratio, though fluctuating a lot, remains in the interval $[1,4]$. This not just confirms
@@
 As expected, in high dimensions the algorithm does not perform as well.
-In \cref{fig:ann_dimension_dependence} we plot the average number of computed distances
+In Figure~\ref{fig:ann_dimension_dependence} we plot the average number of computed distances
 for $d = 2, 5, 10$. While for $d = 2$ the growth is hardly noticeable, for $d = 10$
@@
 Similar experiments were performed on \dtype{perturbed} data set of persistence
-diagrams, which was described in \cref{sec:experiment_spanners}.
+diagrams, which was described in Section~\ref{sec:experiment_spanners}.
 For a given number $n$, we randomly select $n$ diagrams from the data set
 and from the remaining diagrams we randomly choose 20 query points.
-In \cref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
+In Figure~\ref{fig:ann_const_ratio_mcgill} we plot the average ratio of computed
 distances to $\log(n)$ for $\wsdist{1}$ and $\wsdist{2}$ metrics.
@@
 
-%As for different variants of the \brndm algorithm,
-%we note that their performance is almost the same, and the algorithm works
-%significantly better than \textsc{QuasiSorted} variants, but obviously
-%worse than the blind greedy variant. There is a consistent, though small,
-%difference between the variants that do not force lower bounds first and
-%the other two variants of the \brndm. \ignore{ (see \cref{fig:blind_rbr_variants}). }
-
 \begin{figure}[!htbp]
@@
 
-\cref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
+Figure~\ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
 The ideal behavior is demonstrated by the non-blind greedy spanner,
@@
 is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
-linear regression (after taking $\log$).  We give in \cref{tbl:regr_coeff_spanner}
+linear regression (after taking $\log$).  We give in Table~\ref{tbl:regr_coeff_spanner}
 the estimated exponents $\alpha$ for \bgrdy and standard
@@
 
-The plot in \cref{fig:blind_greedy_dimensions}
+The plot in Figure~\ref{fig:blind_greedy_dimensions}
 compares performance of \bgrdy in different dimensions.
@@
 
-The plot in \cref{fig:spanner_eps_dependence}
+The plot in Figure~\ref{fig:spanner_eps_dependence}
 compares the \bgrdy and \textsc{Greedy} algorithms
@@
 We randomly selected 250 images from the dataset.
-The results of the experiment are summarized in \cref{tbl:emd_gtsrb}.
-As in \cref{sec:experiment_spanners}, 
+The results of the experiment are summarized in Table~\ref{tbl:emd_gtsrb}.
+As in Section~\ref{sec:experiment_spanners}, 
 the first two columns give the sparseness.
@@
 
-\section{Proof of \cref{thm:ann_bound}}
+\section{Proof of Theorem~\ref{thm:ann_bound}}
 \label{sec:proof_ann}
@@
 contain at most one point from $P$, 
-\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
+\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\square\]
 \end{proof}
@@
 
-\cref{lem:bound_lemma}
+Lemma~\ref{lem:bound_lemma}
 can be summarized as follows: if $\dist(p_i,q)$ is computed in the algorithm,
@@
 \caption{First two steps of the ANN algorithm. 
-%    First $p_1$ is 
-%chosen as the current candidate, and we must compute $\dist(p_2, q)$. After that the algorithm will not compute
-%distance to any of the points inside the heavily shaded ball or outside the lightly shaded ball that are centered 
-%at $p_2$, because their lower bounds allow us to discard them. Note that the point $p_5$, which
-%is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points $p_6,p_7,p_8,p_9$
-%in the annulus between the dashed and solid circle, which are further from $q$ than $p_5$, will be chosen as $c$. 
-%This shows that in our algorithm the distance from the candidate to $q$ can drop \textit{slower} than in the
-%brute-force algorithm, thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis. 
 The small black ball between the dashed circle and the solid circle has radius $v_1 \eps / (1 + \eps)$; it is the ball that we use in the packing argument, because it is smaller than any of the lightly shaded balls that correspond to points like $p_2$ and $p_4$, that is, the points that do not improve $v$.}
@@
 
-In \cref{fig:ann_illustration} we illustrate this. First, $p_1$ is
+In Figure~\ref{fig:ann_illustration} we illustrate this. First, $p_1$ is
 chosen as the current candidate, and we must compute $\dist(p_2, q)$.
@@
 \begin{proof}
-By the first part of \cref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
+By the first part of Lemma~\ref{lem:bound_lemma}, every point in $p_k,\ldots,p_\ell$
 whose distance to $q$ is queried lies in the ball of radius $(1+\frac{1}{1+\eps})v_k=\frac{2+\eps}{1+\eps} v_k$
 around $c_k$. Moreover, if the distance of two points $p_i$ and $p_j$ with $k\leq i<j\leq\ell$
-is computed, the second part of \cref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
+is computed, the second part of Lemma~\ref{lem:bound_lemma} implies that $\dist(p_i,p_j)> \frac{\eps}{1+\eps}r_i\geq \frac{\eps}{1+\eps}v_\ell$.
 Hence, all points in $p_k,\ldots,p_\ell$ for which the algorithm computes the distance
-have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying \cref{lem:packing_lemma}.
-See \cref{fig:ann_illustration}.
+have a pairwise distance of at least $\frac{\eps}{1+\eps}v_\ell$. The statement follows by applying Lemma~\ref{lem:packing_lemma}.
+See Figure~\ref{fig:ann_illustration}.
 \end{proof}
@@
 the number of computed distances is a constant (since $v_k=v_\ell$). 
-This means that to prove \cref{thm:ann_bound}, it would suffice to show
+This means that to prove Theorem~\ref{thm:ann_bound}, it would suffice to show
 that the candidate changes only a logarithmic number of times in expectation.
-Let us consider \cref{fig:ann_illustration} again. Notice that the point $p_5$, which
+Let us consider Figure~\ref{fig:ann_illustration} again. Notice that the point $p_5$, which
 is closer to $q$ than $p_1$, also will not be a candidate, and at least one of the points 
@@
 candidate to $q$ can drop \textit{slower} than in the brute-force algorithm, 
-thus \cref{thm:ann_bound} does not immediately follow from standard backwards analysis,
+thus Theorem~\ref{thm:ann_bound} does not immediately follow from standard backwards analysis,
 and we need to modify it slightly.
 
-\begin{proof} (of \cref{thm:ann_bound})
+\begin{proof} (of Theorem~\ref{thm:ann_bound})
 %
@@
 some $i\leq j$, and in the sequence $r_1,\ldots,r_{\ell-1}$, $r_k$
-is minimal by construction. Using \cref{lem:sequence_lemma},
+is minimal by construction. Using Lemma~\ref{lem:sequence_lemma},
 the number of distance computations among the points

### 2020-09-20 paper/ijcga-expensive-distances.tex
ijcga draft
Fix typos and small reviewers comments.

--- 2020-09-20 paper/draft_ijcga.tex
@@
-\documentclass{ws-ijcga}
+\documentclass[]{ws-ijcga}
 
@@
 
-We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
-with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
+We propose several heuristics to explore the metric space to obtain accurate lower and upper bounds
+with a small number of distance computations. Once the ratio of upper and lower bound is at most $(1+\eps)$
 for each point pair, the set of all computed distances forms the spanner.
@@
 \end{enumerate}
-The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
+The algorithm for constructing a net-tree is complicated and not easy to implement. Beygelzimer
 et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
@@
 $O(n\log n+n \eps^{-d})$ distances 
-to construct a \espanner in doubling dimension $d$.
+to construct a {\espanner} in doubling dimension $d$.
 The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
@@
 We introduce a new framework for constructing spanners
-which we call \emph{blind spanners}: the idea is to maintain,
+that we call \emph{blind spanners}: the idea is to maintain,
 for every pair of points $(p_i,p_j)$,
@@
 space, on images with Earth Mover's distance, and on persistence diagrams
-with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See Section~\ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
+with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
 Clearly, our model is not meaningful in Euclidean spaces because
@@
 dense spanners. More data from experiments in $\RR^n$ can be found in
-Section~\ref{sec:spanners_euclidean_addendum}.
+\ref{sec:spanners_euclidean_addendum}.
 
@@
 gray value images by considering each image as a distribution on pixels.
-For this (non-standard) variant, we obtain even larger speed-ups~-- see
-Section~\ref{sec:spanners_euclidean_addendum} for details.
+For this (non-standard) variant, we obtain even larger speed-ups~--- see
+\ref{sec:spanners_euclidean_addendum} for details.
 
@@
 
-The full proof is given in Section~\ref{sec:proof_ann},
+The full proof is given in \ref{sec:proof_ann},
 we summarize the main ideas:
@@
 around $p_j$ as described above degenerates, and the packing argument fails.
-Indeed, as the example in Figure~\ref{fig:exact_bad_example} (in Section~\ref{sec:proof_ann}) 
+Indeed, as the example in Figure~\ref{fig:exact_bad_example} (in \ref{sec:proof_ann}) 
 shows, there are point sets where the expected number
@@
 
-The existence of a $2$-approximation algorithm did not help us to significantly
-reduce the number of exact distance computations, although it seems obvious
+The existence of a $2$-approximation algorithm did not help in significantly
+reducing the number of exact distance computations, although it seems obvious
 that knowing the all approximate distances is useful.
@@
 In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
-such that $r_i<r_k$ for all $1\leq i\leq k-1$. We call an element
+such that $r_i > r_k$ for all $1\leq i\leq k-1$. We call an element
 of this form a \emph{minimum} of the sequence. 
@@
 Let $\diag$ denote the line in the plane with the equation $y = x$; we call it the diagonal.
-A persistence diagram is a finite multiset of points $(x, y)$ with $x > y$
-$\Delta$ and all points of $\Delta$ with infinite cardinality.
+A \textit{persistence diagram} is a
+multiset that consists of all points of $\diag$ with infinite multiplicity
+and finitely many points $(x, y)$ with $y > x$, each with finite multiplicity.
 The points with $y > x$ are called \textit{off-diagonal} points,
-and points on $\Delta$ are called \textit{diagonal} points.
-The \emph{$q$-Wasserstein distance} is defined as
+and the points on $\diag$ are called \textit{diagonal} points.
+Let $X$ and $Y$ be two persistence diagrams.
+The \emph{$q$-Wasserstein distance} between $X$ and $Y$ is defined as
 \[
@@
 Equivalent definition, which is better suited for actual computations,
-is as follows. Let $\pi\colon \RR^2 \to \Delta$ be defined
+is as follows. Let $\pi\colon \RR^2 \to \diag$ be defined
 as $(x,y) \mapsto (\frac{x+y}{2}, \frac{x+y}{2})$; we refer to $\pi$ as \textit{diagonal
-projection}. Let $\tX$ be the set $(X\setminus \Delta) \cup \pi(Y \setminus \Delta)$
-and $\tY$ be the set $(Y\setminus \Delta) \cup \pi(X \setminus \Delta)$; in plain words,
+projection}. Let $\tX$ be the set $(X\setminus \diag) \cup \pi(Y \setminus \diag)$
+and $\tY$ be the set $(Y\setminus \diag) \cup \pi(X \setminus \diag)$; in plain words,
 to obtain $\tX$, we add to the off-diagonal points of $X$

### 2021-01-18 ijcga_expensive_distances_sources/ijcga-expensive-distances.tex
Add cleaned sources for final IJCGA version

--- 2020-09-20 paper/ijcga-expensive-distances.tex
@@
 \usepackage{amsfonts}
-%\usepackage{amsthm}
 
@@
 \usepackage{braket}
-%\usepackage{mathptmx}
 
@@
 
-% copied from hera paper - check
 \usepackage{subcaption}
 
-%\newtheorem{theorem}{Theorem}
-%\newtheorem{lemma}[theorem]{Lemma}
-%\newtheorem{proposition}[theorem]{Proposition}
-%\newtheorem{corollary}[theorem]{Corollary}
-%\newtheorem{definition}[theorem]{Definition}
-
-% copied from expensive distances paper
-%
 \usepackage{algorithm}
@@
 
-%\theoremstyle{definition}
-%\newtheorem{mythm}{Theorem}
-%\newtheorem{mylem}{Lemma}
-%\newtheorem{mycor}{Corollary}
-%\newtheorem{myprop}{Proposition}
-%\newtheorem{mydef}{Definition}
-
 \newcommand{\yaxis}{$y$\nobreakdash-axis\xspace}
@@
 
-% Hera stuff
-
-\newcommand{\ignore}[1]{}
-
 \newcommand{\dtype}[1]{{\textit{\small #1}}}
-
 \newcommand{\libraryname}[1]{{\textsc{#1}}\xspace}
-
 \newcommand{\hera}{\libraryname{Hera}}
-\newcommand{\Hera}{\libraryname{Hera}}
-\newcommand{\dionisus}{\libraryname{Dionysus}}
-\newcommand{\Dionisus}{\libraryname{Dionysus}}
 
@@
 % expensive distances stuff
-\input{my_preamble_exp_dist.tex}
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+
+\renewcommand{\S}{\mathbb{S}}
+\newcommand{\dappr}{\tilde{d}}
+\newcommand{\orient}{O}
+\newcommand{\cpro}{\Pi}
+
+\newcommand{\metricspace}{\mathcal{M}}
+\newcommand{\approxmetric}{\gamma}
+\newcommand{\parent}{\mathrm{par}}
+\newcommand{\point}{p}
+\newcommand{\level}{\ell}
+\newcommand{\diam}{\mathrm{diam}}
+\newcommand{\diag}{{\Delta}}
+\newcommand{\pointset}{P}
+\newcommand{\distspace}{\mathcal{M}}
+\newcommand{\dist}{\delta}
+\newcommand{\adist}{\gamma}
+\newcommand{\complexity}{C_{\dist}}
+\newcommand{\doublingdimension}{\Delta}
+
+\newcommand{\espanner}{{$(1+\eps)$-spanner}}
+
+\newcommand{\grdy}{\algname{Greedy}}
+\newcommand{\bgrdy}{\algname{BlindGreedy}}
+\newcommand{\brndm}{\algname{BlindRandom}}
+\newcommand{\bqsgrdy}{\algname{BlindQuasiSortedGreedy}}
+\newcommand{\bqsshaker}{\algname{BlindQuasiSortedShaker}}
+
+\newcommand{\tX}{{\tilde{X}}}
+\newcommand{\tY}{{\tilde{Y}}}
+
+\newcommand{\myparagraph}[1]{\textbf{#1.}}
+
+\def\marrow{\marginpar[\hfill$\longrightarrow$]{$\longleftarrow$}}
+
 \title{Metric Spaces with Expensive Distances}
@@
 
-%\address{University Department, University Name, Address\\
-%City, State ZIP/Zone,Country\,\footnote{State completely without
-%abbreviations, the affiliation and mailing address,
-%including country. Typeset in 8 pt italic.}\\
-%e-mail address
-%}
+\address{Institute of Geometry, Graz University of Technology, Kopernikusgasse 24\\
+8010 Graz, Austria\,\\
+kerber@tugraz.at
+}
 
 \author{Arnur Nigmetov}
-
-%\address{Group, Laboratory, Address\\
-%City, State ZIP/Zone, Country\\
-%e-mail address
-%}
+\address{Lawrence Berkeley National Laboratory, 1 Cyclotron Road\\
+Berkeley, CA 94720, USA\,\\
+anigmetov@lbl.gov
+}
 
@@
 of all distances in the metric space, and it is plausible to assume that such an ordering will simplify
-algorithmic tasks on metric spaces, at least in practice.
+algorithmic tasks on metric spaces, at least in practice. Our main motivation for this variant
+is the Wasserstein distance, for which one can use the auction algorithm by Bertsekas
+to guarantee the relative error provided by the user. From our practice, the algorithm terminates much
+faster, if the user is satisfied with a $2$-approximation; better approximations, e.g., with relative error $0.1$,
+require much more computations.
 
@@
 even though the inputs in our data sets are of moderate size.
-%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
-%computing all $\binom{n}{2}$ distances.
 Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
@@
 The \textit{doubling dimension} of a doubling space is defined as $\log k$
-(%since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
 we always use $\log$ to denote the logarithm with base 2).
@@
 
-\ignore{
-The last option that we consider is based on an additional assumption: while
-computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
-cheap approximation algorithm
-that gives a $2$-approximation of $\dist$ much faster 
-(the output of this approximation algorithm is guaranteed to
-lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
-that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
-We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
-these intervals are not disjoint, but by calling the approximation algorithm we can assign
-each pair $(p_i, p_j)$ to one of the buckets.
-Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
-then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
-that fully contains this interval, thus we need to find $k$ such that
-$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
-and we take $k = \lfloor \log v \rfloor - 1$.
-After we found a bucket for each pair $(p_i, p_j)$, 
-we have some sort of a weak ordering: we cannot say anything
-about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
-are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
-traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
-we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
-we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
-order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
-in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
-We refer to the former option as \bqsgrdy and to the latter one as \bqsshaker.
-}
-
 \section{Experiments on spanners}
@@
 vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.
-
-%We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
-%experimentally the performance of these heuristics. Clearly, for this metric space, our
-%cost model is not meaningful since distance comparisons are cheap;
-%but we picked this environment for controlled experiments.
-%In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
-%By \textit{sparseness} of a graph with $n$
-%vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.
 
@@
 is equal to the spanner size, while for the non-blind greedy spanner,
-this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
+this number is always $\binom{n}{2}$ and for WSPD it is lower bounded
 by the size of the spanner.
@@
 are infeasible for such $n$.
-
-%\ignore{
-%Summing up, the \bgrdy algorithm is the best algorithm for our model, but also \brndm
-%algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
-%first. 
-%If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
-%Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
-%tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
-%Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
-%(\bqsgrdy is worse, requiring all the edges). It would seem plausible that, if we have
-%access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
-%find a working heuristic.
-%}
 
@@
 in that infinite-dimensional space, but have structures
-(e.g. clusters of close-by diagrams) which should be favorable for our approach.
+(e.g., clusters of close-by diagrams) which should be favorable for our approach.
 Moreover, the size of the \bgrdy spanner can itself be considered as a piece
@@
 whether \bgrdy is worth trying.
-%The distances that \bgrdy computes would be computed anyway (and can be re-used),
-%and
 
@@
 experimental evaluation on worst-case examples do not show such a quadratic
-behavior. Therefore, we postpone the theoretical analysis of the spanner construction
+behavior.
+Finally, an anonymous reviewer kindly pointed us to
+the paper \cite{beacons}, which uses the same idea of inferring
+lower and upper bounds for unseen distances from the triangle inequality.
+The authors of \cite{beacons} proved that it is possible
+to approximate a large fraction of pairwise distances.
+However, there is no proof that one gets a spanner in the strict sense,
+and the problem seems hard.
+Therefore, we postpone the theoretical analysis of the spanner construction
 to an extended version of this article.
@@
 about the behavior of \bgrdy spanner in Euclidean space.
-
-\ignore{
-\begin{figure}[ht!]
-    \begin{centering}
-\begin{tikzpicture}
-    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
-        legend pos=outer north east]
-
-    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, lower bound first}
-
-    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
-    \addlegendentry{Blind random, connect first, lower bound first}
-
-\end{axis}
-\end{tikzpicture}
-\end{centering}
-    \caption{Comparison of the four variants of \brndm algorithm.}
-    \label{fig:blind_rbr_variants}
-\end{figure}
-}
 
@@
 while the standard approach with color histograms only defines
-a pseudometric, since any permutaion of pixels does not
+a pseudometric, since any permutation of pixels does not
 change the distribution in the color space.
@@
 \end{enumerate}
-%By definition, the \textit{cost} of transportation plan $C$
 Let $\dist(\cdot, \cdot)$ denote the Euclidean
